---
title: "Longitudinal Association of Obesity and Stress with Type 2 Diabetes"
author: "Dave Bridges"
date: "2024-07-17"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

## Purpose

To test the effect modification of obesity on the stress-diabetes relationships in a longitudinal manner.

```{r data-entry}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures-longitudinal/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

library(readr)
library(dplyr)
library(tidyr)
library(broom)

input.file <- 'data-combined.csv'
combined.data <- read_csv(input.file)

#enrollment specific data
enrollment.datafile <- 'SurveyAnthropometricsBMI.csv'
enrollment.data <- read_csv(enrollment.datafile)

#combined.data <-
#  left_join(combined.data,enrollment.data,by="DeID_PatientID")
```

Loaded in the cleaned data from `r input.file`. This script can be found in `r getwd()` and was most recently run on `r date()`. This dataset has `r dim(combined.data)[1]` values.

The diagnosis of Type 2 Diabetes is shown in the `DeID_Diabetes_Diagnosis` field.  

```{r data-cleaning}
library(forcats)
combined.data <- 
  combined.data %>%
  mutate(BMI_cat= factor(BMI_cat, 
                         levels=c("Underweight",
                                  "Normal",
                                  "Overweight",
                                  'Class I Obese',
                                  'Class II Obese',
                                  'Class III Obese'))) %>%
  mutate(BMI_cat.obese= factor(BMI_cat.obese, 
                               levels=c("Underweight",
                                        "Normal",
                                        "Overweight",
                                        'Obese'))) %>%
  mutate(BMI_cat.Ob.NonOb= factor(BMI_cat.Ob.NonOb, 
                                  levels=c("Non-Obese",
                                           'Obese'))) %>%
  mutate(Stress=relevel(as.factor(High.Stress),ref="Low")) %>% #set low as reference value
    mutate(Race.Ethnicity=relevel(as.factor(Race.Ethnicity),ref="White")) %>% #set white as reference value
  mutate(Stress.quartile=fct_recode(as.factor(Stress.quartile),
                                    "Q1"="(-0.016,4]",
                                    "Q2"="(4,8]",
                                    "Q3"="(8,12]",
                                    "Q4"="(12,16]")) %>%
    mutate(Stress.quartile= factor(Stress.quartile, 
                                  levels=c("Q1","Q2","Q3","Q4"))) 
  
```

# Creating a Longitudinal Dataset

## Baseline characteristics

The baseline is the date at which they did the survey (`DeID_SurveyDate`).  At this stage they needed a PSS score and needed not to be diagnosed with type 2 diabetes before that survey.

```{r removing-existing-t2d}
#Remove baseline diabetes
library(lubridate)
combined.data <-
  combined.data %>%
  mutate(Diabetes.Diagnosis = ymd_hms(DeID_Diabetes_Diagnosis),
         Survey.Date = mdy_hm(DeID_Survey_Date),
         DeID_DeceasedDate = mdy_hm(DeID_DeceasedDate)) %>%
  mutate(Time.to.Diabetes = (Survey.Date %--% Diabetes.Diagnosis)/ddays(1)) %>%
  mutate(Follow.Up = case_when(
    !(is.na(DeID_DeceasedDate))~(Survey.Date %--% DeID_DeceasedDate)/ddays(1),
    is.na(DeID_DeceasedDate)~(Survey.Date %--% today())/ddays(1))) #should be last encounter

combined.data %>%
  filter(Type2Diabetes==1,
         Time.to.Diabetes>0) %>%
  select(Survey.Date,Diabetes.Diagnosis,DeID_DeceasedDate,Time.to.Diabetes,Follow.Up) %>% 
  kable(caption="Patients who were diagnosed with diabetes after their survey")

combined.data %>%
  group_by(Time.to.Diabetes>0) %>%
  count %>%
  kable(caption="Summary of patients and type 2 diabetes.  True is got diabetes after the survey, False is got diabetes before the survey.  NA never got diabetes")

with(combined.data, table(Time.to.Diabetes>0,Type2Diabetes)) %>% kable(caption="Confirmation that every patient with diabetes is categorized into before or after")

non.diabetes.data <- 
  combined.data %>%
  filter(is.na(Time.to.Diabetes) | Time.to.Diabetes>0)

#need death data for censoring
#need BMI data near survey event
#need last encounter
```

From the `r nrow(combined.data)` dataset, `r nrow(non.diabetes.data)` people did not have diabetes at baseline.  The mean follow up time was `r round(mean(combined.data$Follow.Up)/365,2)` years.

## Removing no PSS Score

```{r no-pss}
baseline.participants <-
  non.diabetes.data %>%
  filter(!is.na(Stress_d1))
```

Among the `r nrow(non.diabetes.data)`, `r nrow(baseline.participants)` had a value for PSS score.

# Defining the Intervals

The endpoint is one of:

* Diagnosed with diabetes (time from survey to time from diabetes)
* Last encounter, but still alive (time from survey to last encounter)
* Death (time from survey to date of death)

```{r endpoint}
library(survival)
# event.data <-
#   baseline.participants %>%
#   mutate(status=case_when(Type2Diabetes==1~1,
#                          Type2Diabetes==0~1)) %>%
#   mutate(time=case_when(Type2Diabetes==1~Time.to.Diabetes,
#                         Type2Diabetes==0~All.Time))
# 
# #with(event.data, Surv(time,status))[1:100] #lots of missing All.Time
# 
# #event.data %>% select(All.Time, Time.to.Diabetes, time, Type2Diabetes, status, Survey.Date,DeID_Survey_Date) %>% head
# 
# s1 <- survfit(Surv(time, status) ~ 1, data = event.data)
# library(ggsurvfit)
# 
# survfit2(Surv(time, status) ~ BMI_cat.Ob.NonOb, data = event.data) %>% 
#   ggsurvfit(type="risk") +
#   labs(
#     x = "Days",
#     y = "Diabetes Diagnoses"
#   )  +
#   add_confidence_interval() +
#   add_risktable() +
#   add_quantile(y_value = 0.6, color = "gray50", linewidth = 0.75) +
#   scale_ggsurvfit()
```

# Session Information

```{r sessioninfo}
sessionInfo()
```
