---
title: "Summarizing the Encounter Level Data to Define BMI"
author: "Dave Bridges"
format: html
editor: visual
---

## Purpose

To test the effect modification of obesity on the stress-diabetes relationships. This script collects the the raw data files, processes and merges them. This script can be found in `r getwd()` and was most recently run on `r date()`.

```{r input-data}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

encounters.datafile <- 'EncounterAnthropometricsBMI.csv'
```

The input data file is in `r encounters.datafile`. This includes the BMI for each patient, potentially at multiple time points. This script takes this file and pulls out just the first BMI measure

```{r data-entry}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)

#get encounter data
bmi.data <- read_csv(encounters.datafile) 
```

# Defining BMI Data

We will use the median BMI measure if there are multiple in the encounters file.

```{r bmi-measurements}
bmi.cutoff <- 1000

bmi.data.median <- 
  bmi.data %>%
  filter(!(is.na(BMI))) %>% # remove rows where there is no BMI data
  group_by(DeID_PatientID) %>%
  summarize(BMI = median(BMI,na.rm=T),
            BMI.n = length(BMI)) %>%
  filter(BMI<bmi.cutoff) #remove BMI under cutoff
```

## Analysis of BMI

```{r bmi-analysis}
library(ggplot2)
bmi.data.median %>%
  ggplot(aes(x=BMI)) +
  geom_density() +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position = c(0.1,0.85))

bmi.data.median %>%
  ggplot(aes(x=BMI)) +
  geom_histogram() +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position = c(0.1,0.85))

bmi.data.median %>%
  summarize(mean=mean(BMI,na.rm=T),
             min=min(BMI, na.rm=T),
             max=max(BMI,na.rm=T),
             sd=sd(BMI,na.rm=T),
             n=length(BMI)) %>%
  kable(caption="Summary statistics for the BMI measurements used in this study")
```

# Validation and Checking for Outliers

We filtered out all BMI that the median was \>`r bmi.cutoff`. After this the highest BMI were

```{r BMI-outliers}
bmi.data.median %>%
  arrange(desc(BMI)) %>%
  select(BMI,BMI.n) %>%
  head(50) %>%
  kable(caption="Top 50 median BMI values and number of encounters")
```

# Output

```{r output}
output.file <- 'MedianEncounterAnthropometricsBMI.csv'
write_csv(bmi.data.median, file=output.file)
```

These data were written out to `r output.file`. This is the input file for the other scripts.

# Session Information

```{r sessioninfo}
sessionInfo()
```
