---
title: "Re-analysis of GSE160729 adipocyte snRNAseq"
author: "Dave Bridges"
date: "May 27, 2021"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)
library(tibble)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```

# Purpose

To reanalyse a single cell RNAseq study on adipocyte tissue.  

# Experimental Details

Savari et al used HFD-induced obesity and then did single nuclei isolation.

# Raw Data

Savari et al shared code and Rds friles for annotated adipocytes at https://osf.io/tsjqc/ and https://github.com/JesperGrud/snRNAseq_eWAT

```{r data-input}
input.rds.adipocytes <- 'GSE160729-eWAT_Annotated.Rds'

# code modified from https://github.com/JesperGrud/snRNAseq_eWAT/blob/main/Scripts/Main_05_Adipocytes.md
## Load libraries
library(umap)
library(Seurat) 


## Import data and subset
eWAT <- readRDS(input.rds.adipocytes)
#Amb <- readRDS("Ambient.Rds") #cant find this file on osf
Adipocytes <- subset(eWAT, subset = Annotation == "Adipocyte")
```
Loaded in the annotated Rds file from OSF, renamed as ``input.rds.adipocytes`

Followed the Seurat clustering protocol https://satijalab.org/seurat/articles/pbmc3k_tutorial.html

## Highly Variable Features in Adipocytes

```{r GSE160729-variable-features}

Adipocytes <- FindVariableFeatures(Adipocytes, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(Adipocytes), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(Adipocytes)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```


# Scaling and Dimensional Reduction

```{r GSE160729-scale-pca}
all.genes <- rownames(Adipocytes)
#Adipocytes <- ScaleData(Adipocytes, features = all.genes)
#Adipocytes <- RunPCA(Adipocytes)

#print(Adipocytes[["pca"]], dims = 1:5, nfeatures = 5)
#VizDimLoadings(Adipocytes, dims = 1:2, reduction = "pca")
DimPlot(Adipocytes, reduction = "pca")
#DimHeatmap(Adipocytes, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(Adipocytes, dims = 1:15, cells = 500, balanced = TRUE)

Adipocytes <- JackStraw(Adipocytes, num.replicate = 100)
Adipocytes <- ScoreJackStraw(Adipocytes, dims = 1:20)
JackStrawPlot(Adipocytes, dims=1:15)
ElbowPlot(Adipocytes)
#based on this i think we should use the first 13 PCs
clusters.used <- 3
```

## Cell Clustering

```{r GSE160729-clustering}

#Adipocytes <- FindNeighbors(Adipocytes, dims = 1:clusters.used)
#Adipocytes <- FindClusters(Adipocytes, resolution = 0.3)
#Adipocytes <- RunUMAP(Adipocytes, dims=1:clusters.used)
DimPlot(Adipocytes, reduction='umap')
DimPlot(Adipocytes, reduction="harmony")
DimPlot(Adipocytes, reduction="umap", split.by = 'Diet')
DimPlot(Adipocytes, reduction="umap", split.by = 'Subtype')
```

## Cluster Biomarkers

```{r GSE160729-biomarkers}


# Adipocytes.markers <- FindAllMarkers(Adipocytes, 
#                                      only.pos=T,
#                                      min.pct=0.9,
#                                      logfc.threshold = 0.5)
# 
# Adipocytes.markers %>%
#   group_by(cluster) %>%
#   top_n(n=3, wt=avg_log2FC) %>%
#   kable(caption="Top unique markers per cluster")
gc.genes <- c("Adipoq","Pnpla2","Nr3c1","Hsd11b1","Angptl4","Tsc22d3","Sgk1","Lep","Acaca","Nnat")


```

# Differential Expresion

```{r GSE160729-de-by-hfd}
Idents(Adipocytes) <- Adipocytes$Diet
Adipocyte.Diet.DE <- FindMarkers(Adipocytes, ident.1 = "LFD", ident.2 = "HFD", verbose = FALSE, min.pct=0.3)
Adipocyte.Diet.DE %>%
  arrange(-avg_log2FC) %>%
  head

library(rvest)
tfs <- read_html("http://genome.gsc.riken.jp/TFdb/cdimage/htdocs/tf_list.html") %>%
  html_table() 
tfs <- as.data.frame(tfs[[1]])

Adipocyte.Diet.DE %>%
  rownames_to_column('Gene') %>%
  filter(Gene %in% tfs$Symbol) %>% 
  arrange(avg_log2FC) %>%
  kable(caption="HFD-Differentially Expressed Transcription Factors in Adipocytes")

upregulated.tf <- 
  Adipocyte.Diet.DE %>%
  rownames_to_column('Gene') %>%
  filter(Gene %in% tfs$Symbol) %>% 
  filter(avg_log2FC<0) %>%
  arrange(-avg_log2FC) %>%
  pull(Gene)
  
library(ggplot2)
VlnPlot(Adipocytes, features=upregulated.tf, group.by="Diet", pt.size=0)
FeaturePlot(Adipocytes, features=upregulated.tf[1:2], split.by="Diet", label.size = 2)
```


# Session Information

```{r session-information, echo=T}
sessionInfo()
```
