---
title: "Data Analysis"
author: "Tiange Bu"
date: "2025-04-04"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)
```

## Purpose

The purpose of this experiment is to quantify the differences of physiological, metabolic, and psychological responses in lean and obese mice, under stressed and unstressed conditions.

## Experimental Details

\[Link to the protocol used (permalink preferred) for the experiment and include any notes relevant to your analysis. This might include specifics not in the general protocol such as cell lines, treatment doses etc.\]

High-fat diet (D12492) formula: https://researchdiets.com/formulas/d12492

## Raw Data

\[Describe your raw data files, including what the columns mean (and what units they are in).\]

```{r data-input}
library(readxl) #loads the readr package
filename <- "raw data/Data Collection Sheet.xlsx" #input file(s)

#this loads whatever the file is into a dataframe called exp.data if it exists

property <- read_excel(filename, sheet = "Property") |>
  mutate(Diet = factor(Diet, levels = c("NCD", "HFD")), #relevel diet so NCD is the reference
         Stress = factor(Stress, levels = c("CON", "CUS"))) #relevel stress so CON is the reference

liver_histology <- "raw data/histo_db_liver NAS copy.xlsx"

liver <- read_excel(liver_histology, skip = 5) |>
  rename(Code = `ID`, Steatosis_1 = `Steatosis`, Steatosis_2 = `...4`, Steatosis_3 = `...5`, Steatosis_4 = `...6`, Steatosis_5 = `...7`, Steatosis_average = `Average...8`, Inflammation_1 = `Lobular inflammation`, Inflammation_2 = `...10`, Inflammation_3 = `...11`, Inflammation_4 = `...12`, Inflammation_5 = `...13`, Inflammation_average = `Average...14`, Ballooning_1 = `Hepatocellular balooning`, Ballooning_2 = `...16`, Ballooning_3 = `...17`, Ballooning_4 = `...18`, Ballooning_5 = `...19`, Ballooning_average = `Average...20`) |>
  left_join(property)

inflammation_foci <- "raw data/Inflammation foci count.xlsx"

foci_raw <- read_excel(inflammation_foci) 

#load packages
library(lme4)
library(lmerTest)
library(knitr)
library(broom)
library(broom.mixed)
library(performance)
library(see)
library(ggpattern)
library(mediation)
library(influence.ME)
library(emmeans)
library(pbkrtest)
library(DHARMa)
library(drc)
library(fitdistrplus)
library(AER)
library(car)

```

These data can be found in `r getwd()` in a file named `r ifelse(filename %in% dir(), filename, "no file found")`. This input file was most recently updated on `r ifelse(filename %in% dir(), format(file.info(filename)$mtime, "%Y-%m-%d"), "unknown")`. This script was most recently updated on `r date()`.

## Analysis

\[Describe the analysis as you intersperse code chunks\]

### Liver Histology

```{r liver-histology}
#Steatosis
liver |>
  group_by(Diet, Stress) |>
  summarize(mean_stea = mean(`Steatosis_average`, na.rm = TRUE), 
            se_stea = sd(`Steatosis_average`, na.rm = TRUE)/sqrt(n()),
            .groups = "drop") |>

ggplot(
       aes(x = Diet,
           y = mean_stea,
           ymin = mean_stea - se_stea,
           ymax = mean_stea + se_stea,
           fill = Diet, 
           alpha = Stress)) + 
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(width = 0.2, 
                position = position_dodge(width = 0.9)) +
  labs(y = "Average NAS Score (Steatosis)",
       x = "") +
  geom_hline(yintercept = 0,
             lty = 2) +
  scale_fill_manual(values = c("grey", "black")) +
  scale_alpha_manual(values = c("CON" = 1, "CUS" = 0.8)) +
  theme_classic(base_size = 16)

#Inflammation
liver |>
  group_by(Diet, Stress) |>
  summarize(mean_infl = mean(`Inflammation_average`, na.rm = TRUE), 
            se_infl = sd(`Inflammation_average`, na.rm = TRUE)/sqrt(n()),
            .groups = "drop") |>

ggplot(
       aes(x = Diet,
           y = mean_infl,
           ymin = mean_infl - se_infl,
           ymax = mean_infl + se_infl,
           fill = Diet, 
           alpha = Stress)) + 
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(width = 0.2, 
                position = position_dodge(width = 0.9)) +
  labs(y = "Average NAS Score (Inflammation)",
       x = "") +
  geom_hline(yintercept = 0,
             lty = 2) +
  scale_fill_manual(values = c("grey", "black")) +
  scale_alpha_manual(values = c("CON" = 1, "CUS" = 0.8)) +
  theme_classic(base_size = 16)

#Ballooning
liver |>
  group_by(Diet, Stress) |>
  summarize(mean_ball = mean(`Ballooning_average`, na.rm = TRUE), 
            se_ball = sd(`Ballooning_average`, na.rm = TRUE)/sqrt(n()),
            .groups = "drop") |>

ggplot(
       aes(x = Diet,
           y = mean_ball,
           ymin = mean_ball - se_ball,
           ymax = mean_ball + se_ball,
           fill = Diet, 
           alpha = Stress)) + 
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(width = 0.2, 
                position = position_dodge(width = 0.9)) +
  labs(y = "Average NAS Score (Ballooning)",
       x = "") +
  geom_hline(yintercept = 0,
             lty = 2) +
  scale_fill_manual(values = c("grey", "black")) +
  scale_alpha_manual(values = c("CON" = 1, "CUS" = 0.8)) +
  theme_classic(base_size = 16)

#Total score
liver |>
  group_by(Diet, Stress) |>
  summarize(mean_liver = mean(`Total score`, na.rm = TRUE), 
            se_liver = sd(`Total score`, na.rm = TRUE)/sqrt(n()),
            .groups = "drop") |>

ggplot(
       aes(x = Diet,
           y = mean_liver,
           ymin = mean_liver - se_liver,
           ymax = mean_liver + se_liver,
           fill = Diet, 
           alpha = Stress)) + 
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(width = 0.2, 
                position = position_dodge(width = 0.9)) +
  labs(y = "Average NAS Score (Total)",
       x = "") +
  geom_hline(yintercept = 0,
             lty = 2) +
  scale_fill_manual(values = c("grey", "black")) +
  scale_alpha_manual(values = c("CON" = 1, "CUS" = 0.8)) +
  theme_classic(base_size = 16)

```
#### Statistics About Liver Histology
```{r liver-stats}
liver.lm.total <- lm(`Total score` ~ Diet * Stress + (1|Code), data = liver)
liver.lm.total |>
  tidy(effects = 'fixed') |>
  kable(caption = 'FINAL MODEL: Effect of Diet, Stress, and Diet:Stress Interaction on Total NAS Score (LM)')

liver.mlm.data <- liver |>
  fill(Group, .direction = c("down")) |>
  dplyr::select(-ends_with("average"), -`Total score`, -`Start Date`, -`End Date`, -`glucose`) |>
  group_by(Group, Diet, Stress) |>
  pivot_longer(`Steatosis_1`:`Ballooning_3`, 
               names_sep = "_", 
               names_to = c("Feature", "Image"), 
               values_to = "Score") |>
  pivot_wider(names_from = Feature, values_from = Score) 

liver.mlm.stea <- lmer(Steatosis ~ Diet * Stress + (1|Code), data = liver.mlm.data)
liver.mlm.stea |>
  tidy(effects = 'fixed') |>
  kable(caption = 'FINAL MODEL: Effect of Diet, Stress, and Diet:Stress Interaction on Steatosis Score')


liver.mlm.infl <- lmer(Inflammation ~ Diet * Stress + (1|Code), data = liver.mlm.data)
liver.mlm.infl |>
  tidy(effects = 'fixed') |>
  kable(caption = 'FINAL MODEL: Effect of Diet, Stress, and Diet:Stress Interaction on Inflammation Score')


liver.mlm.ball <- lmer(Ballooning ~ Diet * Stress + (1|Code), data = liver.mlm.data)
liver.mlm.ball |>
  tidy(effects = 'fixed') |>
  kable(caption = 'FINAL MODEL: Effect of Diet, Stress, and Diet:Stress Interaction on Ballooning Score')

```

#### Inflammation Foci
```{r inflammation-foci}
foci <- foci_raw |>
  separate(`Image #`, into = c("Code", "Image #"), sep = "-", convert = TRUE) |>
  left_join(property, by = "Code")

ggplot(foci, mapping = aes(x = `Inflammation Foci`)) +
         geom_histogram(binwidth = 1, fill = "grey", color = "black") +
         labs(x = "Number of Inflammation Foci",
              y = "Frequency") +
         theme_classic(base_size = 16)

foci_mlm <- lmer(`Inflammation Foci` ~ Diet * Stress + (1|Code), data = foci)
foci_mlm |>
  tidy(effects = 'fixed') |>
  kable(caption = 'Effect of Diet, Stress, and Diet:Stress Interaction on Inflammation Foci')


foci_gmlm <- glmer(`Inflammation Foci` ~ Diet * Stress + (1 | Code), data = foci, family = poisson)
fit_pois <- fitdist(foci$`Inflammation Foci`, "pois")
fit_nb <- fitdist(foci$`Inflammation Foci`, "nbinom")
fit_normal <- fitdist(foci$`Inflammation Foci`, "norm")

gofstat(list(fit_pois, fit_nb, fit_normal), 
        fitnames = c("Poisson", "Negative Binomial", "Normal"))

foci_gmlm |>
  tidy(effects = 'fixed') |>
  kable(caption = 'Effect of Diet, Stress, and Diet:Stress Interaction on Inflammation Foci (Poisson GLMM)')
# count data; number of error = mean

infl.inf <- influence(foci_gmlm, group = "Code")
plot(infl.inf, which = "cook")

foci |>
  group_by(Code, Diet, Stress) |>
  summarize(mean_animal_foci = mean(`Inflammation Foci`, na.rm = TRUE)) |>
  ungroup() |>
  group_by(Diet, Stress) |>
  summarize(mean_foci = mean(mean_animal_foci, na.rm = TRUE),
            se_foci = sd(mean_animal_foci, na.rm = TRUE)/sqrt(n()),
            .groups = "drop") |>
  ggplot(
       aes(x = Diet,
           y = mean_foci,
           ymin = mean_foci - se_foci,
           ymax = mean_foci + se_foci,
           fill = Diet,
           alpha = Stress)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(width = 0.2,
                position = position_dodge(width = 0.9)) +
  labs(y = "Average Inflammation Foci Counts",
       x = "") +
  scale_fill_manual(values = c("grey", "black")) +
  scale_alpha_manual(values = c("CON" = 1, "CUS" = 0.8)) +
  theme_classic(base_size = 15)

foci_gmlm <- glmer(`Inflammation Foci` > 0 ~ Diet * Stress + (1 | Code), data = foci, family = binomial)
foci_gmlm |>
  tidy(effects = 'fixed', exponentiate = TRUE) |>
  kable(caption = 'Effect of Diet, Stress, and Diet:Stress Interaction on Presence of Inflammation Foci (Binomial GLMM)')

```

## Session Information

```{r session-information}
sessionInfo()
```
