---
title: "Analysis of DESeq and MEME analysis of GSE236575"
author: "Dave Bridges"
date: "2025-11-20"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
  
knitr:
  opts_chunk:
    fig-path: "figures/"          # folder for all figure files
    dev: ["png", "pdf"]           # both formats for every plot
    dpi: 300                      # ← this is the correct place for resolution
    dev.args:
      png:
        type: "cairo-png"         # better anti-aliasing / font rendering (optional but recommended)
      pdf:
        family: "sans"            # font family (optional)
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

This script analyses the results from a DESeq2 and MEME analysis of GSE236575.  The purpose of this analysis is to identify differentially expressed regions and enriched motifs in that dataset.  See the README.md file in this folder for details on the generation of these files

## Raw Data

Describe your raw data files, including what the columns mean (and what units they are in).

```{r data-input}
library(readr) #loads the readr package
deseq.filename <- "results/deseq2/deseq2_results.txt" #input file(s)
deseq.counts.filename <- "results/deseq2/deseq2_normalized_counts.txt"

deseq.results <- read_tsv(deseq.filename) #reads in the data
```

These data can be found in `r getwd()` in a file named `r deseq.filename`.  This input file was most recently updated on `r  format(file.info(deseq.filename)$mtime, "%Y-%m-%d")`.  This script was most recently updated on `r date()`.

## Analysis

There were `r nrow(deseq.results)` regions analyzed in this dataset.  Of these, `r nrow(deseq.results |> filter(padj < 0.05))` (`r nrow(deseq.results |> filter(padj < 0.05))/nrow(deseq.results)*100`%) were found to be significantly differentially accessible comparing the NCD to HFD adipocytes at an FDR of 0.05.  Of those significant regions, `r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange>0))` (`r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange>0))/nrow(deseq.results |> filter(padj < 0.05))*100`%) were upregulated in HFD adipocytes and (`r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange<0))/nrow(deseq.results |> filter(padj < 0.05))*100`%) were more accessible in NCD adipocytes.

Out of the differentially regulated subset, the HFD adipocytes had an average log2 fold change of `r mean(deseq.results |> filter(padj < 0.05, log2FoldChange>0) |> pull(log2FoldChange))` +/- `r se(deseq.results |> filter(padj < 0.05, log2FoldChange>0) |> pull(log2FoldChange))`, while the NCD adipocytes had an average log2 fold change of `r mean(deseq.results |> filter(padj < 0.05, log2FoldChange<0) |> pull(log2FoldChange))` +/- `r se(deseq.results |> filter(padj < 0.05, log2FoldChange<0) |> pull(log2FoldChange))`.

> **Note on peak cutoffs.** The counts above use the volcano-narrative cutoff
> `padj < 0.05` (no log2FC magnitude threshold): `r nrow(deseq.results |> filter(padj < 0.05, log2FoldChange > 0))` HFD-up
> and `r nrow(deseq.results |> filter(padj < 0.05, log2FoldChange < 0))` CHD-up peaks. The motif-scanned
> *canonical* peak sets (`results/deseq2/HFD_specific_peaks.bed` / `CHD_specific_peaks.bed`
> and everything downstream — AME, HOMER, composite scan, full motif scan) use the
> stricter pipeline cutoff `padj < 0.05 & |log2FC| > 1`, giving
> `r nrow(deseq.results |> filter(padj < 0.05, log2FoldChange > 1))` HFD-specific and
> `r nrow(deseq.results |> filter(padj < 0.05, log2FoldChange < -1))` CHD-specific peaks.
> The composite/AME enrichment ORs reported in `SUMMARY.md` use these stricter sets as
> denominators; the volcano below classifies the full padj < 0.05 set so that peaks with
> 0 < |log2FC| ≤ 1 (which sit in the `shared` BED but are still nominally significant)
> are not silently excluded from the AP-1/GRE display.


### Annotation to known genes

```{r annotation-summary}
library(ChIPseeker)
library(GenomicRanges)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

# BED files from main.nf use UCSC-style chr-prefixed naming (chr1, chrM, ...)
# so deseq.results$chr is already "chr1" etc. — no extra prefixing needed.
hfd.peaks <- deseq.results |> filter(padj<0.05, log2FoldChange>0)
hfd.gr <- with(hfd.peaks, GRanges(seqnames=chr, ranges=IRanges(start+1, end))) # 0-based BED -> 1-based GRanges
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

hfd.annot <- annotatePeak(hfd.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db")
hfd.ann_df <- as.data.frame(hfd.annot)

ncd.peaks <- deseq.results |> filter(padj<0.05, log2FoldChange<0)
ncd.gr <- with(ncd.peaks, GRanges(seqnames=chr, ranges=IRanges(start+1, end)))

ncd.annot <- annotatePeak(ncd.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db")
ncd.ann_df <- as.data.frame(ncd.annot)

all.gr <- with(deseq.results, GRanges(seqnames=chr, ranges=IRanges(start+1, end)))
all.ann_df <- as.data.frame(annotatePeak(all.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db"))
```

These `r nrow(hfd.peaks)` chromatin regions that were differentially opened by HFD were annotated as closest to `r length(unique(hfd.ann_df$SYMBOL))` unique genes, whereas the `r nrow(ncd.peaks)` regions more accessible in NCD were annotated as closest to `r length(unique(ncd.ann_df$SYMBOL))` unique genes.  

### Volcano Plots of Regions

Each HFD-opened peak (padj < 0.05, log2FC > 0) is classified by motif content of
its 500 bp summit window: AP-1 family (35 JASPAR 2024 bZIP motifs; JUN/JUNB/JUND,
FOS/FOSL1/FOSL2, BATF/BATF3, ATF3 and all Jun·Fos heterodimers), and GRE
(NR3C1 MA0113.4 + Pgr/PGR IR3 proxies MA2323.1/MA2327.1). AP-1 calls come from
the full-motif FIMO scan for HFD/CHD-specific peaks and from
`shared_ap1_per_peak.tsv` for shared peaks; GR calls come from the composite scan.

```{r volcano-regions-motif-prep}
library(readr)
library(dplyr)
library(stringr)

ap1_ids <- c(
  "MA0099.4","MA0462.3","MA0476.2","MA0477.3","MA0478.2",
  "MA0488.2","MA0489.3","MA0490.3","MA0491.3","MA0492.2",
  "MA0605.3","MA1126.2","MA1127.1","MA1128.2","MA1129.1",
  "MA1130.2","MA1131.2","MA1132.2","MA1133.2","MA1134.2",
  "MA1135.2","MA1136.1","MA1137.2","MA1138.2","MA1139.2",
  "MA1140.3","MA1141.2","MA1142.2","MA1143.2","MA1144.2",
  "MA1145.2","MA1634.2","MA0835.3","MA1951.2","MA1988.2"
)

gr_ids <- c("MA0113.4", "MA2323.1", "MA2327.1")

ap1_from_full_scan <- function(path) {
  read_tsv(path, show_col_types = FALSE) |>
    mutate(has_ap1 = rowSums(across(any_of(ap1_ids))) > 0) |>
    dplyr::select(peak, has_ap1)
}

gr_from_composite <- function(path) {
  read_tsv(path, show_col_types = FALSE) |>
    mutate(has_gr = rowSums(across(any_of(gr_ids))) > 0) |>
    dplyr::select(peak, has_gr)
}

ap1_all <- bind_rows(
  ap1_from_full_scan("results/motif_analysis/full_motif_scan/HFD_specific/HFD_specific_per_peak_all_motifs.tsv"),
  ap1_from_full_scan("results/motif_analysis/full_motif_scan/CHD_specific/CHD_specific_per_peak_all_motifs.tsv"),
  read_tsv("results/motif_analysis/composite_scan/shared/shared_ap1_per_peak.tsv",
           show_col_types = FALSE)
)

gr_all <- bind_rows(
  gr_from_composite("results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv"),
  gr_from_composite("results/motif_analysis/composite_scan/CHD_specific/CHD_specific_per_peak_motifs.tsv"),
  gr_from_composite("results/motif_analysis/composite_scan/shared/shared_per_peak_motifs.tsv")
)
```

```{r volcano-regions}
library(ggplot2)
# Compute 500-bp peak window + motif joins BEFORE merging in all.ann_df
# (all.ann_df also has a `start` column, which would otherwise rename
# deseq.results$start to start.x and break the mid calculation).
deseq.results.annot <- deseq.results |>
  mutate(
    mid      = (start + end) %/% 2L,
    peak_500 = paste0(chr, ":", mid - 250L, "-", mid + 250L)
  ) |>
  left_join(ap1_all, by = c("peak_500" = "peak")) |>
  left_join(gr_all,  by = c("peak_500" = "peak")) |>
  left_join(all.ann_df, by = c("chr" = "seqnames", "end" = "end")) |>
  mutate(
    has_ap1 = coalesce(has_ap1, FALSE),
    has_gr  = coalesce(has_gr,  FALSE),
    motif_class = case_when(
      padj < 0.05 &  has_ap1 &  has_gr ~ "AP-1 and GRE",
      padj < 0.05 &  has_ap1 & !has_gr ~ "AP-1 only",
      padj < 0.05 & !has_ap1 &  has_gr ~ "GRE only",
      padj < 0.05                      ~ "Neither",
      TRUE                              ~ "Not significant"
    ),
    # Factor order = drawing z-order (first = bottom). guide_legend(reverse=TRUE)
    # then renders the legend top-to-bottom as: AP-1 only, GRE only,
    # AP-1 and GRE, Neither, Not significant.
    motif_class = factor(
      motif_class,
      levels = c("Not significant", "Neither", "AP-1 and GRE", "GRE only", "AP-1 only")
    )
  )

class_colors <- c(
  "AP-1 only"       = "#1f77b4",
  "GRE only"        = "#2ca02c",
  "AP-1 and GRE"    = "#d62728",
  "Neither"         = "#9467bd",
  "Not significant" = "grey75"
)

# Compose legend labels with (% increased, % decreased) within each direction.
sig_peaks <- deseq.results.annot |> filter(padj < 0.05)
n_up <- sum(sig_peaks$log2FoldChange > 0)
n_dn <- sum(sig_peaks$log2FoldChange < 0)

class_summary <- sig_peaks |>
  count(motif_class, direction = ifelse(log2FoldChange > 0, "up", "dn")) |>
  tidyr::pivot_wider(names_from = direction, values_from = n, values_fill = 0L) |>
  mutate(
    pct_up = 100 * up / n_up,
    pct_dn = 100 * dn / n_dn
  )

class_labels <- setNames(levels(deseq.results.annot$motif_class),
                         levels(deseq.results.annot$motif_class))
for (cls in class_summary$motif_class) {
  row <- class_summary |> filter(motif_class == cls)
  class_labels[as.character(cls)] <- sprintf("%s (%.1f%% ↑, %.1f%% ↓)",
                                             cls, row$pct_up, row$pct_dn)
}

ggplot(deseq.results.annot |> arrange(motif_class),
       aes(x = log2FoldChange, y = -log10(pvalue), color = motif_class)) +
  geom_point(alpha = 0.6, size = 1.1) +
  scale_color_manual(values = class_colors, labels = class_labels,
                     name = "Motif content") +
  xlab("Log2 Fold Change (HFD vs NCD)") +
  ylab("-Log10 P-value") +
  ggtitle("Differentially Accessible Regions") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  geom_vline(xintercept = c(-1, 1),    linetype = "dashed", color = "blue") +
  theme_classic(base_size = 16) +
  guides(color = guide_legend(reverse = TRUE,
                              override.aes = list(alpha = 1, size = 3))) +
  theme(legend.position=c(0.26,0.8))
```

```{r volcano-regions-motif-counts}
deseq.results.annot |>
  filter(padj < 0.05) |>
  mutate(direction = ifelse(log2FoldChange > 0, "increased (HFD)", "decreased (HFD)")) |>
  count(direction, motif_class, name = "n_peaks") |>
  group_by(direction) |>
  mutate(pct = round(100 * n_peaks / sum(n_peaks), 1)) |>
  ungroup() |>
  tidyr::pivot_wider(names_from = direction,
                     values_from = c(n_peaks, pct),
                     values_fill = 0) |>
  knitr::kable(caption = "Motif content of differentially accessible peaks by direction (padj < 0.05)")
```

### Composition of significant peaks across log2FC

Stacked histogram (left) of significant peaks (padj < 0.05) by log2FC bin,
coloured by motif class — equivalent information to the volcano but without
overplotting, so the density of AP-1 / AP-1+GRE peaks is directly comparable
between HFD-up (right) and CHD-up (left). The right panel re-normalises each
bin to 100% so the motif-class *fraction* at each fold-change is visible
independent of how many peaks fall in that bin.

```{r volcano-regions-composition, fig.width=12, fig.height=4.5}
library(ggplot2)
library(patchwork)

lfc_breaks <- seq(-6, 6, by = 0.25)
lfc_mid    <- (lfc_breaks[-1] + lfc_breaks[-length(lfc_breaks)]) / 2

comp_data <- deseq.results.annot |>
  filter(padj < 0.05, motif_class != "Not significant") |>
  mutate(motif_class = droplevels(motif_class),
         bin_idx     = cut(log2FoldChange, breaks = lfc_breaks,
                           include.lowest = TRUE, labels = FALSE),
         bin_mid     = lfc_mid[bin_idx]) |>
  filter(!is.na(bin_mid)) |>
  count(bin_mid, motif_class, name = "n_peaks") |>
  group_by(bin_mid) |>
  mutate(pct = 100 * n_peaks / sum(n_peaks)) |>
  ungroup()

p_counts <- ggplot(comp_data,
                   aes(x = bin_mid, y = n_peaks, fill = motif_class)) +
  geom_col(width = 0.25) +
  scale_fill_manual(values = class_colors, name = "Motif content",
                    breaks = c("AP-1 only", "GRE only", "AP-1 and GRE", "Neither")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  xlab("Log2 Fold Change (HFD vs NCD)") +
  ylab("Significant peaks (n)") +
  ggtitle("Peak count by log2FC bin") +
  theme_classic(base_size = 14)

p_frac <- ggplot(comp_data,
                 aes(x = bin_mid, y = pct, fill = motif_class)) +
  geom_col(width = 0.25, position = "stack") +
  scale_fill_manual(values = class_colors, guide = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  xlab("Log2 Fold Change (HFD vs NCD)") +
  ylab("% of significant peaks in bin") +
  ggtitle("Motif-class composition (100% stacked)") +
  theme_classic(base_size = 14)

p_counts + p_frac + plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

### GSEA Analysis of nearest genes

```{r enrichment-hfd}
library(clusterProfiler)

hfd.gobp <- enrichGO(gene          = bitr(unique(hfd.ann_df$SYMBOL), 
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Mm.eg.db)$ENTREZID,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(hfd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(hfd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
library(enrichplot)
emapplot(pairwise_termsim(hfd.gobp))

#transfac
library(msigdbr)
```

#### Gene Transcription Regulation Database 

```{r hfd-gtrd}
gtrd <- msigdbr(
    db_species="MM",
    species = "Mus musculus",
    subcollection="GTRD"
)


hfd.gobp.gtrd <- enricher(
    unique(hfd.ann_df$SYMBOL),                              # your ENTREZ or SYMBOL list
    TERM2GENE = gtrd[, c("gs_name", "gene_symbol")]
)
dotplot(hfd.gobp.gtrd, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(hfd.gobp.gtrd, showCategory=20) + ggtitle("GO-BP Enrichment")
#emapplot(pairwise_termsim(hfd.gobp.gtrd))

```

```{r enrichment-ncd}
ncd.gobp <- enrichGO(gene          = bitr(unique(ncd.ann_df$SYMBOL), 
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Mm.eg.db)$ENTREZID,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(ncd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(ncd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
emapplot(pairwise_termsim(ncd.gobp))
```

## Pioneer + GR Composite Motif Analysis

Tests the sensitization hypothesis from `results/motif_analysis/SUMMARY.md`:
do HFD-opened peaks contain an enriched fraction of pioneer-factor + GR composite enhancers? These would be candidate sites where HFD-driven chromatin remodeling licenses GR binding that would otherwise be inaccessible.

Inputs come from the `COMPOSITE_MOTIF_SCAN` Nextflow process, which uses FIMO to scan HFD-specific and shared peak FASTAs against JASPAR 2024 motifs for FOXA1, FOXA2, CEBPA, CEBPB, FOXO1, and NR3C1.

```{r composite-motif-load}
library(readr)
library(dplyr)
library(stringr)

hfd.motifs    <- read_tsv("results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv")
shared.motifs <- read_tsv("results/motif_analysis/composite_scan/shared/shared_per_peak_motifs.tsv")

# JASPAR ID -> friendly TF name. PGR motifs are included as GR-class proxies
# because NR3C1 (MA0113.4) is often pruned from JASPAR's non-redundant subset
# due to motif similarity with PGR/AR. PGR's IR3 motif is biochemically
# essentially identical to GR's, so we treat PGR signal as has_gr.
motif_map <- c(
  "MA0148.5" = "FOXA1",
  "MA0047.5" = "FOXA2",
  "MA0466.4" = "CEBPB",
  "MA0102.5" = "CEBPA",
  "MA0480.3" = "FOXO1",
  "MA0113.4" = "NR3C1",
  "MA2327.1" = "PGR_h",
  "MA2323.1" = "PGR_m"
)
rename_motif_cols <- function(df) {
  names(df) <- ifelse(names(df) %in% names(motif_map), motif_map[names(df)], names(df))
  df
}
hfd    <- rename_motif_cols(hfd.motifs)
shared <- rename_motif_cols(shared.motifs)

# Classify peaks by presence of pioneer/cooperator and GR motifs.
# A motif is "present" if FIMO found >= 1 occurrence within the 500-bp peak window.
classify_peaks <- function(df) {
  cols <- names(df)
  has <- function(tf) if (tf %in% cols) df[[tf]] > 0 else rep(FALSE, nrow(df))
  df$has_foxa <- has("FOXA1") | has("FOXA2")
  df$has_cebp <- has("CEBPA") | has("CEBPB")
  df$has_foxo <- has("FOXO1")
  # GR-class: NR3C1 directly, or PGR (human/mouse) as IR3 proxy
  df$has_gr   <- has("NR3C1") | has("PGR_h") | has("PGR_m")
  df$foxa_gr  <- df$has_foxa & df$has_gr
  df$cebp_gr  <- df$has_cebp & df$has_gr
  df$foxo_gr  <- df$has_foxo & df$has_gr
  df$any_pioneer    <- df$has_foxa | df$has_cebp | df$has_foxo
  df$any_pioneer_gr <- df$any_pioneer & df$has_gr
  df
}
hfd.c    <- classify_peaks(hfd)
shared.c <- classify_peaks(shared)
```

### Co-occurrence Fisher tests

Is the rate of pioneer+GR co-occurrence enriched in HFD-opened peaks vs the shared background? A significant odds ratio > 1 means HFD-opened chromatin disproportionately contains composite enhancers — the direct prediction of the pioneer-mediated sensitization model.

```{r composite-motif-fisher}
fisher_2x2 <- function(label, hfd_pos, hfd_n, sh_pos, sh_n) {
  tab <- matrix(c(hfd_pos, hfd_n - hfd_pos,
                  sh_pos,  sh_n  - sh_pos), nrow = 2)
  ft <- fisher.test(tab, alternative = "greater")
  tibble(
    comparison    = label,
    hfd_with      = hfd_pos,
    hfd_total     = hfd_n,
    hfd_pct       = round(100 * hfd_pos / hfd_n, 2),
    shared_with   = sh_pos,
    shared_total  = sh_n,
    shared_pct    = round(100 * sh_pos / sh_n, 2),
    odds_ratio    = unname(ft$estimate),
    p_value       = ft$p.value
  )
}

cooc <- bind_rows(
  fisher_2x2("FoxA + GR",  sum(hfd.c$foxa_gr), nrow(hfd.c), sum(shared.c$foxa_gr), nrow(shared.c)),
  fisher_2x2("C/EBP + GR", sum(hfd.c$cebp_gr), nrow(hfd.c), sum(shared.c$cebp_gr), nrow(shared.c)),
  fisher_2x2("FoxO + GR",  sum(hfd.c$foxo_gr), nrow(hfd.c), sum(shared.c$foxo_gr), nrow(shared.c)),
  fisher_2x2("Any pioneer + GR", sum(hfd.c$any_pioneer_gr), nrow(hfd.c), sum(shared.c$any_pioneer_gr), nrow(shared.c))
)
knitr::kable(cooc, digits = c(0,0,0,2,0,0,2,3,4))
```

### Gene annotation of candidate sensitizing enhancers

For HFD-opened peaks containing both a pioneer motif and a GR motif, annotate to the nearest TSS — these are candidate genes whose chromatin opening under HFD may license potentiated GR binding.

```{r composite-motif-annotate}
parse_peaks_to_gr <- function(peak_ids) {
  m <- str_match(peak_ids, "^(.+):(\\d+)-(\\d+)$")
  GRanges(
    seqnames = m[, 2],
    ranges = IRanges(start = as.integer(m[, 3]) + 1, end = as.integer(m[, 4]))
  )
}

annotate_composite <- function(df, field, label) {
  peaks <- df %>% filter(.data[[field]]) %>% pull(peak)
  if (length(peaks) == 0) return(tibble())
  gr <- parse_peaks_to_gr(peaks)
  ann <- annotatePeak(
    gr,
    TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,
    tssRegion = c(-2000, 500),
    annoDb = "org.Mm.eg.db",
    verbose = FALSE
  )
  as.data.frame(ann) %>%
    mutate(composite = label) %>%
    arrange(abs(distanceToTSS))
}

foxa_gr.ann <- annotate_composite(hfd.c, "foxa_gr",  "FoxA + GR")
cebp_gr.ann <- annotate_composite(hfd.c, "cebp_gr",  "C/EBP + GR")
foxo_gr.ann <- annotate_composite(hfd.c, "foxo_gr",  "FoxO + GR")
```

#### Top candidate genes — FoxA + GR composite peaks

FoxA1 is the canonical GR pioneer in steroid-responsive tissues. Genes near FoxA+GR composite enhancers in HFD-opened chromatin are the highest-priority candidates for HFD-driven GR sensitization.

```{r composite-motif-foxa-genes}
foxa_gr.top <- foxa_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  dplyr::select(SYMBOL, distanceToTSS, annotation, seqnames, start, end) %>%
  head(40)
knitr::kable(foxa_gr.top)
```

#### Top candidate genes — C/EBP + GR composite peaks

C/EBPβ is the established adipocyte GR pioneer (Madsen 2014; Siersbæk 2011).

```{r composite-motif-cebp-genes}
cebp_gr.top <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  select(SYMBOL, distanceToTSS, annotation, seqnames, start, end) %>%
  head(40)
knitr::kable(cebp_gr.top)
```

#### Pathway enrichment of FoxA + GR composite genes

```{r composite-motif-foxa-pathway}
foxa_gr.symbols <- foxa_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  pull(SYMBOL) %>%
  unique()

if (length(foxa_gr.symbols) >= 10) {
  foxa_gr.entrez <- bitr(foxa_gr.symbols,
                         fromType = "SYMBOL",
                         toType   = "ENTREZID",
                         OrgDb    = org.Mm.eg.db)$ENTREZID
  foxa_gr.gobp <- enrichGO(gene          = foxa_gr.entrez,
                           OrgDb         = org.Mm.eg.db,
                           keyType       = "ENTREZID",
                           ont           = "BP",
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.05,
                           qvalueCutoff  = 0.05,
                           readable      = TRUE)
  if (!is.null(foxa_gr.gobp) && nrow(as.data.frame(foxa_gr.gobp)) > 0) {
    dotplot(foxa_gr.gobp, showCategory = 20) + ggtitle("FoxA + GR composite genes — GO-BP")
  }
}
```

### Pathway enrichment of C/EBP + GR composite genes

GO Biological Process enrichment for the 310 nearest-TSS genes of HFD-specific C/EBP + GR composite peaks. These are the candidate sensitizing enhancers' target genes — the pathways they regulate should reflect HFD-induced glucocorticoid-responsive biology.

```{r composite-motif-cebp-pathway}
cebp_gr.symbols <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  pull(SYMBOL) %>%
  unique()

if (length(cebp_gr.symbols) >= 10) {
  cebp_gr.entrez <- bitr(cebp_gr.symbols,
                         fromType = "SYMBOL",
                         toType   = "ENTREZID",
                         OrgDb    = org.Mm.eg.db)$ENTREZID
  cebp_gr.gobp <- enrichGO(gene          = cebp_gr.entrez,
                           OrgDb         = org.Mm.eg.db,
                           keyType       = "ENTREZID",
                           ont           = "BP",
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.05,
                           qvalueCutoff  = 0.05,
                           readable      = TRUE)
  if (!is.null(cebp_gr.gobp) && nrow(as.data.frame(cebp_gr.gobp)) > 0) {
    dotplot(cebp_gr.gobp, showCategory = 20) +
      ggtitle("GO-BP: C/EBP + GR composite genes (HFD-specific)")
  } else {
    cat("No significant GO-BP terms at FDR < 0.05.\n")
  }
}
```

### Full list of C/EBP + GR composite genes

All unique nearest genes for the 310 HFD-specific C/EBP + GR composite peaks, sorted by distance to TSS. Also written to a TSV for downstream use.

```{r composite-motif-cebp-full-list}
cebp_gr_list <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  dplyr::select(SYMBOL, distanceToTSS, annotation, seqnames, start, end)

dir.create("results/motif_analysis/composite_scan", recursive = TRUE, showWarnings = FALSE)
write_tsv(cebp_gr_list, "results/motif_analysis/composite_scan/HFD_specific_cebp_gr_genes.tsv")
cat("Unique nearest genes:", nrow(cebp_gr_list), "\n")
knitr::kable(cebp_gr_list,
             caption = "All C/EBP + GR composite-peak nearest genes (HFD-specific)")
```

### CHD-specific negative control

The pioneer-mediated sensitization hypothesis predicts that C/EBP+GR composite enrichment is specific to HFD-driven chromatin remodeling. If the same enrichment also appears in CHD-specific peaks, the signal isn't HFD-induced — it's a property of any condition-specific peak. Run the same Fisher tests on the 776 CHD-specific peaks vs shared background.

```{r composite-motif-chd-control}
chd.motifs <- read_tsv("results/motif_analysis/composite_scan/CHD_specific/CHD_specific_per_peak_motifs.tsv",
                       show_col_types = FALSE)
chd    <- rename_motif_cols(chd.motifs)
chd.c  <- classify_peaks(chd)

cooc.chd <- bind_rows(
  fisher_2x2("FoxA + GR",        sum(chd.c$foxa_gr),        nrow(chd.c), sum(shared.c$foxa_gr),        nrow(shared.c)),
  fisher_2x2("C/EBP + GR",       sum(chd.c$cebp_gr),        nrow(chd.c), sum(shared.c$cebp_gr),        nrow(shared.c)),
  fisher_2x2("FoxO + GR",        sum(chd.c$foxo_gr),        nrow(chd.c), sum(shared.c$foxo_gr),        nrow(shared.c)),
  fisher_2x2("Any pioneer + GR", sum(chd.c$any_pioneer_gr), nrow(chd.c), sum(shared.c$any_pioneer_gr), nrow(shared.c))
)
knitr::kable(cooc.chd, digits = c(0,0,0,2,0,0,2,3,4),
             caption = "Composite motif co-occurrence: CHD-specific vs shared")

cooc.compare <- bind_rows(
  cooc     |> mutate(direction = "HFD-specific"),
  cooc.chd |> mutate(direction = "CHD-specific")
) |>
  dplyr::select(direction, comparison, fg_pct = hfd_pct, shared_pct, odds_ratio, p_value)
knitr::kable(cooc.compare, digits = c(0, 0, 2, 2, 3, 4),
             caption = "HFD vs CHD enrichment side by side")
```

### Sanity check (wider ±50 kb window) on key GR-pathway genes

The earlier nearest-TSS sanity check was too strict — composite enhancers often act tens of kilobases from their target gene. This wider check counts how many HFD-specific composite peaks fall within ±50 kb of each gene's transcriptional unit. Useful for asking "does this gene have a candidate sensitizing enhancer in its regulatory landscape?" rather than "is this gene the closest TSS to a composite peak?"

```{r composite-motif-sanity-wide}
library(TxDb.Mmusculus.UCSC.mm10.knownGene)

gr_pathway_genes <- c("Hsd11b1", "Nr3c1", "Ncoa1", "Ncoa2", "Fkbp5",
                      "Foxa1", "Foxa2", "Cebpb", "Cebpa", "Foxo1",
                      "Pnpla2", "Tsc22d3", "Per1", "Zbtb16", "Klf15", "Angptl4")

sym2entrez <- bitr(gr_pathway_genes,
                   fromType = "SYMBOL",
                   toType   = "ENTREZID",
                   OrgDb    = org.Mm.eg.db)

txdb       <- TxDb.Mmusculus.UCSC.mm10.knownGene
gene_gr    <- genes(txdb, single.strand.genes.only = FALSE)
gene_gr    <- gene_gr[names(gene_gr) %in% sym2entrez$ENTREZID]

# Map ENTREZID -> SYMBOL
entrez2sym <- setNames(sym2entrez$SYMBOL, sym2entrez$ENTREZID)
names(gene_gr) <- entrez2sym[names(gene_gr)]

# Extend each gene's locus by ±50 kb
gene_windows <- resize(gene_gr,
                       width = width(gene_gr) + 100000,
                       fix   = "center")

# Build per-composite peak GRanges (HFD-specific)
foxa_hfd_gr <- parse_peaks_to_gr(hfd.c |> filter(foxa_gr) |> pull(peak))
cebp_hfd_gr <- parse_peaks_to_gr(hfd.c |> filter(cebp_gr) |> pull(peak))
foxo_hfd_gr <- parse_peaks_to_gr(hfd.c |> filter(foxo_gr) |> pull(peak))

count_overlaps_safe <- function(windows, peaks) {
  if (length(peaks) == 0) return(rep(0L, length(windows)))
  as.integer(countOverlaps(windows, peaks))
}

sanity_wide <- tibble(
  gene                = names(gene_windows),
  foxa_gr_within_50kb = count_overlaps_safe(gene_windows, foxa_hfd_gr),
  cebp_gr_within_50kb = count_overlaps_safe(gene_windows, cebp_hfd_gr),
  foxo_gr_within_50kb = count_overlaps_safe(gene_windows, foxo_hfd_gr)
) |>
  arrange(desc(cebp_gr_within_50kb + foxa_gr_within_50kb + foxo_gr_within_50kb))

knitr::kable(sanity_wide,
             caption = "HFD-specific composite peaks within ±50 kb of GR-pathway gene loci")
```

### TF motif enrichment volcano

Visualization of the broader AME enrichment landscape (HFD-specific vs shared). Each point is one JASPAR motif. Highlighting the four TF families relevant to the sensitization hypothesis — FoxA, FoxO, C/EBP, and GR/PGR — shows where they sit in significance and fold enrichment relative to all other tested motifs.

```{r ame-volcano, fig.width=10, fig.height=7}
library(ggrepel)

ame <- read_tsv("results/motif_analysis/ame_results/HFD_vs_shared/ame.tsv",
                comment = "#",
                show_col_types = FALSE) |>
  filter(!is.na(rank))

ame <- ame |>
  mutate(
    log2_enrich     = log2((`%TP` + 0.5) / (`%FP` + 0.5)),
    neg_log10_padj  = -log10(pmax(`adj_p-value`, 1e-300)),
    tf_family = case_when(
      grepl("^FOXA", motif_alt_ID, ignore.case = TRUE) ~ "FoxA",
      grepl("^FOXO", motif_alt_ID, ignore.case = TRUE) ~ "FoxO",
      grepl("^CEBP", motif_alt_ID, ignore.case = TRUE) ~ "C/EBP",
      grepl("^NR3C1$|^Pgr$|^PGR$", motif_alt_ID)       ~ "GR/PGR",
      TRUE                                              ~ "Other"
    )
  )

ggplot(ame, aes(x = log2_enrich, y = neg_log10_padj)) +
  geom_point(data = filter(ame, tf_family == "Other"),
             alpha = 0.3, size = 0.8, color = "grey70") +
  geom_point(data = filter(ame, tf_family != "Other"),
             aes(color = tf_family), alpha = 0.9, size = 2.5) +
  ggrepel::geom_text_repel(
    data = filter(ame, tf_family != "Other"),
    aes(label = motif_alt_ID, color = tf_family),
    size = 3, max.overlaps = 50,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "FoxA"   = "#1f77b4",
    "FoxO"   = "#2ca02c",
    "C/EBP"  = "#d62728",
    "GR/PGR" = "#ff7f0e"
  )) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_classic(base_size = 14) +
  labs(
    x     = "log2(% foreground / % background)",
    y     = "-log10(adj. p-value)",
    title = "TF motif enrichment in HFD-opened peaks (AME, JASPAR 2024)",
    color = "TF family"
  )
```

## Data-driven sensitizer-TF discovery

The candidate-driven approach (FoxA, FoxO, C/EBP — picked from the GR-pioneer literature) failed: C/EBP+GR composite enrichment was the same in HFD-specific (OR=1.42) and CHD-specific (OR=1.43) peaks, so the signal isn't HFD-induced — it's a generic "condition-specific peak vs constitutive" signal. FoxA and FoxO weren't enriched at all.

Pivoting to a fully data-driven approach: find TFs that score on **two independent axes**.

- **Axis 1**: Motif enrichment in HFD-specific vs CHD-specific peaks (from `RUN_AME(HFD_vs_CHD)`). This identifies TFs whose binding sites distinguish HFD-driven from CHD-driven chromatin remodeling, controlling for "condition-specific peak" being enhancer-biased.
- **Axis 2**: Motif co-occurrence with GR-class motifs (NR3C1, PGR) within HFD-specific peaks. For each motif M, test whether GR-motif presence rate is higher in M-containing HFD peaks than in non-M-containing HFD peaks (Fisher exact, conditional on being HFD-specific).

A TF scoring on both axes is a candidate sensitizer pioneer chosen by the data rather than priors.

```{r data-driven-load}
library(readr)
library(dplyr)
library(stringr)
library(tidyr)

# Axis 1: HFD vs CHD AME results
ame.hvc <- read_tsv("results/motif_analysis/ame_results/HFD_vs_CHD/ame.tsv",
                    comment = "#",
                    show_col_types = FALSE) |>
  filter(!is.na(rank)) |>
  mutate(
    log2_enrich_hvc    = log2((`%TP` + 0.5) / (`%FP` + 0.5)),
    neg_log10_padj_hvc = -log10(pmax(`adj_p-value`, 1e-300)),
    motif_id           = motif_ID
  ) |>
  dplyr::select(motif_id, motif_alt_ID, log2_enrich_hvc, neg_log10_padj_hvc,
         pct_HFD = `%TP`, pct_CHD = `%FP`, adj_p_hvc = `adj_p-value`)

# Axis 2: full per-peak motif occurrence matrix in HFD-specific peaks
hfd.all <- read_tsv("results/motif_analysis/full_motif_scan/HFD_specific/HFD_specific_per_peak_all_motifs.tsv",
                    show_col_types = FALSE)
cat("HFD-specific peaks scanned:", nrow(hfd.all),
    "; motifs with >=1 hit somewhere:", ncol(hfd.all) - 1, "\n")
```

### GR co-occurrence within HFD-specific peaks (axis 2)

```{r data-driven-gr-cooccurrence}
# GR-class motif IDs (NR3C1 plus PGR human/mouse as IR3 proxies)
gr_motif_ids <- c("MA0113.4", "MA2327.1", "MA2323.1")
gr_cols      <- intersect(gr_motif_ids, names(hfd.all))
hfd.all$has_gr <- if (length(gr_cols) > 0) {
  rowSums(hfd.all[, gr_cols, drop = FALSE]) > 0
} else {
  rep(FALSE, nrow(hfd.all))
}
cat("HFD peaks with at least one GR-class motif:", sum(hfd.all$has_gr),
    "/", nrow(hfd.all),
    sprintf("(%.1f%%)\n", 100 * mean(hfd.all$has_gr)))

# Fisher test for GR enrichment
# You need the GR motif count in your background (shared/non-HFD) peaks
n_hfd      <- nrow(hfd.all)
n_hfd_gr   <- sum(hfd.all$has_gr)
n_bg       <- nrow(shared.motifs)   # shared peaks — adjust if different
n_bg_gr    <- shared.motifs |>
  mutate(has_gr = if_any(all_of(gr_motif_ids), ~ . > 0)) |>
  summarise(sum(has_gr)) |>
  pull()     # GR motif count in background peaks 

gr.hfd.mat <- matrix(
  c(n_hfd_gr,        n_hfd - n_hfd_gr,
    n_bg_gr,         n_bg - n_bg_gr),
  nrow = 2, byrow = TRUE,
  dimnames = list(c("GR+", "GR-"), c("HFD", "Background"))
)

library(knitr)
library(broom)
kable(gr.hfd.mat, caption="GR motifs vs HFD background")

hfd.gr.ft <- fisher.test(gr.hfd.mat) 

with(hfd.gr.ft,tibble(
  OR      = estimate,
  ci_low  = conf.int[1],
  ci_high = conf.int[2],
  p_value = p.value
)) |> kable(caption="Fisher test for GR enrichment in HFD")

# For each non-GR motif M: Fisher exact for GR present | M present vs M absent.
# Restricted to HFD-specific peaks (so we're testing co-occurrence within
# HFD-opened chromatin, not the marginal enrichment of M itself).
motif_cols <- setdiff(names(hfd.all), c("peak", "has_gr", gr_cols))

gr_cooc <- lapply(motif_cols, function(m) {
  m_pres <- hfd.all[[m]] > 0
  a <- sum(m_pres &  hfd.all$has_gr)   # M+ GR+
  b <- sum(m_pres & !hfd.all$has_gr)   # M+ GR-
  c <- sum(!m_pres &  hfd.all$has_gr)  # M- GR+
  d <- sum(!m_pres & !hfd.all$has_gr)  # M- GR-
  if (a + b < 30 || c + d < 30) {
    return(tibble(motif_id = m, n_with_m = a+b, n_m_and_gr = a,
                  cooc_or = NA_real_, cooc_p = NA_real_))
  }
  ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2), alternative = "greater")
  tibble(motif_id = m, n_with_m = a+b, n_m_and_gr = a,
         cooc_or = unname(ft$estimate),
         cooc_p  = ft$p.value)
}) |> bind_rows() |>
  filter(!is.na(cooc_or)) |>
  mutate(cooc_q   = p.adjust(cooc_p, method = "BH"),
         log2_or  = log2(pmax(cooc_or, 1e-3)),
         neg_log10_q = -log10(pmax(cooc_q, 1e-300)))

cat("Motifs tested for GR co-occurrence:", nrow(gr_cooc), "\n")
cat("Significant at q<0.05:", sum(gr_cooc$cooc_q < 0.05, na.rm=TRUE), "\n")
```

### Two-axis ranking — HFD enrichment × GR co-occurrence

```{r data-driven-twoaxis}
two_axis <- gr_cooc |>
  left_join(ame.hvc, by = "motif_id") |>
  filter(!is.na(log2_enrich_hvc))

# Combined score: only meaningful when BOTH axes are positive
two_axis <- two_axis |>
  mutate(
    combined_score = pmax(log2_enrich_hvc, 0) * pmax(log2_or, 0),
    in_quadrant_TR = log2_enrich_hvc > 0 & log2_or > 0  # top-right quadrant
  ) |>
  arrange(desc(combined_score))

# Strict pass: both axes individually significant. Bonferroni-strength
# correction across ~770 motifs makes this very harsh - typically only
# 0-5 motifs pass. Kept for completeness.
top_strict <- two_axis |>
  filter(in_quadrant_TR, cooc_q < 0.05, adj_p_hvc < 0.05) |>
  arrange(desc(combined_score)) |>
  dplyr::select(motif_id, motif_alt_ID,
         pct_HFD, pct_CHD, log2_enrich_hvc, adj_p_hvc,
         n_with_m, n_m_and_gr, cooc_or, cooc_q,
         combined_score) |>
  head(40)
knitr::kable(top_strict, digits = c(0,0,2,2,3,4,0,0,3,4,3),
             caption = "Strict: HFD-enriched (adj_p<0.05) AND GR-coenriched (q<0.05)")

# Relaxed pass: top-right quadrant, ranked by combined score. No strict
# q thresholds - the q-value distribution is dominated by multiple-testing
# burden across ~770 motifs. Combined-score ranking surfaces the
# biologically meaningful cloud the user can see in the scatter plot.
top_relaxed <- two_axis |>
  filter(in_quadrant_TR) |>
  arrange(desc(combined_score)) |>
  dplyr::select(motif_id, motif_alt_ID,
         pct_HFD, pct_CHD, log2_enrich_hvc, adj_p_hvc,
         n_with_m, n_m_and_gr, cooc_or, cooc_p, cooc_q,
         combined_score) |>
  head(40)
knitr::kable(top_relaxed, digits = c(0,0,2,2,3,4,0,0,3,4,4,3),
             caption = "Relaxed: top 40 in top-right quadrant by combined score (no strict q thresholds)")

# Save full ranked table
dir.create("results/motif_analysis/full_motif_scan", recursive = TRUE, showWarnings = FALSE)
write_tsv(two_axis |> arrange(desc(combined_score)),
          "results/motif_analysis/full_motif_scan/data_driven_sensitizer_ranking.tsv")
```

### TF family aggregation

Looking at family-level signal — even if no single member is significant, multiple members clustering in the top-right is meaningful.

```{r data-driven-family-summary}
# Family classification (broadened to capture more TFs in "Other")
classify_family <- function(name) {
  case_when(
    grepl("^FOXA",  name, ignore.case = TRUE) ~ "FoxA",
    grepl("^FOXO",  name, ignore.case = TRUE) ~ "FoxO",
    grepl("^FOX",   name, ignore.case = TRUE) ~ "Fox (other)",
    grepl("^CEBP",  name, ignore.case = TRUE) ~ "C/EBP",
    grepl("^NR3C1$|^Pgr$|^PGR$|^AR$|^MR$",   name)             ~ "GR/PGR/AR",
    grepl("^MEF2",  name, ignore.case = TRUE) ~ "MEF2",
    grepl("^FOS|^JUN|^BATF|^FRA[0-9]?",  name, ignore.case = TRUE) ~ "AP-1",
    grepl("^ATF[0-9]?$|^ATF$",       name, ignore.case = TRUE) ~ "ATF",
    grepl("^CREB",  name, ignore.case = TRUE) ~ "CREB",
    grepl("^DDIT3$|^CHOP$",          name, ignore.case = TRUE) ~ "DDIT3/CHOP",
    grepl("^NFE2|^NRF|^MAF",         name, ignore.case = TRUE) ~ "NRF/NFE2/MAF",
    grepl("^KLF",   name, ignore.case = TRUE) ~ "KLF",
    grepl("^SP[0-9]$",               name)             ~ "SP",
    grepl("^STAT",  name, ignore.case = TRUE) ~ "STAT",
    grepl("^IRF",   name, ignore.case = TRUE) ~ "IRF",
    grepl("^HIF",   name, ignore.case = TRUE) ~ "HIF",
    grepl("^NF.?KB|^REL",            name, ignore.case = TRUE) ~ "NF-kB",
    grepl("^RXR|^PPAR|^LXR|^NR[0-9]|^HNF4|^ESR|^ERR|^THR",  name, ignore.case = TRUE) ~ "Nuclear receptor",
    grepl("^TEAD",  name, ignore.case = TRUE) ~ "TEAD/Hippo",
    grepl("^SREBF|^SREBP",           name, ignore.case = TRUE) ~ "SREBP",
    grepl("^DMRT",  name, ignore.case = TRUE) ~ "DMRT",
    grepl("^ZNF|^Zfp|^ZFP",          name)             ~ "Zinc finger",
    grepl("^HOX|^DLX|^MSX|^NKX|^LBX|^VAX|^EVX|^PHOX|^PITX|^OTX|^LHX|^EMX|^GBX|^ALX|^ARX|^ISL|^MEIS|^PBX|^CDX|^GSX|^HMX|^GSC|^MIXL|^NOTO|^RAX|^SHOX|^TLX|^UNCX|^VSX|^BARX|^BARHL|^EN[0-9]|^ESX|^PAX|^POU|^PRRX|^SATB|^BSX|^DRGX|^ARID|^BARH",
         name, ignore.case = TRUE) ~ "Homeobox/AT-rich",
    grepl("^SOX",   name, ignore.case = TRUE) ~ "SOX",
    grepl("^GATA",  name, ignore.case = TRUE) ~ "GATA",
    grepl("^TCF|^LEF",               name, ignore.case = TRUE) ~ "TCF/LEF",
    grepl("^CTCF",  name, ignore.case = TRUE) ~ "CTCF",
    grepl("^E2F",   name, ignore.case = TRUE) ~ "E2F",
    grepl("^ETV|^ELF|^ELK|^GABP|^EHF|^FEV|^ERF|^SPDEF|^FLI1|^ERG|^ETS",
         name, ignore.case = TRUE) ~ "ETS",
    TRUE                                                ~ "Other"
  )
}

family_summary <- two_axis |>
  mutate(family = classify_family(motif_alt_ID)) |>
  filter(family != "Other") |>
  group_by(family) |>
  summarize(
    n_motifs        = n(),
    n_top_right     = sum(log2_enrich_hvc > 0 & log2_or > 0, na.rm = TRUE),
    pct_top_right   = round(100 * n_top_right / n_motifs, 1),
    median_axis1    = round(median(log2_enrich_hvc, na.rm = TRUE), 3),
    median_axis2    = round(median(log2_or,        na.rm = TRUE), 3),
    best_axis1      = round(max(log2_enrich_hvc, na.rm = TRUE),   3),
    best_axis2      = round(max(log2_or,        na.rm = TRUE),   3),
    best_combined   = round(max(combined_score, na.rm = TRUE),   3),
    .groups = "drop"
  ) |>
  arrange(desc(best_combined))

knitr::kable(family_summary, caption = "TF family aggregation: how many family members land in the top-right quadrant, and best within-family scores")
```

### Two-axis scatter plot

```{r data-driven-scatter, fig.width=11, fig.height=8}
library(ggrepel)

# Label top hits in top-right quadrant by combined score, regardless of strict
# q thresholds. Multiple-testing correction across 770 motifs is too punishing
# to drive labeling - the user can see structured signal in the scatter that
# the strict filter hides.
label_set <- two_axis |>
  filter(in_quadrant_TR) |>
  arrange(desc(combined_score)) |>
  head(30) |>
  pull(motif_id)

two_axis_plot <- two_axis |>
  mutate(
    label_flag = motif_id %in% label_set,
    family     = classify_family(motif_alt_ID)
  )

ggplot(two_axis_plot, aes(x = log2_enrich_hvc, y = log2_or)) +
  geom_point(data = filter(two_axis_plot, family == "Other"),
             alpha = 0.25, size = 0.8, color = "grey70") +
  geom_point(data = filter(two_axis_plot, family != "Other"),
             aes(color = family), alpha = 0.9, size = 2.2) +
  geom_text_repel(
    data = filter(two_axis_plot, label_flag),
    aes(label = motif_alt_ID, color = family),
    size = 3.2, max.overlaps = 50, show.legend = FALSE
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_classic(base_size = 14) +
  labs(
    x     = "Axis 1: log2(%HFD-specific / %CHD-specific)  →  HFD-enriched",
    y     = "Axis 2: log2(odds ratio for GR co-occurrence in HFD peaks)  →  GR co-enriched",
    title = "Data-driven sensitizer-TF candidates",
    subtitle = "Top-right quadrant: HFD-enriched AND co-occurring with GR motifs in HFD-opened chromatin",
    color = "TF family"
  )
```

### Pathway enrichment of the top candidates

For the top 5 data-driven sensitizer TF candidates, list HFD-specific peaks containing both the candidate motif AND a GR motif, then annotate to nearest TSS.

```{r data-driven-top-candidates-genes, eval=TRUE}
# Top 10 by combined score within the top-right quadrant. Loosened from the
# strict-pass filter (which usually returns only 0-1 motifs) so we annotate
# the structured cloud of HFD-enriched x GR-coenriched TFs.
top5 <- two_axis |>
  filter(in_quadrant_TR) |>
  arrange(desc(combined_score)) |>
  head(10) |>
  pull(motif_id)

#added candidate genes to top5
top5 <- c(top5,"MA0693.4",'MA1128.2') #added vdr and FOSL1/JUN

cat("Top 5 candidates by combined score:\n")
print(two_axis |> filter(motif_id %in% top5) |>
        dplyr::select(motif_id, motif_alt_ID, log2_enrich_hvc, log2_or, combined_score))

# For each top candidate, get the peaks containing both that motif and GR
for (mid in top5) {
  if (!mid %in% names(hfd.all)) next
  cand_peaks <- hfd.all |>
    filter(.data[[mid]] > 0 & has_gr) |>
    pull(peak)
  if (length(cand_peaks) < 10) {
    cat("\n", mid, ": only", length(cand_peaks), "composite peaks - skipping\n")
    next
  }
  cand_gr <- parse_peaks_to_gr(cand_peaks)
  cand_ann <- as.data.frame(annotatePeak(
    cand_gr,
    TxDb     = TxDb.Mmusculus.UCSC.mm10.knownGene,
    tssRegion = c(-2000, 500),
    annoDb    = "org.Mm.eg.db",
    verbose   = FALSE
  ))
  cand_genes <- cand_ann |>
    filter(!is.na(SYMBOL)) |>
    distinct(SYMBOL, .keep_all = TRUE) |>
    arrange(abs(distanceToTSS))
  alt_name <- two_axis$motif_alt_ID[two_axis$motif_id == mid][1]
  cat("\n=== ", mid, " (", alt_name, ") + GR composite peaks ===\n", sep = "")
  cat("  Composite peaks:", length(cand_peaks),
      "; unique nearest genes:", nrow(cand_genes), "\n")
  print(head(cand_genes |> dplyr::select(SYMBOL, distanceToTSS), 20))

  write_tsv(cand_genes,
            sprintf("results/motif_analysis/full_motif_scan/%s_GR_composite_genes.tsv",
                    gsub("[^A-Za-z0-9._-]", "_", paste0(alt_name, "_", mid))))
}
```


## Session Information

```{r session-information}
sessionInfo()
```
