---
title: "Analysis of DESeq and MEME analysis of GSE236575"
author: "Dave Bridges"
date: "2025-11-20"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

This script analyses the results from a DESeq2 and MEME analysis of GSE236575.  The purpose of this analysis is to identify differentially expressed regions and enriched motifs in that dataset.  See the README.md file in this folder for details on the generation of these files

## Raw Data

The DESeq2 analysis was done on the remote server (see `main.nf` for commands).  This script loads in both those results and the raw counts that were used for that analysis.  

```{r data-input}
library(readr) #loads the readr package
deseq.filename <- "results/deseq2/deseq2_results.txt" #input file(s)
deseq.counts.filename <- "results/deseq2/deseq2_normalized_counts.txt"

deseq.results <- read_tsv(deseq.filename) #reads in the data
```

These data can be found in `r getwd()` in a file named `r deseq.filename`.  This input file was most recently updated on `r  format(file.info(deseq.filename)$mtime, "%Y-%m-%d")`.  This script was most recently updated on `r date()`.

## Analysis

There were `r nrow(deseq.results)` regions analyzed in this dataset.  Of these, `r nrow(deseq.results |> filter(padj < 0.05))` (`r nrow(deseq.results |> filter(padj < 0.05))/nrow(deseq.results)*100`%) were found to be significantly differentially accessible comparing the NCD to HFD adipocytes at an FDR of 0.05.  Of those significant regions, `r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange>0))` (`r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange>0))/nrow(deseq.results |> filter(padj < 0.05))*100`%) were upregulated in HFD adipocytes and (`r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange<0))/nrow(deseq.results |> filter(padj < 0.05))*100`%) were more accessible in NCD adipocytes.

Out of the differentially regulated subset, the HFD adipocytes had an average log2 fold change of `r mean(deseq.results |> filter(padj < 0.05, log2FoldChange>0) |> pull(log2FoldChange))` +/- `r se(deseq.results |> filter(padj < 0.05, log2FoldChange>0) |> pull(log2FoldChange))`, while the NCD adipocytes had an average log2 fold change of `r mean(deseq.results |> filter(padj < 0.05, log2FoldChange<0) |> pull(log2FoldChange))` +/- `r se(deseq.results |> filter(padj < 0.05, log2FoldChange<0) |> pull(log2FoldChange))`.


### Annotation to known genes

```{r annotation-summary}
library(ChIPseeker)
library(GenomicRanges)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

# BED files from main.nf use UCSC-style chr-prefixed naming (chr1, chrM, ...)
# so deseq.results$chr is already "chr1" etc. — no extra prefixing needed.
hfd.peaks <- deseq.results |> filter(padj<0.05, log2FoldChange>0)
hfd.gr <- with(hfd.peaks, GRanges(seqnames=chr, ranges=IRanges(start+1, end))) # 0-based BED -> 1-based GRanges
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

hfd.annot <- annotatePeak(hfd.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db")
hfd.ann_df <- as.data.frame(hfd.annot)

ncd.peaks <- deseq.results |> filter(padj<0.05, log2FoldChange<0)
ncd.gr <- with(ncd.peaks, GRanges(seqnames=chr, ranges=IRanges(start+1, end)))

ncd.annot <- annotatePeak(ncd.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db")
ncd.ann_df <- as.data.frame(ncd.annot)

all.gr <- with(deseq.results, GRanges(seqnames=chr, ranges=IRanges(start+1, end)))
all.ann_df <- as.data.frame(annotatePeak(all.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db"))
```

These `r nrow(hfd.peaks)` chromatin regions that were differentially opened by HFD were annotated as closest to `r length(unique(hfd.ann_df$SYMBOL))` unique genes, whereas the `r nrow(ncd.peaks)` regions more accessible in NCD were annotated as closest to `r length(unique(ncd.ann_df$SYMBOL))` unique genes.  

### Volcano Plots of Regions

```{r volcano-regions}
library(ggplot2)
library(ggrepel)
# Join on chromosome AND end so peaks on different chromosomes can't collide.
# all.ann_df$seqnames == deseq.results$chr; all.ann_df$end == deseq.results$end.
deseq.results.annot <- deseq.results |>
  left_join(all.ann_df, by=c("chr"="seqnames", "end"="end"))

ggplot(deseq.results.annot, aes(x=log2FoldChange, y=-log10(pvalue))) +
  geom_point(alpha=0.4) +
  theme_minimal() +
  xlab("Log2 Fold Change (HFD vs NCD)") +
  ylab("-Log10 P-value") +
  ggtitle("Differentially Accessible Regions") +
  ggrepel::geom_text_repel(
    data = deseq.results.annot |>
      filter(padj < 0.05 & abs(log2FoldChange) > 0.5) |>
      arrange(padj) |> head(10),
    aes(label = SYMBOL), size = 5, max.overlaps = Inf) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color="red") +
  geom_vline(xintercept=c(-1, 1), linetype="dashed", color="blue") +
  theme_classic(base_size=16)
```

### GSEA Analysis of nearest genes

```{r enrichment-hfd}
library(clusterProfiler)

hfd.gobp <- enrichGO(gene          = bitr(unique(hfd.ann_df$SYMBOL), 
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Mm.eg.db)$ENTREZID,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(hfd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(hfd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
library(enrichplot)
emapplot(pairwise_termsim(hfd.gobp))

#transfac
library(msigdbr)
```

#### Gene Transcription Regulation Database 

```{r hfd-gtrd}
gtrd <- msigdbr(
    db_species="MM",
    species = "Mus musculus",
    subcollection="GTRD"
)


hfd.gobp.gtrd <- enricher(
    unique(hfd.ann_df$SYMBOL),                              # your ENTREZ or SYMBOL list
    TERM2GENE = gtrd[, c("gs_name", "gene_symbol")]
)
dotplot(hfd.gobp.gtrd, showCategory=20) + ggtitle("GTRD Enrichment")
barplot(hfd.gobp.gtrd, showCategory=20) + ggtitle("GTRD Enrichment")
#emapplot(pairwise_termsim(hfd.gobp.gtrd))

```

```{r enrichment-ncd}
ncd.gobp <- enrichGO(gene          = bitr(unique(ncd.ann_df$SYMBOL), 
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Mm.eg.db)$ENTREZID,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(ncd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(ncd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
emapplot(pairwise_termsim(ncd.gobp))
```

## Pioneer + GR Composite Motif Analysis

Tests the sensitization hypothesis from `results/motif_analysis/SUMMARY.md`:
do HFD-opened peaks contain an enriched fraction of pioneer-factor + GR composite enhancers? These would be candidate sites where HFD-driven chromatin remodeling licenses GR binding that would otherwise be inaccessible.

Inputs come from the `COMPOSITE_MOTIF_SCAN` Nextflow process, which uses FIMO to scan HFD-specific and shared peak FASTAs against JASPAR 2024 motifs for FOXA1, FOXA2, CEBPA, CEBPB, FOXO1, and NR3C1.

```{r composite-motif-load}
library(readr)
library(dplyr)
library(stringr)

hfd.motifs    <- read_tsv("results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv")
shared.motifs <- read_tsv("results/motif_analysis/composite_scan/shared/shared_per_peak_motifs.tsv")

# JASPAR ID -> friendly TF name. PGR motifs are included as GR-class proxies
# because NR3C1 (MA0113.4) is often pruned from JASPAR's non-redundant subset
# due to motif similarity with PGR/AR. PGR's IR3 motif is biochemically
# essentially identical to GR's, so we treat PGR signal as has_gr.
motif_map <- c(
  "MA0148.5" = "FOXA1",
  "MA0047.5" = "FOXA2",
  "MA0466.4" = "CEBPB",
  "MA0102.5" = "CEBPA",
  "MA0480.3" = "FOXO1",
  "MA0113.4" = "NR3C1",
  "MA2327.1" = "PGR_h",
  "MA2323.1" = "PGR_m"
)
rename_motif_cols <- function(df) {
  names(df) <- ifelse(names(df) %in% names(motif_map), motif_map[names(df)], names(df))
  df
}
hfd    <- rename_motif_cols(hfd.motifs)
shared <- rename_motif_cols(shared.motifs)

# Classify peaks by presence of pioneer/cooperator and GR motifs.
# A motif is "present" if FIMO found >= 1 occurrence within the 500-bp peak window.
classify_peaks <- function(df) {
  cols <- names(df)
  has <- function(tf) if (tf %in% cols) df[[tf]] > 0 else rep(FALSE, nrow(df))
  df$has_foxa <- has("FOXA1") | has("FOXA2")
  df$has_cebp <- has("CEBPA") | has("CEBPB")
  df$has_foxo <- has("FOXO1")
  # GR-class: NR3C1 directly, or PGR (human/mouse) as IR3 proxy
  df$has_gr   <- has("NR3C1") | has("PGR_h") | has("PGR_m")
  df$foxa_gr  <- df$has_foxa & df$has_gr
  df$cebp_gr  <- df$has_cebp & df$has_gr
  df$foxo_gr  <- df$has_foxo & df$has_gr
  df$any_pioneer    <- df$has_foxa | df$has_cebp | df$has_foxo
  df$any_pioneer_gr <- df$any_pioneer & df$has_gr
  df
}
hfd.c    <- classify_peaks(hfd)
shared.c <- classify_peaks(shared)
```

### Co-occurrence Fisher tests

Is the rate of pioneer+GR co-occurrence enriched in HFD-opened peaks vs the shared background? A significant odds ratio > 1 means HFD-opened chromatin disproportionately contains composite enhancers — the direct prediction of the pioneer-mediated sensitization model.

```{r composite-motif-fisher}
fisher_2x2 <- function(label, hfd_pos, hfd_n, sh_pos, sh_n) {
  tab <- matrix(c(hfd_pos, hfd_n - hfd_pos,
                  sh_pos,  sh_n  - sh_pos), nrow = 2)
  ft <- fisher.test(tab, alternative = "greater")
  tibble(
    comparison    = label,
    hfd_with      = hfd_pos,
    hfd_total     = hfd_n,
    hfd_pct       = round(100 * hfd_pos / hfd_n, 2),
    shared_with   = sh_pos,
    shared_total  = sh_n,
    shared_pct    = round(100 * sh_pos / sh_n, 2),
    odds_ratio    = unname(ft$estimate),
    p_value       = ft$p.value
  )
}

cooc <- bind_rows(
  fisher_2x2("FoxA + GR",  sum(hfd.c$foxa_gr), nrow(hfd.c), sum(shared.c$foxa_gr), nrow(shared.c)),
  fisher_2x2("C/EBP + GR", sum(hfd.c$cebp_gr), nrow(hfd.c), sum(shared.c$cebp_gr), nrow(shared.c)),
  fisher_2x2("FoxO + GR",  sum(hfd.c$foxo_gr), nrow(hfd.c), sum(shared.c$foxo_gr), nrow(shared.c)),
  fisher_2x2("Any pioneer + GR", sum(hfd.c$any_pioneer_gr), nrow(hfd.c), sum(shared.c$any_pioneer_gr), nrow(shared.c))
)
knitr::kable(cooc, digits = c(0,0,0,2,0,0,2,3,4))
```

### Gene annotation of candidate sensitizing enhancers

For HFD-opened peaks containing both a pioneer motif and a GR motif, annotate to the nearest TSS — these are candidate genes whose chromatin opening under HFD may license potentiated GR binding.

```{r composite-motif-annotate}
parse_peaks_to_gr <- function(peak_ids) {
  m <- str_match(peak_ids, "^(.+):(\\d+)-(\\d+)$")
  GRanges(
    seqnames = m[, 2],
    ranges = IRanges(start = as.integer(m[, 3]) + 1, end = as.integer(m[, 4]))
  )
}

annotate_composite <- function(df, field, label) {
  peaks <- df %>% filter(.data[[field]]) %>% pull(peak)
  if (length(peaks) == 0) return(tibble())
  gr <- parse_peaks_to_gr(peaks)
  ann <- annotatePeak(
    gr,
    TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,
    tssRegion = c(-2000, 500),
    annoDb = "org.Mm.eg.db",
    verbose = FALSE
  )
  as.data.frame(ann) %>%
    mutate(composite = label) %>%
    arrange(abs(distanceToTSS))
}

foxa_gr.ann <- annotate_composite(hfd.c, "foxa_gr",  "FoxA + GR")
cebp_gr.ann <- annotate_composite(hfd.c, "cebp_gr",  "C/EBP + GR")
foxo_gr.ann <- annotate_composite(hfd.c, "foxo_gr",  "FoxO + GR")
```

#### Top candidate genes — FoxA + GR composite peaks

FoxA1 is the canonical GR pioneer in steroid-responsive tissues. Genes near FoxA+GR composite enhancers in HFD-opened chromatin are the highest-priority candidates for HFD-driven GR sensitization.

```{r composite-motif-foxa-genes}
foxa_gr.top <- foxa_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  select(SYMBOL, distanceToTSS, annotation, seqnames, start, end) %>%
  head(40)
knitr::kable(foxa_gr.top)
```

#### Top candidate genes — C/EBP + GR composite peaks

C/EBPβ is the established adipocyte GR pioneer (Madsen 2014; Siersbæk 2011).

```{r composite-motif-cebp-genes}
cebp_gr.top <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  select(SYMBOL, distanceToTSS, annotation, seqnames, start, end) %>%
  head(40)
knitr::kable(cebp_gr.top)
```

#### Pathway enrichment of FoxA + GR composite genes

```{r composite-motif-foxa-pathway}
foxa_gr.symbols <- foxa_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  pull(SYMBOL) %>%
  unique()

if (length(foxa_gr.symbols) >= 10) {
  foxa_gr.entrez <- bitr(foxa_gr.symbols,
                         fromType = "SYMBOL",
                         toType   = "ENTREZID",
                         OrgDb    = org.Mm.eg.db)$ENTREZID
  foxa_gr.gobp <- enrichGO(gene          = foxa_gr.entrez,
                           OrgDb         = org.Mm.eg.db,
                           keyType       = "ENTREZID",
                           ont           = "BP",
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.05,
                           qvalueCutoff  = 0.05,
                           readable      = TRUE)
  if (!is.null(foxa_gr.gobp) && nrow(as.data.frame(foxa_gr.gobp)) > 0) {
    dotplot(foxa_gr.gobp, showCategory = 20) + ggtitle("FoxA + GR composite genes — GO-BP")
  }
}
```

### Sanity check on key GR-pathway genes

Quick look at whether known glucocorticoid-sensitization genes appear in our composite-peak gene lists:

```{r composite-motif-sanity}
gr_pathway_genes <- c("Hsd11b1", "Nr3c1", "Ncoa1", "Ncoa2", "Fkbp5",
                      "Foxa1", "Foxa2", "Cebpb", "Cebpa", "Foxo1",
                      "Tsc22d3", "Per1", "Zbtb16", "Klf15", "Angptl4")
all_composite_genes <- bind_rows(foxa_gr.ann, cebp_gr.ann, foxo_gr.ann) %>%
  filter(!is.na(SYMBOL))
sanity <- tibble(gene = gr_pathway_genes) %>%
  rowwise() %>%
  mutate(
    foxa_gr = gene %in% foxa_gr.ann$SYMBOL,
    cebp_gr = gene %in% cebp_gr.ann$SYMBOL,
    foxo_gr = gene %in% foxo_gr.ann$SYMBOL
  ) %>%
  ungroup()
knitr::kable(sanity)
```

## Interpretation

A brief summary of what the interpretation of these results were

## References

If needed you will need a *.bib file.  See the details at <https://quarto.org/docs/authoring/citations.html>

## Session Information

```{r session-information}
sessionInfo()
```
