---
title: "Analysis of DESeq and MEME analysis of GSE236575"
author: "Dave Bridges"
date: "2025-11-20"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

This script analyses the results from a DESeq2 and MEME analysis of GSE236575.  The purpose of this analysis is to identify differentially expressed regions and enriched motifs in that dataset.  See the README.md file in this folder for details on the generation of these files

## Raw Data

Describe your raw data files, including what the columns mean (and what units they are in).

```{r data-input}
library(readr) #loads the readr package
deseq.filename <- "results/deseq2/deseq2_results.txt" #input file(s)
deseq.counts.filename <- "results/deseq2/deseq2_normalized_counts.txt"

deseq.results <- read_tsv(deseq.filename) #reads in the data
```

These data can be found in `r getwd()` in a file named `r deseq.filename`.  This input file was most recently updated on `r  format(file.info(deseq.filename)$mtime, "%Y-%m-%d")`.  This script was most recently updated on `r date()`.

## Analysis

There were `r nrow(deseq.results)` regions analyzed in this dataset.  Of these, `r nrow(deseq.results |> filter(padj < 0.05))` (`r nrow(deseq.results |> filter(padj < 0.05))/nrow(deseq.results)*100`%) were found to be significantly differentially accessible comparing the NCD to HFD adipocytes at an FDR of 0.05.  Of those significant regions, `r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange>0))` (`r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange>0))/nrow(deseq.results |> filter(padj < 0.05))*100`%) were upregulated in HFD adipocytes and (`r nrow(deseq.results |> filter(padj < 0.05,log2FoldChange<0))/nrow(deseq.results |> filter(padj < 0.05))*100`%) were more accessible in NCD adipocytes.

Out of the differentially regulated subset, the HFD adipocytes had an average log2 fold change of `r mean(deseq.results |> filter(padj < 0.05, log2FoldChange>0) |> pull(log2FoldChange))` +/- `r se(deseq.results |> filter(padj < 0.05, log2FoldChange>0) |> pull(log2FoldChange))`, while the NCD adipocytes had an average log2 fold change of `r mean(deseq.results |> filter(padj < 0.05, log2FoldChange<0) |> pull(log2FoldChange))` +/- `r se(deseq.results |> filter(padj < 0.05, log2FoldChange<0) |> pull(log2FoldChange))`.


### Annotation to known genes

```{r annotation-summary}
library(ChIPseeker)
library(GenomicRanges)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

# BED files from main.nf use UCSC-style chr-prefixed naming (chr1, chrM, ...)
# so deseq.results$chr is already "chr1" etc. — no extra prefixing needed.
hfd.peaks <- deseq.results |> filter(padj<0.05, log2FoldChange>0)
hfd.gr <- with(hfd.peaks, GRanges(seqnames=chr, ranges=IRanges(start+1, end))) # 0-based BED -> 1-based GRanges
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

hfd.annot <- annotatePeak(hfd.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db")
hfd.ann_df <- as.data.frame(hfd.annot)

ncd.peaks <- deseq.results |> filter(padj<0.05, log2FoldChange<0)
ncd.gr <- with(ncd.peaks, GRanges(seqnames=chr, ranges=IRanges(start+1, end)))

ncd.annot <- annotatePeak(ncd.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db")
ncd.ann_df <- as.data.frame(ncd.annot)

all.gr <- with(deseq.results, GRanges(seqnames=chr, ranges=IRanges(start+1, end)))
all.ann_df <- as.data.frame(annotatePeak(all.gr, TxDb=txdb, tssRegion=c(-2000, 500), annoDb="org.Mm.eg.db"))
```

These `r nrow(hfd.peaks)` chromatin regions that were differentially opened by HFD were annotated as closest to `r length(unique(hfd.ann_df$SYMBOL))` unique genes, whereas the `r nrow(ncd.peaks)` regions more accessible in NCD were annotated as closest to `r length(unique(ncd.ann_df$SYMBOL))` unique genes.  

### Volcano Plots of Regions

```{r volcano-regions}
library(ggplot2)
library(ggrepel)
# Join on chromosome AND end so peaks on different chromosomes can't collide.
# all.ann_df$seqnames == deseq.results$chr; all.ann_df$end == deseq.results$end.
deseq.results.annot <- deseq.results |>
  left_join(all.ann_df, by=c("chr"="seqnames", "end"="end"))

ggplot(deseq.results.annot, aes(x=log2FoldChange, y=-log10(pvalue))) +
  geom_point(alpha=0.4) +
  theme_minimal() +
  xlab("Log2 Fold Change (HFD vs NCD)") +
  ylab("-Log10 P-value") +
  ggtitle("Differentially Accessible Regions") +
  ggrepel::geom_text_repel(
    data = deseq.results.annot |>
      filter(padj < 0.05 & abs(log2FoldChange) > 0.5) |>
      arrange(padj) |> head(10),
    aes(label = SYMBOL), size = 5, max.overlaps = Inf) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color="red") +
  geom_vline(xintercept=c(-1, 1), linetype="dashed", color="blue") +
  theme_classic(base_size=16)
```

### GSEA Analysis of nearest genes

```{r enrichment-hfd}
library(clusterProfiler)

hfd.gobp <- enrichGO(gene          = bitr(unique(hfd.ann_df$SYMBOL), 
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Mm.eg.db)$ENTREZID,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(hfd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(hfd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
library(enrichplot)
emapplot(pairwise_termsim(hfd.gobp))

#transfac
library(msigdbr)
```

#### Gene Transcription Regulation Database 

```{r hfd-gtrd}
gtrd <- msigdbr(
    db_species="MM",
    species = "Mus musculus",
    subcollection="GTRD"
)


hfd.gobp.gtrd <- enricher(
    unique(hfd.ann_df$SYMBOL),                              # your ENTREZ or SYMBOL list
    TERM2GENE = gtrd[, c("gs_name", "gene_symbol")]
)
dotplot(hfd.gobp.gtrd, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(hfd.gobp.gtrd, showCategory=20) + ggtitle("GO-BP Enrichment")
#emapplot(pairwise_termsim(hfd.gobp.gtrd))

```

```{r enrichment-ncd}
ncd.gobp <- enrichGO(gene          = bitr(unique(ncd.ann_df$SYMBOL), 
                 fromType = "SYMBOL",
                 toType = "ENTREZID",
                 OrgDb = org.Mm.eg.db)$ENTREZID,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(ncd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
barplot(ncd.gobp, showCategory=20) + ggtitle("GO-BP Enrichment")
emapplot(pairwise_termsim(ncd.gobp))
```

## Pioneer + GR Composite Motif Analysis

Tests the sensitization hypothesis from `results/motif_analysis/SUMMARY.md`:
do HFD-opened peaks contain an enriched fraction of pioneer-factor + GR composite enhancers? These would be candidate sites where HFD-driven chromatin remodeling licenses GR binding that would otherwise be inaccessible.

Inputs come from the `COMPOSITE_MOTIF_SCAN` Nextflow process, which uses FIMO to scan HFD-specific and shared peak FASTAs against JASPAR 2024 motifs for FOXA1, FOXA2, CEBPA, CEBPB, FOXO1, and NR3C1.

```{r composite-motif-load}
library(readr)
library(dplyr)
library(stringr)

hfd.motifs    <- read_tsv("results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv")
shared.motifs <- read_tsv("results/motif_analysis/composite_scan/shared/shared_per_peak_motifs.tsv")

# JASPAR ID -> friendly TF name. PGR motifs are included as GR-class proxies
# because NR3C1 (MA0113.4) is often pruned from JASPAR's non-redundant subset
# due to motif similarity with PGR/AR. PGR's IR3 motif is biochemically
# essentially identical to GR's, so we treat PGR signal as has_gr.
motif_map <- c(
  "MA0148.5" = "FOXA1",
  "MA0047.5" = "FOXA2",
  "MA0466.4" = "CEBPB",
  "MA0102.5" = "CEBPA",
  "MA0480.3" = "FOXO1",
  "MA0113.4" = "NR3C1",
  "MA2327.1" = "PGR_h",
  "MA2323.1" = "PGR_m"
)
rename_motif_cols <- function(df) {
  names(df) <- ifelse(names(df) %in% names(motif_map), motif_map[names(df)], names(df))
  df
}
hfd    <- rename_motif_cols(hfd.motifs)
shared <- rename_motif_cols(shared.motifs)

# Classify peaks by presence of pioneer/cooperator and GR motifs.
# A motif is "present" if FIMO found >= 1 occurrence within the 500-bp peak window.
classify_peaks <- function(df) {
  cols <- names(df)
  has <- function(tf) if (tf %in% cols) df[[tf]] > 0 else rep(FALSE, nrow(df))
  df$has_foxa <- has("FOXA1") | has("FOXA2")
  df$has_cebp <- has("CEBPA") | has("CEBPB")
  df$has_foxo <- has("FOXO1")
  # GR-class: NR3C1 directly, or PGR (human/mouse) as IR3 proxy
  df$has_gr   <- has("NR3C1") | has("PGR_h") | has("PGR_m")
  df$foxa_gr  <- df$has_foxa & df$has_gr
  df$cebp_gr  <- df$has_cebp & df$has_gr
  df$foxo_gr  <- df$has_foxo & df$has_gr
  df$any_pioneer    <- df$has_foxa | df$has_cebp | df$has_foxo
  df$any_pioneer_gr <- df$any_pioneer & df$has_gr
  df
}
hfd.c    <- classify_peaks(hfd)
shared.c <- classify_peaks(shared)
```

### Co-occurrence Fisher tests

Is the rate of pioneer+GR co-occurrence enriched in HFD-opened peaks vs the shared background? A significant odds ratio > 1 means HFD-opened chromatin disproportionately contains composite enhancers — the direct prediction of the pioneer-mediated sensitization model.

```{r composite-motif-fisher}
fisher_2x2 <- function(label, hfd_pos, hfd_n, sh_pos, sh_n) {
  tab <- matrix(c(hfd_pos, hfd_n - hfd_pos,
                  sh_pos,  sh_n  - sh_pos), nrow = 2)
  ft <- fisher.test(tab, alternative = "greater")
  tibble(
    comparison    = label,
    hfd_with      = hfd_pos,
    hfd_total     = hfd_n,
    hfd_pct       = round(100 * hfd_pos / hfd_n, 2),
    shared_with   = sh_pos,
    shared_total  = sh_n,
    shared_pct    = round(100 * sh_pos / sh_n, 2),
    odds_ratio    = unname(ft$estimate),
    p_value       = ft$p.value
  )
}

cooc <- bind_rows(
  fisher_2x2("FoxA + GR",  sum(hfd.c$foxa_gr), nrow(hfd.c), sum(shared.c$foxa_gr), nrow(shared.c)),
  fisher_2x2("C/EBP + GR", sum(hfd.c$cebp_gr), nrow(hfd.c), sum(shared.c$cebp_gr), nrow(shared.c)),
  fisher_2x2("FoxO + GR",  sum(hfd.c$foxo_gr), nrow(hfd.c), sum(shared.c$foxo_gr), nrow(shared.c)),
  fisher_2x2("Any pioneer + GR", sum(hfd.c$any_pioneer_gr), nrow(hfd.c), sum(shared.c$any_pioneer_gr), nrow(shared.c))
)
knitr::kable(cooc, digits = c(0,0,0,2,0,0,2,3,4))
```

### Gene annotation of candidate sensitizing enhancers

For HFD-opened peaks containing both a pioneer motif and a GR motif, annotate to the nearest TSS — these are candidate genes whose chromatin opening under HFD may license potentiated GR binding.

```{r composite-motif-annotate}
parse_peaks_to_gr <- function(peak_ids) {
  m <- str_match(peak_ids, "^(.+):(\\d+)-(\\d+)$")
  GRanges(
    seqnames = m[, 2],
    ranges = IRanges(start = as.integer(m[, 3]) + 1, end = as.integer(m[, 4]))
  )
}

annotate_composite <- function(df, field, label) {
  peaks <- df %>% filter(.data[[field]]) %>% pull(peak)
  if (length(peaks) == 0) return(tibble())
  gr <- parse_peaks_to_gr(peaks)
  ann <- annotatePeak(
    gr,
    TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,
    tssRegion = c(-2000, 500),
    annoDb = "org.Mm.eg.db",
    verbose = FALSE
  )
  as.data.frame(ann) %>%
    mutate(composite = label) %>%
    arrange(abs(distanceToTSS))
}

foxa_gr.ann <- annotate_composite(hfd.c, "foxa_gr",  "FoxA + GR")
cebp_gr.ann <- annotate_composite(hfd.c, "cebp_gr",  "C/EBP + GR")
foxo_gr.ann <- annotate_composite(hfd.c, "foxo_gr",  "FoxO + GR")
```

#### Top candidate genes — FoxA + GR composite peaks

FoxA1 is the canonical GR pioneer in steroid-responsive tissues. Genes near FoxA+GR composite enhancers in HFD-opened chromatin are the highest-priority candidates for HFD-driven GR sensitization.

```{r composite-motif-foxa-genes}
foxa_gr.top <- foxa_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  select(SYMBOL, distanceToTSS, annotation, seqnames, start, end) %>%
  head(40)
knitr::kable(foxa_gr.top)
```

#### Top candidate genes — C/EBP + GR composite peaks

C/EBPβ is the established adipocyte GR pioneer (Madsen 2014; Siersbæk 2011).

```{r composite-motif-cebp-genes}
cebp_gr.top <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  select(SYMBOL, distanceToTSS, annotation, seqnames, start, end) %>%
  head(40)
knitr::kable(cebp_gr.top)
```

#### Pathway enrichment of FoxA + GR composite genes

```{r composite-motif-foxa-pathway}
foxa_gr.symbols <- foxa_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  pull(SYMBOL) %>%
  unique()

if (length(foxa_gr.symbols) >= 10) {
  foxa_gr.entrez <- bitr(foxa_gr.symbols,
                         fromType = "SYMBOL",
                         toType   = "ENTREZID",
                         OrgDb    = org.Mm.eg.db)$ENTREZID
  foxa_gr.gobp <- enrichGO(gene          = foxa_gr.entrez,
                           OrgDb         = org.Mm.eg.db,
                           keyType       = "ENTREZID",
                           ont           = "BP",
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.05,
                           qvalueCutoff  = 0.05,
                           readable      = TRUE)
  if (!is.null(foxa_gr.gobp) && nrow(as.data.frame(foxa_gr.gobp)) > 0) {
    dotplot(foxa_gr.gobp, showCategory = 20) + ggtitle("FoxA + GR composite genes — GO-BP")
  }
}
```

### Pathway enrichment of C/EBP + GR composite genes

GO Biological Process enrichment for the 310 nearest-TSS genes of HFD-specific C/EBP + GR composite peaks. These are the candidate sensitizing enhancers' target genes — the pathways they regulate should reflect HFD-induced glucocorticoid-responsive biology.

```{r composite-motif-cebp-pathway}
cebp_gr.symbols <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  pull(SYMBOL) %>%
  unique()

if (length(cebp_gr.symbols) >= 10) {
  cebp_gr.entrez <- bitr(cebp_gr.symbols,
                         fromType = "SYMBOL",
                         toType   = "ENTREZID",
                         OrgDb    = org.Mm.eg.db)$ENTREZID
  cebp_gr.gobp <- enrichGO(gene          = cebp_gr.entrez,
                           OrgDb         = org.Mm.eg.db,
                           keyType       = "ENTREZID",
                           ont           = "BP",
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.05,
                           qvalueCutoff  = 0.05,
                           readable      = TRUE)
  if (!is.null(cebp_gr.gobp) && nrow(as.data.frame(cebp_gr.gobp)) > 0) {
    dotplot(cebp_gr.gobp, showCategory = 20) +
      ggtitle("GO-BP: C/EBP + GR composite genes (HFD-specific)")
  } else {
    cat("No significant GO-BP terms at FDR < 0.05.\n")
  }
}
```

### Full list of C/EBP + GR composite genes

All unique nearest genes for the 310 HFD-specific C/EBP + GR composite peaks, sorted by distance to TSS. Also written to a TSV for downstream use.

```{r composite-motif-cebp-full-list}
cebp_gr_list <- cebp_gr.ann %>%
  filter(!is.na(SYMBOL)) %>%
  distinct(SYMBOL, .keep_all = TRUE) %>%
  arrange(abs(distanceToTSS)) %>%
  select(SYMBOL, distanceToTSS, annotation, seqnames, start, end)

dir.create("results/motif_analysis/composite_scan", recursive = TRUE, showWarnings = FALSE)
write_tsv(cebp_gr_list, "results/motif_analysis/composite_scan/HFD_specific_cebp_gr_genes.tsv")
cat("Unique nearest genes:", nrow(cebp_gr_list), "\n")
knitr::kable(cebp_gr_list,
             caption = "All C/EBP + GR composite-peak nearest genes (HFD-specific)")
```

### CHD-specific negative control

The pioneer-mediated sensitization hypothesis predicts that C/EBP+GR composite enrichment is specific to HFD-driven chromatin remodeling. If the same enrichment also appears in CHD-specific peaks, the signal isn't HFD-induced — it's a property of any condition-specific peak. Run the same Fisher tests on the 776 CHD-specific peaks vs shared background.

```{r composite-motif-chd-control}
chd.motifs <- read_tsv("results/motif_analysis/composite_scan/CHD_specific/CHD_specific_per_peak_motifs.tsv",
                       show_col_types = FALSE)
chd    <- rename_motif_cols(chd.motifs)
chd.c  <- classify_peaks(chd)

cooc.chd <- bind_rows(
  fisher_2x2("FoxA + GR",        sum(chd.c$foxa_gr),        nrow(chd.c), sum(shared.c$foxa_gr),        nrow(shared.c)),
  fisher_2x2("C/EBP + GR",       sum(chd.c$cebp_gr),        nrow(chd.c), sum(shared.c$cebp_gr),        nrow(shared.c)),
  fisher_2x2("FoxO + GR",        sum(chd.c$foxo_gr),        nrow(chd.c), sum(shared.c$foxo_gr),        nrow(shared.c)),
  fisher_2x2("Any pioneer + GR", sum(chd.c$any_pioneer_gr), nrow(chd.c), sum(shared.c$any_pioneer_gr), nrow(shared.c))
)
knitr::kable(cooc.chd, digits = c(0,0,0,2,0,0,2,3,4),
             caption = "Composite motif co-occurrence: CHD-specific vs shared")

cooc.compare <- bind_rows(
  cooc     |> mutate(direction = "HFD-specific"),
  cooc.chd |> mutate(direction = "CHD-specific")
) |>
  select(direction, comparison, fg_pct = hfd_pct, shared_pct, odds_ratio, p_value)
knitr::kable(cooc.compare, digits = c(0, 0, 2, 2, 3, 4),
             caption = "HFD vs CHD enrichment side by side")
```

### Sanity check (wider ±50 kb window) on key GR-pathway genes

The earlier nearest-TSS sanity check was too strict — composite enhancers often act tens of kilobases from their target gene. This wider check counts how many HFD-specific composite peaks fall within ±50 kb of each gene's transcriptional unit. Useful for asking "does this gene have a candidate sensitizing enhancer in its regulatory landscape?" rather than "is this gene the closest TSS to a composite peak?"

```{r composite-motif-sanity-wide}
library(TxDb.Mmusculus.UCSC.mm10.knownGene)

gr_pathway_genes <- c("Hsd11b1", "Nr3c1", "Ncoa1", "Ncoa2", "Fkbp5",
                      "Foxa1", "Foxa2", "Cebpb", "Cebpa", "Foxo1",
                      "Pnpla2", "Tsc22d3", "Per1", "Zbtb16", "Klf15", "Angptl4")

sym2entrez <- bitr(gr_pathway_genes,
                   fromType = "SYMBOL",
                   toType   = "ENTREZID",
                   OrgDb    = org.Mm.eg.db)

txdb       <- TxDb.Mmusculus.UCSC.mm10.knownGene
gene_gr    <- genes(txdb, single.strand.genes.only = FALSE)
gene_gr    <- gene_gr[names(gene_gr) %in% sym2entrez$ENTREZID]

# Map ENTREZID -> SYMBOL
entrez2sym <- setNames(sym2entrez$SYMBOL, sym2entrez$ENTREZID)
names(gene_gr) <- entrez2sym[names(gene_gr)]

# Extend each gene's locus by ±50 kb
gene_windows <- resize(gene_gr,
                       width = width(gene_gr) + 100000,
                       fix   = "center")

# Build per-composite peak GRanges (HFD-specific)
foxa_hfd_gr <- parse_peaks_to_gr(hfd.c |> filter(foxa_gr) |> pull(peak))
cebp_hfd_gr <- parse_peaks_to_gr(hfd.c |> filter(cebp_gr) |> pull(peak))
foxo_hfd_gr <- parse_peaks_to_gr(hfd.c |> filter(foxo_gr) |> pull(peak))

count_overlaps_safe <- function(windows, peaks) {
  if (length(peaks) == 0) return(rep(0L, length(windows)))
  as.integer(countOverlaps(windows, peaks))
}

sanity_wide <- tibble(
  gene                = names(gene_windows),
  foxa_gr_within_50kb = count_overlaps_safe(gene_windows, foxa_hfd_gr),
  cebp_gr_within_50kb = count_overlaps_safe(gene_windows, cebp_hfd_gr),
  foxo_gr_within_50kb = count_overlaps_safe(gene_windows, foxo_hfd_gr)
) |>
  arrange(desc(cebp_gr_within_50kb + foxa_gr_within_50kb + foxo_gr_within_50kb))

knitr::kable(sanity_wide,
             caption = "HFD-specific composite peaks within ±50 kb of GR-pathway gene loci")
```

### TF motif enrichment volcano

Visualization of the broader AME enrichment landscape (HFD-specific vs shared). Each point is one JASPAR motif. Highlighting the four TF families relevant to the sensitization hypothesis — FoxA, FoxO, C/EBP, and GR/PGR — shows where they sit in significance and fold enrichment relative to all other tested motifs.

```{r ame-volcano, fig.width=10, fig.height=7}
library(ggrepel)

ame <- read_tsv("results/motif_analysis/ame_results/HFD_vs_shared/ame.tsv",
                comment = "#",
                show_col_types = FALSE) |>
  filter(!is.na(rank))

ame <- ame |>
  mutate(
    log2_enrich     = log2((`%TP` + 0.5) / (`%FP` + 0.5)),
    neg_log10_padj  = -log10(pmax(`adj_p-value`, 1e-300)),
    tf_family = case_when(
      grepl("^FOXA", motif_alt_ID, ignore.case = TRUE) ~ "FoxA",
      grepl("^FOXO", motif_alt_ID, ignore.case = TRUE) ~ "FoxO",
      grepl("^CEBP", motif_alt_ID, ignore.case = TRUE) ~ "C/EBP",
      grepl("^NR3C1$|^Pgr$|^PGR$", motif_alt_ID)       ~ "GR/PGR",
      TRUE                                              ~ "Other"
    )
  )

ggplot(ame, aes(x = log2_enrich, y = neg_log10_padj)) +
  geom_point(data = filter(ame, tf_family == "Other"),
             alpha = 0.3, size = 0.8, color = "grey70") +
  geom_point(data = filter(ame, tf_family != "Other"),
             aes(color = tf_family), alpha = 0.9, size = 2.5) +
  ggrepel::geom_text_repel(
    data = filter(ame, tf_family != "Other"),
    aes(label = motif_alt_ID, color = tf_family),
    size = 3, max.overlaps = 50,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "FoxA"   = "#1f77b4",
    "FoxO"   = "#2ca02c",
    "C/EBP"  = "#d62728",
    "GR/PGR" = "#ff7f0e"
  )) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_classic(base_size = 14) +
  labs(
    x     = "log2(% foreground / % background)",
    y     = "-log10(adj. p-value)",
    title = "TF motif enrichment in HFD-opened peaks (AME, JASPAR 2024)",
    color = "TF family"
  )
```

## Interpretation

A brief summary of what the interpretation of these results were

## References

If needed you will need a *.bib file.  See the details at <https://quarto.org/docs/authoring/citations.html>

## Session Information

```{r session-information}
sessionInfo()
```
