---
title: "GSE236575 RNA-seq × ATAC-seq Integration"
subtitle: "TRAP-seq HFD vs CHD — targeted gene analysis, ATAC peak annotation, motif stratification"
author: "Dave Bridges"
date: "2026-05-12"
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show code"
    fig-path: "figures/rnaseq/"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    fig-path: "figures/rnaseq/"
    dev: ["png", "pdf"]
    dpi: 300
    dev.args:
      png:
        type: "cairo-png"
      pdf:
        family: "sans"
---

```{r setup}
#| echo: false
se <- function(x) sd(x, na.rm=TRUE) / sqrt(sum(!is.na(x)))
library(tidyverse)
color_scheme <- c(CHD="#00274c", HFD="#ffcb05")
```

## Dataset overview

### RNA-seq data source

The matched RNA-seq data are from **GSE236578 [TRAP]**, a sub-series of the
super-series GSE236580 ("Adipose tissue retains an epigenetic memory of obesity
after weight loss"; PMID 39558077).  The same AdipER-Cre/NuTRAP mice and eWAT
adipocyte preparations were used for both ATAC-seq (GSE236575) and TRAP-seq
(GSE236578).  TRAP-seq captures actively translated transcripts specifically
from adipocytes, avoiding contamination from stromal-vascular cells.

| Sample ID | GSM | SRR | Condition |
|-----------|-----|-----|-----------|
| CHD_1 | GSM7558266 | SRR25152278 | Chow diet |
| CHD_2 | GSM7558271 | SRR25152274 | Chow diet |
| CHD_3 | GSM7558276 | SRR25152266 | Chow diet |
| HFD_1 | GSM7558267 | SRR25152277 | High-fat diet |
| HFD_2 | GSM7558272 | SRR25152273 | High-fat diet |
| HFD_3 | GSM7558277 | SRR25152265 | High-fat diet |

Library: paired-end 150 bp, template switching (Maxima H Minus RT), unstranded.
Alignment: STAR → mm10/GENCODE vM25, unique mappers only.
Quantification: featureCounts, gene_name, unstranded (-s 0).

### Pipeline output files

All files are generated by the `RNASEQ_DESEQ2` Nextflow process.

```{r file-paths}
deseq_file    <- "results/rnaseq/deseq2/rnaseq_deseq2_results.txt"
targeted_file <- "results/rnaseq/deseq2/rnaseq_targeted_genes.tsv"
norm_file     <- "results/rnaseq/deseq2/rnaseq_normalized_counts.txt"
atac_deseq    <- "results/deseq2/deseq2_results.txt"
```

---

## 1. DESeq2 differential expression — global summary

```{r deseq-load}
library(readr)
library(dplyr)

deseq <- read_tsv(deseq_file, show_col_types=FALSE) |>
  rownames_to_column("gene") |>
  rename(gene = 1)

cat("Total genes tested:", nrow(deseq), "\n")
cat("HFD-up   (padj<0.05, LFC>0.585):",
    sum(!is.na(deseq$padj) & deseq$padj < 0.05 & deseq$log2FoldChange > 0.585), "\n")
cat("HFD-down (padj<0.05, LFC< -0.585):",
    sum(!is.na(deseq$padj) & deseq$padj < 0.05 & deseq$log2FoldChange < -0.585), "\n")
```

```{r volcano-global, fig.width=9, fig.height=6}
library(ggrepel)
library(ggplot2)

tg_genes <- c("Fos","Jun","Junb","Jund","Fosl1","Fosl2","Batf","Batf3","Atf3","Atf4","Ddit3",
              "Vdr","Ncoa1","Ncoa2","Ncoa3","Med1",
              "Hsd11b1","Nr3c1","Fkbp5","Sgk1","Tsc22d3")

deseq_plot <- deseq |>
  mutate(
    sig = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 0.585,
    label = ifelse(gene %in% tg_genes, gene, NA_character_)
  )

ggplot(deseq_plot, aes(log2FoldChange, -log10(pvalue))) +
  geom_point(aes(color=sig), alpha=0.4, size=0.7) +
  scale_color_manual(values=c("FALSE"="grey70","TRUE"="#d62728"), guide="none") +
  geom_text_repel(aes(label=label), size=3, max.overlaps=40,
                  segment.color="grey60") +
  geom_hline(yintercept=-log10(0.05), lty=2, color="grey50") +
  geom_vline(xintercept=c(-0.585, 0.585), lty=2, color="grey50") +
  labs(x="log2FC (HFD / CHD)", y="-log10(p-value)",
       title="TRAP-seq: HFD vs CHD differential expression",
       subtitle="Red = FDR<0.05, |LFC|>0.585; labelled genes = targeted candidates") +
  theme_classic(base_size=13)
```

---

## 2. Targeted gene analysis

### Load and display results

```{r targeted-load}
tg <- read_tsv(targeted_file, show_col_types=FALSE)

# Format for display
tg_display <- tg |>
  mutate(
    across(c(log2FoldChange, lfcSE, bayes_CrI_lo, bayes_CrI_hi), ~round(., 3)),
    across(c(pvalue, padj), ~signif(., 3)),
    across(c(bayes_P_HFD_gt_CHD, bayes_ER), ~round(., 3)),
    across(c(mean_norm_CHD, mean_norm_HFD), ~round(., 1))
  ) |>
  select(category, gene, log2FoldChange, lfcSE, padj, pvalue,
         mean_norm_CHD, mean_norm_HFD, detected,
         bayes_P_HFD_gt_CHD, bayes_ER, bayes_CrI_lo, bayes_CrI_hi)

knitr::kable(tg_display,
  caption="Targeted gene-level TRAP-seq results (HFD vs CHD). Bayes stats use Normal(0,1) prior on log2FC (analytical conjugate update).",
  digits=c(0,0,3,3,3,3,1,1,0,3,2,3,3))
```

### AP-1 candidates — key comparison

```{r ap1-table}
ap1 <- tg |> filter(category == "AP-1") |>
  arrange(desc(log2FoldChange))

knitr::kable(ap1 |> select(gene, log2FoldChange, padj, pvalue,
                            mean_norm_CHD, mean_norm_HFD,
                            bayes_P_HFD_gt_CHD, bayes_ER),
  caption="AP-1 family members — ranked by HFD log2FC",
  digits=c(0,3,3,3,1,1,3,2))
```

### AP-1 forest plot

```{r ap1-forest, fig.width=7, fig.height=5}
ap1_ok <- ap1 |> filter(!is.na(log2FoldChange))

ggplot(ap1_ok, aes(x=reorder(gene, log2FoldChange), y=log2FoldChange)) +
  geom_col(aes(fill=log2FoldChange > 0), width=0.6) +
  geom_errorbar(aes(ymin=bayes_CrI_lo, ymax=bayes_CrI_hi), width=0.25) +
  geom_hline(yintercept=0, lty=2) +
  coord_flip() +
  scale_fill_manual(values=c("TRUE"="#d62728","FALSE"="#1f77b4"),
                    labels=c("TRUE"="HFD up","FALSE"="CHD up"), name=NULL) +
  labs(x=NULL, y="log2FC (HFD / CHD), 95% CrI",
       title="AP-1 family — TRAP-seq HFD vs CHD",
       subtitle="Bars = posterior mean; error bars = 95% CrI (Normal(0,1) prior)") +
  theme_classic(base_size=12)
```

### VDR pathway and GR coactivators

```{r vdr-table}
vdr <- tg |> filter(category == "VDR_pathway") |> arrange(gene)

knitr::kable(vdr |> select(gene, log2FoldChange, padj,
                            mean_norm_CHD, mean_norm_HFD,
                            bayes_P_HFD_gt_CHD, bayes_ER),
  caption="VDR pathway and coactivators",
  digits=c(0,3,3,1,1,3,2))
```

### Demoted candidates (verification)

```{r demoted-table}
dem <- tg |> filter(category == "demoted") |> arrange(gene)

knitr::kable(dem |> select(gene, log2FoldChange, padj,
                            mean_norm_CHD, mean_norm_HFD,
                            bayes_P_HFD_gt_CHD, bayes_ER),
  caption="Demoted candidates — expected to show no significant HFD upregulation",
  digits=c(0,3,3,1,1,3,2))
```

### Artifact check (DMRT family)

```{r dmrt-table}
dmrt <- tg |> filter(category == "artifact_check")

knitr::kable(dmrt |> select(gene, log2FoldChange, padj,
                             mean_norm_CHD, mean_norm_HFD, detected),
  caption="DMRT family — expected to be undetected if motif enrichment was an AT-rich artifact",
  digits=c(0,3,3,1,1,0))
```

### Functional GR targets

```{r gr-target-table}
gr_tg <- tg |> filter(category == "GR_target") |> arrange(desc(log2FoldChange))

knitr::kable(gr_tg |> select(gene, log2FoldChange, padj,
                              mean_norm_CHD, mean_norm_HFD,
                              bayes_P_HFD_gt_CHD, bayes_ER),
  caption="Functional GR targets — expected HFD upregulation if GR signalling is sensitized",
  digits=c(0,3,3,1,1,3,2))
```

---

## 3. ATAC × RNA-seq integration

### Annotate ATAC peaks to nearest gene

```{r atac-annotation}
library(GenomicRanges)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

atac <- read_tsv(atac_deseq, show_col_types=FALSE) |>
  mutate(.row = row_number())
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

# Build GRanges; carry .row so we can left_join after annotation.
# annotatePeak drops peaks on non-standard contigs (chrUn_*, random) so
# positional bind_cols would fail — left_join on .row preserves all peaks
# and gives NA annotations for any that were silently dropped.
atac_gr <- GRanges(
  seqnames = atac$chr,
  ranges   = IRanges(atac$start + 1, atac$end),
  .row     = atac$.row
)

atac_ann <- annotatePeak(atac_gr, TxDb=txdb, tssRegion=c(-5000, 500),
                         annoDb="org.Mm.eg.db", verbose=FALSE)
atac_ann_df <- as.data.frame(atac_ann)

atac_full <- atac |>
  left_join(atac_ann_df |> dplyr::select(.row, SYMBOL, distanceToTSS, annotation),
            by = ".row") |>
  select(-.row)
```

### Classify ATAC peaks and merge with RNA-seq

```{r atac-rnaseq-merge}
# Classify each ATAC peak by DESeq2 direction
atac_full <- atac_full |>
  mutate(
    atac_dir = case_when(
      !is.na(padj) & padj < 0.05 & log2FoldChange >  1 ~ "HFD_up",
      !is.na(padj) & padj < 0.05 & log2FoldChange < -1 ~ "CHD_up",
      TRUE                                              ~ "shared"
    )
  )

# Join nearest-gene RNA-seq results
deseq_slim <- deseq |>
  select(gene, rna_lfc = log2FoldChange, rna_padj = padj) |>
  filter(!is.na(gene))

atac_rna <- atac_full |>
  left_join(deseq_slim, by=c("SYMBOL"="gene"))
```

### Enrichment: are HFD-up ATAC peaks enriched near HFD-upregulated genes?

```{r enrichment-test}
# HFD-upregulated gene set (any LFC>0, padj<0.05)
hfd_genes <- deseq |> filter(!is.na(padj) & padj < 0.05 & log2FoldChange > 0) |> pull(gene)

peak_near_hfd_gene <- atac_rna |>
  mutate(near_hfd_gene = SYMBOL %in% hfd_genes)

# 2×2 contingency: atac_dir (HFD_up vs rest) × near_hfd_gene (T/F)
tab <- with(peak_near_hfd_gene,
  table(
    atac    = ifelse(atac_dir == "HFD_up", "HFD_up", "other"),
    rna_up  = near_hfd_gene
  )
)
cat("Contingency table (ATAC direction × nearest gene HFD-up):\n")
print(tab)

ft <- fisher.test(tab, alternative="greater")
cat(sprintf("\nFisher OR=%.2f, p=%.2e\n", ft$estimate, ft$p.value))
cat(sprintf("95%% CI: [%.2f, %.2f]\n", ft$conf.int[1], ft$conf.int[2]))
```

### Motif stratification: which motif class best predicts paired gene upregulation?

```{r motif-stratification}
# Load composite motif counts from ATAC pipeline
motif_file <- "results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv"

if (file.exists(motif_file)) {
  motif_map <- c("MA0148.5"="FOXA1","MA0047.5"="FOXA2","MA0466.4"="CEBPB",
                 "MA0102.5"="CEBPA","MA0480.3"="FOXO1","MA0113.4"="NR3C1",
                 "MA2327.1"="PGR_h","MA2323.1"="PGR_m")

  hfd_motifs <- read_tsv(motif_file, show_col_types=FALSE)
  names(hfd_motifs) <- ifelse(names(hfd_motifs) %in% names(motif_map),
                              motif_map[names(hfd_motifs)], names(hfd_motifs))

  # Need to also load AP-1 and VDR motifs from the full-motif scan
  full_scan_file <- "results/motif_analysis/full_motif_scan/HFD_specific/HFD_specific_per_peak_all_motifs.tsv"

  if (file.exists(full_scan_file)) {
    full_motifs <- read_tsv(full_scan_file, show_col_types=FALSE)

    # AP-1 motifs in JASPAR 2024 (representative IDs; from FULL_MOTIF_SCAN output)
    ap1_ids  <- grep("^MA0(099|478|099|477|153|162|039|063|099)",
                     names(full_motifs), value=TRUE)
    # VDR motif
    vdr_ids  <- grep("MA0693", names(full_motifs), value=TRUE)

    hfd_motifs$has_ap1 <- if (length(ap1_ids)) rowSums(full_motifs[, ap1_ids, drop=FALSE]) > 0 else FALSE
    hfd_motifs$has_vdr <- if (length(vdr_ids)) rowSums(full_motifs[, vdr_ids, drop=FALSE]) > 0 else FALSE
  } else {
    hfd_motifs$has_ap1 <- FALSE
    hfd_motifs$has_vdr <- FALSE
    message("Full motif scan not found — AP-1/VDR stratification skipped")
  }

  # GR motif
  gr_cols <- intersect(c("NR3C1","PGR_h","PGR_m"), names(hfd_motifs))
  hfd_motifs$has_gr <- if (length(gr_cols)) rowSums(hfd_motifs[, gr_cols, drop=FALSE]) > 0 else FALSE

  # Motif class assignment
  hfd_motifs <- hfd_motifs |>
    mutate(
      motif_class = case_when(
        has_ap1 & has_gr & !has_vdr ~ "AP1_GR",
        has_vdr & has_gr & !has_ap1 ~ "VDR_GR",
        has_ap1 & has_vdr & has_gr  ~ "AP1_VDR_GR",
        has_gr  & !has_ap1 & !has_vdr ~ "GR_only",
        has_ap1 & !has_gr  ~ "AP1_only",
        has_vdr & !has_gr  ~ "VDR_only",
        TRUE                ~ "none"
      )
    )

  # Parse peak coordinates from FASTA-style peak IDs
  peak_coords <- hfd_motifs |>
    mutate(
      chr   = sub(":.*", "", peak),
      start = as.integer(sub(".*:(\\d+)-.*", "\\1", peak)),
      end   = as.integer(sub(".*-", "", peak))
    )

  # Join to ATAC+RNA table via chr/start/end overlap (nearest match by start)
  hfd_peaks_rna <- atac_rna |>
    filter(atac_dir == "HFD_up") |>
    left_join(peak_coords |> select(chr, start, end, motif_class),
              by=c("chr","start","end"))

  # Per motif class: fraction with HFD-up nearest gene
  strat <- hfd_peaks_rna |>
    filter(!is.na(motif_class)) |>
    group_by(motif_class) |>
    summarise(
      n_peaks      = n(),
      n_near_hfd   = sum(SYMBOL %in% hfd_genes, na.rm=TRUE),
      pct_near_hfd = round(100 * n_near_hfd / n_peaks, 1),
      .groups="drop"
    ) |>
    arrange(desc(pct_near_hfd))

  knitr::kable(strat,
    caption="HFD-up ATAC peaks: fraction near HFD-upregulated genes, by motif class")

  # Fisher test: each motif class vs GR_only
  gr_only  <- hfd_peaks_rna |> filter(motif_class == "GR_only")
  gr_n_up  <- sum(gr_only$SYMBOL %in% hfd_genes, na.rm=TRUE)
  gr_n_tot <- nrow(gr_only)

  fisher_vs_gr <- function(class_label) {
    cl <- hfd_peaks_rna |> filter(motif_class == class_label)
    if (nrow(cl) < 5) return(NULL)
    tab <- matrix(c(sum(cl$SYMBOL %in% hfd_genes, na.rm=TRUE), nrow(cl),
                    gr_n_up, gr_n_tot), nrow=2)
    ft  <- fisher.test(tab)
    tibble(motif_class=class_label, n=nrow(cl),
           pct=round(100*tab[1,1]/nrow(cl),1),
           OR=round(ft$estimate,2), p=signif(ft$p.value,3))
  }

  classes <- setdiff(unique(hfd_peaks_rna$motif_class), c("GR_only",NA))
  fisher_res <- bind_rows(lapply(classes, fisher_vs_gr))
  knitr::kable(fisher_res,
    caption="Motif class vs GR-only: enrichment near HFD-up genes (Fisher exact)")
} else {
  cat("Composite motif scan output not found — run ATAC pipeline first.\n")
}
```

---

## 4. Locus tracks

BigWig files generated by `RNASEQ_BIGWIG` (CPM-normalised, 50bp bins) and
`ATAC_BIGWIG` (RPGC-normalised, 10bp bins) are at:

- RNA-seq: `results/rnaseq/bigwig/<sample_id>.CPM.bw`
- ATAC:    `results/bigwig/atac/<sample_id>.RPGC.bw`

Track ranges below are for mm10.  Visualisation uses pyGenomeTracks or IGV.

```{r track-loci}
loci <- tribble(
  ~gene,    ~chr,   ~start,     ~end,
  "Hsd11b1","chr1",  4819895,    4920000,
  "Fkbp5",  "chr4",  99948700,  100065000,
  "Vdr",    "chr15", 97844000,   97908000,
  "Junb",   "chr8",  84786000,   84800000,
  "Fosl2",  "chr5",  32356000,   32400000
)
knitr::kable(loci, caption="Locus coordinates for track visualisation (mm10)")
```

The following code renders condition-averaged normalised coverage using
`wiggletools` mean across replicates, then `bigWigToBedGraph` + summary
stats. For IGV batch rendering, load bigWigs directly from the paths above.

```{r track-wiggle, eval=FALSE}
# Shell commands to merge per-condition BigWigs (requires wiggletools + UCSC tools)
# Run from the GSE236575 directory.

conds <- c("CHD", "HFD")
types <- c(
  list(tag="atac", dir="results/bigwig/atac",    suffix=".RPGC.bw"),
  list(tag="rna",  dir="results/rnaseq/bigwig",  suffix=".CPM.bw")
)

for (cond in conds) {
  for (t in types) {
    bws <- paste(
      file.path(t$dir, paste0(cond, "_", 1:3, t$suffix)),
      collapse=" "
    )
    cmd <- sprintf(
      "wiggletools mean %s | wigToBigWig -fixedStep stdin results/genome/mm10.chrom.sizes %s/%s_%s_mean.bw",
      bws, t$dir, cond, t$tag
    )
    cat(cmd, "\n")
  }
}
```

---

## 5. Decision outputs

### 5.1 AP-1 lead candidate (Junb vs Fosl2)

```{r ap1-decision}
ap1_key <- tg |>
  filter(gene %in% c("Junb","Fosl2")) |>
  select(gene, log2FoldChange, padj, mean_norm_CHD, mean_norm_HFD,
         bayes_P_HFD_gt_CHD, bayes_ER, bayes_CrI_lo, bayes_CrI_hi)

knitr::kable(ap1_key, digits=c(0,3,3,1,1,3,2,3,3),
  caption="Junb vs Fosl2 — head-to-head comparison for AP-1 lead candidate selection")

cat("\n--- AP-1 lead candidate verdict ---\n")
junb  <- ap1_key |> filter(gene=="Junb")
fosl2 <- ap1_key |> filter(gene=="Fosl2")

if (nrow(junb) && nrow(fosl2) && !is.na(junb$log2FoldChange) && !is.na(fosl2$log2FoldChange)) {
  lead <- if (junb$log2FoldChange > fosl2$log2FoldChange) "Junb" else "Fosl2"
  cat(sprintf("Lead AP-1 candidate by HFD log2FC: %s (LFC=%.3f vs %.3f)\n",
              lead, junb$log2FoldChange, fosl2$log2FoldChange))
  cat(sprintf("Bayesian P(HFD>CHD): Junb=%.3f, Fosl2=%.3f\n",
              junb$bayes_P_HFD_gt_CHD, fosl2$bayes_P_HFD_gt_CHD))
} else {
  cat("One or both genes not detected — see targeted gene table.\n")
}
```

### 5.2 DMRT family expression status

```{r dmrt-decision}
dmrt <- tg |> filter(category == "artifact_check") |>
  select(gene, mean_norm_CHD, mean_norm_HFD, detected, log2FoldChange, padj)

knitr::kable(dmrt, caption="DMRT family expression status",
             digits=c(0,1,1,0,3,3))

cat("\n--- DMRT verdict ---\n")
if (all(!dmrt$detected, na.rm=TRUE)) {
  cat("DMRT1-4 are undetected (mean normalised count < 5 in both conditions).\n")
  cat("Verdict: AT-rich motif enrichment is an artifact; DMRT family are not expressed in eWAT adipocytes.\n")
} else {
  detected_dmrts <- dmrt |> filter(detected) |> pull(gene)
  cat("DETECTED:", paste(detected_dmrts, collapse=", "), "\n")
  cat("Verdict: requires manual review — unexpected DMRT expression in adipocytes.\n")
}
```

### 5.3 Coactivator and HSD11B1 expression status

```{r coactivator-decision}
coact <- tg |> filter(gene %in% c("Ncoa1","Ncoa2","Ncoa3","Med1","Hsd11b1","Hsd11b2")) |>
  select(gene, log2FoldChange, padj, mean_norm_CHD, mean_norm_HFD,
         bayes_P_HFD_gt_CHD, bayes_ER)

knitr::kable(coact, digits=c(0,3,3,1,1,3,2),
  caption="Coactivators and HSD11B1 — expression and HFD regulation")

cat("\n--- Coactivator/HSD11B1 verdict ---\n")
for (g in c("Ncoa1","Ncoa2","Ncoa3","Med1","Hsd11b1","Hsd11b2")) {
  row <- tg |> filter(gene == g)
  if (nrow(row) == 0 || is.na(row$log2FoldChange)) {
    cat(sprintf("  %s: not detected\n", g))
  } else {
    direction <- if (row$log2FoldChange > 0) "HFD-up" else "HFD-down"
    sig       <- if (!is.na(row$padj) && row$padj < 0.05) "sig" else "ns"
    cat(sprintf("  %s: LFC=%.3f, padj=%.3f (%s %s), P(HFD>CHD)=%.3f\n",
                g, row$log2FoldChange, row$padj, direction, sig,
                row$bayes_P_HFD_gt_CHD))
  }
}
```

---

## Session information

```{r session-info}
sessionInfo()
```
