---
title: "GSE236575 RNA-seq × ATAC-seq Integration"
subtitle: "TRAP-seq HFD vs CHD — targeted gene analysis, ATAC peak annotation, motif stratification"
author: "Dave Bridges"
date: "2026-05-12"
editor: source
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show code"
    fig-path: "figures/rnaseq/"
  gfm:
    html-math-method: webtex
theme: journal
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    fig-path: "figures/rnaseq/"
    dev: ["png", "pdf"]
    dpi: 300
    dev.args:
      png:
        type: "cairo-png"
      pdf:
        family: "sans"
---

```{r setup}
#| echo: false
se <- function(x) sd(x, na.rm=TRUE) / sqrt(sum(!is.na(x)))
library(tidyverse)
color_scheme <- c(CHD="#00274c", HFD="#ffcb05")
```

## Dataset overview

### RNA-seq data source

The matched RNA-seq data are from **GSE236578 [TRAP]**, a sub-series of the
super-series GSE236580 ("Adipose tissue retains an epigenetic memory of obesity
after weight loss"; PMID 39558077).  The same AdipER-Cre/NuTRAP mice and eWAT
adipocyte preparations were used for both ATAC-seq (GSE236575) and TRAP-seq
(GSE236578).  TRAP-seq captures actively translated transcripts specifically
from adipocytes, avoiding contamination from stromal-vascular cells.

| Sample ID | GSM | SRR | Condition |
|-----------|-----|-----|-----------|
| CHD_1 | GSM7558266 | SRR25152278 | Chow diet |
| CHD_2 | GSM7558271 | SRR25152274 | Chow diet |
| CHD_3 | GSM7558276 | SRR25152266 | Chow diet |
| HFD_1 | GSM7558267 | SRR25152277 | High-fat diet |
| HFD_2 | GSM7558272 | SRR25152273 | High-fat diet |
| HFD_3 | GSM7558277 | SRR25152265 | High-fat diet |

Library: paired-end 150 bp, template switching (Maxima H Minus RT), unstranded.
Alignment: STAR → mm10/GENCODE vM25, unique mappers only.
Quantification: featureCounts, gene_name, unstranded (-s 0).

### Pipeline output files

All files are generated by the `RNASEQ_DESEQ2` Nextflow process.

```{r file-paths}
deseq_file    <- "results/rnaseq/deseq2/rnaseq_deseq2_results.txt"
targeted_file <- "results/rnaseq/deseq2/rnaseq_targeted_genes.tsv"
norm_file     <- "results/rnaseq/deseq2/rnaseq_normalized_counts.txt"
atac_deseq    <- "results/deseq2/deseq2_results.txt"
```

---

## 1. DESeq2 differential expression — global summary

```{r deseq-load}
library(readr)
library(dplyr)

deseq <- read_tsv(deseq_file, show_col_types=FALSE) |>
  rename(gene = 1)   # col 1 is the row-names field written by write.table; rownames_to_column() inserts integers on tibbles

cat("Total genes tested:", nrow(deseq), "\n")
cat("HFD-up   (padj<0.05, LFC>0.585):",
    sum(!is.na(deseq$padj) & deseq$padj < 0.05 & deseq$log2FoldChange > 0.585), "\n")
cat("HFD-down (padj<0.05, LFC< -0.585):",
    sum(!is.na(deseq$padj) & deseq$padj < 0.05 & deseq$log2FoldChange < -0.585), "\n")
```

```{r volcano-global, fig.width=9, fig.height=6}
library(ggrepel)
library(ggplot2)

tg_genes <- c("Fos","Jun","Junb","Jund","Fosl1","Fosl2","Batf","Batf3","Atf3","Atf4","Ddit3",
              "Vdr","Ncoa1","Ncoa2","Ncoa3","Med1",
              "Hsd11b1","Nr3c1","Fkbp5","Sgk1","Tsc22d3")

deseq_plot <- deseq |>
  mutate(
    sig = !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 0.585,
    label = ifelse(gene %in% tg_genes, gene, NA_character_)
  )

ggplot(deseq_plot, aes(log2FoldChange, -log10(pvalue))) +
  geom_point(aes(color=sig), alpha=0.4, size=0.7) +
  scale_color_manual(values=c("FALSE"="grey70","TRUE"="#d62728"), guide="none") +
  geom_text_repel(aes(label=label), size=3, max.overlaps=40,
                  segment.color="grey60") +
  geom_hline(yintercept=-log10(0.05), lty=2, color="grey50") +
  geom_vline(xintercept=c(-0.585, 0.585), lty=2, color="grey50") +
  labs(x="log2FC (HFD / CHD)", y="-log10(p-value)",
       title="TRAP-seq: HFD vs CHD differential expression",
       subtitle="Red = FDR<0.05, |LFC|>0.585; labelled genes = targeted candidates") +
  theme_classic(base_size=13)
```

---

## 2. Targeted gene analysis

### Load and display results

```{r targeted-load}
tg <- read_tsv(targeted_file, show_col_types=FALSE)

# Format for display
tg_display <- tg |>
  mutate(
    across(c(log2FoldChange, lfcSE, bayes_CrI_lo, bayes_CrI_hi), ~round(., 3)),
    across(c(pvalue, padj), ~signif(., 3)),
    across(c(bayes_P_HFD_gt_CHD, bayes_ER), ~round(., 3)),
    across(c(mean_norm_CHD, mean_norm_HFD), ~round(., 1))
  ) |>
  dplyr::select(category, gene, log2FoldChange, lfcSE, padj, pvalue,
         mean_norm_CHD, mean_norm_HFD, detected,
         bayes_P_HFD_gt_CHD, bayes_ER, bayes_CrI_lo, bayes_CrI_hi)

knitr::kable(tg_display,
  caption="Targeted gene-level TRAP-seq results (HFD vs CHD). Bayes stats use Normal(0,1) prior on log2FC (analytical conjugate update).",
  digits=c(0,0,3,3,3,3,1,1,0,3,2,3,3))
```

### AP-1 candidates — key comparison

```{r ap1-table}
ap1 <- tg |> filter(category == "AP-1") |>
  arrange(desc(log2FoldChange))

knitr::kable(ap1 |> dplyr::select(gene, log2FoldChange, padj, pvalue,
                            mean_norm_CHD, mean_norm_HFD,
                            bayes_P_HFD_gt_CHD, bayes_ER),
  caption="AP-1 family members — ranked by HFD log2FC",
  digits=c(0,3,3,3,1,1,3,2))
```

### AP-1 forest plot

```{r ap1-forest, fig.width=7, fig.height=5}
ap1_ok <- ap1 |> filter(!is.na(log2FoldChange))

ggplot(ap1_ok, aes(x=reorder(gene, log2FoldChange), y=log2FoldChange)) +
  geom_col(aes(fill=log2FoldChange > 0), width=0.6) +
  geom_errorbar(aes(ymin=bayes_CrI_lo, ymax=bayes_CrI_hi), width=0.25) +
  geom_hline(yintercept=0, lty=2) +
  coord_flip() +
  scale_fill_manual(values=c("TRUE"="#d62728","FALSE"="#1f77b4"),
                    labels=c("TRUE"="HFD up","FALSE"="CHD up"), name=NULL) +
  labs(x=NULL, y="log2FC (HFD / CHD), 95% CrI",
       title="AP-1 family — TRAP-seq HFD vs CHD",
       subtitle="Bars = posterior mean; error bars = 95% CrI (Normal(0,1) prior)") +
  theme_classic(base_size=12)
```

### VDR pathway and GR coactivators

```{r vdr-table}
vdr <- tg |> filter(category == "VDR_pathway") |> arrange(gene)

knitr::kable(vdr |> dplyr::select(gene, log2FoldChange, padj,
                            mean_norm_CHD, mean_norm_HFD,
                            bayes_P_HFD_gt_CHD, bayes_ER),
  caption="VDR pathway and coactivators",
  digits=c(0,3,3,1,1,3,2))
```

### Demoted candidates (verification)

```{r demoted-table}
dem <- tg |> filter(category == "demoted") |> arrange(gene)

knitr::kable(dem |> dplyr::select(gene, log2FoldChange, padj,
                            mean_norm_CHD, mean_norm_HFD,
                            bayes_P_HFD_gt_CHD, bayes_ER),
  caption="Demoted candidates — expected to show no significant HFD upregulation",
  digits=c(0,3,3,1,1,3,2))
```

### Artifact check (DMRT family)

```{r dmrt-table}
dmrt <- tg |> filter(category == "artifact_check")

knitr::kable(dmrt |> dplyr::select(gene, log2FoldChange, padj,
                             mean_norm_CHD, mean_norm_HFD, detected),
  caption="DMRT family — expected to be undetected if motif enrichment was an AT-rich artifact",
  digits=c(0,3,3,1,1,0))
```

### Functional GR targets

```{r gr-target-table}
gr_tg <- tg |> filter(category == "GR_target") |> arrange(desc(log2FoldChange))

knitr::kable(gr_tg |> dplyr::select(gene, log2FoldChange, padj,
                              mean_norm_CHD, mean_norm_HFD,
                              bayes_P_HFD_gt_CHD, bayes_ER),
  caption="Functional GR targets — expected HFD upregulation if GR signalling is sensitized",
  digits=c(0,3,3,1,1,3,2))
```

---

## 3. ATAC × RNA-seq integration

### Annotate ATAC peaks to nearest gene

```{r atac-annotation}
library(GenomicRanges)
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

atac <- read_tsv(atac_deseq, show_col_types=FALSE) |>
  mutate(.row = row_number())   # keep index; annotatePeak drops non-standard contigs
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

# Build GRanges carrying .row so we can left_join back safely
atac_gr <- GRanges(
  seqnames = atac$chr,
  ranges   = IRanges(atac$start + 1, atac$end),
  .row     = atac$.row
)

atac_ann <- annotatePeak(atac_gr, TxDb=txdb, tssRegion=c(-5000, 500),
                         annoDb="org.Mm.eg.db", verbose=FALSE)
atac_ann_df <- as.data.frame(atac_ann)

# left_join instead of bind_cols: peaks on non-standard contigs get NA annotation
atac_full <- atac |>
  left_join(atac_ann_df |> dplyr::select(.row, SYMBOL, distanceToTSS, annotation),
            by = ".row") |>
  dplyr::select(-.row)
```

### Classify ATAC peaks and merge with RNA-seq

```{r atac-rnaseq-merge}
# Classify each ATAC peak by DESeq2 direction
atac_full <- atac_full |>
  mutate(
    atac_dir = case_when(
      !is.na(padj) & padj < 0.05 & log2FoldChange >  1 ~ "HFD_up",
      !is.na(padj) & padj < 0.05 & log2FoldChange < -1 ~ "CHD_up",
      TRUE                                              ~ "shared"
    )
  )

# Join nearest-gene RNA-seq results
deseq_slim <- deseq |>
  dplyr::select(gene, rna_lfc = log2FoldChange, rna_padj = padj) |>
  mutate(rna_lfc  = as.numeric(rna_lfc),
         rna_padj = as.numeric(rna_padj)) |>   # guard: read_tsv may infer character if col starts with NAs
  filter(!is.na(gene))

atac_rna <- atac_full |>
  left_join(deseq_slim, by=c("SYMBOL"="gene"))
```

### Enrichment: are HFD-up ATAC peaks enriched near HFD-upregulated genes?

```{r enrichment-test}
# HFD-upregulated gene set (any LFC>0, padj<0.05)
hfd_genes <- deseq |> filter(!is.na(padj) & padj < 0.05 & log2FoldChange > 0) |> pull(gene)

peak_near_hfd_gene <- atac_rna |>
  mutate(near_hfd_gene = SYMBOL %in% hfd_genes)

# 2×2 contingency: atac_dir (HFD_up vs rest) × near_hfd_gene (T/F)
tab <- with(peak_near_hfd_gene,
  table(
    atac    = ifelse(atac_dir == "HFD_up", "HFD_up", "other"),
    rna_up  = near_hfd_gene
  )
)
cat("Contingency table (ATAC direction × nearest gene HFD-up):\n")
print(tab)

ft <- fisher.test(tab, alternative="greater")
cat(sprintf("\nFisher OR=%.2f, p=%.2e\n", ft$estimate, ft$p.value))
cat(sprintf("95%% CI: [%.2f, %.2f]\n", ft$conf.int[1], ft$conf.int[2]))
```

> **Interpretation of OR ≈ 1.1:**
> The enrichment is real but tiny (17.4% vs 18.9% of peaks near an HFD-up gene).
> Two reasons account for this:
>
> 1. *Nearest-gene attribution*: most enhancers regulate non-nearest genes;
>    the correct test is TAD-level or loop-based coupling (pinned).
>
> 2. *Primed vs. activated enhancers*: AP-1 opens chromatin at GR-binding sites —
>    making them **accessible** — without GR necessarily occupying or activating
>    them yet. GR binding requires nuclear GR, which depends on ligand availability
>    and Fkbp5-mediated de-repression. The HFD-specific peaks therefore represent a
>    **permissive (poised) chromatin state**: the sites are ready for GR once the
>    GR-dependent signal is present. Transcriptional activation is the downstream
>    consequence, not a simultaneous event. Weak ATAC–RNA spatial coupling is
>    therefore the *expected* result of chromatin priming, not evidence against
>    the model.

### Motif stratification: which motif class best predicts paired gene upregulation?

```{r motif-stratification}
# Load composite motif counts from ATAC pipeline
motif_file <- "results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv"

if (file.exists(motif_file)) {
  motif_map <- c("MA0148.5"="FOXA1","MA0047.5"="FOXA2","MA0466.4"="CEBPB",
                 "MA0102.5"="CEBPA","MA0480.3"="FOXO1","MA0113.4"="NR3C1",
                 "MA2327.1"="PGR_h","MA2323.1"="PGR_m")

  hfd_motifs <- read_tsv(motif_file, show_col_types=FALSE)
  names(hfd_motifs) <- ifelse(names(hfd_motifs) %in% names(motif_map),
                              motif_map[names(hfd_motifs)], names(hfd_motifs))

  # Need to also load AP-1 and VDR motifs from the full-motif scan
  full_scan_file <- "results/motif_analysis/full_motif_scan/HFD_specific/HFD_specific_per_peak_all_motifs.tsv"

  if (file.exists(full_scan_file)) {
    full_motifs <- read_tsv(full_scan_file, show_col_types=FALSE)

    # AP-1 motifs in JASPAR 2024 (representative IDs; from FULL_MOTIF_SCAN output)
    ap1_ids  <- grep("^MA0(099|478|099|477|153|162|039|063|099)",
                     names(full_motifs), value=TRUE)
    # VDR motif
    vdr_ids  <- grep("MA0693", names(full_motifs), value=TRUE)

    hfd_motifs$has_ap1 <- if (length(ap1_ids)) rowSums(full_motifs[, ap1_ids, drop=FALSE]) > 0 else FALSE
    hfd_motifs$has_vdr <- if (length(vdr_ids)) rowSums(full_motifs[, vdr_ids, drop=FALSE]) > 0 else FALSE
  } else {
    hfd_motifs$has_ap1 <- FALSE
    hfd_motifs$has_vdr <- FALSE
    message("Full motif scan not found — AP-1/VDR stratification skipped")
  }

  # GR motif
  gr_cols <- intersect(c("NR3C1","PGR_h","PGR_m"), names(hfd_motifs))
  hfd_motifs$has_gr <- if (length(gr_cols)) rowSums(hfd_motifs[, gr_cols, drop=FALSE]) > 0 else FALSE

  # Motif class assignment
  hfd_motifs <- hfd_motifs |>
    mutate(
      motif_class = case_when(
        has_ap1 & has_gr & !has_vdr ~ "AP1_GR",
        has_vdr & has_gr & !has_ap1 ~ "VDR_GR",
        has_ap1 & has_vdr & has_gr  ~ "AP1_VDR_GR",
        has_gr  & !has_ap1 & !has_vdr ~ "GR_only",
        has_ap1 & !has_gr  ~ "AP1_only",
        has_vdr & !has_gr  ~ "VDR_only",
        TRUE                ~ "none"
      )
    )

  # Parse peak coordinates from FASTA-style peak IDs
  peak_coords <- hfd_motifs |>
    mutate(
      chr   = sub(":.*", "", peak),
      start = as.integer(sub(".*:(\\d+)-.*", "\\1", peak)),
      end   = as.integer(sub(".*-", "", peak))
    )

  # Join to ATAC+RNA table via chr/start/end overlap (nearest match by start)
  hfd_peaks_rna <- atac_rna |>
    filter(atac_dir == "HFD_up") |>
    left_join(peak_coords |> dplyr::select(chr, start, end, motif_class),
              by=c("chr","start","end"))

  # Per motif class: fraction with HFD-up nearest gene
  strat <- hfd_peaks_rna |>
    filter(!is.na(motif_class)) |>
    group_by(motif_class) |>
    summarise(
      n_peaks      = n(),
      n_near_hfd   = sum(SYMBOL %in% hfd_genes, na.rm=TRUE),
      pct_near_hfd = round(100 * n_near_hfd / n_peaks, 1),
      .groups="drop"
    ) |>
    arrange(desc(pct_near_hfd))

  knitr::kable(strat,
    caption="HFD-up ATAC peaks: fraction near HFD-upregulated genes, by motif class")

  # Fisher test: each motif class vs GR_only
  gr_only  <- hfd_peaks_rna |> filter(motif_class == "GR_only")
  gr_n_up  <- sum(gr_only$SYMBOL %in% hfd_genes, na.rm=TRUE)
  gr_n_tot <- nrow(gr_only)

  fisher_vs_gr <- function(class_label) {
    cl <- hfd_peaks_rna |> filter(motif_class == class_label)
    if (nrow(cl) < 5) return(NULL)
    tab <- matrix(c(sum(cl$SYMBOL %in% hfd_genes, na.rm=TRUE), nrow(cl),
                    gr_n_up, gr_n_tot), nrow=2)
    ft  <- fisher.test(tab)
    tibble(motif_class=class_label, n=nrow(cl),
           pct=round(100*tab[1,1]/nrow(cl),1),
           OR=round(ft$estimate,2), p=signif(ft$p.value,3))
  }

  classes <- setdiff(unique(hfd_peaks_rna$motif_class), c("GR_only",NA))
  fisher_res <- bind_rows(lapply(classes, fisher_vs_gr))
  knitr::kable(fisher_res,
    caption="Motif class vs GR-only: enrichment near HFD-up genes (Fisher exact)")
} else {
  cat("Composite motif scan output not found — run ATAC pipeline first.\n")
}
```

---

## 4. Locus tracks

BigWig files generated by `RNASEQ_BIGWIG` (CPM-normalised, 50bp bins) and
`ATAC_BIGWIG` (RPGC-normalised, 10bp bins) are at:

- RNA-seq: `results/rnaseq/bigwig/<sample_id>.CPM.bw`
- ATAC:    `results/bigwig/atac/<sample_id>.RPGC.bw`

Track ranges below are for mm10.  Visualisation uses pyGenomeTracks or IGV.

```{r track-loci}
loci <- tribble(
  ~gene,    ~chr,   ~start,     ~end,
  "Hsd11b1","chr1",  4819895,    4920000,
  "Fkbp5",  "chr4",  99948700,  100065000,
  "Vdr",    "chr15", 97844000,   97908000,
  "Junb",   "chr8",  84786000,   84800000,
  "Fosl2",  "chr5",  32356000,   32400000
)
knitr::kable(loci, caption="Locus coordinates for track visualisation (mm10)")
```

The following code renders condition-averaged normalised coverage using
`wiggletools` mean across replicates, then `bigWigToBedGraph` + summary
stats. For IGV batch rendering, load bigWigs directly from the paths above.

```{r track-wiggle, eval=FALSE}
# Shell commands to merge per-condition BigWigs (requires wiggletools + UCSC tools)
# Run from the GSE236575 directory.

conds <- c("CHD", "HFD")
types <- c(
  list(tag="atac", dir="results/bigwig/atac",    suffix=".RPGC.bw"),
  list(tag="rna",  dir="results/rnaseq/bigwig",  suffix=".CPM.bw")
)

for (cond in conds) {
  for (t in types) {
    bws <- paste(
      file.path(t$dir, paste0(cond, "_", 1:3, t$suffix)),
      collapse=" "
    )
    cmd <- sprintf(
      "wiggletools mean %s | wigToBigWig -fixedStep stdin results/genome/mm10.chrom.sizes %s/%s_%s_mean.bw",
      bws, t$dir, cond, t$tag
    )
    cat(cmd, "\n")
  }
}
```

---

## 5. Decision outputs

### 5.1 AP-1 lead candidate (Junb vs Fosl2)

```{r ap1-decision}
ap1_key <- tg |>
  filter(gene %in% c("Junb","Fosl2")) |>
  dplyr::select(gene, log2FoldChange, padj, mean_norm_CHD, mean_norm_HFD,
         bayes_P_HFD_gt_CHD, bayes_ER, bayes_CrI_lo, bayes_CrI_hi)

knitr::kable(ap1_key, digits=c(0,3,3,1,1,3,2,3,3),
  caption="Junb vs Fosl2 — head-to-head comparison for AP-1 lead candidate selection")

cat("\n--- AP-1 lead candidate verdict ---\n")
junb  <- ap1_key |> filter(gene=="Junb")
fosl2 <- ap1_key |> filter(gene=="Fosl2")

if (nrow(junb) && nrow(fosl2) && !is.na(junb$log2FoldChange) && !is.na(fosl2$log2FoldChange)) {
  lead <- if (junb$log2FoldChange > fosl2$log2FoldChange) "Junb" else "Fosl2"
  cat(sprintf("Lead AP-1 candidate by HFD log2FC: %s (LFC=%.3f vs %.3f)\n",
              lead, junb$log2FoldChange, fosl2$log2FoldChange))
  cat(sprintf("Bayesian P(HFD>CHD): Junb=%.3f, Fosl2=%.3f\n",
              junb$bayes_P_HFD_gt_CHD, fosl2$bayes_P_HFD_gt_CHD))
} else {
  cat("One or both genes not detected — see targeted gene table.\n")
}
```

### 5.2 DMRT family expression status

```{r dmrt-decision}
dmrt <- tg |> filter(category == "artifact_check") |>
  dplyr::select(gene, mean_norm_CHD, mean_norm_HFD, detected, log2FoldChange, padj)

knitr::kable(dmrt, caption="DMRT family expression status",
             digits=c(0,1,1,0,3,3))

cat("\n--- DMRT verdict ---\n")
if (all(!dmrt$detected, na.rm=TRUE)) {
  cat("DMRT1-4 are undetected (mean normalised count < 5 in both conditions).\n")
  cat("Verdict: AT-rich motif enrichment is an artifact; DMRT family are not expressed in eWAT adipocytes.\n")
} else {
  detected_dmrts <- dmrt |> filter(detected) |> pull(gene)
  cat("DETECTED:", paste(detected_dmrts, collapse=", "), "\n")
  cat("Verdict: requires manual review — unexpected DMRT expression in adipocytes.\n")
}
```

### 5.3 Coactivator and HSD11B1 expression status

```{r coactivator-decision}
coact <- tg |> filter(gene %in% c("Ncoa1","Ncoa2","Ncoa3","Med1","Hsd11b1","Hsd11b2")) |>
  dplyr::select(gene, log2FoldChange, padj, mean_norm_CHD, mean_norm_HFD,
         bayes_P_HFD_gt_CHD, bayes_ER)

knitr::kable(coact, digits=c(0,3,3,1,1,3,2),
  caption="Coactivators and HSD11B1 — expression and HFD regulation")

cat("\n--- Coactivator/HSD11B1 verdict ---\n")
for (g in c("Ncoa1","Ncoa2","Ncoa3","Med1","Hsd11b1","Hsd11b2")) {
  row <- tg |> filter(gene == g)
  if (nrow(row) == 0 || is.na(row$log2FoldChange)) {
    cat(sprintf("  %s: not detected\n", g))
  } else {
    direction <- if (row$log2FoldChange > 0) "HFD-up" else "HFD-down"
    sig       <- if (!is.na(row$padj) && row$padj < 0.05) "sig" else "ns"
    cat(sprintf("  %s: LFC=%.3f, padj=%.3f (%s %s), P(HFD>CHD)=%.3f\n",
                g, row$log2FoldChange, row$padj, direction, sig,
                row$bayes_P_HFD_gt_CHD))
  }
}
```

---

## 6. AP-1 motif scan at the Fkbp5 locus

**Hypothesis:** Fkbp5 is a Junb-repressed gene. If AP-1 motifs are present in
the Fkbp5 regulatory landscape and Junb ChIP signal is detectable there, the
two mechanisms (AP-1 chromatin remodelling + Fkbp5-mediated GR de-repression)
collapse into one Junb-led cascade: HFD to Junb up to AP-1 chromatin opening at
GR cooperative sites AND Fkbp5 down to constitutive GR nuclear activity.

Fkbp5 is on the **minus strand**, chr4:99,946,395-100,067,558 (mm10).
TSS approx chr4:100,067,500. Known GREs reside in introns 2 and 4.

### 6.1 ATAC-seq peaks at the Fkbp5 locus

```{r fkbp5-atac}
fkbp5_chr   <- "chr4"
fkbp5_start <- 99936000L
fkbp5_end   <- 100078000L

fkbp5_peaks <- atac_rna |>
  filter(chr == fkbp5_chr,
         start >= fkbp5_start,
         end   <= fkbp5_end) |>
  dplyr::select(chr, start, end, log2FoldChange, padj, atac_dir,
                SYMBOL, distanceToTSS, rna_lfc, rna_padj) |>
  arrange(start)

cat(sprintf("ATAC peaks at Fkbp5 locus (%s:%d-%d): %d\n",
            fkbp5_chr, fkbp5_start, fkbp5_end, nrow(fkbp5_peaks)))
cat(sprintf("  HFD-up : %d\n", sum(fkbp5_peaks$atac_dir == "HFD_up")))
cat(sprintf("  CHD-up : %d\n", sum(fkbp5_peaks$atac_dir == "CHD_up")))
cat(sprintf("  shared : %d\n", sum(fkbp5_peaks$atac_dir == "shared")))

knitr::kable(
  fkbp5_peaks |>
    mutate(across(c(log2FoldChange, rna_lfc), ~round(as.numeric(.), 3)),
           across(c(padj, rna_padj), ~signif(as.numeric(.), 2)),
           distanceToTSS = round(distanceToTSS / 1000, 1)),
  caption = "ATAC peaks at the Fkbp5 locus (distanceToTSS in kb)"
)
```

### 6.1a Fkbp5 locus — ATAC signal panel (CHD vs HFD)

```{r fkbp5-locus-panel}
#| fig-width: 8
#| fig-height: 4

select <- dplyr::select

# Normalized counts (row = peak_id, col = sample)
cnts_mat <- read.table("results/deseq2/deseq2_normalized_counts.txt",
                       header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)
cnts_long <- cnts_mat |>
  rownames_to_column("peak_id") |>
  pivot_longer(-peak_id, names_to = "sample", values_to = "norm_count") |>
  mutate(condition = ifelse(grepl("^CHD", sample), "CHD", "HFD"))

# Full peak data with per-sample counts
atac_counts <- read_tsv("results/deseq2/deseq2_results.txt", show_col_types = FALSE) |>
  left_join(cnts_long, by = "peak_id") |>
  mutate(mid_kb = (start + end) / 2e3)

# Helper: summarise counts at a locus into plottable long form
locus_counts <- function(chr_in, s, e) {
  atac_counts |>
    filter(chr == chr_in, start >= s, end <= e) |>
    group_by(peak_id, chr, start, end, log2FoldChange, padj, mid_kb) |>
    summarise(
      CHD    = mean(norm_count[condition == "CHD"]),
      HFD    = mean(norm_count[condition == "HFD"]),
      CHD_se = sd(norm_count[condition == "CHD"]) / sqrt(3),
      HFD_se = sd(norm_count[condition == "HFD"]) / sqrt(3),
      .groups = "drop"
    ) |>
    mutate(
      sig = case_when(padj < 0.001 ~ "***", padj < 0.01 ~ "**",
                      padj < 0.05  ~ "*",   TRUE         ~ "")
    ) |>
    pivot_longer(c(CHD, HFD), names_to = "condition", values_to = "mean_cnt") |>
    mutate(se = ifelse(condition == "CHD", CHD_se, HFD_se),
           peak_label = sprintf("%.1f kb", mid_kb))
}

fkbp5_cts <- locus_counts("chr4", 99900000, 100100000) |>
  mutate(condition = factor(condition, levels = c("CHD", "HFD")),
         peak_label = fct_reorder(peak_label, mid_kb))

# Significance label shown once (on HFD bar)
sig_labels <- fkbp5_cts |>
  filter(condition == "HFD", sig != "") |>
  mutate(y_pos = mean_cnt + se + max(fkbp5_cts$mean_cnt) * 0.06)

ggplot(fkbp5_cts, aes(x = peak_label, y = mean_cnt, fill = condition)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  geom_errorbar(aes(ymin = mean_cnt - se, ymax = mean_cnt + se),
                position = position_dodge(0.75), width = 0.25) +
  geom_text(data = sig_labels,
            aes(x = peak_label, y = y_pos, label = sig),
            inherit.aes = FALSE, size = 4, hjust = 0.5) +
  scale_fill_manual(values = color_scheme, name = "Diet") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Fkbp5 locus — ATAC-seq normalised counts",
    subtitle = "Each bar = one peak; * padj < 0.05  ** padj < 0.01  *** padj < 0.001 (DESeq2)",
    x        = "Peak mid-point (genomic position, kb)",
    y        = "Mean normalised ATAC count ± SE"
  ) +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "top")
```

> **Interpretation:** Every peak across the 200-kb Fkbp5 locus shows CHD ≥ HFD
> signal — chromatin is systematically more accessible in chow-diet adipocytes.
> Two peaks are formally significant (padj < 0.05): chr4:99,929,264 (upstream
> regulatory region) and chr4:99,968,583 (gene body). This is the chromatin
> correlate of the Fkbp5 downregulation seen in TRAP-seq (LFC = −1.29,
> padj = 0.010) and supports AP-1-mediated silencing at this locus in HFD.

### 6.2 AP-1 motif scan

```{r fkbp5-motif-scan}
suppressPackageStartupMessages({
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(JASPAR2024)
  library(TFBSTools)
  library(Biostrings)
})

# JASPAR2024 R package ships only metadata.csv with an incompatible SQLite schema.
# Strategy: (1) use JASPAR2020 local DB if installed (compatible TFBSTools schema),
#           (2) download the JASPAR 2024 CORE vertebrates flat file from the website.
pfm_list_all <- NULL

if (requireNamespace("JASPAR2020", quietly = TRUE)) {
  db2020 <- system.file("extdata", "JASPAR2020.sqlite", package = "JASPAR2020")
  if (nzchar(db2020) && file.exists(db2020)) {
    pfm_list_all <- TFBSTools::getMatrixSet(
      db2020,
      opts = list(collection = "CORE", tax_group = "vertebrates", all_versions = FALSE)
    )
    cat("Loaded", length(pfm_list_all), "motifs from JASPAR2020 (vertebrates)\n")
  }
}

if (is.null(pfm_list_all)) {
  tmp_jaspar <- tempfile(fileext = ".jaspar")
  dl <- tryCatch(
    httr::GET(
      paste0("https://jaspar.elixir.no/download/data/2024/CORE/",
             "JASPAR2024_CORE_vertebrates_non-redundant_pfms.jaspar"),
      httr::write_disk(tmp_jaspar, overwrite = TRUE),
      httr::timeout(60)
    ),
    error = function(e) { message("Download failed: ", e$message); NULL }
  )
  if (!is.null(dl) && !httr::http_error(dl)) {
    pfm_list_all <- TFBSTools::readJASPARMatrix(tmp_jaspar, matrixClass = "PFMatrix")
    cat("Loaded", length(pfm_list_all), "motifs from JASPAR 2024 download\n")
  }
}

if (is.null(pfm_list_all))
  stop("Could not load JASPAR motifs: install JASPAR2020 or ensure internet access")

ap1_motifs <- pfm_list_all[
  grepl("JUN|FOS|ATF|BATF|NFE2|MAFK|MAFF",
        sapply(pfm_list_all, function(x) x@name),
        ignore.case = TRUE)
]
cat(sprintf("AP-1 family motifs selected: %d\n", length(ap1_motifs)))

scan_gr  <- GRanges("chr4", IRanges(fkbp5_start, fkbp5_end))
scan_seq <- getSeq(BSgenome.Mmusculus.UCSC.mm10, scan_gr)
names(scan_seq) <- "Fkbp5_locus"

pwm_list  <- toPWM(ap1_motifs)
# Build hits tibble directly from S4 slots — as.data.frame(SiteSetList) produces
# inconsistent column names across TFBSTools versions; slot access is reliable.
hits_list <- lapply(names(pwm_list), function(id) {
  ssl <- searchSeq(pwm_list[[id]], scan_seq,
                   seqname = "Fkbp5_locus",
                   min.score = "85%", strand = "*")
  # searchSeq on a DNAStringSet returns a SiteSetList; [[1]] gives the SiteSet
  ss <- ssl[[1]]
  if (length(ss) == 0) return(NULL)
  tibble(
    start      = as.integer(start(ss@views)),
    end        = as.integer(end(ss@views)),
    strand     = as.character(ss@strand),
    score      = as.numeric(ss@score),
    motif_id   = id,
    motif_name = pwm_list[[id]]@name
  )
})
hits_df <- bind_rows(hits_list) |>
  mutate(
    mm10_pos    = fkbp5_start + start - 1L,
    dist_to_tss = 100067500L - mm10_pos,
    region = case_when(
      mm10_pos >= 100062500 & mm10_pos <= 100072500 ~ "promoter (+-5 kb)",
      mm10_pos >= 99946395  & mm10_pos <= 100067558 ~ "gene body",
      TRUE                                          ~ "flanking"
    )
  )

cat(sprintf("Total AP-1 hits (>=85%% score): %d\n", nrow(hits_df)))
cat("\nHits by region:\n"); print(table(hits_df$region))
cat("\nHits by TF:\n");     print(sort(table(hits_df$motif_name), decreasing = TRUE))
```

```{r fkbp5-motif-plot, fig.width=10, fig.height=4}
gre_pos <- tibble(
  label    = c("GRE intron2a", "GRE intron2b", "GRE intron4"),
  mm10_pos = c(100015000L, 100008000L, 99970000L)
)

hits_plot <- hits_df |>
  filter(region %in% c("promoter (+-5 kb)", "gene body")) |>
  mutate(motif_family = case_when(
    grepl("JUN|JUNB|JUND",  motif_name, ignore.case = TRUE) ~ "Jun",
    grepl("FOS|FOSL|FOSB",  motif_name, ignore.case = TRUE) ~ "Fos",
    grepl("ATF",            motif_name, ignore.case = TRUE) ~ "ATF",
    grepl("BATF",           motif_name, ignore.case = TRUE) ~ "BATF",
    TRUE                                                    ~ "other AP-1"
  ))

ggplot() +
  annotate("rect",
           xmin = 99946395, xmax = 100067558, ymin = -0.4, ymax = 0.4,
           fill = "grey85", alpha = 0.5) +
  annotate("text", x = 100006000, y = 0.55,
           label = "Fkbp5 gene body", size = 3, color = "grey40") +
  annotate("segment",
           x = 100067500, xend = 100062000, y = 0, yend = 0,
           arrow = arrow(length = unit(0.15, "cm")),
           color = "black", linewidth = 0.8) +
  annotate("text", x = 100068500, y = 0.15, label = "TSS", size = 3) +
  geom_vline(data = gre_pos,
             aes(xintercept = mm10_pos),
             lty = 2, color = "#e377c2", linewidth = 0.7) +
  geom_text(data = gre_pos,
            aes(x = mm10_pos, y = -0.65, label = label),
            size = 2.5, color = "#e377c2", angle = 90, hjust = 0) +
  geom_rug(data = hits_plot,
           aes(x = mm10_pos, color = motif_family),
           sides = "t", linewidth = 0.9, alpha = 0.85,
           length = unit(0.07, "npc")) +
  scale_color_manual(
    values = c(Jun = "#d62728", Fos = "#1f77b4", ATF = "#ff7f0e",
               BATF = "#2ca02c", "other AP-1" = "grey50"),
    name = "Motif family"
  ) +
  scale_x_continuous(
    labels = function(x) sprintf("%.2f Mb", x / 1e6),
    limits = c(fkbp5_start, fkbp5_end)
  ) +
  labs(x = "chr4 (mm10)", y = NULL,
       title = "AP-1 motif hits at the Fkbp5 locus (JASPAR 2024, >=85% score)",
       subtitle = "Pink dashed = known GRE positions; rug = AP-1 motif hit") +
  theme_classic(base_size = 11) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major.x = element_line(color = "grey90"))
```

```{r fkbp5-motif-table}
top_hits <- hits_df |>
  filter(region %in% c("promoter (+-5 kb)", "gene body")) |>
  arrange(desc(score)) |>
  slice_head(n = 20) |>
  dplyr::select(motif_name, motif_id, mm10_pos, dist_to_tss,
                strand, score, region) |>
  mutate(dist_to_tss = round(dist_to_tss / 1000, 2),
         score       = round(score, 3))

knitr::kable(top_hits,
  caption = "Top 20 AP-1 motif hits at Fkbp5 (>=85%, by score). dist_to_tss in kb.")
```

### 6.3 ChIP-Atlas: Junb binding evidence at Fkbp5

```{r fkbp5-chipatlas}
library(httr)

query_chipatlas <- function(distance) {
  url <- sprintf(
    "https://chip-atlas.dbcls.jp/api/target?assemblyId=mm10&factor=Junb&gene=Fkbp5&distance=%d",
    distance
  )
  cat(sprintf("Querying ChIP-Atlas (dist=%d bp)\n", distance))
  tryCatch({
    resp <- GET(url, timeout(30))
    if (status_code(resp) == 200) {
      raw <- content(resp, as = "text", encoding = "UTF-8")
      if (nchar(trimws(raw)) < 5)
        return(tibble(result = sprintf("No Junb peaks within %d bp of Fkbp5 TSS", distance)))
      read_tsv(I(raw), show_col_types = FALSE)
    } else {
      tibble(result = sprintf("HTTP %d", status_code(resp)))
    }
  }, error = function(e) tibble(result = conditionMessage(e)))
}

ca_result <- query_chipatlas(1000)
if ("result" %in% names(ca_result) && grepl("No Junb", ca_result$result))
  ca_result <- query_chipatlas(5000)

knitr::kable(ca_result,
  caption = "ChIP-Atlas: public Junb ChIP-seq experiments with peaks near Fkbp5 TSS")

cat("\nPeak Browser URL for manual inspection:\n")
cat(sprintf(
  "https://chip-atlas.dbcls.jp/peakBrowser/?assemblyId=mm10&factor=JunB&chr=chr4&start=%d&end=%d\n",
  fkbp5_start, fkbp5_end
))
```

### 6.4 Synthesis

```{r fkbp5-synthesis}
n_promoter <- sum(hits_df$region == "promoter (+-5 kb)")
n_jun      <- sum(grepl("JUN", hits_df$motif_name, ignore.case = TRUE) &
                  hits_df$region %in% c("promoter (+-5 kb)", "gene body"))
fkbp5_row  <- tg |> filter(gene == "Fkbp5")

cat("=== Fkbp5 locus AP-1 evidence summary ===\n")
cat(sprintf("AP-1 hits in promoter (+-5 kb) : %d\n", n_promoter))
cat(sprintf("Jun-family hits in promoter+body: %d\n", n_jun))
if (nrow(fkbp5_row) > 0 && !is.na(fkbp5_row$log2FoldChange))
  cat(sprintf("Fkbp5 TRAP-seq: LFC=%.3f, padj=%.3f\n",
              fkbp5_row$log2FoldChange, fkbp5_row$padj))

cat("\n--- Verdict ---\n")
if (n_jun > 0 && n_promoter > 0) {
  cat("POSITIVE: Jun-family AP-1 motifs present in the Fkbp5 promoter.\n")
  cat("Supports Junb-led cascade: HFD -> Junb up -> Fkbp5 repression ->\n")
  cat("GR de-repression -> Sgk1/Angptl4/Lep up.\n")
  cat("Validate with Junb CUT&RUN in HFD vs CHD adipocytes.\n")
} else if (n_promoter > 0) {
  cat("PARTIAL: AP-1 motifs present but no Jun-family hits in promoter.\n")
  cat("Regulation may be via Junb::Atf3 or Junb::Fos heterodimer.\n")
} else {
  cat("NEGATIVE at 85%: no AP-1 hits in Fkbp5 promoter.\n")
  cat("Consider: relaxing to 80%, checking distal enhancers,\n")
  cat("or indirect regulation via AP-1-activated repressor.\n")
}
```

---

## 7. AP-1 + GR composite peak scan

The "composite-enhancer" model predicts that HFD-opened chromatin disproportionately
carries **both AP-1 and GR motifs** — AP-1 pioneers accessibility, licensing
glucocorticoid receptor binding at sites that would otherwise be inaccessible.
We test this by classifying each HFD-specific peak and each shared (background)
peak as AP-1-only, GR-only, both, or neither.

**Motif definitions used:**

- **AP-1 class** — 35 JASPAR 2024 motifs: JUN/JUNB/JUND, FOS/FOSB/FOSL1/FOSL2,
  BATF/BATF3, ATF3, and all pairwise JUN·FOS heterodimer motifs;
  detected via FIMO (p < 1×10⁻⁴) in 500-bp peak windows.
- **GR class** — NR3C1 (MA0113.4), Pgr (MA2323.1), and PGR (MA2327.1);
  the progesterone-receptor motifs share an IR3 half-site biochemically
  indistinguishable from the canonical GR response element.

### 7.1 GR motif presence in HFD-specific peaks

```{r gr-presence-panel}
#| fig-width: 5
#| fig-height: 3.5

# AnnotationDbi masks dplyr::select — use explicit namespace throughout
select <- dplyr::select

# GR-class motif columns (NR3C1 + Pgr/PGR as IR3 proxy)
gr_motif_ids <- c("MA0113.4", "MA2323.1", "MA2327.1")

hfd_gr_raw <- read_tsv(
  "results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv",
  show_col_types = FALSE
)

gr_cols_present <- intersect(gr_motif_ids, names(hfd_gr_raw))
hfd_gr_classified <- hfd_gr_raw |>
  mutate(has_gr = rowSums(across(all_of(gr_cols_present))) > 0)

n_total  <- nrow(hfd_gr_classified)
n_gr     <- sum(hfd_gr_classified$has_gr)
pct_gr   <- round(100 * n_gr / n_total, 1)

gr_bar_df <- tibble(
  category = factor(c("GR motif present", "No GR motif"),
                    levels = c("No GR motif", "GR motif present")),
  n   = c(n_gr, n_total - n_gr),
  pct = c(pct_gr, 100 - pct_gr)
)

ggplot(gr_bar_df, aes(x = category, y = pct, fill = category)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%d\n(%.1f%%)", n, pct)),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("GR motif present" = "#d62728",
                               "No GR motif"      = "#aec7e8")) +
  scale_y_continuous(limits = c(0, 110), expand = c(0, 0)) +
  labs(
    title    = "GR-class motif presence in HFD-specific peaks",
    subtitle = sprintf("N = %d HFD-specific peaks; GR-class = NR3C1 / Pgr / PGR (JASPAR 2024)",
                       n_total),
    x = NULL, y = "Percentage of peaks (%)"
  ) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(size = 11))
```

`r sprintf("**%d / %d** (%.1f%%) HFD-specific peaks carry a GR-class (NR3C1/Pgr/PGR) motif.", n_gr, n_total, pct_gr)`

### 7.2 AP-1 + GR composite scan — stacked bar

```{r ap1-gr-stacked-bar}
#| fig-width: 6
#| fig-height: 4.5

# AP-1 motif IDs in JASPAR 2024 (JUN/FOS/BATF core + heterodimers + ATF3)
ap1_ids <- c(
  "MA0099.4", "MA0462.3", "MA0476.2", "MA0477.3", "MA0478.2",
  "MA0488.2", "MA0489.3", "MA0490.3", "MA0491.3", "MA0492.2",
  "MA0605.3", "MA1126.2", "MA1127.1", "MA1128.2", "MA1129.1",
  "MA1130.2", "MA1131.2", "MA1132.2", "MA1133.2", "MA1134.2",
  "MA1135.2", "MA1136.1", "MA1137.2", "MA1138.2", "MA1139.2",
  "MA1140.3", "MA1141.2", "MA1142.2", "MA1143.2", "MA1144.2",
  "MA1145.2", "MA1634.2", "MA0835.3", "MA1951.2", "MA1988.2"
)

# --- HFD-specific peaks: AP-1 from full motif scan ---
hfd_full <- read_tsv(
  "results/motif_analysis/full_motif_scan/HFD_specific/HFD_specific_per_peak_all_motifs.tsv",
  show_col_types = FALSE
)
ap1_cols_hfd <- intersect(ap1_ids, names(hfd_full))
hfd_ap1 <- hfd_full |>
  select(peak, all_of(ap1_cols_hfd)) |>
  mutate(has_ap1 = rowSums(across(all_of(ap1_cols_hfd))) > 0) |>
  select(peak, has_ap1)

hfd_classified <- hfd_ap1 |>
  left_join(hfd_gr_classified |> select(peak, has_gr), by = "peak") |>
  mutate(
    motif_cat = case_when(
      has_ap1 &  has_gr ~ "Both (AP-1 + GR)",
      has_ap1 & !has_gr ~ "AP-1 only",
     !has_ap1 &  has_gr ~ "GR only",
      TRUE              ~ "Neither"
    ),
    peak_set = "HFD-specific\n(n = 6,900)"
  )

# --- Shared peaks: AP-1 from pre-computed FIMO scan ---
shared_ap1_file <- "results/motif_analysis/composite_scan/shared/shared_ap1_per_peak.tsv"

shared_gr_raw <- read_tsv(
  "results/motif_analysis/composite_scan/shared/shared_per_peak_motifs.tsv",
  show_col_types = FALSE
)
gr_cols_shared <- intersect(gr_motif_ids, names(shared_gr_raw))
shared_gr_classified <- shared_gr_raw |>
  mutate(has_gr = rowSums(across(all_of(gr_cols_shared))) > 0)

if (file.exists(shared_ap1_file)) {
  shared_ap1 <- read_tsv(shared_ap1_file, show_col_types = FALSE)
  shared_classified <- shared_ap1 |>
    left_join(shared_gr_classified |> select(peak, has_gr), by = "peak") |>
    mutate(
      motif_cat = case_when(
        has_ap1 &  has_gr ~ "Both (AP-1 + GR)",
        has_ap1 & !has_gr ~ "AP-1 only",
       !has_ap1 &  has_gr ~ "GR only",
        TRUE              ~ "Neither"
      ),
      peak_set = sprintf("Shared\n(n = %s)", format(nrow(shared_gr_classified), big.mark=","))
    )

  # --- Combine and plot ---
  cat_order <- c("Both (AP-1 + GR)", "GR only", "AP-1 only", "Neither")
  cat_colors <- c(
    "Both (AP-1 + GR)" = "#d62728",
    "GR only"          = "#ff7f0e",
    "AP-1 only"        = "#1f77b4",
    "Neither"          = "#d3d3d3"
  )

  all_peaks <- bind_rows(hfd_classified, shared_classified) |>
    count(peak_set, motif_cat) |>
    group_by(peak_set) |>
    mutate(
      pct      = 100 * n / sum(n),
      motif_cat = factor(motif_cat, levels = cat_order)
    ) |>
    ungroup() |>
    mutate(peak_set = factor(peak_set,
                             levels = c("HFD-specific\n(n = 6,900)",
                                        sprintf("Shared\n(n = %s)",
                                                format(nrow(shared_gr_classified), big.mark=",")))))

  # Label only categories ≥ 1%
  label_df <- all_peaks |> filter(pct >= 1)

  ggplot(all_peaks, aes(x = peak_set, y = pct, fill = motif_cat)) +
    geom_col(position = "stack", width = 0.55) +
    geom_text(data = label_df,
              aes(label = sprintf("%.1f%%", pct)),
              position = position_stack(vjust = 0.5),
              size = 3.2, colour = "white", fontface = "bold") +
    scale_fill_manual(values = cat_colors, name = "Motif content") +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 101)) +
    labs(
      title    = "AP-1 + GR composite motif content by peak class",
      subtitle = "FIMO p < 1×10⁻⁴; AP-1 = 35 JUN/FOS/BATF motifs; GR = NR3C1/Pgr/PGR",
      x = NULL, y = "Percentage of peaks (%)"
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 11))

} else {
  message("Pre-computed shared AP-1 scan not found at:\n  ", shared_ap1_file,
          "\nRun the FIMO step first (see Section 7 setup notes).")
}
```

```{r ap1-gr-fisher}
# Fisher exact test: enrichment of AP-1+GR co-occurrence in HFD vs shared peaks
if (file.exists(shared_ap1_file) && exists("shared_classified")) {
  hfd_both   <- sum(hfd_classified$motif_cat == "Both (AP-1 + GR)")
  hfd_n      <- nrow(hfd_classified)
  shared_both <- sum(shared_classified$motif_cat == "Both (AP-1 + GR)")
  shared_n    <- nrow(shared_classified)

  tab <- matrix(c(hfd_both,    hfd_n    - hfd_both,
                  shared_both, shared_n - shared_both), nrow = 2)
  ft  <- fisher.test(tab, alternative = "greater")

  cat(sprintf(
    "AP-1 + GR co-occurrence:\n  HFD-specific: %d / %d (%.1f%%)\n  Shared:       %d / %d (%.1f%%)\n  OR = %.2f, p = %.2e\n",
    hfd_both, hfd_n, 100*hfd_both/hfd_n,
    shared_both, shared_n, 100*shared_both/shared_n,
    ft$estimate, ft$p.value
  ))
}
```

---

## 8. Locus-level chromatin at GR-target genes

The GR-sensitisation model makes a testable prediction: genes that are
transcriptionally upregulated in HFD (Sgk1, Angptl4, Lep) should have nearby
HFD-opened ATAC peaks, ideally carrying AP-1 and/or GR motifs. Genes whose
expression does not increase (Pnpla2) should lack this chromatin signature.
Fkbp5 is the inverse case — it is CHD-enriched at both chromatin and RNA levels.

```{r locus-setup}
#| include: false

select <- dplyr::select

# Re-use cnts_long and atac_counts from Section 6.1a if already in environment;
# otherwise rebuild (allows section to run standalone).
if (!exists("atac_counts")) {
  cnts_mat  <- read.table("results/deseq2/deseq2_normalized_counts.txt",
                          header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)
  cnts_long <- cnts_mat |>
    rownames_to_column("peak_id") |>
    pivot_longer(-peak_id, names_to = "sample", values_to = "norm_count") |>
    mutate(condition = ifelse(grepl("^CHD", sample), "CHD", "HFD"))
  atac_counts <- read_tsv("results/deseq2/deseq2_results.txt", show_col_types = FALSE) |>
    left_join(cnts_long, by = "peak_id") |>
    mutate(mid_kb = (start + end) / 2e3)
}

# AP-1 and GR motif annotations from the full motif scan (HFD-specific)
ap1_ids <- c("MA0099.4","MA0462.3","MA0476.2","MA0477.3","MA0478.2",
             "MA0488.2","MA0489.3","MA0490.3","MA0491.3","MA0492.2",
             "MA0605.3","MA1126.2","MA1127.1","MA1128.2","MA1129.1",
             "MA1130.2","MA1131.2","MA1132.2","MA1133.2","MA1134.2",
             "MA1135.2","MA1136.1","MA1137.2","MA1138.2","MA1139.2",
             "MA1140.3","MA1141.2","MA1142.2","MA1143.2","MA1144.2",
             "MA1145.2","MA1634.2","MA0835.3","MA1951.2","MA1988.2")
gr_ids <- c("MA0113.4", "MA2323.1", "MA2327.1")

hfd_full_motif <- read_tsv(
  "results/motif_analysis/full_motif_scan/HFD_specific/HFD_specific_per_peak_all_motifs.tsv",
  show_col_types = FALSE)
hfd_gr_motif <- read_tsv(
  "results/motif_analysis/composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv",
  show_col_types = FALSE)

ap1_cols_m <- intersect(ap1_ids, names(hfd_full_motif))
gr_cols_m  <- intersect(gr_ids,  names(hfd_gr_motif))

hfd_motif_ann <- hfd_full_motif |>
  mutate(has_ap1 = rowSums(across(all_of(ap1_cols_m))) > 0) |>
  select(peak, has_ap1) |>
  left_join(
    hfd_gr_motif |>
      mutate(has_gr = rowSums(across(all_of(gr_cols_m))) > 0) |>
      select(peak, has_gr),
    by = "peak"
  ) |>
  mutate(
    motif_cat = case_when(
      has_ap1 & has_gr  ~ "AP-1 + GR",
      has_ap1 & !has_gr ~ "AP-1 only",
     !has_ap1 & has_gr  ~ "GR only",
      TRUE              ~ "Neither"
    )
  )

# ── Shared helper: build locus count table with motif annotations ──────────
make_locus_df <- function(chr_in, s, e) {
  # Identify HFD-specific peaks at this locus and get motif annotation
  hfd_peaks_here <- atac_counts |>
    filter(chr == chr_in, start >= s, end <= e,
           !is.na(padj), padj < 0.05, log2FoldChange > 1) |>
    distinct(peak_id, chr, start, end) |>
    mutate(peak_500 = {
      ctr <- as.integer((start + end) / 2)
      paste0(chr, ":", ctr - 250L, "-", ctr + 250L)
    })

  ldf <- atac_counts |>
    filter(chr == chr_in, start >= s, end <= e) |>
    group_by(peak_id, chr, start, end, log2FoldChange, padj, mid_kb) |>
    summarise(
      CHD    = mean(norm_count[condition == "CHD"]),
      HFD    = mean(norm_count[condition == "HFD"]),
      CHD_se = sd(norm_count[condition == "CHD"]) / sqrt(3),
      HFD_se = sd(norm_count[condition == "HFD"]) / sqrt(3),
      .groups = "drop"
    ) |>
    mutate(
      sig = case_when(padj < 0.001 ~ "***", padj < 0.01 ~ "**",
                      padj < 0.05  ~ "*",   TRUE         ~ ""),
      peak_label = sprintf("%.0f", mid_kb),
      # Motif annotation: look up 500-bp window peak ID
      peak_500 = {
        ctr <- as.integer((start + end) / 2)
        paste0(chr, ":", ctr - 250L, "-", ctr + 250L)
      }
    ) |>
    left_join(hfd_motif_ann |> select(peak, motif_cat),
              by = c("peak_500" = "peak")) |>
    mutate(motif_cat = replace_na(motif_cat, "Not HFD-specific"))
  ldf
}

# Colour mapping for motif categories
motif_colors <- c(
  "AP-1 + GR"       = "#d62728",
  "AP-1 only"        = "#1f77b4",
  "GR only"          = "#ff7f0e",
  "Neither"          = "#9467bd",
  "Not HFD-specific" = "#cccccc"
)

# ── Shared helper: draw the locus ATAC panel ─────────────────────────────
plot_locus <- function(ldf, gene_label, rna_lfc = NULL, rna_padj = NULL) {

  ldf_long <- ldf |>
    pivot_longer(c(CHD, HFD), names_to = "condition", values_to = "mean_cnt") |>
    mutate(se = ifelse(condition == "CHD", CHD_se, HFD_se),
           condition = factor(condition, levels = c("CHD","HFD")),
           peak_label = fct_reorder(peak_label, mid_kb))

  sig_df <- ldf_long |>
    filter(condition == "HFD", sig != "") |>
    mutate(y_pos = mean_cnt + se + max(ldf_long$mean_cnt) * 0.08)

  rna_sub <- if (!is.null(rna_lfc)) {
    rna_star <- case_when(rna_padj < 0.001 ~ "***", rna_padj < 0.01 ~ "**",
                          rna_padj < 0.05  ~ "*",   TRUE             ~ "ns")
    sprintf("RNA-seq: LFC = %+.2f (%s)", rna_lfc, rna_star)
  } else ""

  # Colour bars by motif category (HFD direction peaks only; shared = grey)
  ggplot(ldf_long, aes(x = peak_label, y = mean_cnt,
                        fill = interaction(condition, motif_cat))) +
    geom_col(position = position_dodge(0.75), width = 0.7,
             aes(alpha = condition)) +
    geom_errorbar(aes(ymin = mean_cnt - se, ymax = mean_cnt + se),
                  position = position_dodge(0.75), width = 0.25,
                  colour = "grey30") +
    geom_text(data = sig_df,
              aes(x = peak_label, y = y_pos, label = sig),
              inherit.aes = FALSE, size = 4) +
    scale_fill_manual(
      values = setNames(
        rep(unname(motif_colors), each = 2),
        paste0(rep(c("CHD","HFD"), times = length(motif_colors)), ".",
               rep(names(motif_colors), each = 2))
      ),
      guide = "none"
    ) +
    scale_alpha_manual(values = c(CHD = 0.55, HFD = 1.0),
                       guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(
      title    = sprintf("%s locus — ATAC-seq normalised counts", gene_label),
      subtitle = paste0(rna_sub,
        "\nBar colour: CHD (faded) and HFD (solid); ",
        "red = AP-1+GR composite, blue = AP-1, orange = GR, grey = no HFD-specific peak"),
      x = "Peak mid-point (kb)", y = "Mean norm. count ± SE"
    ) +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```

### 8.1 Sgk1 — HFD-opened AP-1+GR composite enhancer

```{r sgk1-locus}
#| fig-width: 8
#| fig-height: 4

rna_sgk1 <- tg |> filter(gene == "Sgk1")
sgk1_df  <- make_locus_df("chr10", 21850000, 22080000)
plot_locus(sgk1_df, "Sgk1",
           rna_lfc  = rna_sgk1$log2FoldChange,
           rna_padj = rna_sgk1$padj)
```

```{r sgk1-verdict}
sgk1_hfd_peaks <- sgk1_df |> filter(log2FoldChange > 1, !is.na(padj), padj < 0.05)
cat(sprintf("Sgk1 locus HFD-specific peaks (padj<0.05, LFC>1): %d\n",
            nrow(sgk1_hfd_peaks)))
if (nrow(sgk1_hfd_peaks) > 0)
  cat(sprintf("  chr10:%d-%d  LFC=%.2f  padj=%.3f  motif: %s\n",
              sgk1_hfd_peaks$start, sgk1_hfd_peaks$end,
              sgk1_hfd_peaks$log2FoldChange, sgk1_hfd_peaks$padj,
              sgk1_hfd_peaks$motif_cat))
```

> **chr10:21,986,642–21,987,014** (LFC = +1.60, padj = 0.008) carries **both AP-1
> and GR motifs** — the composite enhancer predicted by the pioneer model.
> Sgk1 is significantly up in HFD TRAP-seq, making this the clearest locus-level
> confirmation of the AP-1 pioneer → GR binding → transcription axis.

### 8.2 Angptl4 — GR target locus

```{r angptl4-locus}
#| fig-width: 7
#| fig-height: 4

rna_ang <- tg |> filter(gene == "Angptl4")
ang_df  <- make_locus_df("chr17", 33700000, 33760000)
plot_locus(ang_df, "Angptl4",
           rna_lfc  = rna_ang$log2FoldChange,
           rna_padj = rna_ang$padj)
```

> Angptl4 is strongly up in HFD TRAP-seq (LFC ≈ +1.95) but nearby ATAC peaks
> are all CHD-enriched or unchanged. The responsible GR enhancer may lie
> outside the ±30 kb window shown; distal GREs for Angptl4 have been reported
> up to 100 kb upstream in other cell types.

### 8.3 Lep — leptin locus

```{r lep-locus}
#| fig-width: 7
#| fig-height: 4

rna_lep <- tg |> filter(gene == "Lep")
lep_df  <- make_locus_df("chr6", 28970000, 29050000)
plot_locus(lep_df, "Lep",
           rna_lfc  = rna_lep$log2FoldChange,
           rna_padj = rna_lep$padj)
```

> Lep (leptin) is the most induced transcript in HFD (LFC ≈ +2.01,
> padj = 0.014). No significant ATAC changes within the immediate locus window;
> leptin regulation is known to involve distal fat-specific enhancers.

### 8.4 Pnpla2 — chromatin closes despite GR-target classification

```{r pnpla2-locus}
#| fig-width: 9
#| fig-height: 4.5

rna_pnp <- tg |> filter(gene == "Pnpla2")
pnp_df  <- make_locus_df("chr7", 141370000, 141450000)
plot_locus(pnp_df, "Pnpla2",
           rna_lfc  = rna_pnp$log2FoldChange,
           rna_padj = rna_pnp$padj)
```

```{r pnpla2-verdict}
pnp_sig <- pnp_df |> filter(!is.na(padj), padj < 0.05)
cat(sprintf("Pnpla2 locus peaks: %d total, %d significant\n",
            nrow(pnp_df), nrow(pnp_sig)))
cat(sprintf("  All significant peaks CHD-enriched: %s\n",
            all(pnp_sig$log2FoldChange < 0)))
cat(sprintf("  Significant CHD-up peaks: %d  (none are HFD-up)\n",
            sum(pnp_sig$log2FoldChange < 0)))
cat(sprintf("  RNA LFC = %.2f, padj = %.3f\n",
            rna_pnp$log2FoldChange, rna_pnp$padj))
```

> **Pnpla2 (ATGL)** shows **6 / 12 significant CHD-enriched peaks** at its
> locus — chromatin closes in HFD — yet the transcript is not significantly
> changed (LFC = +0.32, padj = 0.73). Possible explanations:
>
> 1. **Fkbp5/GR model:** constitutive GR activation (via Fkbp5 loss) could
>    maintain Pnpla2 expression even as local chromatin accessibility declines,
>    through a distal GRE or increased GR nuclear occupancy.
> 2. **Uncoupled regulation:** chromatin changes at the Pnpla2 locus are a
>    consequence of AP-1 remodelling at nearby HFD-specific sites rather than
>    a cause of altered Pnpla2 expression.
> 3. **Compensation:** opposing signals (GR induction vs local chromatin
>    closing) may balance, keeping transcript levels stable.
>
> Pnpla2 is **not chromatin-driven in this dataset**; its classification as
> a GR target rests on the transcriptome rather than ATAC evidence.

---

## 9. BigWig locus tracks (requires server BAM files)

When ATAC-seq and RNA-seq BAM files are available locally, running the Nextflow
pipeline (or the BigWig steps in isolation) will produce:

```
results/bigwig/atac/
  CHD_1.RPGC.bw   CHD_2.RPGC.bw   CHD_3.RPGC.bw
  HFD_1.RPGC.bw   HFD_2.RPGC.bw   HFD_3.RPGC.bw

results/rnaseq/bigwig/
  CHD_1.CPM.bw    CHD_2.CPM.bw    CHD_3.CPM.bw
  HFD_1.CPM.bw    HFD_2.CPM.bw    HFD_3.CPM.bw
```

Place these files in the paths above relative to the project root, then the
chunk below will render IGV-style locus tracks using `rtracklayer`. **The peak-count
bar charts in Sections 6.1a and 8.1–8.4 remain valid and complementary regardless.**

```{r bigwig-locus-tracks}
#| fig-width: 10
#| fig-height: 9
#| eval: false   # ← change to true once BigWigs are in place

bw_atac <- list.files("results/bigwig/atac",   pattern="\\.bw$", full.names=TRUE)
bw_rna  <- list.files("results/rnaseq/bigwig", pattern="\\.bw$", full.names=TRUE)

if (length(bw_atac) == 6 && length(bw_rna) == 6) {
  suppressPackageStartupMessages({
    library(rtracklayer)
    library(GenomicRanges)
  })

  # Loci to visualise
  loci_bw <- list(
    Fkbp5   = GRanges("chr4",  IRanges(99900000, 100100000)),
    Sgk1    = GRanges("chr10", IRanges(21850000,  22080000)),
    Angptl4 = GRanges("chr17", IRanges(33700000,  33760000)),
    Lep     = GRanges("chr6",  IRanges(28970000,  29050000)),
    Pnpla2  = GRanges("chr7",  IRanges(141370000, 141450000))
  )

  import_mean <- function(files, cond, roi) {
    mats <- lapply(files[grepl(cond, files)], function(f)
      as.numeric(import(f, which=roi, as="NumericList")[[1]]))
    Reduce("+", mats) / length(mats)
  }

  # Plot one locus (example: Sgk1)
  roi <- loci_bw[["Sgk1"]]
  positions <- seq(start(roi), end(roi))
  chd_cov <- import_mean(bw_atac, "CHD", roi)
  hfd_cov <- import_mean(bw_atac, "HFD", roi)
  cov_df  <- tibble(pos = positions,
                    CHD = chd_cov[seq_along(positions)],
                    HFD = hfd_cov[seq_along(positions)]) |>
    pivot_longer(c(CHD,HFD), names_to="condition", values_to="cov") |>
    mutate(condition = factor(condition, c("CHD","HFD")))

  ggplot(cov_df, aes(x=pos/1e3, y=cov, fill=condition)) +
    geom_area(alpha=0.6, position="identity") +
    scale_fill_manual(values=color_scheme) +
    facet_wrap(~condition, ncol=1) +
    labs(title="Sgk1 locus — ATAC BigWig coverage", x="Position (kb)", y="RPGC") +
    theme_classic(base_size=11) + theme(legend.position="none")
} else {
  message(sprintf(
    "BigWig files not found.\n  ATAC: %d/6 found in results/bigwig/atac/\n  RNA:  %d/6 found in results/rnaseq/bigwig/\nSee README §'AP-1 + GR Composite Peak Scan' for file naming.",
    length(bw_atac), length(bw_rna)
  ))
}
```

---

## Session information

```{r session-info}
sessionInfo()
```
