---
title: "Effects of Dexamethasone on NCD/HFD Mouse Quadriceps"
author: "Laura Gunder and Dave Bridges"
date: "January 8, 2018"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('black', 'red')

library(ggplot2)
#sets default theme
theme_set(theme_bw())

#to change colors use
#scale_color_manual(values = color.scheme)
#scale_fill_manual(values = color.scheme)
```

# Purpose
to examine the molecular effects of dexamethasone at varying lengths of treatment on obese and non-obese mice 


# Experimental Details


The relevant protocols used were:

* RNA Purification, a modified version of http://bridgeslab.sph.umich.edu/protocols/index.php?title=Preparation_of_RNA_Samples_from_Mouse_Tissues&oldid=1359
* cDNA Synthesis, used 2 ug then diluted cDNA 5X prior to qPCR: http://bridgeslab.sph.umich.edu/protocols/index.php?title=First_Strand_cDNA_Synthesis_(AB_Kit)&oldid=1242
* qPCR: http://bridgeslab.sph.umich.edu/protocols/index.php?title=QPCR&oldid=1252


# Raw Data

```{r datafiles}
sample.mapping.file <- 'sample-key.csv'
run.1.file <- 'Well Results New.csv'
```

The sample mapping is found in `r sample.mapping.file` and the qPCR data can be found in `r run.1.file`.  This analysis can be found in `r getwd()` and was most recently run on `r date()`.

```{r data-input}
library(readr) #loads the readr package
#load sample mapping file
#first define the columns in the sample key
sample.cols <- cols(
  Mouse = col_factor(levels=NULL),
  Diet = col_factor(levels=c("NCD","HFD")),
  Time = col_integer()
)

sample.mapping <- read_csv(sample.mapping.file,
                           col_types = sample.cols) #manually defining NA values as N/A from input file
#load qPCR result files
run.1.data <- read_csv(file=run.1.file, skip=12,
                       col_types = cols(
                         Cq = col_double(),
                         `Target Name` = col_factor(levels=NULL),
                         `Sample Name` = col_factor(levels=NULL)
                       ))


#combinedinto a single data frame

combined.data <- run.1.data

#indicate unreliable data and remove if necessary

bad.samples <- c(2625,2618,2636,2642) # did not amplify
bad.wells.1 <- c('J1','J6','J7','J9','J11','J12','G7','D13','D16','A15','C24','C17','J10','B21') 
bad.wells.2 <- c('P6','G5','G7','G4','G1','G2','G8','G3','J10','O21')
bad.wells.3 <- c('G7','P16','M13','P15')
bad.wells.4 <- c('B3','G2','F20','C1','C3','C2','C4','F3','H23','G10','E7','F14','G4','F24','G3','E15','D1','G16','F1','G12','D8')
bad.wells.5 <- c('A9')
#outlier technical replicates

combined.data.clean <- filter(combined.data, 
                              !(`Experiment Name`=='2017-12-13 lcg.eds'&Well%in%bad.wells.1))
combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-01-10 195156.eds'&Well%in%bad.wells.2)) #F3
combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-01-17 182258.eds'&Well%in%bad.wells.3))
combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-07-09 LG.eds'&Well%in%bad.wells.4))
combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-07-16 lg.eds'&Well%in%bad.wells.5))
#F3

combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-01-10 195156.eds'&`Target Name`=='Pgk1'))
combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-07-09 lg.eds'&`Target Name`=='Foxo3'))

combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-01-10 195156.eds'&`Target Name`=='Foxo3'))

combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-01-17 182258.eds'&`Target Name`=='Foxo1'))

combined.data.clean <- filter(combined.data.clean, 
                              !(`Experiment Name`=='2018-07-09 lg.eds'&`Target Name`=='GDF15'))

#add in fifth plate
#comment out GDF15 from all plates except 1 (see code in 118) 2018-07-09 LG.eds and new plate
#comment out Foxo3 from all plates except 1 (see code in 118) 2018-07-09 LG.eds earlier and new plate

combined.data.clean <- filter(combined.data.clean, !(`Sample Name`%in%bad.samples))

#merge qPCR results into single dataset
qpcr.data <- 
  full_join(x=combined.data.clean, y=sample.mapping, by=c("Sample Name"="Mouse")) %>% # joining tables by sample column
  select(`Experiment Name`,`Sample Name`,Well,`Target Name`,Cq,Time,Diet) #removed extra columns we didnt need

#combine technical replicates
qpcr.results <-
  qpcr.data %>%
  group_by(`Experiment Name`,`Sample Name`, `Target Name`, Time, Diet) %>%
  summarize(Ct.mean = mean(Cq, na.rm=T),
            Ct.min = min(Cq, na.rm=T),
            Ct.max = max(Cq, na.rm=T)) %>%
  mutate(Ct.range = Ct.max-Ct.min)

qpcr.results.all <-
  qpcr.data %>%
  group_by( `Target Name`) %>%
  summarize(Ct.mean = mean(Cq, na.rm=T),
            Ct.min = min(Cq, na.rm=T),
            Ct.max = max(Cq, na.rm=T)) %>%
  mutate(Ct.range = Ct.max-Ct.min)

#kable(filter(qpcr.results, Ct.range>1) %>% filter(`Target Name` != 'Foxo1') %>% arrange(desc(Ct.range)), caption="Samples where Ct range between technical replicates exceeds 1")

#filter(qpcr.data, `Sample Name`==2645&`Target Name`=="GDF15")
```


We removed the samples `r paste(bad.samples, collapse=",")` from the dataset because they did not amplify.  We removed wells `r paste(bad.wells.1, collapse=",")` in the first experiment and `r paste(bad.wells.2, collapse=",")` in the second experiment because they were outliers relative to the other technical replicates. 

# Analysis

```{r qpcr-analysis}
control.gene <- 'Pgk1'
control.time <- 0
control.diet <- "NCD"

all.samples.rplp0 <- as.character(filter(qpcr.results, `Target Name`==control.gene)$`Sample Name`)

sample.results <-
  filter(qpcr.results, `Sample Name` %in% all.samples.rplp0) %>%
  mutate(Quant = 2^-Ct.mean) %>%
  group_by(`Sample Name`) %>%
  mutate(Quant.norm = Quant/Quant[`Target Name`==control.gene]) %>%
  group_by(`Target Name`) %>%
  mutate(Rel.norm = Quant.norm/mean(Quant.norm[Diet==control.diet&Time==control.time], na.rm=T))

summary.results <-
  sample.results %>%
  group_by(`Target Name`, Diet,Time) %>%
  summarize(Expression=mean(Rel.norm, na.rm=T),
            Expression.se = se(Rel.norm)) 

output_file <- "Analysed Control Gene Data.csv"
write_csv(summary.results, output_file)
```

All mRNA levels were adjusted to **`r control.gene`**, then normalized to **`r control.diet`** at time **`r control.time`**.  Summarized data was written out to `r output_file`.

# Foxo1 versus Foxo3a

```{r foxo-comp}
foxo.data <-
  sample.results %>%
  filter(`Target Name` %in% c('Foxo1','Foxo3')) %>%
  group_by(`Target Name`) %>%
  summarize(Ct = mean(Ct.mean, na.rm=T)) %>%
  mutate(DCt = Ct-Ct[`Target Name`=='Foxo1']) %>%
  mutate(Quant = 2^-DCt)

kable(foxo.data, caption="Relative amplification of Foxo1 and Foxo3")
```

# Plots of changes in gene expression

```{r qpcr-lineplot-all-points}
library(ggplot2)
p <- ggplot(filter(sample.results,`Target Name` != 'GDF15'),
            aes(y=Rel.norm, x=Time))
p + geom_smooth(aes(color=Diet), se=F) +
  geom_point(aes(color=Diet)) +
  labs(y="Relative Expression",x="Time (days of Dexamethasone)") +
  facet_grid(~`Target Name`) +
  scale_color_manual(values = color.scheme) +
  scale_fill_manual(values = color.scheme)
```

```{r qpcr-lineplot-all}
p <- ggplot(summary.results,
            aes(y=Expression, x=Time, ymin=Expression - Expression.se, ymax=Expression + Expression.se,))
p + geom_line(aes(color=Diet)) + 
  geom_errorbar(aes(color=Diet)) +
  labs(y="Relative Expression",x="Time (days of Dexamethasone)") +
  facet_grid(~`Target Name`)  + 
  scale_color_manual(values = color.scheme) +
  scale_fill_manual(values = color.scheme)
```

```{r qpcr-lineplot-atrogenes}
p <- ggplot(filter(summary.results, `Target Name` %in% c('Nr3c1','Foxo1','Foxo3','Fbxo32','Trim63')),
            aes(y=Expression, x=Time, ymin=Expression - Expression.se, ymax=Expression + Expression.se,))
p + geom_line(aes(color=Diet)) + 
  geom_errorbar(aes(color=Diet)) +
  labs(y="Relative Expression",x="Time (days of Dexamethasone)") +
  facet_grid(~`Target Name`) + 
  scale_color_manual(values = color.scheme) +
  scale_fill_manual(values = color.scheme)

```

## Statistics

### Baseline Effects of Diet

First tested if there was an effect at baseline for expression of *Trim63* and *Fbxo32*.  

```{r baseline}
library(car)

baseline.stats <- 
  sample.results %>%
  filter(`Target Name` %in% c('Nr3c1','Foxo1','Foxo3','Fbxo32','Trim63','GDF15')) %>%
  filter(Time==0) %>%
  group_by(`Target Name`) %>%
  summarize(HFD.Effect = mean(Rel.norm[Diet=="HFD"])/mean(Rel.norm[Diet=="NCD"]),
            Shapiro.p = min(shapiro.test(Rel.norm[Diet=="HFD"])$p.value,
                            shapiro.test(Rel.norm[Diet=="NCD"])$p.value),
            Levene.p = leveneTest(Rel.norm~Diet)$"Pr(>F)"[1],
            Wilcox.p = wilcox.test(Rel.norm~Diet, var.equal=T)$p.value,
            Welch.p = t.test(Rel.norm~Diet, var.equal=F)$p.value,
            Student.p = t.test(Rel.norm~Diet, var.equal=T)$p.value)

kable(baseline.stats, caption="Pairwise statistics for baseline effects of diet.")

```

### Effects of Dexamethasone Over Time

First did pairwise tests at each time point.

```{r dex-time-stats}
dex.stats <- 
  sample.results %>%
  filter(`Target Name` %in% c('Nr3c1','Foxo3','Fbxo32','Trim63')) %>%
  filter(Time!=0) %>%
  group_by(Time,`Target Name`) %>%
  summarize(HFD.Effect = mean(Rel.norm[Diet=="HFD"])/mean(Rel.norm[Diet=="NCD"]),
            Shapiro.p.HFD = shapiro.test(Rel.norm[Diet=="HFD"])$p.value,
            Shaprio.p.NCD = shapiro.test(Rel.norm[Diet=="NCD"])$p.value,
            Shapiro.p = min(shapiro.test(Rel.norm[Diet=="HFD"])$p.value,
                            shapiro.test(Rel.norm[Diet=="NCD"])$p.value),
            Levene.p = leveneTest(Rel.norm~Diet)$"Pr(>F)"[1],
            Wilcox.p = wilcox.test(Rel.norm~Diet, var.equal=T)$p.value,
            Welch.p = t.test(Rel.norm~Diet, var.equal=F)$p.value,
            Student.p = t.test(Rel.norm~Diet, var.equal=T)$p.value)

kable(dex.stats, caption="Pairwise statistics for effects of diet in dexamethasone treated animals.")
```

Also analysed this with a linear model with diet and time as covariates, allowing for an interaction.  

```{r time.lm}
library(broom)
lm(Rel.norm ~ Diet * Time, data=filter(sample.results, `Target Name`=='Fbxo32')) %>% tidy %>% kable(caption="Linear model for effects of time and diet on Fbxo32")
lm(Rel.norm ~ Diet * Time, data=filter(sample.results, `Target Name`=='Trim63')) %>% tidy %>% kable(caption="Linear model for effects of time and diet on Trim63")
lm(Rel.norm ~ Diet * Time, data=filter(sample.results, `Target Name`=='Nr3c1')) %>% tidy %>% kable(caption="Linear model for effects of time and diet on Nr3c1")
lm(Rel.norm ~ Diet * Time, data=filter(sample.results, `Target Name`=='Foxo1')) %>% tidy %>% kable(caption="Linear model for effects of time and diet on Foxo1")
lm(Rel.norm ~ Diet * Time, data=filter(sample.results, `Target Name`=='Foxo3')) %>% tidy %>% kable(caption="Linear model for effects of time and diet on Foxo3")

```

# Interpretation

None of these time courses have a significant interaction between time and diet


# Session Information

```{r session-information, echo=T}
sessionInfo()
```