---
title: "Analysis of Water and Dexamethasone Intake in NCD/HFD Mice"
author: "Laura Gunder and Dave Bridges"
date: "January 6, 2019"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')

# for knitcitations
library("knitcitations")
cleanbib()
options("citation_format" = "pandoc")
```

# Purpose

# Experimental Details

```{r dex-conc}
dex.conc <- 3.78 #ug/mL based on 0.054g/L and 70mg/g of dex powder
```

Dexamethasone was dissolved in water at a concentration of `r dex.conc` ug/mL and animals were given *ad libitum* access.  Water bottle volumes were determined weekly.

# Raw Data

The input file contains tracked water and dexamethasone amounts per week.  The data includes the staring and ending volumes each week and a calculation of animals per week.

```{r data-input}
library(readr) #loads the readr package
filename <- '../../data/raw/Dex and Water Intake NCD HFD.csv' #make this a separate line, you can use any variable you want

#this loads whatever the file is into a dataframe called exp.data if it exists
exp.data <- 
  read_csv(filename, col_types = cols(
  `CAGE ID` = col_factor(levels=NULL),
  WEEK = col_integer(),
  TREATMENT = col_factor(levels=NULL),
  DIET = col_factor(levels=c("NCD","HFD")),
  `NUMBER OF ANIMALS IN CAGE` = col_double(),
  `START ML` = col_double(),
  `END ML` = col_double(),
  `WEEKLY AMOUNT CONSUMED` = col_double(),
  COHORT = col_factor(levels=NULL),
  `ML/ANIMAL/DAY` = col_double(),
  X11 = col_character()
))

```

These data can be found in **`r getwd()`** in a file named **`r ifelse(filename %in% dir(), filename, "no file found")`**.  This script was most recently updated on **`r date()`**.

# Analysis

## Weekly Analysis

```{r weekly-lineplot}
library(forcats)
exp.data <-
  exp.data %>%
  #filter(WEEK<=2) %>%
  filter(COHORT %in% c('LG','TC')) %>%
  mutate(Dexamethasone = ifelse(TREATMENT=="DEX", `ML/ANIMAL/DAY`*dex.conc, 0)) %>%
  rename(Fluid=`ML/ANIMAL/DAY`) %>%
  mutate(Diet = fct_recode(DIET, "Normal Chow Diet" = "NCD", "High Fat Diet" = "HFD"),
         Treatment = fct_recode(TREATMENT, "Water"="WATER", "Dexamethasone" = "DEX"))

library(ggplot2)

p <- ggplot(exp.data, aes(y=Fluid, x=WEEK))
p + geom_point(aes(col=Diet)) +
  facet_grid(~Treatment) +
  geom_smooth(method="loess", aes(col=Diet)) +
  labs(y="Fluid Intake (mL/day/mouse)", title="Total Water Consumption")


weekly.summary.data <-
  exp.data %>%
  group_by(Diet, Treatment, WEEK) %>%
  summarize(Intake.mean = mean(Dexamethasone,na.rm=T),
            Intake.se = se(Dexamethasone)) 
  
  
ymax <- with(weekly.summary.data, max(Intake.mean + Intake.se, na.rm=T))
plot <- with(filter(weekly.summary.data, Treatment=="Dexamethasone"&Diet=="High Fat Diet"),
     plot(WEEK,Intake.mean,
            pch=19,
            type="l",
            las=1,
          col="red",
          lty=2,
          ylim=c(0,ymax),
            ylab="Dexamethasone Intake (ug/day/mouse)",
            xlab="Weeks of Dexamethasone Treatment"))

with(filter(weekly.summary.data, Treatment=="Dexamethasone"&Diet=="Normal Chow Diet"),lines(WEEK, Intake.mean, col="red",lty=1))

with(filter(weekly.summary.data, Treatment=="Dexamethasone"&Diet=="High Fat Diet"), superpose.eb(WEEK,Intake.mean,Intake.se, col="red"))
with(filter(weekly.summary.data, Treatment=="Dexamethasone"&Diet=="Normal Chow Diet"), superpose.eb(WEEK,Intake.mean,Intake.se, col="red"))

legend("topleft", c("Normal Chow Diet","High Fat Diet"), lty=c(1,2), bty="n", col="red")
```

## Average Analysis

```{r average-lineplot}
aggregated.data <- 
  exp.data %>%
  filter(WEEK<=2) %>%
  #filter(COHORT %in% c('LG','TC')) %>%
  group_by(`CAGE ID`, Diet,Treatment) %>%
  summarize(Intake = mean(Dexamethasone)) %>%
  group_by(Diet, Treatment) %>%
  filter(Treatment=="Dexamethasone") %>%
  summarize(Intake.mean = mean(Intake),
            Intake.se = se(Intake),
            Shapiro = shapiro.test(Intake)$p.value,
            n = length(Intake)) 
kable(aggregated.data, caption="Average dexamethasone intake per group, averaged accross the experiment")


aggregated.fluid.data <- 
  exp.data %>%
  filter(WEEK<=2) %>%
  #filter(COHORT %in% c('LG','TC')) %>%
  group_by(`CAGE ID`, Diet,Treatment) %>%
  summarize(Intake = mean(Fluid, na.rm=T)) %>%
  group_by(Diet, Treatment) %>%
  summarize(Intake.mean = mean(Intake, na.rm=T),
            Intake.se = se(Intake),
            #Shapiro = shapiro.test(Intake)$p.value,
            n = length(Intake)) 
kable(aggregated.fluid.data, caption="Average fluid intake per group, averaged accross the experiment")

agg.data.ymax <- with(aggregated.data, max(Intake.mean+Intake.se, na.rm=T))

plot <- with(filter(aggregated.data, Treatment=="Dexamethasone"),
             barplot(Intake.mean,
                     las=1,
                     xlab="Diet",
                     ylim=c(0,agg.data.ymax),
                     names.arg=c("Normal Chow Diet","High Fat Diet"),
                     ylab="Dexamethasone Intake (ug/mouse/day"))

with(filter(aggregated.data, Treatment=="Dexamethasone"),superpose.eb(plot, Intake.mean, Intake.se))

library(car)
stats.data <- 
  exp.data %>%
  filter(WEEK<=2) %>%
  #filter(COHORT %in% c('LG','TC')) %>%
  group_by(`CAGE ID`, Diet,Treatment) %>%
  summarize(Intake = mean(Dexamethasone)) %>%
  filter(Treatment=="Dexamethasone")

library(broom)
leveneTest(Intake~Diet, data=stats.data) %>% tidy %>% kable(caption="Levene's test")
t.test(Intake~Diet, data=stats.data, var.equal=T) %>% tidy %>% kable(caption="Student's t-test")
```


# Session Information

```{r session-information, echo=T}
sessionInfo()
```


