---
title: "Generation of UCSC formatted BigBed files"
author: "Dave Bridges"
date: "2025-11-22"
editor: source
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

Convert into excel files into BED files

## Experimental Details


## Raw Data

Data from Ying *et al.,* 2010 at http://dx.doi.org/10.1371/journal.pone.0015188

```{r data-input}
library(readxl) #loads the readr package
filename <- "pone.0015188.s001.xls" #input file(s)

exp.data <- read_excel(filename, skip=1)
#BiocManager::install("rtracklayer")
library(rtracklayer)
bed <- exp.data |>
  mutate(chrom= `chromosome number`,
         chromsStart = startpeak - 1,  # convert to 0-based
         chromEnd   = endpeak,
         name = `user defined string`,
         score = 0,
         strand = '.') |>
  select(chrom, chromsStart, chromEnd, name, score, strand) |>
  na.omit()

bed.filename <- "pone.0015188.s001.bed"
write_tsv(bed, bed.filename, col_names = FALSE)
```

Had to convert this from mm9 to mm10 by liftover at https://genome.ucsc.edu/cgi-bin/hgLiftOver.  

Then converted to bigbed with this command

```{r GSE236575-bed-files}
read_tsv('../GSE236575/results/motif_analysis/bed_files/HFD_specific_chr.bed', 
         col_names = c('chrom','chromStart','chromEnd')) |>
  mutate(name = paste0('HFD_specific_', row_number()),
         score = 0,
         strand = '.') |>
  write_tsv('GSE236575_HFD_specific.bed', col_names = FALSE)

read_tsv('../GSE236575/results/motif_analysis/bed_files/CHD_specific_chr.bed', 
         col_names = c('chrom','chromStart','chromEnd')) |>
  mutate(name = paste0('NCD_specific_', row_number()),
         score = 0,
         strand = '.') |>
  write_tsv('GSE236575_CHD_specific.bed', col_names = FALSE)

read_tsv('../GSE236575/results/motif_analysis/bed_files/shared_chr.bed', 
         col_names = c('chrom','chromStart','chromEnd')) |>
  mutate(name = paste0('shared_', row_number()),
         score = 0,
         strand = '.') |>
  write_tsv('GSE236575_shared.bed', col_names = FALSE)

```

## Create BigBed files

` bedToBigBed -type=bed6 -sort pone.0015188.s001_mm10.bed mm10.chrom.sizes pone.0015188.s001.bb`
`bedToBigBed -type=bed6 -sort GSE236575_shared.bed mm10.chrom.sizes GSE236575_shared.bb`
`bedToBigBed -type=bed6 -sort GSE236575_CHD_specific.bed mm10.chrom.sizes GSE236575_CHD_specific.bb`
`bedToBigBed -type=bed6 -sort GSE236575_HFD_specific.bed mm10.chrom.sizes GSE236575_HFD_specific.bb`

## Peak Overlap

```{r peak-overlap}
library(rtracklayer)
library(GenomicRanges)
library(dplyr)

# 1. Read your two BED files (automatically handles .bed, .bed.gz, etc.)
hfd.atac_peaks  <- import("GSE236575_HFD_specific.bed")   # your ATAC peaks
ncd.atac_peaks  <- import("GSE236575_CHD_specific.bed")   # your ATAC peaks
chip_peaks  <- import("pone.0015188.s001_mm10.bed")

# 2. Make sure they are proper GRanges with names (optional but very useful)
# If your BED files have a 4th column with peak IDs, they will be used as names
# Otherwise we give them unique IDs:
if (is.null(names(hfd.atac_peaks)))  names(hfd.atac_peaks) <- paste0("ATAC_", 1:length(hfd.atac_peaks))
if (is.null(names(hfd.atac_peaks)))  names(ncd.atac_peaks) <- paste0("ATAC_", 1:length(ncd.atac_peaks))
if (is.null(names(chip_peaks)))  names(chip_peaks) <- paste0("ChIP_", 1:length(chip_peaks))

# 3. Find overlaps
# hits = which ChIP-seq peaks overlap at least one ATAC-seq peak
hfd.hits <- findOverlaps(chip_peaks, hfd.atac_peaks, type = "any", select = "all")
ncd.hits <- findOverlaps(chip_peaks, ncd.atac_peaks, type = "any", select = "all")

# 4. Extract the overlapping ChIP-seq peaks
hfd.chip_peaks_that_overlap_ATAC <- chip_peaks[queryHits(hfd.hits)]
ncd.chip_peaks_that_overlap_ATAC <- chip_peaks[queryHits(ncd.hits)]

# 5. How many ChIP-seq peaks overlap ATAC?
cat(sprintf("%d out of %d ChIP-seq peaks (%.1f%%) overlap at least one ATAC-seq peak from HFD\n",
      length(unique(queryHits(hfd.hits))),
      length(chip_peaks),
      100 * length(unique(queryHits(hfd.hits))) / length(chip_peaks)))

cat(sprintf("%d out of %d ChIP-seq peaks (%.1f%%) overlap at least one ATAC-seq peak from NCD\n",
      length(unique(queryHits(ncd.hits))),
      length(chip_peaks),
      100 * length(unique(queryHits(ncd.hits))) / length(chip_peaks)))

```

## Session Information

```{r session-information}
sessionInfo()
```
