\documentclass[a4paper,12pt]{letter}
\usepackage{times}

% Some of the article customisations are relevant for this class

\name{Dave Bridges} % To be used for the return address on the envelope
\signature{Dave Bridges, Ph.D. \\Assistant Professor \\ UTHSC} % Goes after the closing (ie at the end of the letter, with space for a signature)
%\address{Address \\ of \\ Sender}
% Alternatively, these may be set on an individual basis within each letter environment.

\makelabels % this command prints envelope labels on the final page of the document

\begin{document}
\begin{letter}{Prof. Adrian J L Clark \\
Editor in Chief\\
Journal of Molecular Endocrinology }

\opening{} % eg Hello.

Prof. Clark,

Thank you for considering our manuscript, and providing us with the second round reviewer feedback.  As detailed in our response we have modified our manuscript, addressing the second reviewer's comments, as the first reviewer was satisfied by the previous response.  As the changes were minor and all comprehensively addressed, we believe that this manuscript is now ready for publication.

The changes include an explanation of the statistical barriers set for this manuscript as well as several minor clarifications.  In the last point there is a disagreement between us and the reviewer as to the relevance of Figure 7.  We believe that this is important negative data worthy of inclusion, but if necessary we will move this to the supplement.  If you have any further questions for us, please do not hesitate to ask.


\closing{Regards} % eg Regards,

%\cc{} % people this letter is cc-ed to
%\encl{} % list of anything enclosed
%\ps{} % any post scriptums. ``PS'' labels must be put in manually

\end{letter}
\end{document}
