---
title: "Experimental GR occupancy × HFD differential chromatin"
subtitle: "Is measured adipocyte GR binding enriched in HFD-specific open chromatin (GSE236575)?"
author: "davebrid"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
library(GenomicRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(stringr)
library(purrr)
library(ggplot2)

# AnnotationDbi/GenomicRanges can mask dplyr::select — pin it (see MEMORY)
select <- dplyr::select

set.seed(1)

# ── Paths ────────────────────────────────────────────────────────────────────
atac_dir <- "../snATACseq/GSE236575/results/deseq2"   # GSE236575 ATAC peak classes (mm10)
peak_dir <- "results/peaks"                            # this project's GR peaks
out_dir  <- "results/gr_enrichment"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

# Question

The GSE236575 composite-enhancer model proposed that AP-1 (led by Junb) opens
chromatin on HFD, licensing **predicted** GR binding (NR3C1 motif; Fisher OR = 1.28).
Here we replace predicted GR sites with **experimentally measured** GR occupancy
(re-analyzed adipocyte GR ChIP-seq) and ask:

> Is GR binding enriched in HFD-specific open chromatin, relative to a background of
> shared (non-differential) peaks?

We test this against several GR peak sets of increasing stringency, report both a
frequentist (Fisher) and a Bayesian (beta-binomial posterior OR + 95% CrI) result,
and guard against the peak-width confound with a width-adjusted logistic model.
CHD-specific peaks serve as a negative contrast (the model predicts no GR enrichment
in chromatin that *closes* on HFD).

# Load peaks

```{r readers}
# All BEDs and narrowPeaks are 0-based half-open → +1 to the start for 1-based GRanges.
read_bed3 <- function(path) {
  readr::read_tsv(path, col_names = c("chr", "start", "end"),
                  col_types = "cii", progress = FALSE) |>
    with(GRanges(chr, IRanges(start + 1L, end)))
}

read_narrowpeak <- function(path) {
  cn <- c("chr","start","end","name","score","strand",
          "signal","pval","qval","peak")
  readr::read_tsv(path, col_names = cn,
                  col_types = "ciicicdddi", progress = FALSE) |>
    mutate(strand = "*") |>
    with(GRanges(chr, IRanges(start + 1L, end),
                 signal = signal, qval = qval))
}
```

```{r load-atac}
atac <- list(
  HFD_specific = read_bed3(file.path(atac_dir, "HFD_specific_peaks.bed")),
  CHD_specific = read_bed3(file.path(atac_dir, "CHD_specific_peaks.bed")),
  shared       = read_bed3(file.path(atac_dir, "shared_peaks.bed"))
)
map_int(atac, length)
```

```{r load-gr}
# Pipeline consensus sets (union of IP narrowPeaks, merged). hg38 human is the
# lifted mm10 version.
gr_union <- list(
  eWAT_union  = read_bed3(file.path(peak_dir, "GSE64458_eWAT/GSE64458_eWAT_GR_consensus.bed")),
  iWAT_union  = read_bed3(file.path(peak_dir, "GSE64458_iWAT/GSE64458_iWAT_GR_consensus.bed"))
)

# Human lifted set (use it only if the .mm10.bed has been synced)
human_mm10 <- file.path(peak_dir, "GSE163061_human/GSE163061_human_GR_consensus.mm10.bed")
if (file.exists(human_mm10)) {
  gr_union$human_union <- read_bed3(human_mm10)
} else {
  message("Human lifted consensus not found yet: ", human_mm10,
          " — sync it from the server to include the tier-3 set.")
}
```

```{r build-reproducible}
# Strain-reproducible mouse GR sets. The two eWAT/iWAT IPs are different strains
# (B6 vs 129), so peaks present in BOTH are high-confidence GR sites — a much more
# specific set than the permissive union (which is ~110k peaks). Defined
# symmetrically: regions of either strain's peaks that overlap the other strain.
reproducible <- function(a_path, b_path) {
  a <- read_narrowpeak(a_path); b <- read_narrowpeak(b_path)
  GenomicRanges::reduce(c(IRanges::subsetByOverlaps(a, b),
                          IRanges::subsetByOverlaps(b, a)))
}

gr_repro <- list(
  eWAT_repro = reproducible(
    file.path(peak_dir, "GSE64458_eWAT/GSE64458_B6_eWAT_GR_peaks.narrowPeak"),
    file.path(peak_dir, "GSE64458_eWAT/GSE64458_129_eWAT_GR_peaks.narrowPeak")),
  iWAT_repro = reproducible(
    file.path(peak_dir, "GSE64458_iWAT/GSE64458_B6_iWAT_GR_peaks.narrowPeak"),
    file.path(peak_dir, "GSE64458_iWAT/GSE64458_129_iWAT_GR_peaks.narrowPeak"))
)

gr_sets <- c(gr_union, gr_repro)
tibble(gr_set = names(gr_sets), n_peaks = map_int(gr_sets, length))
```

# Overlap & enrichment

```{r enrichment-fns}
# Beta-binomial posterior OR: independent Beta priors on P(GR-bound) in
# foreground vs background; OR posterior by Monte-Carlo. Beta(0.5,0.5) Jeffreys
# prior (weakly informative).
posterior_or <- function(k_fg, n_fg, k_bg, n_bg, a = 0.5, b = 0.5, ndraw = 2e5) {
  p_fg <- rbeta(ndraw, a + k_fg, b + n_fg - k_fg)
  p_bg <- rbeta(ndraw, a + k_bg, b + n_bg - k_bg)
  or   <- (p_fg / (1 - p_fg)) / (p_bg / (1 - p_bg))
  tibble(
    p_fg       = k_fg / n_fg,
    p_bg       = k_bg / n_bg,
    or_post    = median(or),
    cri_lo     = unname(quantile(or, 0.025)),
    cri_hi     = unname(quantile(or, 0.975)),
    p_or_gt1   = mean(or > 1)
  )
}

# One foreground-vs-background test for a given GR set.
test_enrichment <- function(gr, fg, bg, gr_name, fg_name) {
  k_fg <- sum(overlapsAny(fg, gr)); n_fg <- length(fg)
  k_bg <- sum(overlapsAny(bg, gr)); n_bg <- length(bg)
  ft   <- fisher.test(matrix(c(k_fg, n_fg - k_fg, k_bg, n_bg - k_bg), 2))
  bind_cols(
    tibble(gr_set = gr_name, foreground = fg_name,
           k_fg = k_fg, n_fg = n_fg, k_bg = k_bg, n_bg = n_bg,
           or_fisher = unname(ft$estimate),
           fisher_lo = ft$conf.int[1], fisher_hi = ft$conf.int[2],
           p_fisher  = ft$p.value),
    posterior_or(k_fg, n_fg, k_bg, n_bg) |> select(-p_fg, -p_bg)
  )
}
```

```{r run-enrichment}
# HFD-specific (primary) and CHD-specific (negative contrast), each vs shared.
grid <- tidyr::expand_grid(
  gr_name = names(gr_sets),
  fg_name = c("HFD_specific", "CHD_specific")
)

enrich <- pmap_dfr(grid, function(gr_name, fg_name) {
  test_enrichment(gr_sets[[gr_name]], atac[[fg_name]], atac$shared,
                  gr_name, fg_name)
})

enrich <- enrich |>
  mutate(pct_fg = 100 * k_fg / n_fg,
         pct_bg = 100 * k_bg / n_bg) |>
  arrange(fg_name, desc(or_post))

readr::write_tsv(enrich, file.path(out_dir, "gr_enrichment_results.tsv"))
enrich |>
  select(gr_set, foreground, pct_fg, pct_bg,
         or_fisher, p_fisher, or_post, cri_lo, cri_hi, p_or_gt1)
```

# Forest plot

```{r forest, fig.width=8, fig.height=4.5}
enrich |>
  mutate(gr_set = factor(gr_set, levels = rev(names(gr_sets)))) |>
  ggplot(aes(or_post, gr_set, colour = foreground)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey50") +
  geom_pointrange(aes(xmin = cri_lo, xmax = cri_hi),
                  position = position_dodge(width = 0.5)) +
  scale_x_log10() +
  labs(x = "Posterior odds ratio (GR-bound | peak class vs shared), 95% CrI",
       y = NULL, colour = "Foreground",
       title = "Experimental GR occupancy enrichment",
       subtitle = "HFD-specific should be > 1; CHD-specific is the negative contrast") +
  theme_minimal(base_size = 12)

ggsave(file.path(out_dir, "gr_enrichment_forest.pdf"), width = 8, height = 4.5)
```

# Width-adjusted robustness check

GR-overlap probability rises with peak width, and HFD-specific vs shared peaks may
differ in width — a confound the 2×2 test ignores. Refit as logistic regression
with `log10(width)` as a covariate; the class coefficient is the width-adjusted OR.

```{r width-adjusted}
peak_df <- function(gr, cls) {
  tibble(class = cls, width = width(gr),
         # GR membership filled per set below
         start = start(gr), gr = gr)
}

adjusted_or <- function(gr_set, gr_name) {
  fg <- atac$HFD_specific; bg <- atac$shared
  df <- bind_rows(
    tibble(class = "HFD_specific", width = width(fg),
           bound = overlapsAny(fg, gr_set)),
    tibble(class = "shared", width = width(bg),
           bound = overlapsAny(bg, gr_set))
  ) |>
    mutate(class = relevel(factor(class), ref = "shared"),
           lw = log10(width))
  m <- glm(bound ~ class + lw, family = binomial, data = df)
  co <- summary(m)$coefficients["classHFD_specific", ]
  tibble(gr_set = gr_name,
         or_adj = exp(co["Estimate"]),
         adj_lo = exp(co["Estimate"] - 1.96 * co["Std. Error"]),
         adj_hi = exp(co["Estimate"] + 1.96 * co["Std. Error"]),
         p_adj  = co["Pr(>|z|)"])
}

adj <- imap_dfr(gr_sets, ~ adjusted_or(.x, .y))
readr::write_tsv(adj, file.path(out_dir, "gr_enrichment_width_adjusted.tsv"))
adj
```

# Interpretation

```{r interpretation, results='asis'}
prim <- enrich |> filter(gr_set == "eWAT_repro", foreground == "HFD_specific")
cat(sprintf(
"- **Primary (in-vivo eWAT, strain-reproducible):** %.1f%% of HFD-specific peaks are GR-bound vs %.1f%% of shared peaks. Posterior OR = %.2f (95%% CrI %.2f–%.2f), P(OR>1) = %.3f; Fisher p = %.2g.\n",
  prim$pct_fg, prim$pct_bg, prim$or_post, prim$cri_lo, prim$cri_hi,
  prim$p_or_gt1, prim$p_fisher))
```

**Notes / caveats**

- *Union vs reproducible:* the union GR sets (~110k mouse peaks) are over-permissive
  and will shrink any OR toward 1 (GR appears "everywhere"); the strain-reproducible
  set is the specificity-appropriate primary readout. Compare both.
- *Tissue match:* eWAT is the GSE236575 depot; iWAT is a secondary depot; human is
  cross-species (lifted, ~20% of peaks recovered) → supporting evidence only.
- *Next:* layer per-peak AP-1 status (reuse the GSE236575 `composite_scan` AP-1 calls)
  to get measured **AP-1 + GR** co-occupancy, then feed that term into the composite
  Junb-nomination posterior (replacing the predicted-GR Fisher OR = 1.28).
```
