---
title: "Experimental GR occupancy × HFD differential chromatin"
subtitle: "Is measured adipocyte GR binding enriched in HFD-specific open chromatin (GSE236575)?"
author: "davebrid"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false   # inputs (synced peak BEDs) change without code changes — caching
                 # silently serves stale results; analysis runs in <1 min anyway
---

```{r setup}
library(GenomicRanges)
library(GenomeInfoDb)
library(rtracklayer)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(stringr)
library(purrr)
library(ggplot2)

select <- dplyr::select        # AnnotationDbi/GenomicRanges mask dplyr::select (see MEMORY)
set.seed(1)

atac_dir <- "../snATACseq/GSE236575/results/deseq2"               # GSE236575 ATAC classes (mm10)
gse_dir  <- "../snATACseq/GSE236575/results/motif_analysis"       # motif / composite calls
peak_dir <- "results/peaks"                                       # this project's GR peaks
chain    <- "hg38ToMm10.over.chain"                               # local liftOver chain (human fallback)
out_dir  <- "results/gr_enrichment"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

# Question and headline

The GSE236575 composite-enhancer model proposed that AP-1 (led by Junb) opens
chromatin on HFD, licensing **predicted** GR binding (NR3C1 motif; Fisher OR = 1.28).
Here we replace predicted GR sites with **experimentally measured** GR occupancy
(re-analyzed adipocyte GR ChIP-seq) and ask whether GR binding is enriched in
HFD-specific open chromatin vs a background of shared (non-differential) peaks.

**Headline (see below):** measured GR occupancy is *depleted*, not enriched, at
HFD-specific peaks — robustly, across two independent datasets (in-vivo mouse eWAT
**chow**, and human adipocyte **dex**), and after adjusting for peak width. GR
instead concentrates on constitutive (**shared**) and **CHD-specific** chromatin.

This does **not** overturn the AP-1-priming model so much as *qualify* it:

- It **supports mechanism 1** (Fkbp5/GR brake-release via *pre-existing* GREs):
  GR-bound chromatin that *closes* on HFD (CHD-specific) is exactly the Fkbp5-locus
  behavior.
- It **argues against a genome-wide mechanism 2** (GR binding the newly AP-1-opened
  sites). The motif *potential* is there, but isn't realized as occupancy in any GR
  ChIP we have.
- **Key caveat:** neither dataset is condition-matched (mouse = chow, can't capture
  HFD-induced binding; human = cross-species dex in vitro). So this bounds, rather
  than closes, the question — and the *signal inside the depletion* (last section) is
  where the biology is.

# Load peaks

```{r readers}
# BEDs / narrowPeaks are 0-based half-open → +1 to start for 1-based GRanges.
clean <- function(gr) keepStandardChromosomes(gr, pruning.mode = "coarse")

read_bed3 <- function(path) {
  readr::read_tsv(path, col_names = c("chr","start","end"),
                  col_types = "cii", progress = FALSE) |>
    with(GRanges(chr, IRanges(start + 1L, end))) |> clean()
}

read_narrowpeak <- function(path) {
  cn <- c("chr","start","end","name","score","strand","signal","pval","qval","peak")
  readr::read_tsv(path, col_names = cn, col_types = "ciicicdddi", progress = FALSE) |>
    with(GRanges(chr, IRanges(start + 1L, end), signal = signal, qval = qval)) |> clean()
}

# "chr1:100-600" → GRanges (motif-table peak ids). NOT cleaned, so it stays 1:1
# with the motif table rows (mcols attached downstream); overlaps are unaffected.
parse_peak <- function(x) {
  m <- str_match(x, "^(.*):([0-9]+)-([0-9]+)$")
  GRanges(m[, 2], IRanges(as.integer(m[, 3]), as.integer(m[, 4])))
}
```

```{r load-atac}
atac <- list(
  HFD_specific = read_bed3(file.path(atac_dir, "HFD_specific_peaks.bed")),
  CHD_specific = read_bed3(file.path(atac_dir, "CHD_specific_peaks.bed")),
  shared       = read_bed3(file.path(atac_dir, "shared_peaks.bed"))
)
map_int(atac, length)
```

```{r load-gr}
# Pipeline consensus sets (union of IP narrowPeaks, merged).
gr_union <- list(
  eWAT_union = read_bed3(file.path(peak_dir, "GSE64458_eWAT/GSE64458_eWAT_GR_consensus.bed")),
  iWAT_union = read_bed3(file.path(peak_dir, "GSE64458_iWAT/GSE64458_iWAT_GR_consensus.bed"))
)

# Strain-reproducible mouse sets: the eWAT/iWAT IPs are different strains (B6 vs 129),
# so peaks present in BOTH are high-confidence GR sites — far more specific than the
# ~110k-peak union. Defined symmetrically (regions of either strain overlapping the other).
reproducible <- function(a_path, b_path) {
  a <- read_narrowpeak(a_path); b <- read_narrowpeak(b_path)
  GenomicRanges::reduce(c(IRanges::subsetByOverlaps(a, b), IRanges::subsetByOverlaps(b, a)))
}
gr_repro <- list(
  eWAT_repro = reproducible(
    file.path(peak_dir, "GSE64458_eWAT/GSE64458_B6_eWAT_GR_peaks.narrowPeak"),
    file.path(peak_dir, "GSE64458_eWAT/GSE64458_129_eWAT_GR_peaks.narrowPeak")),
  iWAT_repro = reproducible(
    file.path(peak_dir, "GSE64458_iWAT/GSE64458_B6_iWAT_GR_peaks.narrowPeak"),
    file.path(peak_dir, "GSE64458_iWAT/GSE64458_129_iWAT_GR_peaks.narrowPeak"))
)

# Human (tier-3, cross-species): prefer the pipeline-lifted .mm10.bed; else lift the
# hg38 consensus locally with the chain (merged fragments).
human_set <- function() {
  mm10 <- file.path(peak_dir, "GSE163061_human/GSE163061_human_GR_consensus.mm10.bed")
  if (file.exists(mm10)) return(read_bed3(mm10))
  hg38 <- file.path(peak_dir, "GSE163061_human/GSE163061_human_GR_consensus.bed")
  if (file.exists(hg38) && file.exists(chain)) {
    message("Lifting human hg38 consensus locally via rtracklayer …")
    lifted <- unlist(liftOver(read_bed3(hg38), import.chain(chain)))
    return(clean(GenomicRanges::reduce(lifted)))
  }
  message("No human consensus available — skipping the tier-3 set."); NULL
}
hs <- human_set()

gr_sets <- c(gr_union, gr_repro)
if (!is.null(hs)) gr_sets$human_dex <- hs
tibble(gr_set = names(gr_sets), n_peaks = map_int(gr_sets, length))
```

# Overlap & enrichment

```{r enrichment-fns}
# Beta-binomial posterior OR: independent Beta(0.5,0.5) (Jeffreys) priors on
# P(GR-bound) in foreground vs background; OR posterior by Monte Carlo.
posterior_or <- function(k_fg, n_fg, k_bg, n_bg, a = 0.5, b = 0.5, ndraw = 2e5) {
  p_fg <- rbeta(ndraw, a + k_fg, b + n_fg - k_fg)
  p_bg <- rbeta(ndraw, a + k_bg, b + n_bg - k_bg)
  or   <- (p_fg / (1 - p_fg)) / (p_bg / (1 - p_bg))
  tibble(or_post = median(or), cri_lo = unname(quantile(or, 0.025)),
         cri_hi = unname(quantile(or, 0.975)), p_or_gt1 = mean(or > 1))
}

test_enrichment <- function(gr, fg, bg, gr_name, fg_name) {
  k_fg <- sum(overlapsAny(fg, gr)); n_fg <- length(fg)
  k_bg <- sum(overlapsAny(bg, gr)); n_bg <- length(bg)
  ft   <- fisher.test(matrix(c(k_fg, n_fg - k_fg, k_bg, n_bg - k_bg), 2))
  bind_cols(
    tibble(gr_set = gr_name, foreground = fg_name,
           pct_fg = 100 * k_fg / n_fg, pct_bg = 100 * k_bg / n_bg,
           or_fisher = unname(ft$estimate), p_fisher = ft$p.value),
    posterior_or(k_fg, n_fg, k_bg, n_bg)
  )
}
```

```{r run-enrichment}
enrich <- tidyr::expand_grid(gr_name = names(gr_sets),
                             fg_name = c("HFD_specific", "CHD_specific")) |>
  pmap_dfr(function(gr_name, fg_name)
    test_enrichment(gr_sets[[gr_name]], atac[[fg_name]], atac$shared, gr_name, fg_name)) |>
  arrange(foreground, desc(or_post))

readr::write_tsv(enrich, file.path(out_dir, "gr_enrichment_results.tsv"))
enrich
```

```{r forest, fig.width=8, fig.height=5}
enrich |>
  mutate(gr_set = factor(gr_set, levels = rev(names(gr_sets)))) |>
  ggplot(aes(or_post, gr_set, colour = foreground)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey50") +
  geom_pointrange(aes(xmin = cri_lo, xmax = cri_hi),
                  position = position_dodge(width = 0.6)) +
  scale_x_log10() +
  labs(x = "Posterior OR (GR-bound | class vs shared), 95% CrI", y = NULL,
       colour = "Foreground", title = "Experimental GR occupancy enrichment",
       subtitle = "All HFD-specific ORs < 1 (depletion); CHD-specific trends > 1") +
  theme_minimal(base_size = 12)
ggsave(file.path(out_dir, "gr_enrichment_forest.pdf"), width = 8, height = 5)
```

## Width-adjusted robustness

GR-overlap probability rises with peak width and the ATAC classes differ in width —
refit as logistic regression with `log10(width)` so the class term is width-adjusted.

```{r width-adjusted}
adjusted_or <- function(gr_set, gr_name) {
  df <- bind_rows(
    tibble(class = "HFD_specific", width = width(atac$HFD_specific),
           bound = overlapsAny(atac$HFD_specific, gr_set)),
    tibble(class = "shared", width = width(atac$shared),
           bound = overlapsAny(atac$shared, gr_set))
  ) |> mutate(class = relevel(factor(class), ref = "shared"), lw = log10(width))
  co <- summary(glm(bound ~ class + lw, binomial, df))$coefficients["classHFD_specific", ]
  tibble(gr_set = gr_name, or_adj = exp(co[1]),
         adj_lo = exp(co[1] - 1.96 * co[2]), adj_hi = exp(co[1] + 1.96 * co[2]),
         p_adj = co[4])
}
adj <- imap_dfr(gr_sets, adjusted_or)
readr::write_tsv(adj, file.path(out_dir, "gr_enrichment_width_adjusted.tsv"))
adj
```

# Signal inside the depletion: characterizing GR-bound HFD peaks

Globally GR avoids HFD-specific chromatin, but a minority of HFD peaks *are* GR-bound.
Are those genuine (motif-supported) GR sites, and what do they regulate? We use the
in-vivo eWAT GR set (tissue-matched) and the existing GSE236575 motif/composite calls.

```{r load-motif}
mot <- readr::read_tsv(
  file.path(gse_dir, "composite_scan/HFD_specific/HFD_specific_per_peak_motifs.tsv"),
  show_col_types = FALSE, progress = FALSE)
hfd <- parse_peak(mot$peak)
hfd$pred_GR       <- mot$MA0113.4 > 0                          # NR3C1 predicted motif
hfd$meas_GR_repro <- overlapsAny(hfd, gr_sets$eWAT_repro)
hfd$meas_GR_union <- overlapsAny(hfd, gr_sets$eWAT_union)
hd <- as_tibble(mcols(hfd))
```

## Measured GR binding tracks the predicted GR motif

If the (few) GR-bound HFD sites are real, predicted NR3C1 motif should predict measured
binding even within the depleted set.

```{r motif-validation}
tab <- table(predicted_GR_motif = hd$pred_GR, measured_GR_bound = hd$meas_GR_repro)
ft  <- fisher.test(tab)
tab
sprintf("predicted GR motif → measured GR binding: OR = %.2f (95%% CI %.2f–%.2f, p = %.2g)",
        ft$estimate, ft$conf.int[1], ft$conf.int[2], ft$p.value)
```

## AP-1 + GR composite peaks that are measured-GR-bound

The 1,078 AP-1+GR composite HFD peaks are the model's strongest candidates. Which are
actually GR-occupied, and what genes do they sit near?

```{r composite-subset}
ng <- readr::read_tsv(
  file.path(gse_dir, "composite_scan/HFD_specific_ap1_gr_nearest_genes.tsv"),
  show_col_types = FALSE, progress = FALSE)
comp <- parse_peak(ng$peak)
comp$SYMBOL <- ng$SYMBOL; comp$dist <- ng$distanceToTSS
comp$meas_GR_repro <- overlapsAny(comp, gr_sets$eWAT_repro)
comp$meas_GR_union <- overlapsAny(comp, gr_sets$eWAT_union)
cd <- as_tibble(mcols(comp))

sprintf("Composite peaks GR-bound: reproducible %d/%d (%.1f%%); union %d/%d (%.1f%%)",
        sum(cd$meas_GR_repro), nrow(cd), 100*mean(cd$meas_GR_repro),
        sum(cd$meas_GR_union), nrow(cd), 100*mean(cd$meas_GR_union))

# Is the canonical composite target Sgk1 GR-bound?
cd |> filter(SYMBOL == "Sgk1") |> select(SYMBOL, dist, meas_GR_repro, meas_GR_union)

# Measured-GR-bound (reproducible) composite peaks, nearest TSS first
gr_bound_composite <- cd |> filter(meas_GR_repro) |>
  arrange(abs(dist)) |> select(SYMBOL, dist)
readr::write_tsv(gr_bound_composite, file.path(out_dir, "gr_bound_composite_peaks.tsv"))
gr_bound_composite
```

## Is the composite set itself more GR-bound than other HFD peaks?

```{r composite-enrichment}
# composite membership for all HFD peaks (overlap with the 1,078 composite intervals)
hd$composite <- overlapsAny(hfd, comp)
ct <- table(composite = hd$composite, measured_GR = hd$meas_GR_repro)
ft2 <- fisher.test(ct)
ct
sprintf("AP-1+GR composite → measured GR binding (within HFD peaks): OR = %.2f (p = %.2g)",
        ft2$estimate, ft2$p.value)
```

# Interpretation

```{r interpretation, results='asis'}
prim <- enrich |> filter(gr_set == "eWAT_repro", foreground == "HFD_specific")
cat(sprintf(
"- **Primary (in-vivo eWAT, strain-reproducible):** %.1f%% of HFD-specific vs %.1f%% of shared peaks are GR-bound → posterior OR = %.2f (95%% CrI %.2f–%.2f), P(OR>1) = %.3f. GR is **depleted** at HFD-opened chromatin.\n",
  prim$pct_fg, prim$pct_bg, prim$or_post, prim$cri_lo, prim$cri_hi, prim$p_or_gt1))
cat(sprintf(
"- **But binding is real where it occurs:** within HFD peaks, predicted NR3C1 motif predicts measured GR occupancy (OR = %.1f). So the model's GR-motif sites are partly genuinely bound even in chow — the depletion is about *condition*, not artifact.\n",
  unname(ft$estimate)))
```

**Takeaways for the Bayesian TF model**

- The measured-GR term **does not boost** the predicted-GR enrichment (Fisher OR 1.28);
  globally it points the other way. Feed it in as a *qualifying* term: GR occupancy
  concentrates on pre-existing/closing chromatin (mechanism 1), not the HFD-opened
  AP-1 program (mechanism 2).
- The defensible positive claim is **local, not genome-wide**: a motif-validated subset
  of AP-1+GR composite HFD peaks is GR-bound even in chow (`gr_bound_composite_peaks.tsv`),
  and predicted GR motif → measured binding (OR ≈ 3). These are the Sgk1-like candidates
  worth following.
- **To settle mechanism 2** we need condition-matched GR ChIP (in-vivo HFD or dex
  adipose). Until then, the integrated nomination should weight measured GR modestly
  and flag the condition caveat.

**Caveats:** mouse GR = chow (no HFD/dex induction); human GR = cross-species dex in
vitro (lifted, ~20% recovery). union sets (~110k peaks) are over-permissive and
shrink ORs toward 1 — the strain-reproducible set is the specificity-appropriate readout.
```
