---
title: "Integration with GSE26123"
author: "Dave Bridges"
date: "February 18, 2015"
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
```

This analysis looked at the subcutaneous insulin and insulin/dex treated adipose explants described in:

Lee M-J, Gong D-W, Burkey BF, Fried SK (2011) Pathways regulated by glucocorticoids in omental and subcutaneous human adipose tissues: a microarray study. Am J Physiol Endocrinol Metab 300: E571–E580. doi:10.1152/ajpendo.00231.2010.

We did not examine omental fat.  This script was generated using GEO2R

```{r geo2r}
# Version info: R 2.14.1, Biobase 2.15.3, GEOquery 2.23.2, limma 3.10.1
# R scripts generated  Wed Feb 18 12:30:15 EST 2015

################################################################
#   Differential expression analysis with limma
library(Biobase)
library(GEOquery)
library(limma)

# load series and platform data from GEO

gset <- getGEO("GSE26123", GSEMatrix =TRUE)
if (length(gset) > 1) idx <- grep("GPL8300", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

# make proper column names to match toptable 
fvarLabels(gset) <- make.names(fvarLabels(gset))

# group names for all samples
sml <- c("G1","G0","X","X","G1","G0","X","X");

# eliminate samples marked as "X"
sel <- which(sml != "X")
sml <- sml[sel]
gset <- gset[ ,sel]

# log2 transform
ex <- exprs(gset)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0) ||
          (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
if (LogC) { ex[which(ex <= 0)] <- NaN
  exprs(gset) <- log2(ex) }

# set up the data and proceed with analysis
fl <- as.factor(sml)
gset$description <- fl
design <- model.matrix(~ description + 0, gset)
colnames(design) <- levels(fl)
fit <- lmFit(gset, design)
cont.matrix <- makeContrasts(G1-G0, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=20000000)

# load NCBI platform annotation
gpl <- annotation(gset)
platf <- getGEO(gpl, AnnotGPL=TRUE)
ncbifd <- data.frame(attr(dataTable(platf), "table"))

# replace original platform annotation
tT <- tT[setdiff(colnames(tT), setdiff(fvarLabels(gset), "ID"))]
tT <- merge(tT, ncbifd, by="ID")
tT <- tT[order(tT$P.Value), ]  # restore correct order

tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","t","B","logFC","Gene.symbol","Gene.title"))
#write.table(tT, file=stdout(), row.names=F, sep="\t")

################################################################
#   Boxplot for selected GEO samples
library(Biobase)
library(GEOquery)

# load series and platform data from GEO

gset <- getGEO("GSE26123", GSEMatrix =TRUE)
if (length(gset) > 1) idx <- grep("GPL8300", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

# group names for all samples in a series
sml <- c("G1","G0","X","X","G1","G0","X","X")

# eliminate samples marked as "X"
sel <- which(sml != "X")
sml <- sml[sel]
gset <- gset[ ,sel]

# order samples by group
ex <- exprs(gset)[ , order(sml)]
sml <- sml[order(sml)]
fl <- as.factor(sml)
labels <- c("Insulin","Insulin and Dex")

# set parameters and draw the plot
palette(c("#dfeaf4","#f4dfdf", "#AABBCC"))
dev.new(width=4+dim(gset)[[2]]/5, height=6)
par(mar=c(2+round(max(nchar(sampleNames(gset)))/2),4,2,1))
title <- paste ("GSE26123", '/', annotation(gset), " selected samples", sep ='')
boxplot(ex, boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=fl)
legend("topleft", labels, fill=palette(), bty="n")

#sig genes
alpha <- 0.25
sig <- droplevels(subset(tT, adj.P.Val<alpha))
```

According to our re-analysis of these data, there were no genes that were significant at a p-value of 0.05.  Using a less stingent p-value of 0.25 showed `r table(tT$adj.P.Val<0.25)['TRUE']` significantly differentially expressed genes.  These genes are listed below:

```{r output-table, results='asis'}
library(xtable)
print(xtable(sig[,c('Gene.symbol','logFC','P.Value','adj.P.Val')]), type='html')
```

# Session Information
```{r sessionInfo}
sessionInfo()
```