DESeq Analysis of Cushing and Acromegaly Patient Samples with Outlier Removed
===============================================================================

```{r exclusion, echo=FALSE}
remove_patient <- 29
```

The counts tables were generated previously using the **filtered_counts_table.Rmd** script.  It removes the outlier data point from the analysis, which was patient **`r remove_patient`**.
This script requires a transcript counts table.  There is also a sample mapping file called **patient_sample_mapping.csv** which links the diagnosis the the samples.  This file was most recently processed on ```r date()```.

```{r loading-data, echo=FALSE, message=FALSE, warning=FALSE}
#required packages
library(DESeq2 ,quietly=TRUE) #for deseq analysis
library(knitcitations) #for generation of citations
write.bibtex(c('base',names(sessionInfo()$otherPkgs)), file = "references.bib", append=T, create_key=F) #writes references to bibtex file
bib <- read.bibtex("references.bib") #loads the bib file


transcript.counts <- read.csv("../data/processed/filtered_transcript_counts_table.csv", row.names="X")
#set the conditions
mapping <- read.csv("../data/raw/patient_sample_mapping.csv")
mapping$samplename <- paste("sample", mapping$sample.., sep="")
mapping$group <- relevel(mapping$group, ref="non-functioning")
levels(mapping$group) <- c("Control", "Acromegaly", "Cushing's")

filtered.mapping <- mapping[mapping$patient.. != remove_patient,]
filtered.transcript.counts <- transcript.counts[,colnames(transcript.counts) != mapping[mapping$patient.. == remove_patient,]$samplename]
```

```{r cds-construction, echo=FALSE, message=FALSE}
#this command removes the cushings samples
acromegaly.mapping <- droplevels(filtered.mapping[filtered.mapping$group != "Cushing's",])
cds.acromegaly = DESeqDataSetFromMatrix( countData=filtered.transcript.counts[,acromegaly.mapping$samplename], 
                                  colData=acromegaly.mapping,
                                  design=~group )

#this command removes the acromegaly samples
cushing.mapping <- droplevels(filtered.mapping[filtered.mapping$group != "Acromegaly"&filtered.mapping$samplename %in% colnames(filtered.transcript.counts),])
cds.cushing = DESeqDataSetFromMatrix( 
  countData=filtered.transcript.counts[,cushing.mapping$samplename], 
  colData=cushing.mapping,
  design=~group )
```

These data were analysed in  by DESeq2 `r citep(bib[["DESeq2"]])`. We did not remove lower expressing transcripts because we pre-filtered the data to examine only one transcript per gene.  

Full Analysis
--------------

```{r deseq-analysis, echo=FALSE, message=FALSE}
acromegaly.cds <- DESeq(cds.acromegaly)
cushing.cds <- DESeq(cds.cushing)

plotMA(acromegaly.cds,ylim=c(-2,2),main="DESeq2 Analysis of Acromegaly Samples")
plotMA(cushing.cds,ylim=c(-2,2),main="DESeq2 Analysis of Cushing Samples")

acromegaly.results <- results(acromegaly.cds)
cushing.results <- results(cushing.cds)

write.csv(acromegaly.results, "../data/processed/acromegaly_deseq2_results.csv")
write.csv(cushing.results, "../data/processed/cushing_deseq2_results.csv")
```



Annotation
-------------

This step annotates the data tables with the official gene symbols.

```{r annotation, echo=FALSE, message=FALSE}
#required packages
library(knitcitations, quietly=TRUE) #for generation of citations
library(biomaRt,quietly=TRUE)
write.bibtex(c('base',names(sessionInfo()$otherPkgs)), file = "references.bib", append=T, create_key=F) #writes references to bibtex file
bib <- read.bibtex("references.bib") #loads the bib file
#download gene name annotation data for results sets.
ensembl <- useMart("ensembl",dataset = 'hsapiens_gene_ensembl')
transcriptInfo <- getBM(
    attributes=c('ensembl_transcript_id', 'ensembl_gene_id','external_gene_id'), 
    values = 'ensembl_transcript_id',
    mart=ensembl)
#merge results files with transcript info files.
annotated.cushing.results <- merge(cushing.results, transcriptInfo, by.x=0, by.y="ensembl_transcript_id")
annotated.acromegaly.results <- merge(acromegaly.results, transcriptInfo, by.x=0, by.y="ensembl_transcript_id")

#write annotated results files.
write.csv(annotated.cushing.results, "../data/processed/Annotated DESeq2 Results - Cushing.csv")
write.csv(annotated.acromegaly.results, "../data/processed/Annotated DESeq2 Results - Acromegaly.csv")
```

The data was annotated from Ensembl data, using the biomaRt package `r citep(bib[[c("biomaRt1","biomaRt2")]])`.

Differentially Expressed Genes
--------------------------------
```{r differentially-expressed, dev=c('png','pdf'), echo=FALSE, fig.show='asis'}
acromegaly.diff.transcripts <- subset(as.data.frame(annotated.acromegaly.results), padj <= 0.05) 
with(acromegaly.diff.transcripts, plot(log2FoldChange, padj, pch=16, main="Differentially Expressed Transcripts in Acromegaly Patients"))
acromegaly.up.transcripts <- subset(acromegaly.diff.transcripts, log2FoldChange > 0)
acromegaly.up.genes <- unique(acromegaly.up.transcripts$external_gene_id)
write(as.character(acromegaly.up.genes), file = "../data/processed/Acromegaly Upregulated Genes.txt")

acromegaly.down.transcripts <- subset(acromegaly.diff.transcripts, log2FoldChange < 0)
acromegaly.down.genes <- unique(acromegaly.down.transcripts$external_gene_id)
write(as.character(acromegaly.down.genes), file = "../data/processed/Acromegaly Downregulated Genes.txt")

cushing.diff.transcripts <- subset(as.data.frame(annotated.cushing.results) , padj <= 0.05) 
with(cushing.diff.transcripts, plot(log2FoldChange, padj, pch=16, main="Differentially Expressed Transcripts in Cushing's Patients"))
cushing.up.transcripts <- subset(cushing.diff.transcripts, log2FoldChange > 0)
cushing.up.genes <- unique(cushing.up.transcripts$external_gene_id)
write(as.character(cushing.up.genes), file = "../data/processed/Cushing Upregulated genes.txt")

cushing.down.transcripts <- subset(cushing.diff.transcripts, log2FoldChange < 0)
cushing.down.genes <- unique(cushing.down.transcripts$external_gene_id)
write(as.character(cushing.down.genes), file = "../data/processed/Cushing Downregulated Genes.txt")
```

### Acromegaly

There were **`r dim(acromegaly.diff.transcripts)[1]`** differentially expressed transcripts from the acromegaly patients, with **`r table(acromegaly.diff.transcripts$log2FoldChange > 0)[1]`** transcripts downregulated and **`r table(acromegaly.diff.transcripts$log2FoldChange > 0)[2]`** transcripts upregulated.

This corresponds to **`r length(acromegaly.up.genes)`** upregulated genes and **`r length(acromegaly.down.genes)`** downregulated genes.

### Cushing's Disease

There were **`r dim(cushing.diff.transcripts)[1]`** differentially expressed transcripts from the acromegaly patients, with **`r table(cushing.diff.transcripts$log2FoldChange > 0)[1]`** transcripts downregulated and **`r table(cushing.diff.transcripts$log2FoldChange > 0)[2]`** transcripts upregulated.

This corresponds to **`r length(cushing.up.genes)`** upregulated genes and **`r length(cushing.down.genes)`** downregulated genes.


Bibiography
------------
```{r bibliography, echo=FALSE, results='asis'}
bibliography("html") 
```

Session Information
-------------------

For the R session, the package versions were:
```{r}
sessionInfo()
```
