---
title: "Rates of Diabetes and Hba1c Prior to Procedures"
author: "Dave Bridges and Trey Carr"
date: "March 3, 2025"
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  pdf: default
#theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To load in the participant data about their Cushing's diagnoses and the procedures.  

## Experimental Details

```{r input-files}
encounter.datafile <- "EncounterAll.csv" 
encounter.bmi.datafile <- "EncounterAnthropometricsBMI.csv" 
diagnosis.datafile <- "DiagnosesComprehensiveAll.csv" 
charlson.datafile <- "ComorbiditiesCharlsonComprehensive.csv"
elixhauser.datafile <- "ComorbiditiesElixhauserComprehensive.csv"
LabResults.datafile <- "LabResults.csv"
cushings.datafile <- "CushingsDataClean.csv"

library(readr) #for loading csv files
library(dplyr) #for data cleaning
library(lubridate) #for date cleaning
library(knitr) #for kables
cushings.data <- read_csv(cushings.datafile) %>%
  rename(ProcedureDate = "DeID_AdmitDate_Clean_procedure")#loaded in data about when their cushings procedures were 
encounter.bmi.data <- read_csv(encounter.bmi.datafile)

encounter.data <- read_csv(encounter.datafile) |>
  mutate(DeID_AdmitDate_Clean = as_date(mdy_hm(DeID_AdmitDate))) %>% #clean and format into just dates not times
  left_join(encounter.bmi.data, ,by=c("DeID_PatientID","DeID_EncounterID")) %>% #add in BMI data for each encounter 
  left_join(cushings.data,by="DeID_PatientID") %>% # added in time of cushings diagnosis
  mutate(TimePoint = case_when(ProcedureDate <= DeID_AdmitDate_Clean ~ "Before",
                               ProcedureDate > DeID_AdmitDate_Clean ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = ProcedureDate - DeID_AdmitDate_Clean)

diagnosis.data <- read_csv(diagnosis.datafile) %>%
  left_join(encounter.data,by=c("DeID_PatientID","DeID_EncounterID"))
charlson.data <- read_csv(charlson.datafile) %>%
  left_join(encounter.data,by=c("DeID_PatientID","DeID_EncounterID"))
elixhauser.data <- read_csv(elixhauser.datafile) %>%
  left_join(encounter.data,by=c("DeID_PatientID","DeID_EncounterID"))
LabResults.data <- read_csv(LabResults.datafile) %>%
  left_join(encounter.data,by=c("DeID_PatientID","DeID_EncounterID"))

total.participants <-
  distinct(encounter.data, DeID_PatientID) %>%
  nrow()

```

## Raw Data

Relevant patient data is in these files:

##Go back and describe each data file. Elixhauser, charlson, labresults

* **Diagnoses** are in `r diagnosis.datafile`.  This includes patient ID, encouter ID, TermCodeMapped (ICD code).
* The `r encounter.datafile` contains metadata about encounters including dates and BMIs


This files was most recently updated on `r format(file.info(diagnosis.datafile)$mtime, "%Y-%m-%d")`.  This script was most recently updated on `r date()`.

## Lab Results

There are `r total.participants` unique patients in this data set.

```{r lab-measures-summary}
LabResults.data |>
  count(RESULT_CODE) |>
  arrange(-n) |>
  kable(caption="Total number of lab results for cases.")
```


### Type 2 Diabetes Lab Results

#### Glucose Measurements

```{r glucose}
glucose.data <-
  LabResults.data %>%
  filter(RESULT_CODE %in% c("GLUC","GLUC_WB")) %>%
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(DeID_AdmitDate_Clean) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese")) 

glucose.summary <-
  glucose.data %>%
  group_by(TimePoint) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

glucose.summary %>% kable(caption="Summary of glucose levels before and after procedures")
```

We found `r nrow(glucose.data)` patients that had a lab result for a HbA1c test. 

Now looking at the effect of BMI

```{r before_t2d_glucose_bmi}
glucose.summary.bmi <-
  glucose.data %>%
  group_by(TimePoint,Obesity) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

glucose.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  kable(caption="Summary of Hb1ac levels before and after procedures")

library(ggplot2)
glucose.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  ggplot(aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=126, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="Glucose (mg/dL)",
       x="") +
  theme_classic(base_size=16)

glucose.data%>% 
  filter(TimePoint=="Before") %>%
  ggplot(aes(y=value,
           x=BMI)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept=126, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="Glucose (mg/dL)",
       x="") +
  theme_classic(base_size=16)

library(broom)
library(mgcv)
gam(value~s(BMI),
   data=glucose.data %>% filter(TimePoint=="Before"),
   REML=TRUE) -> gam.glucose

lm(value~BMI,
   data=glucose.data %>% filter(TimePoint=="Before")) -> lm.glucose

AIC(lm.glucose, gam.glucose) %>% kable(caption="Glucose for linear and nonlinear models,lower AIC indicates better model fit")

gam.glucose %>%
  tidy %>%
  kable()

lm.glucose %>%
  tidy %>%
  kable(caption="Linear model for glucose prior to intervention")

plot(gam.glucose, pages = 1, residuals = TRUE)
```
 
```{r before_t2d_hba1c}
hba1c.data <-
  LabResults.data %>%
  filter(RESULT_CODE %in% c("A1C", "XXA1C", "A1C EX", "X0022")) %>%
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(DeID_AdmitDate_Clean) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese")) 

hba1c.summary <-
  hba1c.data %>%
  group_by(TimePoint) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

hba1c.summary %>% kable(caption="Summary of Hb1ac levels before and after procedures")
```

We found `r nrow(hba1c.data)` patients that had a lab result for a HbA1c test. 

Now looking at the effect of BMI

```{r before_t2d_hba1c_bmi}
hba1c.summary.bmi <-
  hba1c.data %>%
  group_by(TimePoint,Obesity) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

hba1c.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  kable(caption="Summary of Hb1ac levels before and after procedures")

library(ggplot2)
hba1c.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  ggplot(aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=6.5, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="HbA1c (%)",
       x="") +
  theme_classic(base_size=16)

hba1c.data%>% 
  filter(TimePoint=="Before") %>%
  ggplot(aes(y=value,
           x=BMI)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept=6.5, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="HbA1c (%)",
       x="") +
  theme_classic(base_size=16)

library(broom)
library(mgcv)
gam(value~s(BMI)+AgeInYears,
   data=hba1c.data %>% filter(TimePoint=="Before"),
   REML=TRUE) -> gam.hba1c

lm(value~BMI+AgeInYears,
   data=hba1c.data %>% filter(TimePoint=="Before")) -> lm.hba1c

AIC(lm.hba1c, gam.hba1c) %>% kable(caption="AIC for linear and nonlinear models,lower AIC indicates better model fit")

gam.hba1c %>%
  tidy %>%
  kable()

plot(gam.hba1c, pages = 1, residuals = TRUE)
```

Used a generalized additive model with a splined term for BMI. The EDF for this was `r summary(gam.hba1c)$s.table[, "edf"]`, where $edf > 1$ suggests nonlinearity, this was a better model fit based on the AIC.

### ALT
 
```{r before_alt}
alt.data <-
  LabResults.data %>%
  filter(RESULT_CODE %in% c("ALT")) %>%
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(DeID_AdmitDate_Clean) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese")) 

alt.summary <-
  alt.data %>%
  group_by(TimePoint) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

alt.summary %>% kable(caption="Summary of ALT levels before and after procedures")
```

We found `r nrow(alt.data)` patients that had a lab result for a ALT test. 


### AST
 
```{r before_ast}
ast.data <-
  LabResults.data %>%
  filter(RESULT_CODE %in% c("AST")) %>%
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(DeID_AdmitDate_Clean) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese")) 

ast.summary <-
  ast.data %>%
  group_by(TimePoint) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

ast.summary %>% kable(caption="Summary of AST levels before and after procedures")
```

We found `r nrow(ast.data)` patients that had a lab result for a AST test. 




Now looking at the effect of BMI

```{r before_alt_bmi}
alt.summary.bmi <-
  alt.data %>%
  group_by(TimePoint,Obesity) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

alt.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  kable(caption="Summary of ALT levels before and after procedures")

alt.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  ggplot(aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=30, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="ALT (IU/L)",
       x="") +
  theme_classic(base_size=16)

alt.data%>% 
  filter(TimePoint=="Before") %>%
  ggplot(aes(y=value,
           x=BMI)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept=30, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="ALT (IU/L)",
       x="") +
  theme_classic(base_size=16)

library(broom)
library(mgcv)
gam(value~s(BMI)+AgeInYears,
   data=alt.data %>% filter(TimePoint=="Before"),
   REML=TRUE) -> gam.alt

lm(value~BMI+AgeInYears,
   data=alt.data %>% filter(TimePoint=="Before")) -> lm.alt

AIC(lm.alt, gam.alt) %>% kable(caption="AIC for linear and nonlinear models,lower AIC indicates better model fit for ALT levels")

gam.alt %>%
  tidy %>%
  kable()

plot(gam.alt, pages = 1, residuals = TRUE)
```

Used a generalized additive model with a splined term for BMI. The EDF for this was `r summary(gam.alt)$s.table[, "edf"]`, where $edf > 1$ suggests nonlinearity, this was a better model fit based on the AIC.

### LDL Cholesterol
 
```{r before_ldl}
ldl.data <-
  LabResults.data %>%
  filter(RESULT_CODE %in% c("LDLC", "DLDL")) %>%
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(DeID_AdmitDate_Clean) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese")) 

ldl.summary <-
  ldl.data %>%
  group_by(TimePoint) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

ldl.summary %>% kable(caption="Summary of LDL-C levels before and after procedures")
```

We found `r nrow(ldl.data)` patients that had a lab result for a LDL-C test. 

Now looking at the effect of BMI

```{r before_ldl_bmi}
ldl.summary.bmi <-
  ldl.data %>%
  group_by(TimePoint,Obesity) %>%
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            n=length(value))

ldl.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  kable(caption="Summary of Hb1ac levels before and after procedures")

library(ggplot2)
ldl.summary.bmi %>% 
  filter(TimePoint=="Before") %>%
  filter(!(is.na(Obesity))) %>%
  ggplot(aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity)) +
  geom_bar(stat='identity') +
  geom_errorbar(width=0.5) +
  geom_hline(yintercept=160, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="LDL (mg/dL)",
       x="") +
  theme_classic(base_size=16)

ldl.data%>% 
  filter(TimePoint=="Before") %>%
  ggplot(aes(y=value,
           x=BMI)) +
  geom_point() +
  geom_smooth(method="lm") +
  geom_hline(yintercept=160, lty="dashed") +
  labs(title="Cushing's Patients Only",
       y="LDL (mg/dL)",
       x="") +
  theme_classic(base_size=16)

library(broom)
library(mgcv)
gam(value~s(BMI)+AgeInYears,
   data=ldl.data %>% filter(TimePoint=="Before"),
   REML=TRUE) -> gam.ldl

lm(value~BMI+AgeInYears,
   data=ldl.data %>% filter(TimePoint=="Before")) -> lm.ldl

AIC(lm.ldl, gam.ldl) %>% kable(caption="AIC for linear and nonlinear models,lower AIC indicates better model fit")

gam.ldl %>%
  tidy %>%
  kable()

plot(gam.ldl, pages = 1, residuals = TRUE)
```

Used a generalized additive model with a splined term for BMI. The EDF for this was `r summary(gam.ldl)$s.table[, "edf"]`, where $edf > 1$ suggests nonlinearity, this was a better model fit based on the AIC.


## Session Information

```{r session-information}
sessionInfo()
```
