---
title: "Defining the Diagnosis and Surgery Dates"
author: "Dave Bridges and Trey Carr"
date: "March 3, 2025"
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  pdf: default
#theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To load in the participant data about their Cushing's diagnoses and the procedures.  This is an important landmark for defining effects before or after.

## Experimental Details

```{r input-files}
encounter.datafile <- "EncounterAll.csv" 
procedure.datafile <- "ProceduresComprehensive.csv" 
diagnosis.datafile <- "DiagnosesComprehensiveAll.csv" 

library(readr) #for loading csv files
library(dplyr) #for data cleaning
library(lubridate) #for date cleaning
library(knitr) #for kables
encounter.data <- read_csv(encounter.datafile) |>
  mutate(DeID_AdmitDate_Clean = as_date(mdy_hm(DeID_AdmitDate))) #clean and format into just dates not times
procedure.data <- read_csv(procedure.datafile)
diagnosis.data <- read_csv(diagnosis.datafile)
```

## Raw Data

Relevant patient data is in these files:

* **Procedures** are in `r procedure.datafile`.  This includes patient ID, encouter ID, TermCodeMapped (ICD code).
* **Diagnoses** are in `r diagnosis.datafile`.  This includes patient ID, encouter ID, TermCodeMapped (ICD code).
* The `r encounter.datafile` contains metadata about encounters including dates and BMIs



Describe your raw data files, including what the columns mean (and what units they are in).


This files was most recently updated on `r format(file.info(procedure.datafile)$mtime, "%Y-%m-%d")`.  This script was most recently updated on `r date()`.

## Data Cleaning

### Procedures

Procedures are defined by ICD codes. 

```{r procedure-codes}
procedure_codes <- c("07.6", "07.61", "07.62", "07.63", "07.64", "07.65", "07.68", "07.69", "07.15", "07.72", "0GB00ZZ", "0GT00ZZ", "0GB03ZZ", "0GB04ZZ", "0GT04ZZ", "0GC00ZZ", "0GC03ZZ", "0GC04ZZ", "0GT00ZZ", "0GT04ZZ")
```

We used these ICD codes to define the procedures `r paste(procedure_codes,collapse=",")`.

```{r procedure-cleaning}
cushings.procedures <-
  procedure.data |>
  filter(TermCodeMapped %in% procedure_codes)

#several procedures are not in procedure codes
```

#### Unmapped Procedures

Several procedures are not in our procedure codes. 

```{r unmapped-procedures}
procedure.data |>
  filter(!(TermCodeMapped %in% procedure_codes)) |>
  group_by(TermNameMapped) |>
  count() |>
  arrange(desc(n)) |>
  kable(caption="Unmapped procedures and number of times they occured")

```

These all seem like reasonable procedures so I am going to keep all of them in the dataset.

#### All Procedures

```{r all-procedures}
procedure.data |>
  group_by(TermNameMapped) |>
  count() |>
  arrange(desc(n)) |>
  kable(caption="All procedures and number of times they occured")

first.cushings.procedure <-
  procedure.data |>
    left_join(encounter.data, by=c("DeID_PatientID","DeID_EncounterID")) |> #added in data about encounters
  arrange(DeID_PatientID,DeID_AdmitDate_Clean) |>
  distinct(DeID_PatientID, .keep_all = T)
```

#### Summary of Cushing's Procedures in the Dataset

First checked which year the procedure first occured.

```{r cushings-summary-year}
first.cushings.procedure |>
  summarize(Year = year(DeID_AdmitDate_Clean)) |>
  group_by(Year) |>
  count() -> procedures.by.year

procedures.by.year |>
  kable(caption="Diagnoses per year")

library(ggplot2)
procedures.by.year |>
  ggplot(aes(x=Year,
         y=n)) +
  geom_bar(stat='identity') +
  labs(y="First-Time Cushing's Procedures",
       x="") +
  theme_classic(base_size=16)
```

Next evaluated the age at first diagnosis

```{r cushings-summary-age}
first.cushings.procedure |>
  group_by(AgeInYears) |>
  count() -> procedures.by.age

#cushings.by.age |>
#  kable(caption="Diagnoses per age")

library(ggplot2)
procedures.by.age |>
  ggplot(aes(x=AgeInYears,
         y=n)) +
  geom_bar(stat='identity') +
  labs(y="First-Time Cushing's Procedures",
       x="") +
  theme_classic(base_size=16)
```
The average age of the first recorded diagnosis of Cushings is `r mean(first.cushings.procedure$AgeInYears,na.rm=T)` $\pm$ `r sd(first.cushings.procedure$AgeInYears,na.rm=T)` (mean $\pm$ SD).

### Diagnoses

Procedures are defined by ICD codes.  We used E24.0  for ICD10 coding and 255.0  for ICD9 coding

```{r diagnosis-cleaning}
library(dplyr)
cushings.diagnosis.data <-
  diagnosis.data |>
  filter(TermCodeMapped %in% c('255.0','E24.0')) |>
  left_join(encounter.data, by=c("DeID_PatientID","DeID_EncounterID")) |> #added in data about encounters
  arrange(DeID_PatientID,DeID_AdmitDate_Clean) |>
  distinct(DeID_PatientID, .keep_all = T) #onbly take the first diagnosis date for each patient
```

#### Summary of Cushing's Diagnoses in the Dataset

First checked which year the diagnoses first occured.

```{r procedure-summary-year}
library(knitr)
cushings.diagnosis.data |>
  summarize(Year = year(DeID_AdmitDate_Clean)) |>
  group_by(Year) |>
  count() -> cushings.by.year

cushings.by.year |>
  kable(caption="Diagnoses per year")

library(ggplot2)
cushings.by.year |>
  ggplot(aes(x=Year,
         y=n)) +
  geom_bar(stat='identity') +
  labs(y="First-Time Cushing's Diagnoses",
       x="") +
  theme_classic(base_size=16)
```

Next evaluated the age at first diagnosis

```{r procedure-summary-age}
cushings.diagnosis.data |>
  group_by(AgeInYears) |>
  count() -> cushings.by.age

#cushings.by.age |>
#  kable(caption="Diagnoses per age")

library(ggplot2)
cushings.by.age |>
  ggplot(aes(x=AgeInYears,
         y=n)) +
  geom_bar(stat='identity') +
  labs(y="First-Time Cushing's Diagnoses",
       x="") +
  theme_classic(base_size=16)
```

The average age of the first recorded diagnosis of Cushings is `r mean(cushings.diagnosis.data$AgeInYears,na.rm=T)` $\pm$ `r sd(cushings.diagnosis.data$AgeInYears,na.rm=T)` (mean $\pm$ SD).

### Matching Procedures to Diagnoses

```{r procedure-diagnosis-matching}
combined.treatment.data <- left_join(first.cushings.procedure,cushings.diagnosis.data,
                                     by=c("DeID_PatientID"),
                                     suffix=c("_procedure","_diagnosis")) |>
  mutate(TreatmentInterval = DeID_AdmitDate_Clean_procedure - DeID_AdmitDate_Clean_diagnosis)

combined.treatment.data.clean <-
  combined.treatment.data |>
  filter(!is.na(TreatmentInterval)) |> #remove patients with missing dates
  filter(TreatmentInterval >=0 ) #remove patients with negative dates
```

Out of `r nrow(combined.treatment.data)` participants, there were `r sum(combined.treatment.data$TreatmentInterval <0, na.rm=T)` patients with errors in this data (treatment appears to have occured before diagnosis).  A few were missing procedure or diagnosis dates (n=`r sum(is.na(combined.treatment.data$TreatmentInterval))`).  The remaining `r sum(combined.treatment.data$TreatmentInterval >=0, na.rm=T)` were reasonable.  Surprisingly `r sum(combined.treatment.data.clean$TreatmentInterval ==0, na.rm=T)` patients had diagnoses and procedures on the same day.

The average treatment interval, excluding these was `r mean(combined.treatment.data.clean$TreatmentInterval,na.rm=T)` $\pm$ `r sd(combined.treatment.data.clean$TreatmentInterval,na.rm=T)` days

```{r treatment-interval-histogram}
combined.treatment.data |>
  ggplot(aes(x=TreatmentInterval)) +
  geom_histogram(binwidth=1) +
  theme_classic(base_size=16)
```
```{r output-file}
output_file <- 'CushingsDataClean.csv'

combined.treatment.data.clean |>
  select(DeID_PatientID, 
         DeID_AdmitDate_Clean_procedure,
         DeID_AdmitDate_Clean_diagnosis,
         AgeInYears_procedure,
         TreatmentInterval) |>
  write_csv(output_file)
```

A file containing diagnosis dates, age and treatment intervals were written into `r output_file`.  This can be used to anchor other diagnoses and lab results to before or after their Cushing's treatments.

## Session Information

```{r session-information}
sessionInfo()
```
