---
title: "Propensity Mapped Comparason Between Cases and Controls"
author: "Dave Bridges and Trey Carr"
date: "April 9, 2025"
format: 
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
    fig-path: "figures/"
  pdf: default
#theme: journal
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To create a propensity mapped subgroup for lab results.

## Data Entry

```{r data-entry}
case.demographics.filename <- 'CaseDemographics.csv'
control.demographics.filename <- 'ControlDemographics.csv'

case.lab.results.filename <- 'LabResults.csv'
control.lab.results.filename <- '../controls/LabResults.csv'

bp.filename.cases <- 'NursingStandardVitalSigns.csv'
bp.filename.controls <- '../controls/NursingStandardVitalSigns.csv'

case.encounter.datafile <- "EncounterAll.csv" 
case.encounter.bmi.datafile <- "EncounterAnthropometricsBMI.csv" 
control.encounter.datafile <- "../controls/EncounterAll.csv" 
control.encounter.bmi.datafile <- "../controls/EncounterAnthropometricsBMI.csv" 

library(readr)
library(dplyr)

#labs dont have dates/ages
case.encounters <- read_csv(case.encounter.datafile)
case.encounters.bmi <- read_csv(case.encounter.bmi.datafile)
case.encounters.all <- left_join(case.encounters,case.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))

control.encounters <- read_csv(control.encounter.datafile)
control.encounters.bmi <- read_csv(control.encounter.bmi.datafile)
control.encounters.all <- left_join(control.encounters,control.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))


case.demographics <- read_csv(case.demographics.filename)
control.demographics <- read_csv(control.demographics.filename)

case.labs <- read_csv(case.lab.results.filename) |>
  left_join(case.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
    left_join(case.demographics, by="DeID_PatientID") # added demographic data
control.labs <- read_csv(control.lab.results.filename) |> 
  left_join(control.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
  left_join(control.demographics, by="DeID_PatientID") # added demographic data
#remove large objects not needed further
#rm(control.encounters,control.encounters.all,control.encounters.bmi)

#combined into one dataset
library(lubridate)

#this is used to filter for the window between when a lab result is allowable relative to the procedure.
procedure.result.interval <- 30

lab.results <- bind_rows(control.labs |> mutate(Cushings=0),
                         case.labs |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
lab.results.filtered <-
  lab.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
complete.data.race.ethnicity.age <-
  lab.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
complete.data  <-
  complete.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories

#number of controls per case
fold <- 10
```

Note the controls dont have ages or BMIs because it depends on when the lab results were obtained, so had to merge in with encounter data.

### Participant Inclusion Data

We began with `r nrow(lab.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(lab.results |> filter(Cushings==1))-nrow(lab.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - lab.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics}
lab.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary

library(knitr)
missing.demographic.summary |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data")
```

Our final dataset therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** Cushing's patients control patients.

## Propensity Mapping

Used the R package `mapit` to match based on Age, Gender, Race, and Ethnicity.  We wanted exact matches for everything except age, where we wanted the closest case that had a measurement.  We picked `r fold` matched controls for each case


### Hba1c

```{r hba1c-propensity}
library(MatchIt)
m.out.a1c <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("A1C", "XXA1C", "A1C EX", "X0022")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.a1c, standardize = TRUE)  # Balance statistics[5]
plot(m.out.a1c, type = "jitter")         # Propensity score distribution[5]

matched_data.a1c <- match.data(m.out.a1c)

hba1c.data <- 
  matched_data.a1c |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 

hba1c.summary <-
  hba1c.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

hba1c.summary.iii <-
  hba1c.data %>%
  group_by(Cushings,ObesityIII) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

library(ggplot2)
ggplot(hba1c.summary,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Hba1c (Percent)") +
  theme_classic()

ggplot(hba1c.summary.iii,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=ObesityIII,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Hba1c (Percent)",x="") +
  theme_classic()

ggplot(hba1c.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Hba1c (Percent)") +
  theme_classic()


library(knitr)
hba1c.summary %>% kable(caption="Summary of Hb1ac levels by obesity and pre-existing Cushing's")
library(broom)
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Hba1c")

lm(value ~ Cushings + ObesityIII + Cushings:ObesityIII,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Class III Obesity and Cushings on Hba1c")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Hba1c")
```

### Glucose

```{r glucose-propensity}
m.out.glucose <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("GLUC","GLUC_WB")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.glucose, standardize = TRUE)  # Balance statistics[5]
plot(m.out.glucose, type = "jitter")         # Propensity score distribution[5]

matched_data.glucose <- match.data(m.out.glucose)

glucose.data <- 
  matched_data.glucose |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 

glucose.summary <-
  glucose.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

ggplot(glucose.summary,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Glucose (mg/dL)") +
  theme_classic()

ggplot(glucose.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Glucuose (mg/dL)") +
  theme_classic()



glucose.summary %>% kable(caption="Summary of glucose levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Glucose")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Glucose")
```

### ALT

```{r alt-propensity}
m.out.alt <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("ALT")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

summary(m.out.alt, standardize = TRUE)  # Balance statistics[5]
plot(m.out.alt, type = "jitter")  

matched_data.alt <- match.data(m.out.alt)

alt.data <- 
  matched_data.alt |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity)) 

alt.summary <-
  alt.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

library(ggplot2)
ggplot(alt.summary,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="ALT (mg/dL)") +
  theme_classic()

ggplot(alt.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="ALT (mg/dL)") +
  theme_classic()


library(knitr)
alt.summary %>% kable(caption="Summary of ALT levels by obesity and pre-existing Cushing's")
library(broom)
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on ALT")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Hba1c")
```

### LDL Cholesterol

```{r ldl-propensity}
m.out.ldl <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("LDLC","DLDL")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

summary(m.out.ldl, standardize = TRUE)  # Balance statistics[5]
plot(m.out.ldl, type = "jitter")  

matched_data.ldl <- match.data(m.out.ldl)

ldl.data <- 
  matched_data.ldl |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity)) 

ldl.summary <-
  ldl.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

ggplot(ldl.summary,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="LDL Cholesterol (mg/dL)") +
  theme_classic()

ggplot(ldl.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="LDL Cholesterol (mg/dL)") +
  theme_classic()

ldl.summary %>% kable(caption="Summary of LDL-C levels by obesity and pre-existing Cushing's")

lm(value ~ Cushings + Obesity + Cushings:Obesity,data=ldl.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on LDL-C")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=ldl.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on LDL-C")
```

## Blood Pressure

Blood pressure data is in a separate file (`r bp.filename.cases` and `r bp.filename.controls`).  We have mean noninvasive blood pressure, as well as systolic and diastolic.  These data, similar to lab results need to be mapped to encounter data 

```{r blood-pressure-data-cleaning}
bp.data.cases <- read_csv(bp.filename.cases) |>
  left_join(case.encounters.all) |>
  left_join(case.demographics, by="DeID_PatientID") 

bp.data.controls <- read_csv(bp.filename.controls) |>
  left_join(control.encounters.all) |>
  left_join(control.demographics, by="DeID_PatientID") 

bp.results <- bind_rows(bp.data.controls |> mutate(Cushings=0),
          bp.data.cases |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
bp.results.filtered <-
  bp.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
bp.data.race.ethnicity.age <-
  bp.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
bp.data  <-
  bp.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories
```

### Inclusion and Exclusion for Blood Pressure Data

We began with `r nrow(bp.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(bp.results |> filter(Cushings==1))-nrow(bp.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - bp.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(bp.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(bp.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics-bp}
bp.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary.bp

library(knitr)
missing.demographic.summary.bp |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data, for blood pressure measures")
```

Our final dataset for blood pressure therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** control patients.


### Blood Pressure 

```{r bp-propensity}
m.out.bp <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = bp.data, 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.bp, standardize = TRUE)  # Balance statistics[5]
plot(m.out.bp, type = "jitter")         # Propensity score distribution[5]

matched_data <- match.data(m.out.bp)

bp.data <- 
  matched_data |>
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### Mean Arterial Pressure

```{r bp-map}
#MAP
bp.summary.map <-
  bp.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(BPMeanNonInvasive,na.rm=T),
            se=se(BPMeanNonInvasive),
            sd=sd(BPMeanNonInvasive,na.rm=T),
            n=length(BPMeanNonInvasive))

ggplot(bp.summary.map,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Mean; mmHg)") +
  theme_classic()

ggplot(bp.data,
       aes(y=BPMeanNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Mean; mmHg)") +
  theme_classic()



bp.summary.map %>% kable(caption="Summary of blood pressure levels by obesity and pre-existing Cushing's (Mean)")
lm(BPMeanNonInvasive ~ Cushings + Obesity + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Blood Pressure (Mean)")

lm(BPMeanNonInvasive ~ Cushings + BMI + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Blood Pressure (Mean)")
```


#### Systolic Blood Pressure

```{r bp-systolic}
#systolic
bp.summary.sys <-
  bp.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(BPSysNonInvasive,na.rm=T),
            se=se(BPSysNonInvasive),
            sd=sd(BPSysNonInvasive,na.rm=T),
            n=length(BPSysNonInvasive))

ggplot(bp.summary.sys,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Systolic; mmHg)") +
  theme_classic()

ggplot(bp.data,
       aes(y=BPSysNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Systolic; mmHg)") +
  theme_classic()



bp.summary.sys %>% kable(caption="Summary of blood pressure levels by obesity and pre-existing Cushing's (Systolic)")
lm(BPSysNonInvasive ~ Cushings + Obesity + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Blood Pressure (Systolic)")

lm(BPSysNonInvasive ~ Cushings + BMI + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Blood Pressure (Systolic)")
```

#### Diastolic Arterial Pressure

```{r bp-diastolic}
#MAP
bp.summary.dia <-
  bp.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(BPDiaNonInvasive,na.rm=T),
            se=se(BPDiaNonInvasive),
            sd=sd(BPDiaNonInvasive,na.rm=T),
            n=length(BPDiaNonInvasive))

ggplot(bp.summary.dia,
       aes(y=mean,
           ymin=mean-se,
           ymax=mean+se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Diastolic; mmHg)") +
  theme_classic()

ggplot(bp.data,
       aes(y=BPDiaNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Diastolic; mmHg)") +
  theme_classic()



bp.summary.dia %>% kable(caption="Summary of blood pressure levels by obesity and pre-existing Cushing's (Diastoic)")
lm(BPDiaNonInvasive ~ Cushings + Obesity + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Blood Pressure (Diastolic)")

lm(BPDiaNonInvasive ~ Cushings + BMI + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Blood Pressure (Diastolic)")
```


####PRISMA Diagram
Here we will graphically illustrate how we arrived at the listed n number of people for each lab result.

```{r diagram}
labresult_name<- c("Glucose","Hba1c","ALT","LDL","Blood Pressure","Mean Arterial Pressure","Systolic Arterial Pressure","Diastolic Arterial Pressure")

n_people<- c(3639, 20, 326, 13, 19054, 23349, 23349, 23349 )

summary_table<- data.frame(Lab_name=labresult_name, N_people=n_people)

kable(summary_table)



```

## Diagnostic Data on Propensity Mapping

```{r propensity-mapping-diagnostics}
# List of matchit objects by outcome
outcomes <- list(`Blood Pressure` = m.out.bp, Glucose=m.out.glucose, HbA1c= m.out.a1c, `LDL-C` = m.out.ldl)

# Function to extract age summary from a single m.out object
extract_age_summary <- function(m.out, outcome_name) {
  matched_data <- match.data(m.out)
  pre_summary <- m.out$model$data %>% 
    group_by(Cushings) %>% 
    summarise(
      Mean = mean(AgeInYears, na.rm = TRUE),
      SE = sd(AgeInYears, na.rm = TRUE)/sqrt(n())
    ) %>% 
    mutate(Stage = "Pre-matching")
  
  post_summary <- matched_data %>% 
    group_by(Cushings) %>% 
    summarise(
      Mean = mean(AgeInYears, na.rm = TRUE),
      SE = sd(AgeInYears, na.rm = TRUE)/sqrt(n())
    ) %>% 
    mutate(Stage = "Post-matching")
  
  bind_rows(pre_summary, post_summary) %>% 
    mutate(Outcome = outcome_name) %>%
    select(Outcome, Stage, Cushings, Mean, SE)
}

# Apply to all outcomes and combine
age_table <- bind_rows(lapply(names(outcomes), function(x) extract_age_summary(outcomes[[x]], x)))

age_table_wide <- 
  age_table |>
  mutate(Cushings = case_when(Cushings==0~"Control",
                                Cushings==1~"Cushing")) |>
  pivot_wider(names_from=Cushings,
              values_from = c(Mean,SE))

extract_age_smd <- function(m.out, outcome_name){
  bal <- summary(m.out)$sum.matched
  matched_col <- grep("Matched", colnames(bal), value = TRUE)
  pre_col <- grep("Std.*Mean.*Diff.*", colnames(bal), value = TRUE)[1]
  
  data.frame(
    Outcome = outcome_name,
    Stage = c("Pre-matching", "Post-matching"),
    SMD_Age = c(bal["AgeInYears", pre_col], bal["AgeInYears", matched_col])
  )
}

smd_table <- bind_rows(lapply(names(outcomes), function(x) extract_age_smd(outcomes[[x]], x)))

full_join(age_table_wide,smd_table,by=c("Outcome","Stage")) -> propensity.diagnostics.table

propensity.diagnostics.table |>
  kable(caption="Propensity Matching Summary for Age.")

propensity.diagnostics.table.pub <-
  propensity.diagnostics.table |>
  mutate(Control= paste0(round(Mean_Control,2)," +/- ",round(SE_Control,2)),
         `Cushing's`= paste0(round(Mean_Cushing,2)," +/- ",round(SE_Cushing,2))) |>
  rename(`Standardized Mean Difference`=SMD_Age) |>
  mutate(`Standardized Mean Difference`=round(`Standardized Mean Difference`,4)) |>
  select(Outcome,Stage,Control,`Cushing's`,`Standardized Mean Difference`)

kable(propensity.diagnostics.table.pub,caption="Publication Version of Propensity Mapping Summary")

write_csv(propensity.diagnostics.table.pub,"Propensity Matching Summary - Age.csv")

```

### Love Plots

Generated Love plots using ggplot for each outcome.

```{r love-plots}
library(cobalt)
library(ggplot2)

plots <- lapply(names(outcomes), function(name) {
  # Generate Love plot, force only SMDs
  gg <- love.plot(
    outcomes[[name]],
    stats = "mean.diffs",  # only standardized mean differences
    threshold = 0.1,
    abs = TRUE,
    stars = "none"         # avoids mixing with raw differences
  )
  
  # gg is now a ggplot object directly
  gg + 
    theme_classic(base_size=6) +
    labs(
      title = name,
      x = "Standardized Mean Difference"
    )
})

# Example: display HbA1c plot
plots[[1]]
library(cowplot)

combined_plot <- plot_grid(
  plotlist = plots,   # your list of ggplot objects
  ncol = 2,           # number of columns
  labels = names(plots)  # optional: label each subplot with the outcome name
)

# Display
combined_plot
```

## Session Information

```{r session-information}
sessionInfo()
```
