---
title: "Propensity Mapped Comparison Between Cases and Controls"
author: "Dave Bridges and Trey Carr"
date: "April 9, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To create a propensity mapped subgroup for lab results.

## Data Entry

```{r data-entry}
case.demographics.filename <- 'CaseDemographics.csv'
control.demographics.filename <- 'ControlDemographics.csv'

case.lab.results.filename <- 'LabResults.csv'
control.lab.results.filename <- '../controls/LabResults.csv'

bp.filename.cases <- 'NursingStandardVitalSigns.csv'
bp.filename.controls <- '../controls/NursingStandardVitalSigns.csv'

case.encounter.datafile <- "EncounterAll.csv" 
case.encounter.bmi.datafile <- "EncounterAnthropometricsBMI.csv" 
control.encounter.datafile <- "../controls/EncounterAll.csv" 
control.encounter.bmi.datafile <- "../controls/EncounterAnthropometricsBMI.csv" 

library(readr)
library(dplyr)

#labs dont have dates/ages
case.encounters <- read_csv(case.encounter.datafile)
case.encounters.bmi <- read_csv(case.encounter.bmi.datafile)
case.encounters.all <- left_join(case.encounters,case.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))

control.encounters <- read_csv(control.encounter.datafile)
control.encounters.bmi <- read_csv(control.encounter.bmi.datafile)
control.encounters.all <- left_join(control.encounters,control.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))


case.demographics <- read_csv(case.demographics.filename)
control.demographics <- read_csv(control.demographics.filename)

case.labs <- read_csv(case.lab.results.filename) |>
  left_join(case.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
    left_join(case.demographics, by="DeID_PatientID") # added demographic data
control.labs <- read_csv(control.lab.results.filename) |> 
  left_join(control.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
  left_join(control.demographics, by="DeID_PatientID") # added demographic data
#remove large objects not needed further
#rm(control.encounters,control.encounters.all,control.encounters.bmi)

#combined into one dataset
library(lubridate)

#this is used to filter for the window between when a lab result is allowable relative to the procedure.
procedure.result.interval <- 365 #within a year

lab.results <- bind_rows(control.labs |> mutate(Cushings=0),
                         case.labs |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
lab.results.filtered <-
  lab.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
complete.data.race.ethnicity.age <-
  lab.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
complete.data  <-
  complete.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories

#number of controls per case
fold <- 10
```

Note the controls dont have ages or BMIs because it depends on when the lab results were obtained, so had to merge in with encounter data.

### Participant Inclusion Data

We began with `r nrow(lab.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(lab.results |> filter(Cushings==1))-nrow(lab.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - lab.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics}
lab.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary

library(knitr)
missing.demographic.summary |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data")
```

Our final dataset therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** Cushing's patients control patients.

## Propensity Mapping

Used the R package `mapit` to match based on Age, Gender, Race, and Ethnicity.  We wanted exact matches for everything except age, where we wanted the closest case that had a measurement.  We picked `r fold` matched controls for each case

### Total Population

```{r total-population-propensity}
library(MatchIt)
m.out.all <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data,
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.all, standardize = TRUE)  # Balance statistics[5]
plot(m.out.all, type = "jitter")         # Propensity score distribution[5]

matched_data.all <- match.data(m.out.all)

master.data <- 
  matched_data.all |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity)) 
```

#### Demographics of the Total Population

```{r total-population-demographics}
library(janitor)
master.summary.gender <-
  master.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

master.summary.race.ethnicity <-
  master.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

master.summary.quant <-
  master.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


master.summary.table <-
  bind_rows(master.summary.quant,
          master.summary.gender,
          master.summary.race.ethnicity)

library(kableExtra)
master.summary.table |>
  kable(caption="Demographic Summary for Master List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(master.summary.table,"Demographic Summary - Master Sample.csv")

#statistics for BMI
master.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for entire sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

library(broom)
wilcox.test(BMI~Cushings,master.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for entire sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(master.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the master sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

Specific points described in the manuscript:

```{r total-population-details}
complete.data |> 
  distinct(DeID_PatientID, .keep_all=T) |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_totals() |>
  rename('Controls'=`0`,
         'Cases'=`1`) -> unmatched.totals.race
  
race.totals <- 
  filter(unmatched.totals.race, RaceEthnicity=="Total") |> 
  select(Cases,Controls) |>
  as.integer()
white.totals<- 
  filter(unmatched.totals.race, RaceEthnicity=="White") |> 
  select(Cases,Controls) |>
  as.integer()

asian.totals<- 
  filter(unmatched.totals.race, RaceEthnicity=="Asian") |> 
  select(Cases,Controls) |>
  as.integer()

black.totals<- 
  filter(unmatched.totals.race, RaceEthnicity=="Black") |> 
  select(Cases,Controls) |>
  as.integer()

hl.totals<- 
  filter(unmatched.totals.race, RaceEthnicity=="Hispanic or Latino") |> 
  select(Cases,Controls) |>
  as.integer()

nonwhite.totals <- race.totals-white.totals
nonasian.totals <- race.totals-asian.totals
nonblack.totals <- race.totals-black.totals
nonhl.totals <- race.totals-hl.totals

# for obesity status using matched data

master.data |> 
  tabyl(Obesity,Cushings) |>
  adorn_totals() |>
  rename('Controls'=`0`,
         'Cases'=`1`) -> totals.obesity

# for gender differences
complete.data |> 
  distinct(DeID_PatientID, .keep_all=T) |>
  tabyl(GenderCode,Cushings) |>
  adorn_totals() |>
  rename(`Total Population`=`0`,
         `Cushing's Disease`=`1`) -> unmatched.totals.gender

male.totals<- 
  filter(unmatched.totals.gender, GenderCode=="M") |> 
  select(`Cushing's Disease`,`Total Population`) |>
  as.integer()

female.totals<- 
  filter(unmatched.totals.gender, GenderCode=="F") |> 
  select(`Cushing's Disease`,`Total Population`) |>
  as.integer()
```

- Prior to matching by Gender, we noted that participants with Cushing's disease were more likely to identify as Female (**`r female.totals[1]/race.totals[1]*100`%** than the overall Michigan Medicine participant population (**`r female.totals[2]/race.totals[2]*100`%** **p=`r chisq.test(rbind(female.totals, male.totals))$p.value`**)

```{r gender-matching}
unmatched.totals.gender |>
  mutate(Percent.Cases = `Cushing's Disease`/`Cushing's Disease`[GenderCode=="Total"]*100,
         Percent.Controls = `Total Population`/`Total Population`[GenderCode=="Total"]*100) |>
  pivot_longer(cols=starts_with('Percent'),
               names_to = 'Population',
               values_to = 'Percent') |>
  filter(GenderCode!="Total") -> gender.unmatched.summary

gender.unmatched.summary |>
  filter(Population=="Percent.Cases") |>
  mutate(Label= paste0(GenderCode, " - ", round(Percent,0), "%")) |>
  select(Label,GenderCode,`Cushing's Disease`) |>
  ggplot(aes(x = 2, y = `Cushing's Disease`, fill = GenderCode)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(label = Label),
            position = position_stack(vjust = 0.5),
            color = "white", size = 5) +
  xlim(0.5, 2.5) +
  scale_fill_grey() +
  theme_classic(base_size = 16) +
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA)
  ) 

gender.unmatched.summary |>
  filter(Population=="Percent.Cases") |>
  mutate(Label= paste0(GenderCode, " - ", round(Percent,0), "%")) |>
  select(Label,GenderCode,`Cushing's Disease`) |>
  ggplot(aes(x = 2, y = `Cushing's Disease`, fill = GenderCode)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(label = Label),
            position = position_stack(vjust = 0.5),
            color = "white", size = 5) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values=color_scheme) +
  theme_classic(base_size = 16) +
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA)
  ) 

gender.unmatched.summary |>
  ggplot(aes(y=Percent,
             x=reorder(as.factor(GenderCode),Percent),
             fill=Population)) +
  geom_bar(stat='identity',position="dodge") +
  theme_classic(base_size=16) +
  scale_fill_grey(labels=c("Cushing's Disease","Total Population")) +
  labs(y="Percent",x="",fill="") +
  theme(legend.position=c(0.15,0.95),
        legend.background = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA))

gender.unmatched.summary |>
  ggplot(aes(y=Percent,
             x=reorder(as.factor(GenderCode),Percent),
             fill=Population)) +
  geom_bar(stat='identity',position="dodge") +
  theme_classic(base_size=16) +
  scale_fill_manual(labels=c("Cushing's Disease","Total Population"),values=color_scheme) +
  labs(y="Percent",x="",fill="") +
  theme(legend.position=c(0.15,0.95),
        legend.background = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA))
```

- Prior to matching for Race and Ethnicity, we noted that participants with Cushing’s disease were more likely to identify as Non-Hispanic White (**`r white.totals[1]/race.totals[1]*100`%** than the overall Michigan Medicine participant population (**`r white.totals[2]/race.totals[2]*100`%** Non-Hispanic White, **p=`r chisq.test(rbind(white.totals, nonwhite.totals))$p.value`**)
- Correspondingly, participants with Cushing’s disease were less likely to identify as Asian (**p=`r chisq.test(rbind(asian.totals, nonasian.totals))$p.value`**) or Black. (**p=`r chisq.test(rbind(black.totals, nonblack.totals))$p.value`**).

```{r race-matching}
race.matching <- tibble(
  Race = c("White", "Black", "Asian","Hispanic or Latino"),
  `Total Population` = c(white.totals[2]/race.totals[2]*100,
                         black.totals[2]/race.totals[2]*100, 
                         asian.totals[2]/race.totals[2]*100,
                         hl.totals[2]/race.totals[2]*100),
  `Cushing's Disease` = c(white.totals[1]/race.totals[1]*100, 
              black.totals[1]/race.totals[1]*100, 
              asian.totals[1]/race.totals[1]*100,
              hl.totals[1]/race.totals[1]*100)) 

race.matching |>
  pivot_longer(cols=c(2:3), names_to = "Population", values_to = "Percent") |>
  mutate(Population=relevel(as.factor(Population),ref="Total Population")) |>
  ggplot(aes(y=Percent,
             x=reorder(as.factor(Race),-Percent),
             fill=Population)) +
  geom_bar(stat='identity',position="dodge") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  labs(y="Percent",x="",fill="") +
  theme(legend.position=c(0.75,0.75))

race.matching |>
  pivot_longer(cols=c(2:3), names_to = "Population", values_to = "Percent") |>
  mutate(Population=relevel(as.factor(Population),ref="Total Population")) |>
  ggplot(aes(y=Percent,
             x=reorder(as.factor(Race),-Percent),
             fill=Population)) +
  geom_bar(stat='identity',position="dodge") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  labs(y="Percent",x="",fill="") +
  theme(legend.position=c(0.75,0.75))
```

- As we expected, participants with Cushing’s disease had a higher BMI than the controls, with **`r filter(totals.obesity, Obesity=="Obese")['Cases']/filter(totals.obesity, Obesity=="Total")['Cases']*100`**% of participants with Cushing’s disease having a BMI above 30 kg/m2, compared to only **`r filter(totals.obesity, Obesity=="Obese")['Controls']/filter(totals.obesity, Obesity=="Total")['Controls']*100`%** controls (p=**`r chisq.test(as.matrix(rbind(totals.obesity |> filter(Obesity=="Obese") |> select(Cases,Controls),totals.obesity |> filter(Obesity=="Non-Obese") |> select(Cases,Controls))))$p.value`**).  

```{r obesity-matching}
totals.obesity |>
  mutate(Percent.Cases = Cases/Cases[Obesity=="Total"]*100,
         Percent.Controls = Controls/Controls[Obesity=="Total"]*100) |>
  pivot_longer(cols=starts_with('Percent'),
               names_to = 'Population',
               values_to = 'Percent') |>
  filter(Obesity!="Total") |>
  ggplot(aes(y=Percent,
             x=reorder(as.factor(Obesity),Percent),
             fill=Population)) +
  geom_bar(stat='identity',position="dodge") +
  theme_classic(base_size=16) +
  scale_fill_grey(labels=c("Cushing's Disease","Total Population")) +
  labs(y="Percent",x="",fill="") +
  theme(legend.position=c(0.15,0.95),
        legend.background = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA))

totals.obesity |>
  mutate(Percent.Cases = Cases/Cases[Obesity=="Total"]*100,
         Percent.Controls = Controls/Controls[Obesity=="Total"]*100) |>
  pivot_longer(cols=starts_with('Percent'),
               names_to = 'Population',
               values_to = 'Percent') |>
  filter(Obesity!="Total") |>
  ggplot(aes(y=Percent,
             x=reorder(as.factor(Obesity),Percent),
             fill=Population)) +
  geom_bar(stat='identity',position="dodge") +
  theme_classic(base_size=16) +
  scale_fill_manual(labels=c("Cushing's Disease","Total Population"),values=color_scheme) +
  labs(y="Percent",x="",fill="") +
  theme(legend.position=c(0.15,0.95),
        legend.background = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA))
```

#### Stratified Demographics for the Entire Population

```{r total-population-stratified-demographics}
# first check for normality of Age and BMI
master.data |>
  group_by(Obesity,Cushings)|>
  summarize(Age = shapiro.test(sample(AgeInYears,3000,replace=T))$p.value,
            BMI = shapiro.test(sample(BMI,3000,replace=T))$p.value,
            n= length(Cushings)) -> master.data.dist

master.data.dist |>
  kable(caption="Shapiro-Wilk Tests for Distribution of Quantitative Variables",
        digits=c(1,1,99,99,1))|>
  kable_styling(full_width = FALSE, position = "center")

#kruskal wallis testst since they are not normally distributed

master.quant.kw <-
  master.data |> 
  mutate(ObesityCushings=paste0(Obesity,Cushings)) |>
  summarize(BMI.kw=kruskal.test(BMI~ObesityCushings)$p.value,
            Age.kw=kruskal.test(AgeInYears~ObesityCushings)$p.value)

master.2x2.quant.summary <- 
  master.data |>
  group_by(Obesity, Cushings) |>
  summarize_at(.vars = c('AgeInYears', 'BMI'),
               .funs = list(mean = mean, sd = sd,n=length)) |>
  mutate(Age = paste0(round(AgeInYears_mean, 2), " +/- ", round(AgeInYears_sd, 2)),
         BMI = paste0(round(BMI_mean, 2), " +/- ", round(BMI_sd, 2)),
         n = as.character(BMI_n)) |>
  select(Obesity, Cushings, BMI, Age,n) |>
  ungroup() |>
  pivot_longer(
    cols = c(n, BMI, Age),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  # Create group column combining Obesity and Cushings
  mutate(group = paste(Obesity, Cushings, sep = "_")) %>%
  # Pivot wider to spread group into columns, keeping Variable as row identifier
  pivot_wider(
    id_cols = Variable,  # Explicitly specify Variable as the row identifier
    names_from = group,
    values_from = Value,
    names_sort = TRUE
  ) |>
  mutate(p.value=c(NA,as.numeric(master.quant.kw))) |>
  rename("Control, BMI < 30"=`Non-Obese_0`,
         "Cushing's, BMI < 30"=`Non-Obese_1`,
         "Control, BMI > 30" = `Obese_0`,
         "Cushing's, BMI > 30" = Obese_1)


# Create the group column (replacing hyphen with underscore for clean column names)
master.data <- master.data %>%
  mutate(group = paste(gsub("-", "_", Obesity), Cushings, sep = "_"))

# Define the categorical variables to iterate over
cat_vars <- c("GenderName", "RaceEthnicity")

# Initialize an empty list to store tables for each variable
master.tables_list <- list()

# Loop over each categorical variable
for (var in cat_vars) {
  # Compute counts
  counts_df <- master.data %>%
    group_by(!!sym(var), group) %>%
    summarise(count = n(), .groups = "drop") %>%
    # Pivot to wide format
    pivot_wider(
      names_from = group,
      values_from = count,
      values_fill = 0,
      names_sort = TRUE
    ) %>%
    rename(Variable = !!sym(var))

  # Calculate group-wise (column-wise) percentages
  counts_df <- counts_df %>%
    mutate(across(c(Non_Obese_0, Non_Obese_1, Obese_0, Obese_1),
                  ~ sprintf("%d (%.1f%%)", .x, .x / sum(.x, na.rm = TRUE) * 100),
                  .names = "{.col}"))

  # Calculate total count for sorting (sum of numeric part of formatted strings)
  counts_df <- counts_df %>%
    mutate(total_count = rowSums(
      sapply(select(., Non_Obese_0, Non_Obese_1, Obese_0, Obese_1),
             function(x) as.numeric(gsub(" .*", "", x)))
    )) %>%
    # Sort by total count in descending order
    arrange(desc(total_count)) %>%
    # Select desired columns
    select(Variable, Non_Obese_0, Non_Obese_1, Obese_0, Obese_1)

  # Compute chi-squared test p-value
  cont_table <- table(master.data[[var]], master.data$group)
  chi_test <- chisq.test(cont_table)
  p_val <- chi_test$p.value

  # Add p-value to the first row
  counts_df <- counts_df %>%
    mutate(p.value = if_else(row_number() == 1, p_val, NA_real_))

  # Add to list
  master.tables_list[[var]] <- counts_df
}


# Combine all variable tables into one
master.2x2.counts.table <- bind_rows(master.tables_list) |>
  rename("Control, BMI < 30"=`Non_Obese_0`,
         "Cushing's, BMI < 30"=`Non_Obese_1`,
         "Control, BMI > 30" = `Obese_0`,
         "Cushing's, BMI > 30" = Obese_1)
  
master.2x2.table <-
  bind_rows(master.2x2.quant.summary,
            master.2x2.counts.table) 

kable(master.2x2.table,
      caption="Summary of stratified variables") |>
  kable_styling(full_width = FALSE, position = "center")

master.2x2.table |> write_csv(na="",file="Stratified Demographic Summary - Master Sample.csv")
```

```{r stratified-demographics-details}
master.data |>
  mutate(Obesity = fct_recode(Obesity, "NonObese"="Non-Obese")) |>
  tabyl(GenderName, Obesity, Cushings)  -> matched.table.gender

master.data |>
  mutate(Obesity = fct_recode(Obesity, "NonObese"="Non-Obese")) |>
  tabyl(RaceEthnicity, Obesity, Cushings)  -> matched.table.raceethnicity

matched.table.raceethnicity[[2]] |>
  adorn_totals() |>
  mutate(Pct=NonObese/(Obese+NonObese)*100)-> matched.table.raceethnicity.cushings

matched.table.raceethnicity.cushings |> filter(RaceEthnicity=="Asian") |> select(Obese,NonObese) -> matched.asian.cushings

matched.table.raceethnicity.cushings |> filter(RaceEthnicity=="Hispanic or Latino") |> select(Obese,NonObese) -> matched.hl.cushings

matched.table.raceethnicity.cushings |> filter(RaceEthnicity=="Total") |> select(Obese,NonObese) -> matched.total.cushings

matched.nonasian.cushings <- matched.total.cushings - matched.asian.cushings

matched.nonhl.cushings <- matched.total.cushings - matched.hl.cushings
```

Details specified in the manuscript:

- For example, among participants with a $BMI > 30 mg/m^2$, participants with Cushing’s disease had a higher BMI than controls (p=`r wilcox.test(BMI~Cushings, data=master.data |> filter(Obesity=="Obese"))$p.value`).

```{r stratfied-demographics-obesity}
master.data |>
  group_by(Cushings,Obesity) |>
  summarize(BMI.mean = mean(BMI,na.rm=T),
            BMI.se = se(BMI),
            BMI.n = length(!is.na(BMI))) -> matched.bmi.summary

matched.bmi.summary |>
  ggplot(aes(x=Obesity,
             y=BMI.mean,
             fill=as.factor(Cushings))) +
  geom_bar(stat='identity',position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin=BMI.mean - BMI.se,
                    ymax=BMI.mean + BMI.se),
                position = position_dodge(width = 0.9),
                width = 0.5) +
  theme_classic(base_size=16) +
  scale_fill_grey(name="",labels=c("Control","Cushing's Disease")) +
  labs(y="BMI (kg/m2)",x="")

matched.bmi.summary |>
  ggplot(aes(x=Obesity,
             y=BMI.mean,
             fill=as.factor(Cushings))) +
  geom_bar(stat='identity',position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin=BMI.mean - BMI.se,
                    ymax=BMI.mean + BMI.se),
                position = position_dodge(width = 0.9),
                width = 0.5) +
  theme_classic(base_size=16) +
  scale_fill_manual(name="",labels=c("Control","Cushing's Disease"),values = color_scheme) +
  labs(y="BMI (kg/m2)",x="")
```

- Females with Cushing’s disease were more likely to have a BMI over 30 kg/m2 (`r matched.table.gender[[2]] |> mutate(Pct = Obese/(NonObese+Obese)*100) |> filter(GenderName=="Female")|> pull(Pct)`% of females with Cushing's disease had a $BMI>30$) than males (`r matched.table.gender[[2]] |> mutate(Pct = Obese/(NonObese+Obese)*100) |> filter(GenderName=="Male")|> pull(Pct)`%), p=`r chisq.test(matched.table.gender[[2]] |> select(-GenderName) |> as.matrix() )$p.value`).  

```{r stratfied-demographics-gender}
matched.table.gender[[2]] |> 
  mutate(Pct = Obese/(NonObese+Obese)*100) |>
  ggplot(aes(x=GenderName,
             y=Pct,
             fill=GenderName)) +
  geom_bar(stat='identity') +
  theme_classic(base_size=16) +
  scale_fill_grey(name="") +
  labs(y="Percent with BMI > 30 kg/m2",x="") +
  theme(legend.position="none")

matched.table.gender[[2]] |> 
  mutate(Pct = Obese/(NonObese+Obese)*100) |>
  ggplot(aes(x=GenderName,
             y=Pct,
             fill=GenderName)) +
  geom_bar(stat='identity') +
  theme_classic(base_size=16) +
  scale_fill_manual(name="",values = color_scheme)  +
  labs(y="Percent with BMI > 30 kg/m2",x="") +
  theme(legend.position="none")
```

- Among participants with Cushing’s disease, Asian (`r matched.table.raceethnicity.cushings |> filter(RaceEthnicity=="Asian")|> pull(Pct)`%) and Hispanic or Latino (`r matched.table.raceethnicity.cushings |> filter(RaceEthnicity=="Hispanic or Latino")|> pull(Pct)`%) participants were more likely have a BMI  under 30 kg/m2 compared to the overall average (`r matched.table.raceethnicity.cushings |> filter(RaceEthnicity=="Total")|> pull(Pct)`%, p=`r fisher.test(as.matrix(rbind(matched.asian.cushings,matched.nonasian.cushings)))$p.value` and `r fisher.test(as.matrix(rbind(matched.hl.cushings,matched.nonhl.cushings)))$p.value` respectively).
- The average age of participants with Cushing’s disease with or without obesity was similar.


### Hba1c

```{r hba1c-propensity}
m.out.a1c <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("A1C", "XXA1C", "A1C EX", "X0022")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.a1c, standardize = TRUE)  # Balance statistics[5]
plot(m.out.a1c, type = "jitter")         # Propensity score distribution[5]

matched_data.a1c <- match.data(m.out.a1c)

hba1c.data <- 
  matched_data.a1c |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### Demographics of Participants with HbA1c Values

```{r aic-demographics}
hba1c.summary.gender <-
  hba1c.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

hba1c.summary.race.ethnicity <-
  hba1c.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

hba1c.summary.quant <-
  hba1c.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


hba1c.summary.table <-
  bind_rows(hba1c.summary.quant,
          hba1c.summary.gender,
          hba1c.summary.race.ethnicity)

hba1c.summary.table |>
  kable(caption="Demographic Summary for HbA1c List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(hba1c.summary.table,"Demographic Summary - HbA1c Sample.csv")

#statistics for BMI
hba1c.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for HbA1c sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

wilcox.test(BMI~Cushings,hba1c.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for HbA1c sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(hba1c.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the HbA1c sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

```{r hba1c-analyses}
hba1c.summary <-
  hba1c.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

hba1c.summary.iii <-
  hba1c.data %>%
  group_by(Cushings,ObesityIII) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

library(ggplot2)
ggplot(hba1c.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  scale_fill_grey(labels=c("Controls","Cases"),name="") +
  labs(y="Hba1c (Percent)",x="") +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.1,0.9))

ggplot(hba1c.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  scale_fill_manual(labels=c("Controls","Cases"),name="",values=color_scheme) +
  labs(y="Hba1c (Percent)",x="") +
  theme_classic(base_size=16) +
  theme(legend.position=c(0.1,0.9))

ggplot(hba1c.summary.iii,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=ObesityIII,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Hba1c (Percent)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(hba1c.summary.iii,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=ObesityIII,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Hba1c (Percent)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(hba1c.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Hba1c (Percent)",x="") +
  scale_fill_grey() +
  theme_classic(base_size=16) +
  theme(legend.position="none")

ggplot(hba1c.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Hba1c (Percent)",x="") +
  scale_fill_manual(values=color_scheme) +
  theme_classic(base_size=16) +
  theme(legend.position="none")


library(knitr)
hba1c.summary %>% kable(caption="Summary of Hb1ac levels by obesity and pre-existing Cushing's")
library(broom)
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Hba1c")

lm(value ~ Cushings + ObesityIII + Cushings:ObesityIII,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Class III Obesity and Cushings on Hba1c")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Hba1c")
```

#### HbA1c Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r aic-statistics}
library(emmeans)
#used fully adjusted model
lm.hba1c <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = hba1c.data)

## 1. Adjusted means
emm.hba1c <- emmeans(lm.hba1c, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "HbA1c",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.hba1c <- contrast(emmeans(lm.hba1c, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "HbA1c",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.hba1c <- tidy(lm.hba1c) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "HbA1c",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.hba1c %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.hba1c <- means_wide %>%
  left_join(contrast.hba1c, by = c("Outcome", "Obesity"))

hba1c.summary %>% kable(caption="Summary of HbA1c levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on HbA1c")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on HbA1c")
```

### Glucose

```{r glucose-propensity}
m.out.glucose <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("GLUC","GLUC_WB")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.glucose, standardize = TRUE)  # Balance statistics[5]
plot(m.out.glucose, type = "jitter")         # Propensity score distribution[5]

matched_data.glucose <- match.data(m.out.glucose)

glucose.data <- 
  matched_data.glucose |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### Demographics of Participants with Glucose Values

```{r glucose-demographics}
glucose.summary.gender <-
  glucose.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

glucose.summary.race.ethnicity <-
  glucose.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

glucose.summary.quant <-
  glucose.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


glucose.summary.table <-
  bind_rows(glucose.summary.quant,
          glucose.summary.gender,
          glucose.summary.race.ethnicity)

glucose.summary.table |>
  kable(caption="Demographic Summary for Glucose List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(glucose.summary.table,"Demographic Summary - Glucose Sample.csv")

#statistics for BMI
glucose.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for Glucose sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

wilcox.test(BMI~Cushings,glucose.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for Glucose sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(glucose.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the Glucose sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

```{r glucose-analysis}
glucose.summary <-
  glucose.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

ggplot(glucose.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Glucose (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(glucose.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Glucose (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(glucose.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Glucuose (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(glucose.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Glucuose (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")
```

#### Glucose Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r glucose-statistics}
#used fully adjusted model
lm.glucose <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = glucose.data)

## 1. Adjusted means
emm.glucose <- emmeans(lm.glucose, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Glucose",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.glucose <- contrast(emmeans(lm.glucose, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Glucose",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.glucose <- tidy(lm.glucose) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Glucose",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.glucose %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.glucose <- means_wide %>%
  left_join(contrast.glucose, by = c("Outcome", "Obesity"))

glucose.summary %>% kable(caption="Summary of glucose levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Glucose")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Glucose")
```

### Liver Enzymes

#### ALT

```{r alt-propensity}
m.out.alt <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("ALT")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

summary(m.out.alt, standardize = TRUE)  # Balance statistics[5]
plot(m.out.alt, type = "jitter")  

matched_data.alt <- match.data(m.out.alt)

alt.data <- 
  matched_data.alt |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))
```

##### Demographics of Participants with ALT Values

```{r alt-demographics}
alt.summary.gender <-
  alt.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

alt.summary.race.ethnicity <-
  alt.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

alt.summary.quant <-
  alt.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


alt.summary.table <-
  bind_rows(alt.summary.quant,
          alt.summary.gender,
          alt.summary.race.ethnicity)

alt.summary.table |>
  kable(caption="Demographic Summary for ALT List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(alt.summary.table,"Demographic Summary - ALT Sample.csv")

#statistics for BMI
alt.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for ALT sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

wilcox.test(BMI~Cushings,alt.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for ALT sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(alt.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the ALT sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

```{r alt-analysis}
alt.summary <-
  alt.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

library(ggplot2)
ggplot(alt.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="ALT (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(alt.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="ALT (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(alt.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="ALT (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(alt.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="ALT (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")


library(knitr)
alt.summary %>% kable(caption="Summary of ALT levels by obesity and pre-existing Cushing's")
library(broom)
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on ALT")
```

##### ALT Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r alt-statistics}
#used fully adjusted model
lm.alt <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = alt.data)

## 1. Adjusted means
emm.alt <- emmeans(lm.alt, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "ALT",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.alt <- contrast(emmeans(lm.alt, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "ALT",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.alt <- tidy(lm.alt) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "ALT",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.alt %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.alt <- means_wide %>%
  left_join(contrast.alt, by = c("Outcome", "Obesity"))

alt.summary %>% kable(caption="Summary of ALT levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on ALT")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on ALT")
```

#### AST

```{r ast-propensity}
m.out.ast <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("AST")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

summary(m.out.ast, standardize = TRUE)  # Balance statistics[5]
plot(m.out.ast, type = "jitter")  

matched_data.ast <- match.data(m.out.ast)

ast.data <- 
  matched_data.ast |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))
```

##### Demographics of Participants with AST Values

```{r ast-demographics}
ast.summary.gender <-
  ast.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

ast.summary.race.ethnicity <-
  ast.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

ast.summary.quant <-
  ast.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


ast.summary.table <-
  bind_rows(ast.summary.quant,
          ast.summary.gender,
          ast.summary.race.ethnicity)

ast.summary.table |>
  kable(caption="Demographic Summary for AST List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(ast.summary.table,"Demographic Summary - AST Sample.csv")

#statistics for BMI
ast.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for AST sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

wilcox.test(BMI~Cushings,ast.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for AST sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(ast.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the AST sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

```{r ast-analysis}
ast.summary <-
  ast.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

library(ggplot2)
ggplot(ast.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="AST (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(ast.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="AST (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(ast.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="AST (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(ast.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="AST (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")


library(knitr)
ast.summary %>% kable(caption="Summary of AST levels by obesity and pre-existing Cushing's")
library(broom)
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=ast.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on AST")
```

##### AST Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r ast-statistics}
#used fully adjusted model
lm.ast <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = ast.data)

## 1. Adjusted means
emm.ast <- emmeans(lm.ast, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "AST",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.ast <- contrast(emmeans(lm.ast, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "AST",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.ast <- tidy(lm.ast) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "AST",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.ast %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.ast <- means_wide %>%
  left_join(contrast.ast, by = c("Outcome", "Obesity"))

ast.summary %>% kable(caption="Summary of AST levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=ast.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on AST")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=ast.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on AST")
```


### LDL Cholesterol

```{r ldl-propensity}
m.out.ldl <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("LDLC","DLDL")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

summary(m.out.ldl, standardize = TRUE)  # Balance statistics[5]
plot(m.out.ldl, type = "jitter")  

matched_data.ldl <- match.data(m.out.ldl)

ldl.data <- 
  matched_data.ldl |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity)) 
```

#### Demographics of Participants with LDL-C Values

```{r ldl-demographics}
ldl.summary.gender <-
  ldl.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

ldl.summary.race.ethnicity <-
  ldl.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

ldl.summary.quant <-
  ldl.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


ldl.summary.table <-
  bind_rows(ldl.summary.quant,
          ldl.summary.gender,
          ldl.summary.race.ethnicity)

ldl.summary.table |>
  kable(caption="Demographic Summary for LDL-C List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(ldl.summary.table,"Demographic Summary - LDL-C Sample.csv")

#statistics for BMI
ldl.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for LDL-C sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

wilcox.test(BMI~Cushings,ldl.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for LDL-C sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(ldl.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the LDL-C sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

```{r ldl-analysis}
ldl.summary <-
  ldl.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(value,na.rm=T),
            se=se(value),
            sd=sd(value,na.rm=T),
            n=length(value))

ggplot(ldl.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="LDL Cholesterol (mg/dL)", x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(ldl.summary,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="LDL Cholesterol (mg/dL)", x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(ldl.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="LDL Cholesterol (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(ldl.data,
       aes(y=value,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="LDL Cholesterol (mg/dL)",x="") +
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ldl.summary %>% kable(caption="Summary of LDL-C levels by obesity and pre-existing Cushing's")

lm(value ~ Cushings + Obesity + Cushings:Obesity,data=ldl.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on LDL-C")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=ldl.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on LDL-C")
```

#### LDL-C Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.  

**Important**: This model unlike all the otehrs was not adjusted for RaceEthnicity, as all participants with a LDL-C value were of the same Race/Ethnicity (White).

```{r ldl-statistics}
#used fully adjusted model
lm.ldl <- lm(value ~ AgeInYears + GenderName + Cushings * Obesity, data = ldl.data)

## 1. Adjusted means
emm.ldl <- emmeans(lm.ldl, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "LDL-C",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.ldl <- contrast(emmeans(lm.ldl, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "LDL-C",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.ldl <- tidy(lm.ldl) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "LDL-C",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.ldl %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.ldl <- means_wide %>%
  left_join(contrast.ldl, by = c("Outcome", "Obesity"))

ldl.summary %>% kable(caption="Summary of LDL-C levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=ldl.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on LDL-C")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=ldl.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on LDL-C")
```

### Blood Pressure

Blood pressure data is in a separate file (`r bp.filename.cases` and `r bp.filename.controls`).  We have mean noninvasive blood pressure, as well as systolic and diastolic.  These data, similar to lab results need to be mapped to encounter data 

```{r blood-pressure-data-cleaning}
bp.data.cases <- read_csv(bp.filename.cases) |>
  left_join(case.encounters.all) |>
  left_join(case.demographics, by="DeID_PatientID") 

bp.data.controls <- read_csv(bp.filename.controls) |>
  left_join(control.encounters.all) |>
  left_join(control.demographics, by="DeID_PatientID") 

bp.results <- bind_rows(bp.data.controls |> mutate(Cushings=0),
          bp.data.cases |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
bp.results.filtered <-
  bp.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
bp.data.race.ethnicity.age <-
  bp.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
bp.data  <-
  bp.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories
```

#### Inclusion and Exclusion for Blood Pressure Data

We began with `r nrow(bp.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(bp.results |> filter(Cushings==1))-nrow(bp.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - bp.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(bp.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(bp.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics-bp}
bp.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary.bp

library(knitr)
missing.demographic.summary.bp |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data, for blood pressure measures")
```

Our final dataset for blood pressure therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** control patients.


#### Blood Pressure Propensity Mapping

```{r bp-propensity}
m.out.bp <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = bp.data, 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

summary(m.out.bp, standardize = TRUE)  # Balance statistics[5]
plot(m.out.bp, type = "jitter")         # Propensity score distribution[5]

matched_data <- match.data(m.out.bp)

bp.data <- 
  matched_data |>
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) |>
  mutate(MAP_imputed = if_else(
    is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive),
    (BPSysNonInvasive + 2*BPDiaNonInvasive)/3,
    BPMeanNonInvasive
  ))

# Calculate agreement statistics
map_agreement <- bp.data %>%
  filter(!is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive)) %>%
  mutate(MAP_calc = (BPSysNonInvasive + 2*BPDiaNonInvasive)/3,
         difference = BPMeanNonInvasive - MAP_calc) %>%
  summarise(
    mean_diff = round(mean(difference), 2),
    sd_diff = round(sd(difference), 2),
    correlation = round(cor(BPMeanNonInvasive, MAP_calc), 3)
  )

# Count how many will be imputed
n_imputed <- bp.data %>%
  filter(is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive)) %>%
  nrow()
```

Mean arterial pressure was directly measured in the majority of cases. When MAP was missing but systolic and diastolic BP were available (n=`r n_imputed`), MAP was calculated as (SBP + 2×DBP)/3. The calculated values showed high agreement with measured MAP (r=`r map_agreement$correlation`, mean difference=`r map_agreement$mean_diff` ± `r map_agreement$sd_diff` mmHg) in cases where both were available.


#### Demographics of Participants with Blood Pressure Values

```{r bp-demographics}
bp.summary.gender <-
  bp.data |>
  tabyl(GenderName,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=GenderName)|>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

bp.summary.race.ethnicity <-
  bp.data |>
  tabyl(RaceEthnicity,Cushings) |>
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns(position = "front") %>%
  mutate(across(starts_with("."), ~ paste0(.x))) %>%
  mutate(across(starts_with("."), ~ gsub("(.+)% \\((.+)\\)", "\\2 (\\1%)", .x))) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=RaceEthnicity) |>
  mutate(Cushings_count = as.numeric(str_extract(`Cushing's`, "^\\d+"))) |>
  arrange(desc(Cushings_count)) |>
  select(-Cushings_count)

bp.summary.quant <-
  bp.data |>
  group_by(Cushings) %>%
  summarise(
    Age_Mean = mean(AgeInYears, na.rm = TRUE),
    Age_SD = sd(AgeInYears),
    BMI_Mean = mean(BMI, na.rm = TRUE),
    BMI_SD = sd(BMI)
  ) %>%
  mutate(
    Age = sprintf("%.2f ± %.2f", Age_Mean, Age_SD),
    BMI = sprintf("%.2f ± %.2f", BMI_Mean, BMI_SD)
  ) %>%
  select(Cushings, Age, BMI)  |>
  pivot_longer(cols=c('Age','BMI')) |>
  pivot_wider(names_from=Cushings) |>
  rename(Controls=`0`,
         `Cushing's`=`1`,
         Group=name)


bp.summary.table <-
  bind_rows(bp.summary.quant,
          bp.summary.gender,
          bp.summary.race.ethnicity)

bp.summary.table |>
  kable(caption="Demographic Summary for Blood Pressure List") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(bp.summary.table,"Demographic Summary - Blood Pressure Sample.csv")

#statistics for BMI
bp.data |>
  group_by(Cushings) |>
  summarize(shapiro.p=shapiro.test(sample(BMI,3000,replace=T))$p.value) |>
  kable(caption="Shapiro Wilk test for BMI for Blood Pressure sample", digits=c(1,99)) |>
  kable_styling(full_width = FALSE, position = "center")

wilcox.test(BMI~Cushings,bp.data) |>
  tidy() |>
  kable(caption="Mann Whitney test for effect of Cushing's on BMI for Blood Pressure sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")

chisq.test(x=as.numeric(separate(bp.summary.gender, `Cushing's`, sep=" ", into=c("n","Pct")) |>
                          pull(n)),p=c(0.5,0.5)) |>
  tidy() |>
  kable(caption="Chi-squared test for equal proportions of males and females in the Blood Pressure sample",
        digits=c(1,99,1,1)) |>
  kable_styling(full_width = FALSE, position = "center")
```

#### Mean Arterial Pressure

```{r bp-map}
#MAP
bp.summary.map <-
  bp.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(MAP_imputed,na.rm=T),
            se=se(MAP_imputed),
            sd=sd(MAP_imputed,na.rm=T),
            n=length(MAP_imputed))

ggplot(bp.summary.map,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Mean; mmHg)", x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(bp.summary.map,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Mean; mmHg)", x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(bp.data,
       aes(y=MAP_imputed,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Mean; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(bp.data,
       aes(y=MAP_imputed,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Mean; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

bp.summary.map %>% kable(caption="Summary of blood pressure levels by obesity and pre-existing Cushing's (Mean)")
lm(MAP_imputed ~ Cushings + Obesity + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Blood Pressure (Mean)")

lm(MAP_imputed ~ Cushings + BMI + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Blood Pressure (Mean)")
```

##### MAP Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r map-statistics}
#used fully adjusted model
lm.map <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.map <- emmeans(lm.map, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.map <- contrast(emmeans(lm.map, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.map <- tidy(lm.map) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.map %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.map <- means_wide %>%
  left_join(contrast.map, by = c("Outcome", "Obesity"))
```


#### Systolic Blood Pressure

```{r bp-systolic}
#systolic
bp.summary.sys <-
  bp.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(BPSysNonInvasive,na.rm=T),
            se=se(BPSysNonInvasive),
            sd=sd(BPSysNonInvasive,na.rm=T),
            n=length(BPSysNonInvasive))

ggplot(bp.summary.sys,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Systolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(bp.summary.sys,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Systolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(bp.data,
       aes(y=BPSysNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Systolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(bp.data,
       aes(y=BPSysNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Systolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")



bp.summary.sys %>% kable(caption="Summary of blood pressure levels by obesity and pre-existing Cushing's (Systolic)")
lm(BPSysNonInvasive ~ Cushings + Obesity + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Blood Pressure (Systolic)")

lm(BPSysNonInvasive ~ Cushings + BMI + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Blood Pressure (Systolic)")
```

##### Systolic Blood Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r sbp-statistics}
#used fully adjusted model
lm.sbp <- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.sbp <- emmeans(lm.sbp, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.sbp <- contrast(emmeans(lm.sbp, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.sbp <- tidy(lm.sbp) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.sbp %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.sbp <- means_wide %>%
  left_join(contrast.sbp, by = c("Outcome", "Obesity"))
```

#### Diastolic Arterial Pressure

```{r bp-diastolic}
#MAP
bp.summary.dia <-
  bp.data %>%
  group_by(Cushings,Obesity) |>
  summarize(mean=mean(BPDiaNonInvasive,na.rm=T),
            se=se(BPDiaNonInvasive),
            sd=sd(BPDiaNonInvasive,na.rm=T),
            n=length(BPDiaNonInvasive))

ggplot(bp.summary.dia,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Diastolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(bp.summary.dia,
       aes(y=mean,
           ymin=mean- se,
           ymax=mean+ se,
           x=Obesity,
           fill=as.factor(Cushings))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(
    position = position_dodge(width = 0.9), # must match geom_col
    width = 0.5
  ) +
  labs(y="Blood Pressure (Diastolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

ggplot(bp.data,
       aes(y=BPDiaNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Diastolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_grey() +
  theme(legend.position="none")

ggplot(bp.data,
       aes(y=BPDiaNonInvasive,
           x=BMI,
           col=as.factor(Cushings))) +
  geom_point() +
  stat_smooth(method="loess", se=F) +
  labs(y="Blood Pressure (Diastolic; mmHg)",x="") +
  geom_hline(yintercept = 60, linetype = "solid", color = "gray30") +
  coord_cartesian(ylim = c(60, NA)) +  # Use this instead of scale_y_continuous
  theme_classic(base_size=16) +
  scale_fill_manual(values=color_scheme) +
  theme(legend.position="none")

bp.summary.dia %>% kable(caption="Summary of blood pressure levels by obesity and pre-existing Cushing's (Diastoic)")
lm(BPDiaNonInvasive ~ Cushings + Obesity + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Blood Pressure (Diastolic)")

lm(BPDiaNonInvasive ~ Cushings + BMI + Cushings:Obesity,data=bp.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Blood Pressure (Diastolic)")
```

##### Diastolic Blood Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r dbp-statistics}
#used fully adjusted model
lm.dbp <- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.dbp <- emmeans(lm.dbp, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.dbp <- contrast(emmeans(lm.dbp, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.dbp <- tidy(lm.dbp) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.dbp %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.dbp <- means_wide %>%
  left_join(contrast.dbp, by = c("Outcome", "Obesity"))
```


## Diagnostic Data on Propensity Mapping

```{r propensity-mapping-diagnostics}
# List of matchit objects by outcome
outcomes <- list(
  `Any Outcome` = m.out.all,
  `Blood Pressure` = m.out.bp, 
  Glucose = m.out.glucose, 
  ALT = m.out.alt,
  AST = m.out.ast,
  HbA1c = m.out.a1c, 
  `LDL-C` = m.out.ldl
)

# Function to extract age means ± SE per group
extract_age_summary <- function(m.out, outcome_name) {
  matched_data <- match.data(m.out)
  
  # Pre-matching summary
  pre_summary <- m.out$model$data %>% 
    group_by(Cushings) %>% 
    summarise(
      Mean = mean(AgeInYears, na.rm = TRUE),
      SE = sd(AgeInYears, na.rm = TRUE)/sqrt(n())
    ) %>% 
    mutate(Stage = "Pre-matching")
  
  # Post-matching summary
  post_summary <- matched_data %>% 
    group_by(Cushings) %>% 
    summarise(
      Mean = mean(AgeInYears, na.rm = TRUE),
      SE = sd(AgeInYears, na.rm = TRUE)/sqrt(n())
    ) %>% 
    mutate(Stage = "Post-matching")
  
  bind_rows(pre_summary, post_summary) %>% 
    mutate(Outcome = outcome_name) %>%
    select(Outcome, Stage, Cushings, Mean, SE)
}

# Function to extract SMD for AgeInYears
extract_age_smd <- function(m.out, outcome_name){
  s <- summary(m.out, standardize = TRUE)
  
  bal_all <- s$sum.all      # pre-matching balance
  bal_matched <- s$sum.matched  # post-matching balance
  
  data.frame(
    Outcome = outcome_name,
    Stage = c("Pre-matching", "Post-matching"),
    SMD_Age = c(
      bal_all["AgeInYears","Std. Mean Diff."],
      bal_matched["AgeInYears","Std. Mean Diff."]
    )
  )
}

# Apply functions to all outcomes
age_table <- bind_rows(lapply(names(outcomes), function(x) extract_age_summary(outcomes[[x]], x)))
smd_table <- bind_rows(lapply(names(outcomes), function(x) extract_age_smd(outcomes[[x]], x)))

# Merge BEFORE pivot
full_table <- full_join(age_table, smd_table, by = c("Outcome", "Stage"))

# Format for display
full_table_formatted <- full_table %>%
  mutate(
    Cushings = case_when(Cushings == 0 ~ "Control",
                         Cushings == 1 ~ "Cushing"),
    Stage = factor(Stage, levels = c("Pre-matching", "Post-matching")),
    across(c(Mean, SE, SMD_Age), ~round(., 2))
  ) %>%
  pivot_wider(
    names_from = Cushings,
    values_from = c(Mean, SE)
  ) %>%
  arrange(Outcome, Stage)
library(kableExtra)

# Display table with kableExtra
full_table_formatted %>%
  kable(caption = "Propensity Matching Summary for Age (means ± SE and SMD).") %>%
  kable_styling(full_width = FALSE, position = "center")

propensity.diagnostics.table.pub <-
  full_table_formatted |>
  mutate(Controls= paste0(round(Mean_Control,2)," +/- ",round(SE_Control,2)),
         `Cushing's`= paste0(round(Mean_Cushing,2)," +/- ",round(SE_Cushing,2))) |>
  rename(`Standardized Mean Difference`=SMD_Age) |>
  mutate(`Standardized Mean Difference`=round(`Standardized Mean Difference`,4)) |>
  select(Outcome,Stage,Controls,`Cushing's`,`Standardized Mean Difference`)

kable(propensity.diagnostics.table.pub,caption="Publication Version of Propensity Mapping Summary") |>
  kable_styling(full_width = FALSE, position = "center")


write_csv(propensity.diagnostics.table.pub,"Propensity Matching Summary - Age.csv")

```

### Love Plots

Generated Love plots using ggplot for each outcome.

```{r love-plots}
library(cobalt)
library(ggplot2)

plots <- lapply(names(outcomes), function(name) {
  # Generate Love plot, force only SMDs
  gg <- love.plot(
    outcomes[[name]],
    stats = "mean.diffs",  # only standardized mean differences
    threshold = 0.1,
    abs = TRUE,
    stars = "none"         # avoids mixing with raw differences
  )
  
  # gg is now a ggplot object directly
  gg + 
    theme_classic(base_size=6) +
    labs(
      title = name,
      x = "Standardized Mean Difference",
      color = "Stage",
    ) +
    scale_color_grey() +
    theme(legend.position = c(0.75,0.5))
})

# Example: display HbA1c plot
plots[[1]]
library(cowplot)

combined_plot <- plot_grid(
  plotlist = plots,   # your list of ggplot objects
  ncol = 2,           # number of columns
  labels = names(plots)  # optional: label each subplot with the outcome name
)

# Display
combined_plot

plots <- lapply(names(outcomes), function(name) {
  # Generate Love plot, force only SMDs
  gg <- love.plot(
    outcomes[[name]],
    stats = "mean.diffs",  # only standardized mean differences
    threshold = 0.1,
    abs = TRUE,
    stars = "none"         # avoids mixing with raw differences
  )
  
  # gg is now a ggplot object directly
  gg + 
    theme_classic(base_size=6) +
    labs(
      title = name,
      x = "Standardized Mean Difference",
      color = "Stage",
    ) +
    scale_color_manual(values=color_scheme) +
    theme(legend.position = c(0.75,0.5))
})

# Example: display HbA1c plot
plots[[1]]
library(cowplot)

combined_plot <- plot_grid(
  plotlist = plots,   # your list of ggplot objects
  ncol = 2,           # number of columns
  labels = names(plots)  # optional: label each subplot with the outcome name
)

# Display
combined_plot
```

## Summary of Interaction Effects

```{r interaction-effects-summary}
table_summary <- bind_rows(
  table.glucose,
  table.hba1c,
  table.alt,
  table.ast,
  table.map,
  table.sbp,
  table.dbp
) |>
  select(-contrast)

table_summary |>
  kable(caption="Summary of contrasts for all outcomes") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(table_summary,"Summary of Contrasts - All Outcomes.csv")

interaction_summary <- bind_rows(
  interaction_row.glucose,
  interaction_row.hba1c,
  interaction_row.alt,
  interaction_row.ast,
  interaction_row.map,
  interaction_row.sbp,
  interaction_row.dbp
) |>
  select(Outcome,Interaction_CI,p.value) |>
  rename(`Interaction Estimate`=Interaction_CI)

interaction_summary |>
  kable(caption="Summary of interactions for all outcomes as mean with 95% confidence interval") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(interaction_summary,"Summary of Interactions - All Outcomes.csv")
```

## Session Information

```{r session-information}
sessionInfo()
```
