---
title: "Sensitivity Analyses for Cushings x Obesity Project"
author: "Dave Bridges and Trey Carr"
date: "November 21, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To perform sensitivity analyses for different models and populations

## Data Entry

```{r data-entry}
case.demographics.filename <- 'CaseDemographics.csv'
control.demographics.filename <- 'ControlDemographics.csv'

case.lab.results.filename <- 'LabResults.csv'
control.lab.results.filename <- '../controls/LabResults.csv'

bp.filename.cases <- 'NursingStandardVitalSigns.csv'
bp.filename.controls <- '../controls/NursingStandardVitalSigns.csv'

case.encounter.datafile <- "EncounterAll.csv" 
case.encounter.bmi.datafile <- "EncounterAnthropometricsBMI.csv" 
control.encounter.datafile <- "../controls/EncounterAll.csv" 
control.encounter.bmi.datafile <- "../controls/EncounterAnthropometricsBMI.csv" 

library(readr)
library(dplyr)
library(broom)

#labs dont have dates/ages
case.encounters <- read_csv(case.encounter.datafile)
case.encounters.bmi <- read_csv(case.encounter.bmi.datafile)
case.encounters.all <- left_join(case.encounters,case.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))

control.encounters <- read_csv(control.encounter.datafile)
control.encounters.bmi <- read_csv(control.encounter.bmi.datafile)
control.encounters.all <- left_join(control.encounters,control.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))


case.demographics <- read_csv(case.demographics.filename)
control.demographics <- read_csv(control.demographics.filename)

case.labs <- read_csv(case.lab.results.filename) |>
  left_join(case.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
    left_join(case.demographics, by="DeID_PatientID") # added demographic data
control.labs <- read_csv(control.lab.results.filename) |> 
  left_join(control.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
  left_join(control.demographics, by="DeID_PatientID") # added demographic data
#remove large objects not needed further
#rm(control.encounters,control.encounters.all,control.encounters.bmi)

#combined into one dataset
library(lubridate)

#this is used to filter for the window between when a lab result is allowable relative to the procedure.
procedure.result.interval <- 365 #within a year

lab.results <- bind_rows(control.labs |> mutate(Cushings=0),
                         case.labs |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
lab.results.filtered <-
  lab.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
complete.data.race.ethnicity.age <-
  lab.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
complete.data  <-
  complete.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories

#number of controls per case
fold <- 10
```

We began with `r nrow(lab.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(lab.results |> filter(Cushings==1))-nrow(lab.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - lab.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics}
lab.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary

library(knitr)
missing.demographic.summary |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data")
```

Our final dataset therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** Cushing's patients control patients.

## Propensity Mapping

Used the R package `mapit` to match based on Age, Gender, Race, and Ethnicity.  We wanted exact matches for everything except age, where we wanted the closest case that had a measurement.  We picked `r fold` matched controls for each case

### Population Definitions

We generated participant pools for all potential matched participants, then specifically for each outcome

```{r total-population-propensity}
library(MatchIt)
m.out.all <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data,
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.all <- match.data(m.out.all)

master.data <- 
  matched_data.all |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity)) 
```


### Hba1c

```{r hba1c-propensity}
m.out.a1c <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("A1C", "XXA1C", "A1C EX", "X0022")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.a1c <- match.data(m.out.a1c)

hba1c.data <- 
  matched_data.a1c |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### HbA1c Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r aic-statistics}
library(emmeans)
#used fully adjusted model
lm.hba1c <- lm(value ~ RaceEthnicity + AgeInYears + GenderName + Cushings * Obesity, data = hba1c.data)

## 1. Adjusted means
emm.hba1c <- emmeans(lm.hba1c, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "HbA1c",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.hba1c <- contrast(emmeans(lm.hba1c, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "HbA1c",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.hba1c <- tidy(lm.hba1c) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "HbA1c",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.hba1c %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.hba1c <- means_wide %>%
  left_join(contrast.hba1c, by = c("Outcome", "Obesity"))

lm(value ~ Cushings + BMI + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on HbA1c")
```

### Glucose

```{r glucose-propensity}
m.out.glucose <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("GLUC","GLUC_WB")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.glucose <- match.data(m.out.glucose)

glucose.data <- 
  matched_data.glucose |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### Glucose Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r glucose-statistics}
#used fully adjusted model
lm.glucose <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +  Cushings * Obesity, data = glucose.data)

## 1. Adjusted means
emm.glucose <- emmeans(lm.glucose, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Glucose",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.glucose <- contrast(emmeans(lm.glucose, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Glucose",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.glucose <- tidy(lm.glucose) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Glucose",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.glucose %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.glucose <- means_wide %>%
  left_join(contrast.glucose, by = c("Outcome", "Obesity"))

lm(value ~ Cushings + BMI + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Glucose")
```

### Liver Enzymes

#### ALT

```{r alt-propensity}
m.out.alt <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("ALT")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches
matched_data.alt <- match.data(m.out.alt)

alt.data <- 
  matched_data.alt |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))
```


##### ALT Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r alt-statistics}
#used fully adjusted model
lm.alt <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +  Cushings * Obesity, data = alt.data)

## 1. Adjusted means
emm.alt <- emmeans(lm.alt, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "ALT",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.alt <- contrast(emmeans(lm.alt, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "ALT",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.alt <- tidy(lm.alt) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "ALT",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.alt %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.alt <- means_wide %>%
  left_join(contrast.alt, by = c("Outcome", "Obesity"))

lm(value ~ Cushings + BMI + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on ALT")
```

#### AST

```{r ast-propensity}
m.out.ast <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("AST")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

matched_data.ast <- match.data(m.out.ast)

ast.data <- 
  matched_data.ast |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))
```

##### AST Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r ast-statistics}
#used fully adjusted model
lm.ast <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = ast.data)

## 1. Adjusted means
emm.ast <- emmeans(lm.ast, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "AST",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.ast <- contrast(emmeans(lm.ast, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "AST",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.ast <- tidy(lm.ast) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "AST",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.ast %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.ast <- means_wide %>%
  left_join(contrast.ast, by = c("Outcome", "Obesity"))

lm(value ~ Cushings + BMI + Cushings:Obesity,data=ast.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on AST")
```

### Blood Pressure

Blood pressure data is in a separate file (`r bp.filename.cases` and `r bp.filename.controls`).  We have mean noninvasive blood pressure, as well as systolic and diastolic.  These data, similar to lab results need to be mapped to encounter data 

```{r blood-pressure-data-cleaning}
bp.data.cases <- read_csv(bp.filename.cases) |>
  left_join(case.encounters.all) |>
  left_join(case.demographics, by="DeID_PatientID") 

bp.data.controls <- read_csv(bp.filename.controls) |>
  left_join(control.encounters.all) |>
  left_join(control.demographics, by="DeID_PatientID") 

bp.results <- bind_rows(bp.data.controls |> mutate(Cushings=0),
          bp.data.cases |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
bp.results.filtered <-
  bp.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
bp.data.race.ethnicity.age <-
  bp.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
bp.data  <-
  bp.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories
```

#### Inclusion and Exclusion for Blood Pressure Data

We began with `r nrow(bp.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(bp.results |> filter(Cushings==1))-nrow(bp.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - bp.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(bp.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(bp.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics-bp}
bp.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary.bp

library(knitr)
missing.demographic.summary.bp |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data, for blood pressure measures")
```

Our final dataset for blood pressure therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** control patients.


#### Blood Pressure Propensity Mapping

```{r bp-propensity}
m.out.bp <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = bp.data, 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data <- match.data(m.out.bp)

bp.data <- 
  matched_data |>
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) |>
  mutate(MAP_imputed = if_else(
    is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive),
    (BPSysNonInvasive + 2*BPDiaNonInvasive)/3,
    BPMeanNonInvasive
  ))

# Calculate agreement statistics
map_agreement <- bp.data %>%
  filter(!is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive)) %>%
  mutate(MAP_calc = (BPSysNonInvasive + 2*BPDiaNonInvasive)/3,
         difference = BPMeanNonInvasive - MAP_calc) %>%
  summarise(
    mean_diff = round(mean(difference), 2),
    sd_diff = round(sd(difference), 2),
    correlation = round(cor(BPMeanNonInvasive, MAP_calc), 3)
  )

# Count how many will be imputed
n_imputed <- bp.data %>%
  filter(is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive)) %>%
  nrow()
```

Mean arterial pressure was directly measured in the majority of cases. When MAP was missing but systolic and diastolic BP were available (n=`r n_imputed`), MAP was calculated as (SBP + 2×DBP)/3. The calculated values showed high agreement with measured MAP (r=`r map_agreement$correlation`, mean difference=`r map_agreement$mean_diff` ± `r map_agreement$sd_diff` mmHg) in cases where both were available.

#### Mean Arterial Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r map-statistics}
#used fully adjusted model
lm.map <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.map <- emmeans(lm.map, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.map <- contrast(emmeans(lm.map, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.map <- tidy(lm.map) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.map %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.map <- means_wide %>%
  left_join(contrast.map, by = c("Outcome", "Obesity"))
```


#### Systolic Blood Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r sbp-statistics}
#used fully adjusted model
lm.sbp <- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + GenderName + Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.sbp <- emmeans(lm.sbp, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.sbp <- contrast(emmeans(lm.sbp, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.sbp <- tidy(lm.sbp) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.sbp %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.sbp <- means_wide %>%
  left_join(contrast.sbp, by = c("Outcome", "Obesity"))
```

#### Diastolic Arterial Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r dbp-statistics}
#used fully adjusted model
lm.dbp <- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + GenderName + Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.dbp <- emmeans(lm.dbp, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.dbp <- contrast(emmeans(lm.dbp, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.dbp <- tidy(lm.dbp) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.dbp %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.dbp <- means_wide %>%
  left_join(contrast.dbp, by = c("Outcome", "Obesity"))
```

## Summary of Interaction Effects

```{r interaction-effects-summary}
create_lm_summary_row <- function(model_name, lm.glucose = NULL, lm.hba1c = NULL, 
                                   lm.alt = NULL, lm.ast = NULL, 
                                   lm.map = NULL, lm.sbp = NULL, lm.dbp = NULL, 
                                   coef_name = NULL) {
  
  # Helper function to extract and format results from a single model
  format_estimate <- function(model, coef_name = NULL) {
    if (is.null(model)) return(NA_character_)
    
    # Get coefficient summary
    coef_summary <- summary(model)$coefficients
    
    # If coef_name specified, use that row, otherwise use second row (first predictor)
    if (!is.null(coef_name)) {
      if (!coef_name %in% rownames(coef_summary)) return(NA_character_)
      row_idx <- which(rownames(coef_summary) == coef_name)
    } else {
      row_idx <- 2  # Default to first predictor (row 2, since row 1 is intercept)
    }
    
    estimate <- coef_summary[row_idx, "Estimate"]
    se <- coef_summary[row_idx, "Std. Error"]
    p_value <- coef_summary[row_idx, "Pr(>|t|)"]
    
    # Calculate 95% CI
    ci_lower <- estimate - 1.96 * se
    ci_upper <- estimate + 1.96 * se
    
    # Format the string
    result <- sprintf("%.2f [%.2f, %.2f]", estimate, ci_lower, ci_upper)
    
    # Add asterisk if p < 0.05
    if (p_value < 0.05) {
      result <- paste0(result, "*")
    }
    
    return(result)
  }
  
  # Create the row
  tibble(
    `Primary outcome` = model_name,
    Glucose = format_estimate(lm.glucose, coef_name),
    HbA1c = format_estimate(lm.hba1c, coef_name),
    ALT = format_estimate(lm.alt, coef_name),
    AST = format_estimate(lm.ast, coef_name),
    MAP = format_estimate(lm.map, coef_name),
    SBP = format_estimate(lm.sbp, coef_name),
    DBP = format_estimate(lm.dbp, coef_name)
  )
}

row1 <- create_lm_summary_row(
  model_name = "Main Analysis",
  lm.glucose = lm.glucose,
  lm.hba1c = lm.hba1c,
  lm.alt = lm.alt,
  lm.ast = lm.ast,
  lm.map = lm.map,
  lm.sbp = lm.sbp,
  lm.dbp = lm.dbp,
  coef_name = "Cushings:ObesityObese"
)
```

## Alternate Models

### Modifying Covariates

#### Crude models

This includes models with no covariates other than Cushings $\times$ Obesity

```{r crude-models}
lm.hba1c.crude <- lm(value ~ Cushings * Obesity, data = hba1c.data )
lm.glucose.crude <- lm(value ~ Cushings * Obesity, data = glucose.data )
lm.alt.crude <- lm(value ~ Cushings * Obesity, data = alt.data )
lm.ast.crude <- lm(value ~ Cushings * Obesity, data = ast.data )
lm.map.crude <- lm(MAP_imputed ~ Cushings * Obesity, data = bp.data )
lm.sbp.crude <- lm(BPSysNonInvasive ~ Cushings * Obesity, data = bp.data )
lm.dbp.crude <- lm(BPDiaNonInvasive ~ Cushings * Obesity, data = bp.data )

crude.models <- create_lm_summary_row(
  model_name = "Unadjusted Models",
  lm.glucose = lm.glucose.crude,
  lm.hba1c = lm.hba1c.crude,
  lm.alt = lm.alt.crude,
  lm.ast = lm.ast.crude,
  lm.map = lm.map.crude,
  lm.sbp = lm.sbp.crude,
  lm.dbp = lm.dbp.crude,
  coef_name = "Cushings:ObesityObese"
)
```

#### Gender adjusted only

```{r gender-only-models}
lm.hba1c.gender.only <- lm(value ~ GenderCode + Cushings * Obesity, data = hba1c.data )
lm.glucose.gender.only <- lm(value ~ GenderCode + Cushings * Obesity, data = glucose.data )
lm.alt.gender.only <- lm(value ~ GenderCode + Cushings * Obesity, data = alt.data )
lm.ast.gender.only <- lm(value ~ GenderCode + Cushings * Obesity, data = ast.data )
lm.map.gender.only <- lm(MAP_imputed ~ GenderCode + Cushings * Obesity, data = bp.data )
lm.sbp.gender.only <- lm(BPSysNonInvasive ~ GenderCode + Cushings * Obesity, data = bp.data )
lm.dbp.gender.only <- lm(BPDiaNonInvasive ~ GenderCode + Cushings * Obesity, data = bp.data )

gender.only.models <- create_lm_summary_row(
  model_name = "Adjusting for Gender Only",
  lm.glucose = lm.glucose.gender.only,
  lm.hba1c = lm.hba1c.gender.only,
  lm.alt = lm.alt.gender.only,
  lm.ast = lm.ast.gender.only,
  lm.map = lm.map.gender.only,
  lm.sbp = lm.sbp.gender.only,
  lm.dbp = lm.dbp.gender.only,
  coef_name = "Cushings:ObesityObese"
)
```

#### Age adjusted only

```{r age-only-models}
lm.hba1c.age.only <- lm(value ~ AgeInYears + Cushings * Obesity, data = hba1c.data )
lm.glucose.age.only <- lm(value ~ AgeInYears + Cushings * Obesity, data = glucose.data )
lm.alt.age.only <- lm(value ~ AgeInYears + Cushings * Obesity, data = alt.data )
lm.ast.age.only <- lm(value ~ AgeInYears + Cushings * Obesity, data = ast.data )
lm.map.age.only <- lm(MAP_imputed ~ AgeInYears + Cushings * Obesity, data = bp.data )
lm.sbp.age.only <- lm(BPSysNonInvasive ~ AgeInYears + Cushings * Obesity, data = bp.data )
lm.dbp.age.only <- lm(BPDiaNonInvasive ~ AgeInYears + Cushings * Obesity, data = bp.data )

age.only.models <- create_lm_summary_row(
  model_name = "Adjusting for Age Only",
  lm.glucose = lm.glucose.age.only,
  lm.hba1c = lm.hba1c.age.only,
  lm.alt = lm.alt.age.only,
  lm.ast = lm.ast.age.only,
  lm.map = lm.map.age.only,
  lm.sbp = lm.sbp.age.only,
  lm.dbp = lm.dbp.age.only,
  coef_name = "Cushings:ObesityObese"
)
```

#### Race/ethnicity adjusted only

```{r race-only-models}
lm.hba1c.race.only <- lm(value ~ RaceEthnicity + Cushings * Obesity, data = hba1c.data )
lm.glucose.race.only <- lm(value ~ RaceEthnicity + Cushings * Obesity, data = glucose.data )
lm.alt.race.only <- lm(value ~ RaceEthnicity + Cushings * Obesity, data = alt.data )
lm.ast.race.only <- lm(value ~ RaceEthnicity + Cushings * Obesity, data = ast.data )
lm.map.race.only <- lm(MAP_imputed ~ RaceEthnicity + Cushings * Obesity, data = bp.data )
lm.sbp.race.only <- lm(BPSysNonInvasive ~ RaceEthnicity + Cushings * Obesity, data = bp.data )
lm.dbp.race.only <- lm(BPDiaNonInvasive ~ RaceEthnicity + Cushings * Obesity, data = bp.data )

race.only.models <- create_lm_summary_row(
  model_name = "Adjusting for Race and Ethnicity Only",
  lm.glucose = lm.glucose.race.only,
  lm.hba1c = lm.hba1c.race.only,
  lm.alt = lm.alt.race.only,
  lm.ast = lm.ast.race.only,
  lm.map = lm.map.race.only,
  lm.sbp = lm.sbp.race.only,
  lm.dbp = lm.dbp.race.only,
  coef_name = "Cushings:ObesityObese"
)
```

### Sex Stratified models

```{r sex-stratified}
lm.hba1c.m <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = hba1c.data |> filter(GenderCode=="M"))
lm.glucose.m <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = glucose.data |> filter(GenderCode=="M"))
lm.alt.m <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = alt.data |> filter(GenderCode=="M"))
lm.ast.m <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = ast.data |> filter(GenderCode=="M"))
lm.map.m <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(GenderCode=="M"))
lm.sbp.m <- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(GenderCode=="M"))
lm.dbp.m <- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(GenderCode=="M"))

males.only <- create_lm_summary_row(
  model_name = "Male Patients",
  lm.glucose = lm.glucose.m,
  lm.hba1c = lm.hba1c.m,
  lm.alt = lm.alt.m,
  lm.ast = lm.ast.m,
  lm.map = lm.map.m,
  lm.sbp = lm.sbp.m,
  lm.dbp = lm.dbp.m,
  coef_name = "Cushings:ObesityObese"
)



lm.hba1c.f <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = hba1c.data |> filter(GenderCode=="F"))
lm.glucose.f <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = glucose.data |> filter(GenderCode=="F"))
lm.alt.f <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = alt.data |> filter(GenderCode=="F"))
lm.ast.f <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = ast.data |> filter(GenderCode=="F"))
lm.map.f <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(GenderCode=="F"))
lm.sbp.f <- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(GenderCode=="F"))
lm.dbp.f <- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(GenderCode=="F"))

females.only <- create_lm_summary_row(
  model_name = "Female Patients",
  lm.glucose = lm.glucose.f,
  lm.hba1c = lm.hba1c.f,
  lm.alt = lm.alt.f,
  lm.ast = lm.ast.f,
  lm.map = lm.map.f,
  lm.sbp = lm.sbp.f,
  lm.dbp = lm.dbp.f,
  coef_name = "Cushings:ObesityObese"
)
```

### Different Treatment of BMI

#### BMI after group-based stratification

```{r bmi-within-group}
lm.hba1c.bmi.within.group <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity + BMI, data = hba1c.data )
lm.glucose.bmi.within.group <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity + BMI, data = glucose.data )
lm.alt.bmi.within.group <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity + BMI, data = alt.data )
lm.ast.bmi.within.group <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity + BMI, data = ast.data )
lm.map.bmi.within.group <- lm(MAP_imputed ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = bp.data )
lm.sbp.bmi.within.group <- lm(BPSysNonInvasive ~ GenderCode + RaceEthnicity + Cushings * Obesity + BMI, data = bp.data )
lm.dbp.bmi.within.group <- lm(BPDiaNonInvasive ~ GenderCode + RaceEthnicity + Cushings * Obesity + BMI, data = bp.data )

bmi.within.group.models <- create_lm_summary_row(
  model_name = "Adjusting for BMI within Group",
  lm.glucose = lm.glucose.bmi.within.group,
  lm.hba1c = lm.hba1c.bmi.within.group,
  lm.alt = lm.alt.bmi.within.group,
  lm.ast = lm.ast.bmi.within.group,
  lm.map = lm.map.bmi.within.group,
  lm.sbp = lm.sbp.bmi.within.group,
  lm.dbp = lm.dbp.bmi.within.group,
  coef_name = "Cushings:ObesityObese"
)
```

#### BMI as linear

In sensitivity analyses modeling BMI as a continuous predictor (using either a linear term or restricted cubic splines), the obesity effect and its interaction with Cushing’s syndrome were summarized by contrasting predicted outcomes at two BMI values: the observed mean BMI among participants with BMI $geq$≥30 kg/m$^2$ and the observed mean BMI among participants with BMI $<$30 kg/m$^2$. Predictions were generated from the fitted model while averaging over all other covariates. This contrast yields a clinically interpretable estimate directly comparable to the coefficient obtained from models that use a dichotomous BMI $geq$≥30 kg/m$^2$ indicator. Results are reported as mean differences with 95% confidence intervals, with statistical significance denoted by * ($p < 0.05$).

```{r bmi-linear}
lm.hba1c.bmi.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * BMI, data = hba1c.data )
lm.glucose.bmi.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * BMI, data = glucose.data )
lm.alt.bmi.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * BMI, data = alt.data )
lm.ast.bmi.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * BMI, data = ast.data )
lm.map.bmi.linear <- lm(MAP_imputed ~ GenderCode + RaceEthnicity + Cushings * BMI, data = bp.data )
lm.sbp.bmi.linear <- lm(BPSysNonInvasive ~ GenderCode + RaceEthnicity + Cushings * BMI, data = bp.data )
lm.dbp.bmi.linear <- lm(BPDiaNonInvasive ~ GenderCode + RaceEthnicity + Cushings * BMI, data = bp.data )


#calculating contrasts between lean and obese for nonlinear and linear BMI models
library(emmeans)

bmi_obese_vs_lean_contrast <- function(model, data = NULL) {
  
  if (is.null(data)) data <- model.frame(model)
  
  # Compute mean BMI for <30 and >=30
  means_bmi <- data %>%
    filter(!is.na(BMI)) %>%
    mutate(obese = BMI >= 30) %>%
    group_by(obese) %>%
    summarise(mean_bmi = mean(BMI), .groups = "drop")
  
  bmi_lean  <- means_bmi$mean_bmi[means_bmi$obese == FALSE]
  bmi_obese <- means_bmi$mean_bmi[means_bmi$obese == TRUE]
  
  # Use emtrends for numeric BMI
  emm <- emtrends(
    model,
    specs = ~ Cushings,
    var = "BMI",
    at = list(BMI = c(bmi_lean, bmi_obese)),
    infer = c(TRUE, TRUE)
  )
  
  # Contrast obese vs lean BMI
  contr <- contrast(
    emm,
    method = list("Obese – Lean" = c(1, -1)),
    infer = c(TRUE, TRUE)
  )
  
  s <- as.data.frame(summary(contr))
  
  # Detect CI column names
  possible_lower <- c("lower.CL","asymp.LCL","LCL","lower")
  possible_upper <- c("upper.CL","asymp.UCL","UCL","upper")
  low  <- intersect(possible_lower, names(s))[1]
  high <- intersect(possible_upper, names(s))[1]
  
  # Build formatted string
  formatted <- sprintf("%.2f [%.2f, %.2f]%s",
                       s$estimate,
                       s[[low]],
                       s[[high]],
                       ifelse(s$p.value < 0.05, "*", ""))
  
  # Return only the formatted string
  return(formatted)
}



bmi_obese_vs_lean_contrast(lm.glucose.bmi.linear, data = glucose.data) 

bmi.linear.models <- data.frame(
  Glucose = bmi_obese_vs_lean_contrast(lm.glucose.bmi.linear, data = glucose.data) ,
  HbA1c = bmi_obese_vs_lean_contrast(lm.hba1c.bmi.linear, data = hba1c.data) ,
  ALT = bmi_obese_vs_lean_contrast(lm.alt.bmi.linear, data = alt.data),
  AST = bmi_obese_vs_lean_contrast(lm.ast.bmi.linear, data = ast.data),
  MAP = bmi_obese_vs_lean_contrast(lm.map.bmi.linear, data = bp.data),
  SBP = bmi_obese_vs_lean_contrast(lm.sbp.bmi.linear, data = bp.data),
  DBP = bmi_obese_vs_lean_contrast(lm.dbp.bmi.linear, data = bp.data)
)
```

#### BMI as splined nonlinear covariate

Used the splines package.  `ns(BMI, df = 4)` fits a natural (restricted) cubic spline with 4 total degrees of freedom, which places 3 interior knots (by default at the 5th, 50th, and 95th percentiles of BMI) and forces the function to be linear beyond the outermost knots. This creates a smooth, flexible curve that can bend up to three times but remains well-behaved and avoids wild extrapolations at the extremes.

Decades of epidemiological research (Framingham, Nurses’ Health, ARIC, UK Biobank, *etc.*) have repeatedly shown that 4 df are sufficient to capture the true non-linear shapes of BMI with outcomes like diabetes, cardiovascular events, and mortality, without overfitting noise. It reliably reproduces the typical pattern of slowly rising risk at low-to-normal BMI, steepening in the overweight/obese range, and occasional flattening at very high BMI, while keeping confidence intervals honest in the tails where data are sparse. Using fewer df ($\leq$3) often forces an oversimplified shape that misses real curvature, whereas more than 5 df rarely improves fit meaningfully and risks spurious wiggles unless you have thousands of observations.

```{r bmi-non-linear}
library(splines)
lm.hba1c.bmi.non.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = hba1c.data )
lm.glucose.bmi.non.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = glucose.data )
lm.alt.bmi.non.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = alt.data )
lm.ast.bmi.non.linear <- lm(value ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = ast.data )
lm.map.bmi.non.linear <- lm(MAP_imputed ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = bp.data )
lm.sbp.bmi.non.linear <- lm(BPSysNonInvasive ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = bp.data )
lm.dbp.bmi.non.linear <- lm(BPDiaNonInvasive ~ GenderCode + RaceEthnicity + Cushings * ns(BMI, df = 4), data = bp.data )

library(emmeans)
library(dplyr)
library(splines)

bmi_obese_vs_lean_contrast_spline <- function(model, data = NULL) {
  
  if (is.null(data)) data <- model.frame(model)
  
  # 1. Compute mean BMI values within lean vs obese groups
  means_bmi <- data %>%
    filter(!is.na(BMI)) %>%
    mutate(obese = BMI >= 30) %>%
    group_by(obese) %>%
    summarise(mean_bmi = mean(BMI), .groups = "drop")
  
  bmi_lean  <- means_bmi$mean_bmi[!means_bmi$obese]
  bmi_obese <- means_bmi$mean_bmi[means_bmi$obese]
  
  # 2. Build reference grid explicitly so splines reconstruct correctly
  rg <- ref_grid(
    model,
    data = data,
    at = list(BMI = c(bmi_lean, bmi_obese)),
    cov.reduce = FALSE
  )
  
  # 3. Predicted means for Cushing × BMI combinations
  emm <- emmeans(rg, ~ Cushings * BMI)
  
  # 4. Contrast vectors (must match row order in emm)
  # Order is:
  #   Cushing=0,BMI=lean
  #   Cushing=0,BMI=obese
  #   Cushing=1,BMI=lean
  #   Cushing=1,BMI=obese
  obesity_main     <- c(-1,  1, -1,  1)
  cushing_interact <- c( 0,  0, -1,  1) - c(-1,  1,  0,  0)
  
  contr_list <- list(
    Obesity_main     = obesity_main,
    Cushing_interact = cushing_interact
  )
  
  contr <- contrast(emm, method = contr_list, infer = TRUE)
  s <- as.data.frame(summary(contr))
  
  # Keep only the interaction contrast
  s <- s[s$contrast == "Cushing_interact", ]
  
  # Detect CI columns
  ci_lower <- intersect(c("lower.CL","asymp.LCL","LCL","lower"), names(s))[1]
  ci_upper <- intersect(c("upper.CL","asymp.UCL","UCL","upper"), names(s))[1]
  
  # Format for results table
  formatted <- sprintf(
    "%.2f [%.2f, %.2f]%s",
    s$estimate,
    s[[ci_lower]],
    s[[ci_upper]],
    ifelse(s$p.value < 0.05, "*", "")
  )
  
  return(formatted)
}


bmi.non.linear.models <- data.frame(
  Glucose = bmi_obese_vs_lean_contrast_spline(lm.glucose.bmi.non.linear, data = glucose.data) ,
  HbA1c = bmi_obese_vs_lean_contrast_spline(lm.hba1c.bmi.non.linear, data = hba1c.data) ,
  ALT = bmi_obese_vs_lean_contrast_spline(lm.alt.bmi.non.linear, data = alt.data),
  AST = bmi_obese_vs_lean_contrast_spline(lm.ast.bmi.non.linear, data = ast.data),
  MAP = bmi_obese_vs_lean_contrast_spline(lm.map.bmi.non.linear, data = bp.data),
  SBP = bmi_obese_vs_lean_contrast_spline(lm.sbp.bmi.non.linear, data = bp.data),
  DBP = bmi_obese_vs_lean_contrast_spline(lm.dbp.bmi.non.linear, data = bp.data)
)
```

#### BMI as categories

##### BMI Class I

```{r classI-stratified}
lm.hba1c.ci <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = hba1c.data |> filter(BMI<=35))
lm.glucose.ci <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = glucose.data |> filter(BMI<=35))
lm.alt.ci <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = alt.data |> filter(BMI<=35))
lm.ast.ci <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = ast.data |> filter(BMI<=35))
lm.map.ci <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI<=35))
lm.sbp.ci<- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI<=35))
lm.dbp.ci<- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI<=35))

classI.obesity.only <- create_lm_summary_row(
  model_name = "Class I Obesity",
  lm.glucose = lm.glucose.ci,
  lm.hba1c = lm.hba1c.ci,
  lm.alt = lm.alt.ci,
  lm.ast = lm.ast.ci,
  lm.map = lm.map.ci,
  lm.sbp = lm.sbp.ci,
  lm.dbp = lm.dbp.ci,
  coef_name = "Cushings:ObesityObese"
)
```

##### BMI Class II

```{r classII-stratified}
lm.hba1c.cii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = hba1c.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))
lm.glucose.cii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = glucose.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))
lm.alt.cii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = alt.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))
lm.ast.cii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = ast.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))
lm.map.cii <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))
lm.sbp.cii<- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))
lm.dbp.cii<- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI < 30 | (BMI >= 35 & BMI < 40)))

classII.obesity.only <- create_lm_summary_row(
  model_name = "Class II Obesity",
  lm.glucose = lm.glucose.cii,
  lm.hba1c = lm.hba1c.cii,
  lm.alt = lm.alt.cii,
  lm.ast = lm.ast.cii,
  lm.map = lm.map.cii,
  lm.sbp = lm.sbp.cii,
  lm.dbp = lm.dbp.cii,
  coef_name = "Cushings:ObesityObese"
)
```

##### BMI Class III

```{r classIII-stratified}
lm.hba1c.ciii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = hba1c.data |> filter(BMI < 30 | BMI >= 40))
lm.glucose.ciii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = glucose.data |> filter(BMI < 30 | BMI >= 40))
lm.alt.ciii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = alt.data |> filter(BMI < 30 | BMI >= 40))
lm.ast.ciii <- lm(value ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = ast.data |> filter(BMI < 30 | BMI >= 40))
lm.map.ciii <- lm(MAP_imputed ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI < 30 | BMI >= 40))
lm.sbp.ciii<- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI < 30 | BMI >= 40))
lm.dbp.ciii<- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + Cushings * Obesity, data = bp.data |> filter(BMI < 30 | BMI >= 40))

classIII.obesity.only <- create_lm_summary_row(
  model_name = "Class III Obesity",
  lm.glucose = lm.glucose.ciii,
  lm.hba1c = lm.hba1c.ciii,
  lm.alt = lm.alt.ciii,
  lm.ast = lm.ast.ciii,
  lm.map = lm.map.ciii,
  lm.sbp = lm.sbp.ciii,
  lm.dbp = lm.dbp.ciii,
  coef_name = "Cushings:ObesityObese"
)
```


#### Matching Participants by BMI

```{r matching-bmi}
m.out.a1c.bmi <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity + BMI, 
                 data = complete.data |> filter(RESULT_CODE %in% c("A1C", "XXA1C", "A1C EX", "X0022")) |> filter(!is.na(BMI)) |> filter(is.finite(BMI)), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.a1c.bmi <- match.data(m.out.a1c.bmi)

hba1c.data.bmi <- 
  matched_data.a1c.bmi |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 

m.out.glucose.bmi <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity + BMI, 
                 data = complete.data |> filter(RESULT_CODE %in% c("GLUC","GLUC_WB"))|> filter(!is.na(BMI)) |> filter(is.finite(BMI)), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.glucose.bmi <- match.data(m.out.glucose.bmi)

glucose.data.bmi <- 
  matched_data.glucose.bmi |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity))

m.out.alt.bmi <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity + BMI, 
                 data = complete.data |> filter(RESULT_CODE %in% c("ALT"))|> filter(!is.na(BMI)) |> filter(is.finite(BMI)), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches
matched_data.alt.bmi <- match.data(m.out.alt.bmi)

alt.data.bmi <- 
  matched_data.alt.bmi |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))

m.out.ast.bmi <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity + BMI, 
                 data = complete.data |> filter(RESULT_CODE %in% c("AST"))|> filter(!is.na(BMI)) |> filter(is.finite(BMI)), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches
matched_data.ast.bmi <- match.data(m.out.ast.bmi)

ast.data.bmi <- 
  matched_data.ast.bmi |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))

m.out.bp.bmi <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity + BMI, 
                 data = bp.data |> filter(!is.na(BMI)) |> filter(is.finite(BMI)) |> select(-weights,-distance, -subclass), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.bp.bmi <- match.data(m.out.bp.bmi)

bp.data.bmi <- 
  matched_data.bp.bmi |>
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) |>
  mutate(MAP_imputed = if_else(
    is.na(BPMeanNonInvasive) & !is.na(BPSysNonInvasive) & !is.na(BPDiaNonInvasive),
    (BPSysNonInvasive + 2*BPDiaNonInvasive)/3,
    BPMeanNonInvasive
  ))

lm.hba1c.bmi.matched <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = hba1c.data.bmi )
lm.glucose.bmi.matched <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = glucose.data.bmi )
lm.alt.bmi.matched <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = alt.data.bmi )
lm.ast.bmi.matched <- lm(value ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = ast.data.bmi )
lm.map.bmi.matched <- lm(MAP_imputed ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = bp.data.bmi )
lm.sbp.bmi.matched <- lm(BPSysNonInvasive ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = bp.data.bmi )
lm.dbp.bmi.matched <- lm(BPDiaNonInvasive ~ GenderCode + RaceEthnicity + Cushings * Obesity, data = bp.data.bmi )

bmi.matched.models <- create_lm_summary_row(
  model_name = "Matching for BMI",
  lm.glucose = lm.glucose.bmi.matched,
  lm.hba1c = lm.hba1c.bmi.matched,
  lm.alt = lm.alt.bmi.matched,
  lm.ast = lm.ast.bmi.matched,
  lm.map = lm.map.bmi.matched,
  lm.sbp = lm.sbp.bmi.matched,
  lm.dbp = lm.dbp.bmi.matched,
  coef_name = "Cushings:ObesityObese"
)
```


## Summary of Sensitivity Analyses

```{r summary-sensitivity}
summary.sensitivity.data <- bind_rows(row1,
                                      crude.models,
                                      age.only.models, 
                                      gender.only.models, 
                                      race.only.models, 
                                      males.only, 
                                      females.only,
                                      bmi.within.group.models,
                                      #bmi.linear.models |> mutate(bmi.linear.models,`Primary outcome`="BMI (Linear)"),
                                      bmi.non.linear.models|> mutate(bmi.linear.models,`Primary outcome`="BMI (Non-Linear)"),
                                      bmi.matched.models,
                                      classI.obesity.only,
                                      classII.obesity.only,
                                      classIII.obesity.only
                                      ) |>
  rename(Model = `Primary outcome`)

library(kableExtra)
summary.sensitivity.data |>
  kable(caption="Summary of sensitivity analyses") |>
  kable_styling(full_width = FALSE, position = "center")

summary.sensitivity.data |> write_csv("Sensitivity Analyses.csv")
```

Changes in matching
- n = 1:1, 1:5
- match by calendar year, not possible with existing data (recency bias)
- imputed missing data
- NB You should avoid imputing HbA1c because its missingness is plausibly missing not at random (MNAR): clinicians order HbA1c selectively when they suspect hyperglycaemia or known diabetes, so the probability a value is observed is directly related to the (unobserved) true HbA1c. Imputing under a MAR assumption will therefore produce biased estimates (and with only ~20 observed cases the imputations will be highly unstable), so treat HbA1c as a complete-case outcome or, if you must address missingness, use explicit MNAR sensitivity models (e.g., pattern-mixture or selection models) and report how much those assumptions would need to change to alter your conclusions.
- Gender term
- Stratified by adrenal/pituitary
- Including "non-treated"

## Session Information

```{r session-information}
sessionInfo()
```
