---
title: "Sensitivity Analyses for Cushings x Obesity Project"
author: "Dave Bridges and Trey Carr"
date: "November 21, 2025"
format:
  html:
    toc: true
    toc-location: right
    keep-md: true
    code-fold: true
    code-summary: "Show the code"
  pdf: default
knitr:
  opts_chunk:
    fig.path: "figures/"
    dev: ["png", "pdf"]  # Remove !expr, just use array syntax
    fig.keep: "all"
execute:
  echo: true
  warning: false
---

```{r global_options}
# hide this code chunk
#| echo: false
#| message: false

# defines the se function
se <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(length(x))
}

#load these packages, nearly always needed
library(tidyverse)

# sets maize and blue color scheme
color_scheme <- c("#00274c", "#ffcb05")
```

## Purpose

To perform sensitivity analyses for different models and populations

## Data Entry

```{r data-entry}
case.demographics.filename <- 'CaseDemographics.csv'
control.demographics.filename <- 'ControlDemographics.csv'

case.lab.results.filename <- 'LabResults.csv'
control.lab.results.filename <- '../controls/LabResults.csv'

bp.filename.cases <- 'NursingStandardVitalSigns.csv'
bp.filename.controls <- '../controls/NursingStandardVitalSigns.csv'

case.encounter.datafile <- "EncounterAll.csv" 
case.encounter.bmi.datafile <- "EncounterAnthropometricsBMI.csv" 
control.encounter.datafile <- "../controls/EncounterAll.csv" 
control.encounter.bmi.datafile <- "../controls/EncounterAnthropometricsBMI.csv" 

library(readr)
library(dplyr)

#labs dont have dates/ages
case.encounters <- read_csv(case.encounter.datafile)
case.encounters.bmi <- read_csv(case.encounter.bmi.datafile)
case.encounters.all <- left_join(case.encounters,case.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))

control.encounters <- read_csv(control.encounter.datafile)
control.encounters.bmi <- read_csv(control.encounter.bmi.datafile)
control.encounters.all <- left_join(control.encounters,control.encounters.bmi,
                                 by = join_by(DeID_PatientID, DeID_EncounterID))


case.demographics <- read_csv(case.demographics.filename)
control.demographics <- read_csv(control.demographics.filename)

case.labs <- read_csv(case.lab.results.filename) |>
  left_join(case.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
    left_join(case.demographics, by="DeID_PatientID") # added demographic data
control.labs <- read_csv(control.lab.results.filename) |> 
  left_join(control.encounters.all, by = join_by(DeID_PatientID, DeID_EncounterID)) |> #added encounter info
  left_join(control.demographics, by="DeID_PatientID") # added demographic data
#remove large objects not needed further
#rm(control.encounters,control.encounters.all,control.encounters.bmi)

#combined into one dataset
library(lubridate)

#this is used to filter for the window between when a lab result is allowable relative to the procedure.
procedure.result.interval <- 365 #within a year

lab.results <- bind_rows(control.labs |> mutate(Cushings=0),
                         case.labs |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
lab.results.filtered <-
  lab.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
complete.data.race.ethnicity.age <-
  lab.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
complete.data  <-
  complete.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories

#number of controls per case
fold <- 10
```

We began with `r nrow(lab.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(lab.results |> filter(Cushings==1))-nrow(lab.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r lab.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - lab.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics}
lab.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary

library(knitr)
missing.demographic.summary |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data")
```

Our final dataset therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** Cushing's patients control patients.

## Propensity Mapping

Used the R package `mapit` to match based on Age, Gender, Race, and Ethnicity.  We wanted exact matches for everything except age, where we wanted the closest case that had a measurement.  We picked `r fold` matched controls for each case

### Population Definitions

We generated participant pools for all potential matched participants, then specifically for each outcome

```{r total-population-propensity}
library(MatchIt)
m.out.all <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data,
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.all <- match.data(m.out.all)

master.data <- 
  matched_data.all |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity)) 
```


### Hba1c

```{r hba1c-propensity}
m.out.a1c <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("A1C", "XXA1C", "A1C EX", "X0022")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.a1c <- match.data(m.out.a1c)

hba1c.data <- 
  matched_data.a1c |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### HbA1c Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r aic-statistics}
library(emmeans)
#used fully adjusted model
lm.hba1c <- lm(value ~ RaceEthnicity + AgeInYears + GenderName + Cushings * Obesity, data = hba1c.data)

## 1. Adjusted means
emm.hba1c <- emmeans(lm.hba1c, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "HbA1c",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.hba1c <- contrast(emmeans(lm.hba1c, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "HbA1c",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.hba1c <- tidy(lm.hba1c) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "HbA1c",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.hba1c %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.hba1c <- means_wide %>%
  left_join(contrast.hba1c, by = c("Outcome", "Obesity"))

hba1c.summary %>% kable(caption="Summary of HbA1c levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on HbA1c")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=hba1c.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on HbA1c")
```

### Glucose

```{r glucose-propensity}
m.out.glucose <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("GLUC","GLUC_WB")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data.glucose <- match.data(m.out.glucose)

glucose.data <- 
  matched_data.glucose |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### Glucose Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r glucose-statistics}
#used fully adjusted model
lm.glucose <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = glucose.data)

## 1. Adjusted means
emm.glucose <- emmeans(lm.glucose, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Glucose",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.glucose <- contrast(emmeans(lm.glucose, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Glucose",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.glucose <- tidy(lm.glucose) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Glucose",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.glucose %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.glucose <- means_wide %>%
  left_join(contrast.glucose, by = c("Outcome", "Obesity"))

glucose.summary %>% kable(caption="Summary of glucose levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on Glucose")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=glucose.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on Glucose")
```

### Liver Enzymes

#### ALT

```{r alt-propensity}
m.out.alt <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("ALT")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches
matched_data.alt <- match.data(m.out.alt)

alt.data <- 
  matched_data.alt |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))
```


##### ALT Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r alt-statistics}
#used fully adjusted model
lm.alt <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +  Cushings * Obesity, data = alt.data)

## 1. Adjusted means
emm.alt <- emmeans(lm.alt, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "ALT",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.alt <- contrast(emmeans(lm.alt, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "ALT",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.alt <- tidy(lm.alt) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "ALT",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.alt %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.alt <- means_wide %>%
  left_join(contrast.alt, by = c("Outcome", "Obesity"))

alt.summary %>% kable(caption="Summary of ALT levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on ALT")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=alt.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on ALT")
```

#### AST

```{r ast-propensity}
m.out.ast <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = complete.data |> filter(RESULT_CODE %in% c("AST")), 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),
                 ratio=fold) # Force exact matches

matched_data.ast <- match.data(m.out.ast)

ast.data <- 
  matched_data.ast |>
  mutate(value = as.numeric(VALUE)) %>% #forced into numeric form  
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  filter(!is.na(Obesity))
```

##### AST Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r ast-statistics}
#used fully adjusted model
lm.ast <- lm(value ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = ast.data)

## 1. Adjusted means
emm.ast <- emmeans(lm.ast, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "AST",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.ast <- contrast(emmeans(lm.ast, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "AST",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.ast <- tidy(lm.ast) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "AST",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.ast %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.ast <- means_wide %>%
  left_join(contrast.ast, by = c("Outcome", "Obesity"))

ast.summary %>% kable(caption="Summary of AST levels by obesity and pre-existing Cushing's")
lm(value ~ Cushings + Obesity + Cushings:Obesity,data=ast.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and Cushings on AST")

lm(value ~ Cushings + BMI + Cushings:Obesity,data=ast.data) |> 
  tidy() |> 
  kable(caption="2x2 ANOVA with Interaction for effects of Obesity and BMI on AST")
```

### Blood Pressure

Blood pressure data is in a separate file (`r bp.filename.cases` and `r bp.filename.controls`).  We have mean noninvasive blood pressure, as well as systolic and diastolic.  These data, similar to lab results need to be mapped to encounter data 

```{r blood-pressure-data-cleaning}
bp.data.cases <- read_csv(bp.filename.cases) |>
  left_join(case.encounters.all) |>
  left_join(case.demographics, by="DeID_PatientID") 

bp.data.controls <- read_csv(bp.filename.controls) |>
  left_join(control.encounters.all) |>
  left_join(control.demographics, by="DeID_PatientID") 

bp.results <- bind_rows(bp.data.controls |> mutate(Cushings=0),
          bp.data.cases |> mutate(Cushings=1)) |>  #combined, new column indicating cushings
  mutate(cushings_procedure = ymd(cushings_procedure),
         DeID_AdmitDate = mdy_hm(DeID_AdmitDate)) |> 
  mutate(TimePoint = case_when(cushings_procedure<= DeID_AdmitDate ~ "Before",
                               cushings_procedure > DeID_AdmitDate ~ "After",
                               .default="Unknown")) %>% # defined each encounter as before or after
  mutate(ProcedureFollowUp = cushings_procedure - DeID_AdmitDate) #used to define point between the lab result and the procedure.

#filtered out lab results outside of our interval
bp.results.filtered <-
  bp.results |>
  filter(ProcedureFollowUp<procedure.result.interval | is.na(ProcedureFollowUp)) |>
  filter(TimePoint!="After")
  
#filtered out if the age, gender, race, ethnicity was NA for a patient.
bp.data.race.ethnicity.age <-
  bp.results.filtered |>
  filter(!if_any(c(AgeInYears, GenderCode, RaceCode, EthnicityCode), is.na)) #removed if they had missing data for age, gender, race or ethnicity

#this is the complete dataset  
bp.data  <-
  bp.data.race.ethnicity.age |>
  mutate(RaceEthnicity = case_when(EthnicityCode=="HL"~"Hispanic or Latino",
                                   RaceCode=="C"~"White",
                                   RaceCode=="AA"~"Black",
                                   RaceCode=="A"~"Asian",
                                   .default="Other")) #defined race/ethnicity categories
```

#### Inclusion and Exclusion for Blood Pressure Data

We began with `r nrow(bp.results |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls who had lab data, and **`r nrow(bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID))`** cases with Cushing's Disease.

* **Had Laboratory Results**: We also removed `r nrow(bp.results |> filter(Cushings==1))-nrow(bp.results.filtered |> filter(Cushings==1))` lab results from Cushing's patients because they were outside of our acceptable window of `r procedure.result.interval` days before their procedure.  This eliminated a total of **`r bp.results |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow() - bp.results.filtered |> filter(Cushings==1) |> distinct(DeID_PatientID) |> nrow()`** Cushings patients from our dataset.

* **Missing Demographic Data** After filtering out missing demographic data we had `r nrow(bp.data |> filter(Cushings==1) |> distinct(DeID_PatientID))` cases and `r nrow(bp.data |> filter(Cushings==0) |> distinct(DeID_PatientID))` controls with lab results.  The specific pieces of missing information are found here:

```{r mising-demographics-bp}
bp.results.filtered |>
  filter(Cushings==1) |> #only for cushings data
  group_by(DeID_PatientID) %>%
  summarise(
    Age_na = sum(is.na(AgeInYears)),
    Gender_na = sum(is.na(GenderCode)),
    Race_na = sum(is.na(RaceCode)),
    Ethnicity_na = sum(is.na(EthnicityCode)),
    Any_na = any(is.na(AgeInYears) | is.na(GenderCode) | is.na(RaceCode) |is.na(EthnicityCode))) |>
  ungroup() |>
    pivot_longer(cols = ends_with("_na"),
                 names_to = "field",
                 values_to = "has_na") |>
  filter(has_na!=0) %>%
  count(field, name = "n_patients_with_NA")-> missing.demographic.summary.bp

library(knitr)
missing.demographic.summary.bp |> arrange(desc(n_patients_with_NA)) |> kable(caption="Summary of exclusions from Cushing's patient pool due to missing demographic data, for blood pressure measures")
```

Our final dataset for blood pressure therefore included **`r nrow(complete.data |> filter(Cushings==1) |> distinct(DeID_PatientID))`** Cushing's patients, and we had the ability to draw from **`r nrow(complete.data |> filter(Cushings==0) |> distinct(DeID_PatientID))`** control patients.


#### Blood Pressure Propensity Mapping

```{r bp-propensity}
m.out.bp <- matchit(Cushings ~ AgeInYears + GenderCode + RaceEthnicity, 
                 data = bp.data, 
                 method = "nearest", 
                 exact = c("GenderCode", "RaceEthnicity"),# Force exact matches
                 ratio=fold) 

matched_data <- match.data(m.out.bp)

bp.data <- 
  matched_data |>
  arrange(desc(DeID_AdmitDate)) %>% #sort by date
  distinct(DeID_PatientID,.keep_all = T) %>% #unique cases
  mutate(Obesity = if_else(BMI>30, "Obese","Non-Obese"))  |>
  mutate(ObesityIII = if_else(BMI>40, "Class III Obese","All Others"))  |>
  filter(!is.na(Obesity)) 
```

#### Mean Arterial Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r map-statistics}
#used fully adjusted model
lm.map <- lm(BPMeanNonInvasive ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.map <- emmeans(lm.map, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.map <- contrast(emmeans(lm.map, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.map <- tidy(lm.map) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Mean Arterial Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.map %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.map <- means_wide %>%
  left_join(contrast.map, by = c("Outcome", "Obesity"))
```


#### Systolic Blood Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r sbp-statistics}
#used fully adjusted model
lm.sbp <- lm(BPSysNonInvasive ~ RaceEthnicity + AgeInYears + GenderName +                     Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.sbp <- emmeans(lm.sbp, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.sbp <- contrast(emmeans(lm.sbp, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.sbp <- tidy(lm.sbp) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Systolic Blood Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.sbp %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.sbp <- means_wide %>%
  left_join(contrast.sbp, by = c("Outcome", "Obesity"))
```

#### Diastolic Arterial Pressure Statistics

Used the fully adjusted model to test the interaction between Cushings's diagnoses and obesity.

```{r dbp-statistics}
#used fully adjusted model
lm.dbp <- lm(BPDiaNonInvasive ~ RaceEthnicity + AgeInYears + GenderName + Cushings * Obesity, data = bp.data)

## 1. Adjusted means
emm.dbp <- emmeans(lm.dbp, ~ Cushings | Obesity) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Mean_CI = sprintf("%.2f (%.2f–%.2f)", emmean, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, Cushings, Mean_CI)

## 2. Within-stratum Cushing–Control difference
contrast.dbp <- contrast(emmeans(lm.dbp, ~ Cushings | Obesity),
                             method = "revpairwise") %>%
  summary(infer = TRUE) %>%
  as.data.frame() %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Diff_CI = sprintf("%.2f (%.2f–%.2f)", estimate, lower.CL, upper.CL)
  ) %>%
  select(Outcome, Obesity, contrast, Diff_CI, p.value)

## 3. Interaction term
interaction_row.dbp <- tidy(lm.dbp) %>%
  filter(term == "Cushings:ObesityObese") %>%
  mutate(
    Outcome = "Diastolic Blood Pressure",
    Interaction_CI = sprintf("%.2f (%.2f–%.2f)",
                             estimate,
                             estimate - 1.96 * std.error,
                             estimate + 1.96 * std.error)
  ) %>%
  select(Outcome, term, estimate, std.error, statistic, p.value, Interaction_CI)

## 4. Combine into one publication-style table:
# Pivot adjusted means to wide format (Control vs Cushing side by side)
means_wide <- emm.dbp %>%
  mutate(Cushings = ifelse(Cushings == 1, "Cushing", "Control")) %>%
  pivot_wider(names_from = Cushings, values_from = Mean_CI)

# Merge means with differences
table.dbp <- means_wide %>%
  left_join(contrast.dbp, by = c("Outcome", "Obesity"))
```

## Summary of Interaction Effects

```{r interaction-effects-summary}
table_summary <- bind_rows(
  table.glucose,
  table.hba1c,
  table.alt,
  table.ast,
  table.map,
  table.sbp,
  table.dbp
) |>
  select(-contrast)

table_summary |>
  kable(caption="Summary of contrasts for all outcomes") |>
  kable_styling(full_width = FALSE, position = "center")

write_csv(table_summary,"Summary of Contrasts - All Outcomes.csv")

interaction_summary <- bind_rows(
  interaction_row.glucose,
  interaction_row.hba1c,
  interaction_row.alt,
  interaction_row.ast,
  interaction_row.map,
  interaction_row.sbp,
  interaction_row.dbp
) |>
  select(Outcome,Interaction_CI,p.value) |>
  rename(`Interaction Estimate`=Interaction_CI)

interaction_summary |>
  kable(caption="Summary of interactions for all outcomes as mean with 95% confidence interval") |>
  kable_styling(full_width = FALSE, position = "center")
```

## Session Information

```{r session-information}
sessionInfo()
```
