---
title: "Summarizing the Diagnosis Data to Extract Diabetes Diagnoses"
author: "Dave Bridges"
date: "May 25, 2023"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

## Purpose

The Elixhauser comorbidity index uses a combination of type 1 and type 2 diabetes with and without co-morbidities to generate its data.  We want to separately identify type 2 diabetes diagnoses. This script can be found in `r getwd()` and was most recently run on `r date()`.

```{r input-data}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

diagnosis.datafile <- 'ClarityMedicalHistory.csv'
encounters.datafile <- 'EncounterAll.csv'
```

The encounter datafile is in `r encounters.datafile` with the diagnosis data from ClarityMedicalHistory in thee file `r `diagnosis.datafile`. This latter file includes the participant, encounter, date, ICD lexicon (ICD9 or ICD10) the actual code and the name for the code.

```{r data-entry}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(lubridate)

#get diagnosis data
diagnosis.data <- read_csv(diagnosis.datafile) 
  
diagnosis.data <- 
  diagnosis.data %>%
  mutate(DeID_ProblemObservationDate=mdy_hm(DeID_ProblemObservationDate))
```

For ICD10 codes (see https://edit.cms.gov/files/document/valid-icd-10-list.xlsx), all Type 2 Diabetes codes start with *E11* and include the phrase "Type 2"

For ICD9 codes (see https://edit.cms.gov/files/document/valid-icd-9-list.xlsx), all type 2 diabetes codes contain the words type II and Diabetes though in different orders

```{r icd-codes}
diabetes.diagnoses.icd10 <- 
  diagnosis.data %>%
  filter(grepl('E11.', DiagnosisICDCodeMapped)) #includes all ICD10 code E11

diabetes.diagnoses.icd9 <-
  diagnosis.data %>%
  filter(grepl('250.',DiagnosisICDCodeMapped)) %>% #includes ICD9 code 250.x
  filter(grepl('type II',DiagnosisICDNameMapped)) #subset of all diseases marked type II, removes type 1

#combine the two
diabetes.diagnoses <-
  bind_rows(diabetes.diagnoses.icd9,
            diabetes.diagnoses.icd10)

```

This generates a list of all diabetes related diagnoses for all participants.  We want to minimize this to generate a list of all participants who do or do not have a diabetes diagnosis (at any time).

```{r any-diagnosis}
#this generates a list of all participants with at least one ID
diabetic.participants <-
  diabetes.diagnoses %>%
  arrange(DeID_ProblemObservationDate) %>% #sort by first observation
  select(DeID_ProblemObservationDate,DeID_PatientID) %>% #keep just these columns
  rename(DeID_Diabetes_Diagnosis=DeID_ProblemObservationDate) %>%
  distinct(DeID_PatientID,.keep_all=T)  
```

There was a total of `r nrow(diabetic.participants)` that we identified as having type 2 diabetes

# Comparison to Elixhauser Comorbidities Index

```{r data-entry-elixhauser}
comorbidity.datafile <- 'ComorbiditiesElixhauserComprehensive.csv'
comorbidity.data <- read_csv(comorbidity.datafile)

comorbidity.diabetics <-
  comorbidity.data %>%
  filter(DiabetesUncomplicated=='1'|DiabetesComplicated=='1') %>%
  distinct(DeID_PatientID)
```

From the Elixhauser data there were `r nrow(comorbidity.diabetics)` people with diabetes combining those with and without complications.  This may be either type 1 or type 2.

## What percent of all elixhauser diabetic participants had type 2 diabetes?

```{r comorbidity-comparison-elixhauser-complete}
comorbidity.diabetics$DeID_PatientID %in% diabetic.participants$DeID_PatientID %>%
  table %>%
  kable(caption="Number of Elixhauser diabetics identified by IC9/10 codes, these could be type 1")

diabetic.participants$DeID_PatientID %in% comorbidity.diabetics$DeID_PatientID %>%
  table %>%
  kable(caption="Number of ICD9/10 diabetics identified by Elixhauser.  These are perhaps false negatives in Elixhauser")
```

# Output

```{r output}
combined.data <-
  comorbidity.data %>%
  full_join(diabetic.participants,by='DeID_PatientID') %>%
  mutate(Type2Diabetes = case_when(DeID_PatientID %in% diabetic.participants$DeID_PatientID ~ '1',
                                   TRUE~'0'))
  
output.file <- 'ComorbidityDataAnnotated.csv'
write_csv(combined.data, file=output.file)
```

These data were written out to `r output.file`. This is the input file for the other scripts.

# Session Information

```{r sessioninfo}
sessionInfo()
```
