---
title: "Summarizing the Encounter Level Data to Define BMI"
author: "Dave Bridges"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

## Purpose

To test the effect modification of obesity on the stress-diabetes relationships. This script collects the the raw data files, processes and merges them. This script can be found in `r getwd()` and was most recently run on `r date()`.

```{r input-data}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

encounters.datafile <- 'EncounterAnthropometricsBMI.csv'
encounters.datefile <- 'EncounterAll.csv'
mgi.survey.datafile <- 'MGIPatientReportedSurveys.csv'
```

The input data file is in `r encounters.datafile`. This includes the BMI for each patient, potentially at multiple time points. This script takes this file and pulls out just the first BMI measure

```{r data-entry}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(lubridate)

#get encounter data
bmi.data.raw <- read_csv(encounters.datafile) 
encounter.data <- read_csv(encounters.datefile)
survey.data <- read_csv(mgi.survey.datafile)

library(lubridate)
#annotating bmi data with encounters and survey data
bmi.data <- 
  left_join(survey.data, encounter.data, by=c('DeID_PatientID')) %>%
  left_join(bmi.data.raw,by=c('DeID_PatientID','DeID_EncounterID')) %>%
  mutate(EncounterDate = mdy_hm(DeID_AdmitDate))%>%
  mutate(EncounterYear = year(EncounterDate)) %>%
  mutate(Survey_Date = mdy_hm(DeID_Survey_Date))
```

# Finding BMI close to Survey Data

```{r survey-encounter-match}
survey.bmi.data <-
  bmi.data %>%
  filter(!is.na(BMI)) %>%
  mutate(Time.from.Survey = abs((Survey_Date %--% EncounterDate)/ddays(1))) %>%
  group_by(DeID_PatientID) %>%
  filter(Time.from.Survey == min(Time.from.Survey,na.rm=T)) %>%
  rename(Survey.BMI = BMI)

range <- 365
range.bmi.data <-
  survey.bmi.data %>%
  filter(Time.from.Survey < range)
```

There was `r `nrow survey.bmi.data` participants with BMI data.  After filtering out those without a BMI within `r range` days from the survey we were left with `r `nrow range.bmi.data` participants.


```{r output}
output.file <- 'SurveyAnthropometricsBMI.csv'
write_csv(range.bmi.data, file=output.file)
```

These data were written out to `r output.file`. This is the input file for the other scripts.

# Session Information

```{r sessioninfo}
sessionInfo()
```
