---
title: "Association of and Stress with Type 2 Diabetes"
author: "Dave Bridges"
date: "July 29, 2025"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
---

## Purpose

To test the effect modification of stress on diabetes risk, including multivariable analyses.

```{r data-entry}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=TRUE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)

library(readr)
library(dplyr)
library(tidyr)
library(broom)

input.file <- 'data-combined.csv'
combined.data <- read_csv(input.file)

```

Loaded in the cleaned data from `r input.file`. This script can be found in `r getwd()` and was most recently run on `r date()`. This dataset has `r dim(combined.data)[1]` values.

```{r data-cleaning}
library(forcats)
combined.data <- 
  combined.data %>%
  mutate(BMI_cat= factor(BMI_cat, 
                         levels=c("Underweight",
                                  "Normal",
                                  "Overweight",
                                  'Class I Obese',
                                  'Class II Obese',
                                  'Class III Obese'))) %>%
  mutate(BMI_cat.obese= factor(BMI_cat.obese, 
                               levels=c("Underweight",
                                        "Normal",
                                        "Overweight",
                                        'Obese'))) %>%
  mutate(BMI_cat.Ob.NonOb= factor(BMI_cat.Ob.NonOb, 
                                  levels=c("Non-Obese",
                                           'Obese'))) %>%
  mutate(Stress=relevel(as.factor(High.Stress),ref="Low")) %>% #set low as reference value
    mutate(Race.Ethnicity=relevel(as.factor(Race.Ethnicity),ref="White")) %>% #set white as reference value
  mutate(Stress.quartile=fct_recode(as.factor(Stress.quartile),
                                    "Q1"="(-0.016,4]",
                                    "Q2"="(4,8]",
                                    "Q3"="(8,12]",
                                    "Q4"="(12,16]")) %>%
    mutate(Stress.quartile= factor(Stress.quartile, 
                                  levels=c("Q1","Q2","Q3","Q4"))) 
  
```

# Effects of Stress on Diabetes Risk

Started with a crude model of:

$$\log \left( \frac{P(\text{Diabetes} = 1 \mid \text{Stress})}{1 - P(\text{Diabetes} = 1 \mid \text{Stress})} \right)
= \beta_0 + \beta_1 \times \text{Stress}$$

```{r stress-diabetes-univariable}
with(combined.data, table(Type2Diabetes,Stress)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) -> diabetes.stress.counts

library(ggplot2)

ggplot(diabetes.stress.counts,
       aes(y=Percent,
           fill=Stress,
           x=Stress)) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Stress",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.counts, caption="Diabetes rates by stress category")

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress, 
    family="binomial",
    data=combined.data) -> stress.glm1

library(broom)
stress.glm1 %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Adding in Obesity as a Covariate

Next added in obesity (obese vs non-obese) as a covariate

```{r stress-diabetes-obesity}
with(combined.data, table(Type2Diabetes,Stress,BMI_cat.Ob.NonOb)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) -> diabetes.stress.obesity.counts

ggplot(diabetes.stress.obesity.counts,
       aes(y=Percent,
           x=BMI_cat.Ob.NonOb,
           fill=Stress)) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Stress",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.obesity.counts, caption="Diabetes rates by stress and obesity")

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress+BMI_cat.Ob.NonOb, 
    family="binomial",
    data=combined.data) -> stress.glm2

library(broom)
stress.glm2 %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress and obesity on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Adding in Gender as a Covariate

```{r stress-diabetes-obesity-gender}
with(combined.data, table(Type2Diabetes,Stress,BMI_cat.Ob.NonOb,Gender)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) -> diabetes.stress.obesity.gender.counts

ggplot(diabetes.stress.obesity.gender.counts,
       aes(y=Percent,
           x=BMI_cat.Ob.NonOb,
           fill=Stress)) +
  facet_grid(.~Gender) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Gender",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.obesity.counts, caption="Diabetes rates by stress, obesity and gender")

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress+BMI_cat.Ob.NonOb+Gender, 
    family="binomial",
    data=combined.data) -> stress.glm3

stress.glm3 %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, and gender on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Adding in Age as a Covariate

```{r stress-diabetes-obesity-age}
with(combined.data, table(Type2Diabetes,Stress,BMI_cat.Ob.NonOb,Gender,Age.group)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) -> diabetes.stress.obesity.age.counts

ggplot(diabetes.stress.obesity.age.counts,
       aes(y=Percent,
           x=BMI_cat.Ob.NonOb,
           fill=Stress)) +
  facet_grid(.~Age.group) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Age",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.obesity.counts, caption="Diabetes rates by stress, obesity and gender")

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress+BMI_cat.Ob.NonOb+Gender+Age.group, 
    family="binomial",
    data=combined.data) -> stress.glm4

stress.glm4 %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, gender, and age on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Adding in Race/Ethnicity as a Covariate

```{r stress-diabetes-obesity-race}
with(combined.data, table(Type2Diabetes,Stress,BMI_cat.Ob.NonOb,Race.Ethnicity)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) -> diabetes.stress.obesity.race.counts

ggplot(diabetes.stress.obesity.race.counts,
       aes(y=Percent,
           x=BMI_cat.Ob.NonOb,
           fill=Stress)) +
  facet_grid(.~Race.Ethnicity) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Age",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.obesity.counts, caption="Diabetes rates by stress, obesity and gender")

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress+BMI_cat.Ob.NonOb+Gender+Age.group+Race.Ethnicity, 
    family="binomial",
    data=combined.data) -> stress.glm5

stress.glm5 %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, gender, age, and race/ethnicity on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Adding in Neighborhood SES as a Covariate

Used `disadvantage13_17_qrtl` as the variable

```{r stress-diabetes-obesity-ses}
with(combined.data, table(Type2Diabetes,Stress,BMI_cat.Ob.NonOb,disadvantage13_17_qrtl)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) -> diabetes.stress.obesity.ses.counts

ggplot(diabetes.stress.obesity.ses.counts,
       aes(y=Percent,
           x=BMI_cat.Ob.NonOb,
           fill=Stress)) +
  facet_grid(.~disadvantage13_17_qrtl) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Age",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.obesity.counts, caption="Diabetes rates by stress, obesity and gender")

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress+BMI_cat.Ob.NonOb+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data) -> stress.glm6

stress.glm6 %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, gender, age, and neighborhood SES on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

# Combined Summary of Multivariable Models

```{r combined-multivariable-models}
combined.model.data <-
  bind_rows(tidy(stress.glm1, exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
              mutate(Model="Unadjusted"),
            tidy(stress.glm2,exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
              mutate(Model="+ Obesity"),
            tidy(stress.glm3,exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
              mutate(Model="+ Gender"),
            tidy(stress.glm4,exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
              mutate(Model="+ Age"),
            tidy(stress.glm5,exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
              mutate(Model="+ Race/Ethnicity"),
            tidy(stress.glm6,exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
              mutate(Model="+ Neighborhood SES")) %>%
  filter(term=="StressHigh") %>%
  mutate(`OR [95% CI]` = paste(round(estimate,3),
                          " [",
                          round(conf.low,3),
                          "-",
                          round(conf.high,3),
                          "]",
                          sep="")) %>%
  select(Model,`OR [95% CI]`,p.value)

kable(combined.model.data,
      caption="Summary of multivariable models",
      digits =c(0,0,99))

write_csv(combined.model.data, 
          "Multivariable Analysis of Stress-Diabetes Associations.csv")
```

# Effect Modification by Obesity

Used the fully adjusted models to evaluate effect modification by obesity status

```{r stress-diabetes-obesity-effect-modification}
with(combined.data, table(Type2Diabetes,Stress,BMI_cat.Ob.NonOb)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) %>%
  mutate(Odds = Percent/100 / (1 - Percent/100))-> diabetes.stress.obesity.counts

# Reshape to wide format for easier OR calculation
diabetes.stress.obesity.counts.wide <- diabetes.stress.obesity.counts %>%
  select(Stress, BMI_cat.Ob.NonOb, Odds) %>%
  tidyr::pivot_wider(names_from = Stress, values_from = Odds)

# Compute odds ratios for stress within each obesity group
odds.ratios <- diabetes.stress.obesity.counts.wide %>%
  mutate(OR = High / Low) %>%
  mutate(Interaction_OR = if_else(BMI_cat.Ob.NonOb == "Obese", 
                                  OR / OR[BMI_cat.Ob.NonOb == "Non-Obese"], 
                                  NA_real_))

kable(odds.ratios, caption="Crude odds and odds ratios")

ggplot(diabetes.stress.obesity.counts,
       aes(y=Percent,
           x=BMI_cat.Ob.NonOb,
           fill=Stress)) +
  geom_bar(stat='identity',position='dodge') +
  labs(y="Prevalence of Type 2 Diabetes",
       title="Age",
       x="") +
  theme_classic() +
  scale_fill_grey() +
  theme(text=element_text(size=16),
        axis.text.x=element_text(angle=90,vjust=0.5,hjust=1),
        legend.position.inside = c(0.1,0.85))

kable(diabetes.stress.obesity.counts, caption="Diabetes rates by stress, obesity and gender")
  

#adding in age and gender as covariates as a modifier
glm(Type2Diabetes~Stress+BMI_cat.Ob.NonOb+Gender+Age.group+disadvantage13_17_qrtl+Stress:BMI_cat.Ob.NonOb, 
    family="binomial",
    data=combined.data) -> stress.em

stress.em %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, gender, age, and neighborhood SES with effect modification by obesity status on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Stratified Fully Adjusted Models in Support of Effect Modification

```{r stratified-models}
glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat.Ob.NonOb=="Non-Obese")) -> stress.em.non.obese

stress.em.non.obese %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Participants with a BMI <30  Logistic regression of stress, gender, age, and neighborhood SES with effect modification by obesity status on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))

glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat.Ob.NonOb=="Obese")) -> stress.em.obese

stress.em.obese %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Participants with obesity.  Logistic regression of stress, gender, age, and neighborhood SES with effect modification by obesity status on diabetes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```
# Summary Table of Interaction Estimates

```{r interaction-estimates}
interaction.summary.table <-
  diabetes.stress.obesity.counts %>%
  mutate(`Cases/Total`=paste0(Diabetes,"/",Total)) %>%
  select(BMI_cat.Ob.NonOb,Stress,`Cases/Total`) %>%
  pivot_wider(names_from=Stress, values_from=`Cases/Total`) %>%
  rename(BMI = BMI_cat.Ob.NonOb,
         `Lower Stress`=Low,
         `Higher Stress`=High) %>%
  mutate(`OR [95% CI]`=c(
    stress.em.non.obese %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate),
    stress.em.obese %>% tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
      select(estimate, conf.low, conf.high) %>%
      mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate)))

kable(interaction.summary.table, caption="Summary Table for Interaction Effects")

write_csv(interaction.summary.table,"Obesity-Stress Interaction Table.csv")
```

# Interaction Effects Across Obesity Categories

Repeated this analysis across all BMI categories

```{r stress-obesity-bmi-categories}
glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat=="Underweight")) -> stress.em.1

glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat=="Normal")) -> stress.em.2

glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat=="Overweight")) -> stress.em.3

glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat=="Class I Obese")) -> stress.em.4

glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat=="Class II Obese")) -> stress.em.5

glm(Type2Diabetes~Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data %>% filter(BMI_cat=="Class III Obese")) -> stress.em.6

diabetes.stress.obesity.class.counts <-
  with(combined.data, table(Type2Diabetes,Stress,BMI_cat)) %>% 
  data.frame %>%
  pivot_wider(names_from=Type2Diabetes,
              values_from = Freq) %>%
  rename(Diabetes=`1`,
         NonDiabetes=`0`) %>%
  mutate(Total=Diabetes+NonDiabetes) %>%
  mutate(Percent=Diabetes/Total*100) %>%
  mutate(Odds = Percent/100 / (1 - Percent/100))

interaction.summary.table.class <-
  diabetes.stress.obesity.class.counts %>%
  mutate(`Cases/Total`=paste0(Diabetes,"/",Total)) %>%
  select(BMI_cat,`Cases/Total`,Stress) %>%
  pivot_wider(names_from=Stress, values_from=`Cases/Total`) %>%
  rename(BMI = BMI_cat,
         `Lower Stress`=Low,
         `Higher Stress`=High) %>%
  mutate(`OR [95% CI]`=c(
    stress.em.1 %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate),
    stress.em.2 %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate),
    stress.em.3 %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate),
    stress.em.4 %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate),
    stress.em.5 %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate),
    stress.em.6 %>% 
           tidy(exponentiate = T, conf.int = T, conf.level=0.95) %>% filter(term=="StressHigh") %>%
           select(estimate, conf.low, conf.high) %>%
           mutate(Estimate = paste0(round(estimate,2), " [", round(conf.low,2),"-",round(conf.high,2),"]")) %>%
      pull(Estimate)))

kable(interaction.summary.table.class, caption="Summary Table for Interaction Effects")

write_csv(interaction.summary.table.class, "Obesity-Stress Interaction Table by Class.csv")

glm(Type2Diabetes~Stress+BMI_cat+Gender+Age.group+disadvantage13_17_qrtl+Stress:BMI_cat, 
    family="binomial",
    data=combined.data) -> stress.em.class

stress.em.class %>% anova %>% tidy %>% kable(caption="Omnibus ANOVA for all six obesity classes")

stress.em.class %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, gender, age, and neighborhood SES with effect modification by obesity status on diabetes.  This analysis uses all six obesity classes (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

## Nonlinear BMI-Diabetes Relationship

Rather than categories i used a nonlinear spline relationship using actual BMI.  We modeled the association between body mass index (BMI) and diabetes risk using natural cubic splines to flexibly capture potential nonlinear relationships. Specifically, BMI was included as a natural cubic spline with 3 degrees of freedom `(ns(BMI, df = 3))`. Natural cubic splines allow the relationship between BMI and the log-odds of diabetes to vary smoothly across the range of BMI values without imposing a strict linear form, while constraining the function to be linear beyond the boundary knots to ensure stable estimates at extreme BMI values.

```{r diabetes-stress-nonlinear}
library(splines)
glm(Type2Diabetes~ns(BMI,df=3)*Stress+Gender+Age.group+disadvantage13_17_qrtl, 
    family="binomial",
    data=combined.data) -> stress.em.nl

stress.em.nl %>% anova %>% tidy %>% kable(caption="Omnibus ANOVA for all a nonlinear obesity relationship")

stress.em.nl %>%
  tidy(exponentiate=T, conf.int = TRUE, conf.level = 0.95) %>%
  kable(caption="Logistic regression of stress, obesity, gender, age, and neighborhood SES with effect modification by obesity status on diabetes.  This analysis uses a nonlinear BMI-diabetes relationship (exponentiated estimates)", digits =c(0,2,3,2,99,3,3))
```

# Session Information

```{r sessioninfo}
sessionInfo()
```
