---
title: "Combined Analysis of HFD/NCD Dexamethasone Serum Glycerol Levels"
author: "Innocence Harvey and Dave Bridges"
date: "September 17, 2015"
output:
  html_document:
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: no
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures makde will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('#00274c', '#ffcb05')
```


```{r data-entry}
filename <- '../../data/raw/HFD vs Chow Glycerol Assay.xlsx'
sheetname <- 'Exported'
library(xlsx)
combined_data <- read.xlsx(filename, sheetName = sheetname)
combined_data$Diet <- relevel(as.factor(combined_data$Diet), ref='Normal Chow Diet')
combined_data$Treatment <- relevel(as.factor(combined_data$Treatment), ref='Water')

library(dplyr)
library(tidyr)
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))
tg.summary <-
  combined_data %>%
  group_by(Diet, Treatment) %>%
  summarize(Glycerol.mean = mean(Glycerol, na.rm=T),
            Glycerol.se = se(Glycerol),
            Glycerol.n = length(Glycerol))
```
This script uses the files in `r filename` and used the worksheet `r sheetname`. These data are located in `r getwd()` and this script was most recently run on `r date()`.

```{r ncd-hfd-glycerol-barplot}
plot.data <- select(tg.summary, Diet,Treatment, Glycerol.mean) %>% spread(Diet, Glycerol.mean)
plot.data.se <- select(tg.summary, Diet,Treatment, Glycerol.se) %>% spread(Diet, Glycerol.se)

plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
        ylab="Glycerol in Serum (ug/ul)",col=palette()[1:2],
        las=1, ylim=c(0,max(tg.summary$Glycerol.mean + tg.summary$Glycerol.se)))

superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
legend("topleft", levels(combined_data$Treatment), fill=palette()[1:2], bty="n")
```

We observed a `r (filter(tg.summary, Diet=="Normal Chow Diet")$Glycerol.mean[2]-filter(tg.summary, Diet=="Normal Chow Diet")$Glycerol.mean[1])/filter(tg.summary, Diet=="Normal Chow Diet")$Glycerol.mean[2]*100`% increase in glycerol levels in the chow mice, but a `r filter(tg.summary, Diet=="High Fat Diet")$Glycerol.mean[2]/filter(tg.summary, Diet=="High Fat Diet")$Glycerol.mean[1]` fold increase in the HFD animals.

# Statistics

```{r statistics}
tg.aov <- aov(Glycerol ~ Diet + Treatment + Diet:Treatment, data=combined_data)
tg.tukey <- TukeyHSD(tg.aov)
```

From a two-way ANOVA with an interaction, the interaction term was significant (p=`r summary(tg.aov)[[1]]$"Pr(>F)"[3]`).

```{r aov-table-glycerol}
kable(summary(tg.aov)[[1]], caption="ANOVA for Glycerol Levels", digits=c(2,2,2,2,5))
```

Tukey's tests (pairwise  comparasons are):

```{r tukey-table}
kable(tg.tukey$`Diet:Treatment`, caption="Tukey's HSD tests for Serum Glycerol Levels", digits=c(2,2,2,10))
```

# Session Information

```{r session-info}
sessionInfo()
```