---
title: "Combined Analysis of HFD/NCD Dexamethasone WAT ATGL Levels"
author: "Innocence Harvey and Dave Bridges"
date: "February 11, 2016"
output:
  html_document:
    highlight: tango
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    highlight: tango
    keep_tex: yes
    number_sections: yes
    toc: yes
---

```{r global_options, include=FALSE}
library(knitr)
#figures made will go to directory called figures, will make them as both png and pdf files 
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)
# set echo and message to TRUE if you want to display code blocks and code output respectively

knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

#load these packages, nearly always needed
library(tidyr)
library(dplyr)

# sets maize and blue color scheme
color.scheme <- c('black', 'red')

library(ggplot2)
#sets default theme
theme_set(theme_bw())
```


```{r data-entry}
filename <- '../../data/processed/aGRKO HFD Male IWAT Westerns.csv'
library(readr)
combined_data <- read_csv(filename)
combined_data$Genotype <- relevel(as.factor(combined_data$Genotype), ref='WildType')
combined_data$Treatment <- relevel(as.factor(combined_data$Treatment), ref='Water')

#relabeled factors
library(forcats)
combined_data$Treatment <- fct_recode(combined_data$Treatment, "Water" = "Control") 
combined_data$Genotype <- fct_recode(combined_data$Genotype, "WildType" = "WT") 
combined_data$Genotype <- fct_recode(combined_data$Genotype, "Knockout" = "aGRKO") 

shapiro.p <- function(x) shapiro.test(x)$p.value
mean.na <- function(x) mean(x, na.rm=T)
 
blot.data <-
  combined_data %>%
  mutate(ATGL.norm = ATGL/Revert,
         CGI58.norm = Adhb/Revert,
         HSL.norm = Hsl/Revert,
         Perlipin.norm = Perilipin/Revert,
         pPerilipin.norm = pPerilipin/Perilipin,
         pHSL_563.norm = `pHsl 563`/Hsl,
         pHSL_660.norm = `pHsl 660`/Hsl) %>%
  select(Genotype,Treatment, ends_with('norm')) %>%
  mutate(ATGL.rel=ATGL.norm/mean(ATGL.norm[Genotype=="WildType"&Treatment=="Water"],na.rm=T))
         # HSL.rel=HSL.norm/mean(HSL.norm[Genotype=="Normal Chow Genotype"&Treatment=="Water"],na.rm=T),
         # CGI58.rel=CGI58.norm/mean(CGI58.norm[Genotype=="Normal Chow Genotype"&Treatment=="Water"],na.rm=T),
         # Perlipin.rel=Perlipin.norm/mean(Perlipin.norm[Genotype=="Normal Chow Genotype"&Treatment=="Water"],na.rm=T),
         # pHSL_563.rel=pHSL_563.norm/mean(pHSL_563.norm[Genotype=="Normal Chow Genotype"&Treatment=="Water"],na.rm=T),
         # pHSL_660.rel=pHSL_660.norm/mean(pHSL_660.norm[Genotype=="Normal Chow Genotype"&Treatment=="Water"],na.rm=T),
         # pPerilipin.rel=pPerilipin.norm/mean(pPerilipin.norm[Genotype=="Normal Chow Genotype"&Treatment=="Water"],na.rm=T)) 

blot.summary <-
  blot.data %>%
  group_by(Genotype, Treatment) %>% 
  select(-ends_with('norm')) %>%
  summarize_if(is.numeric, funs(mean.na,se,length))

blot.summary %>% select(ends_with('mean.na'),Treatment) %>% 
  rename(ATGL = ATGL.rel_mean.na, 
         HSL = HSL.rel_mean.na, 
         CGI58= CGI58.rel_mean.na,
         Perilipin=Perlipin.rel_mean.na,
         pHSL_563=pHSL_563.rel_mean.na,
         pHSL_660=pHSL_660.rel_mean.na,
         pPerilipin=pPerilipin.rel_mean.na) %>% 
  gather(Protein,Levels,-Genotype,-Treatment) %>% 
  kable(caption="Normalized Average Levels")
```


This script uses the files in `r filename`. These data are located in `r getwd()` and this script was most recently run on `r date()`.

# ATGL

```{r atgl-wat-barplot}
plot.data <- blot.summary %>% select(Genotype,Treatment,ATGL.rel_mean.na) %>% spread(Genotype,ATGL.rel_mean.na)
plot.data.se <- blot.summary %>% select(Genotype,Treatment,ATGL.rel_se) %>% spread(Genotype,ATGL.rel_se)

ymax <- 3


plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
        ylab="ATGL (Relative Units)",col=palette()[1:2],
        las=1, ylim=c(0,ymax))

superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
```

Blots were normalized to total protein (Revert) staining.  We observed a `r blot.summary$ATGL.rel_mean.na[2]` fold increase in NCD animals due to dexamethasone and a `r blot.summary$ATGL.rel_mean.na[4]/blot.summary$ATGL.rel_mean.na[3]` fold increase in HFD animals.

## ATGL Statistics

```{r atgl-statistics}
tg.aov <- aov(ATGL.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
library(broom)
kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for ATGL Levels")
```

# HSL

```{r hsl-wat-barplot}
# plot.data <- blot.summary %>% select(Genotype,Treatment,HSL.rel_mean.na) %>% spread(Genotype,HSL.rel_mean.na)
# plot.data.se <- blot.summary %>% select(Genotype,Treatment,HSL.rel_se) %>% spread(Genotype,HSL.rel_se)
# 
# ymax <- 1.2
# 
# plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
#         ylab="HSL (Relative Units)",col=palette()[1:2],
#         las=1, ylim=c(0,ymax))
# 
# superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
# #legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
# ```
# 
# Blots were normalized to total protein (Revert) staining.  We observed a `r blot.summary$HSL.rel_mean.na[2]` fold increase in NCD animals due to dexamethasone and a `r blot.summary$HSL.rel_mean.na[4]/blot.summary$HSL.rel_mean.na[3]` fold increase in HFD animals.
# 
# ## HSL Statistics
# 
# ```{r hsl-statistics}
# tg.aov <- aov(HSL.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
# kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for HSL Levels")
# ```
# 
# # pHSL-563
# 
# ```{r hsl-563-wat-barplot}
# plot.data <- blot.summary %>% select(Genotype,Treatment,pHSL_563.rel_mean.na) %>% spread(Genotype,pHSL_563.rel_mean.na)
# plot.data.se <- blot.summary %>% select(Genotype,Treatment,pHSL_563.rel_se) %>% spread(Genotype,pHSL_563.rel_se)
# 
# ymax <- 2.2
# 
# plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
#         ylab="pHSL-Ser 563 (Relative Units)",col=palette()[1:2],
#         las=1, ylim=c(0,ymax))
# 
# superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
# #legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
# ```
# 
# Blots were normalized to total protein (Revert) staining.  We observed a `r blot.summary$pHSL_563.rel_mean.na[2]` fold increase in NCD animals due to dexamethasone and a `r blot.summary$pHSL_563.rel_mean.na[4]/blot.summary$pHSL_563.rel_mean.na[3]` fold increase in HFD animals.
# 
# ## pHSL-563 Statistics
# 
# ```{r hsl-563-statistics}
# tg.aov <- aov(pHSL_563.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
# kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for HSL pSer 563 Levels")
# ```
# 
# # pHSL-660
# 
# ```{r hsl-660-wat-barplot}
# plot.data <- blot.summary %>% select(Genotype,Treatment,pHSL_660.rel_mean.na) %>% spread(Genotype,pHSL_660.rel_mean.na)
# plot.data.se <- blot.summary %>% select(Genotype,Treatment,pHSL_660.rel_se) %>% spread(Genotype,pHSL_660.rel_se)
# 
# ymax <- 2.2
# 
# plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
#         ylab="pHSL-Ser 660 (Relative Units)",col=palette()[1:2],
#         las=1, ylim=c(0,ymax))
# 
# superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
# #legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
```

Blots were normalized to total protein (Revert) staining.  We observed a `r blot.summary$pHSL_660.rel_mean.na[2]` fold increase in NCD animals due to dexamethasone and a `r blot.summary$pHSL_660.rel_mean.na[4]/blot.summary$pHSL_660.rel_mean.na[3]` fold increase in HFD animals.

## pHSL-660 Statistics

```{r hsl-660-statistics}
tg.aov <- aov(pHSL_660.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for HSL pSer 563 Levels")
```

# Perilipin

```{r perilipin-wat-barplot}
# plot.data <- blot.summary %>% select(Genotype,Treatment,Perlipin.rel_mean.na) %>% spread(Genotype,Perlipin.rel_mean.na)
# plot.data.se <- blot.summary %>% select(Genotype,Treatment,Perlipin.rel_se) %>% spread(Genotype,Perlipin.rel_se)
# 
# ymax <- 8
# 
# plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
#         ylab="Total Perilipin (Relative Units)",col=palette()[1:2],
#         las=1, ylim=c(0,ymax))
# 
# superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
# #legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
# ```
# 
# ## Perilipin Statistics
# 
# ```{r perilipin-statistics}
# tg.aov <- aov(Perlipin.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
# kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for Total Perilipin Levels")
# ```
# 
# # pPerilipin
# 
# ```{r p-perilipin-wat-barplot}
# plot.data <- blot.summary %>% select(Genotype,Treatment,pPerilipin.rel_mean.na) %>% spread(Genotype,pPerilipin.rel_mean.na)
# plot.data.se <- blot.summary %>% select(Genotype,Treatment,pPerilipin.rel_se) %>% spread(Genotype,pPerilipin.rel_se)
# 
# ymax <- 1.2
# 
# plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
#         ylab="pPerilipin (Relative Units)",col=palette()[1:2],
#         las=1, ylim=c(0,ymax))
# 
# superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
# #legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
# ```
# 
# ## pPerilipin Statistics
# 
# ```{r p-perilipin-statistics}
# tg.aov <- aov(pPerilipin.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
# kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for Perilipin Phosphorylation Levels")
# ```
# 
# # ADRB5 or CGI-58
# 
# ```{r cgi58-wat-barplot}
# plot.data <- blot.summary %>% select(Genotype,Treatment,CGI58.rel_mean.na) %>% spread(Genotype,CGI58.rel_mean.na)
# plot.data.se <- blot.summary %>% select(Genotype,Treatment,CGI58.rel_se) %>% spread(Genotype,CGI58.rel_se)
# 
# ymax <- 1.5
# 
# plot <- barplot(as.matrix(plot.data[2:3]), beside=T, 
#         ylab="CGI-58 (Relative Units)",col=palette()[1:2],
#         las=1, ylim=c(0,ymax))
# 
# superpose.eb(plot, as.matrix(plot.data[2:3]), as.matrix(plot.data.se[2:3]))
# #legend("topleft", levels(combined_data$Treatment), fill=palette(), bty="n")
```

Blots were normalized to total protein (Revert) staining.  We observed a `r blot.summary$CGI58.rel_mean.na[2]` fold increase in NCD animals due to dexamethasone and a `r blot.summary$CGI58.rel_mean.na[4]/blot.summary$CGI58.rel_mean.na[3]` fold increase in HFD animals.

## CGI-58 Statistics

```{r cgi58-statistics}
# tg.aov <- aov(CGI58.rel ~ Genotype + Treatment + Genotype:Treatment, data=blot.data)
# kable(tidy(tg.aov), caption="2-Way ANOVA with Interaction for CGI-58 Levels")
```



# Session Information

```{r session-info}
sessionInfo()
```