---
title: "Analysis of Tissue Weights from HFD/Dexamethasone Studies"
author: "Innocence Harvey and Dave Bridges"
date: "April 7, 2017"
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
  htmin_notebook: toc:yes
  htmin_document:
    keep_md: yes
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.path='figures/',
               echo=FALSE, warning=FALSE, message=FALSE,dev=c('png','pdf'))
options(scipen = 2, digits = 3)


knitr::knit_hooks$set(inline = function(x) {
  knitr:::format_sci(x, 'md')
})


superpose.eb <- function (x, y, ebl, ebu = ebl, length = 0.08, ...)
  arrows(x, y + ebu, x, y - ebl, angle = 90, code = 3,
  length = length, ...)

  
se <- function(x) sd(x, na.rm=T)/sqrt(length(x))

library(tidyr)
library(dplyr)

color.scheme <- c("black","red")
```

# Data Entry

```{r data-entry}
datafile <- '../../data/raw/HFD and Chow Tissue Weights.csv'
library(readr)
data <- read_csv(datafile)
library(forcats)

#set reference levels
data$Treatment <- relevel(as.factor(data$Group), ref="Control")
data$Diet <- relevel(as.factor(data$Diet), ref="Chow")
data$Status <- relevel(as.factor(data$Status), ref="Fasted")

#recoded factor names
data$Treatment <- fct_recode(data$Treatment, "Water" = "Control") 
data$Treatment <- fct_recode(data$Treatment, "Dexamethasone" = "Dex") 
data$Diet <- fct_recode(data$Diet, "NCD" = "Chow") 
```

This script generates figures from the tissue weights found in `r datafile`.  This file is located in `r getwd()` and was most recently updated on `r date()`.

# Number of Animals

```{r n-calculation}
summary.data.numbers <-
  data %>%
  group_by(Status, Diet, Treatment) %>%
  summarize(Number=length(iWAT))

kable(summary.data.numbers, caption="Number of mice in each group")
```

# Summary Data

The remainder of this analysis is only for Fasted animals

```{r analysis}
mean.na <- function(x) mean(x, na.rm=T)
shapiro.p <- function(x) shapiro.test(x)$p.value

summary.data <-
  filter(data, Status=="Fasted") %>%
  group_by(Diet, Treatment) %>%
  select(iWAT,eWAT,TS,Quad) %>%
  summarize_each(funs(mean.na,se, shapiro.p))

kable(select(summary.data, Treatment, ends_with('mean.na')), caption="Averaged Values")
```

# Muscle Barplots

```{r muscle-weight-barplot}

par(mfrow=c(1,2))
plot.data <- summary.data %>% select(Treatment, Quad_mean.na) %>% spread(Diet, Quad_mean.na) 
plot.data.se <- summary.data %>% select(Treatment, Quad_se) %>% spread(Diet, Quad_se)

ymax <- max(plot.data[2:3]+max(plot.data.se[2:3]))
plot <- barplot(as.matrix(plot.data[2:3]),
                beside=T, las=1,
                col=color.scheme,
                ylim=c(0,ymax),
                ylab="Tissue Weight (mg)",
                main="Quadriceps")
superpose.eb(plot,
             as.matrix(plot.data[2:3]),
             as.matrix(plot.data.se[2:3]))
#for TS muscle

plot.data <- summary.data %>% select(Treatment, TS_mean.na) %>% spread(Diet, TS_mean.na) 
plot.data.se <- summary.data %>% select(Treatment, TS_se) %>% spread(Diet, TS_se)

ymax <- max(plot.data[2:3]+max(plot.data.se[2:3]))
plot <- barplot(as.matrix(plot.data[2:3]),
                beside=T, las=1,
                col=color.scheme,
                ylim=c(0,ymax),
                ylab="Tissue Weight (mg)",
                main="Triceps surae")
superpose.eb(plot,
             as.matrix(plot.data[2:3]),
             as.matrix(plot.data.se[2:3]))

```

# Adipose Barplots

```{r adipose-weight-barplot}

par(mfrow=c(1,2))
plot.data <- summary.data %>% select(Treatment, iWAT_mean.na) %>% spread(Diet, iWAT_mean.na) 
plot.data.se <- summary.data %>% select(Treatment, iWAT_se) %>% spread(Diet, iWAT_se)

ymax <- max(plot.data[2:3]+max(plot.data.se[2:3]))
plot <- barplot(as.matrix(plot.data[2:3]),
                beside=T, las=1,
                col=color.scheme,
                ylim=c(0,ymax),
                ylab="Tissue Weight (mg)",
                main="Inguinal Adipose Tissue")
superpose.eb(plot,
             as.matrix(plot.data[2:3]),
             as.matrix(plot.data.se[2:3]))
#for TS muscle

plot.data <- summary.data %>% select(Treatment, eWAT_mean.na) %>% spread(Diet, eWAT_mean.na) 
plot.data.se <- summary.data %>% select(Treatment, eWAT_se) %>% spread(Diet, eWAT_se)

ymax <- max(plot.data[2:3]+max(plot.data.se[2:3]))
plot <- barplot(as.matrix(plot.data[2:3]),
                beside=T, las=1,
                col=color.scheme,
                ylab="Tissue Weight (mg)",
                ylim=c(0,ymax),
                main="Epididymal Adipose Tissue")
superpose.eb(plot,
             as.matrix(plot.data[2:3]),
             as.matrix(plot.data.se[2:3]))

```

# Analysis of Adipose Tissue Weights

We observed reductions in both iWAT and eWAT with HFD/Dexamethasone.  This included a `r  (filter(summary.data, Diet=="HFD")$iWAT_mean.na[1]-filter(summary.data, Diet=="HFD")$iWAT_mean.na[2])/filter(summary.data, Diet=="HFD")$iWAT_mean.na[1]*100` reduction in iWAT mass and a `r  (filter(summary.data, Diet=="HFD")$eWAT_mean.na[1]-filter(summary.data, Diet=="HFD")$eWAT_mean.na[2])/filter(summary.data, Diet=="HFD")$eWAT_mean.na[1]*100` reduction in eWAT mass.

# Session Information
```{r sessionInfo}
sessionInfo()
```
